### 2024-02-02 paper.Rmd
corrections

---
title: "Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists"
author: "Ronald Visser"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction

ABM application increased in Archaeology

No formal training

EU project

# Method

Tutorials developed

Testing on conferences

Survey before and after the workshop

Qualtrics

R + packages

Open data: Github / Zenodo

Website

# Results

## Tutorials

## Before tutorial

## After tutorial

## Final tutorials

## Dissemination of tutorials

# Conclusion

Lessons learned

Future aspects

### 2024-02-02 Analyse_post_workshop_survey.R

@@ -1,4 +1,7 @@
-library(tidyverse)
+library(ggplot2)
+library(dplyr)
+library(lubridate)
+library(forcats)
 
 post_surveys <- read.csv("data/ABM_post_workshop_survey.csv")
 
@@ -17,6 +20,9 @@ post_surveys$Workshop[date(post_surveys$Start.Date)>="2024-01-22"]  <- "202401_A
 post_surveys$How.would.you.rate.the.teaching.material.[post_surveys$How.would.you.rate.the.teaching.material.=="Sufficent"] <- "Sufficient"
 post_surveys$How.would.you.rate.the.workshop.in.general.[post_surveys$How.would.you.rate.the.workshop.in.general.=="Sufficent"] <- "Sufficient"
 post_surveys$How.would.you.rate.the.trainers.teachers.[post_surveys$How.would.you.rate.the.trainers.teachers.=="Sufficent"] <- "Sufficient"
+
+drop_privacy <- c("IP.Address","Recipient.Last.Name","Recipient.First.Name","Recipient.Email","External.Data.Reference", "Location.Latitude", "Location.Longitude","User.Language")
+post_surveys <- post_surveys[,!(names(post_surveys) %in% drop_privacy)]
 
 write.csv(post_surveys, "data/post_workshop_surveys.csv")
 

### 2024-02-02 Analyse_pre_workshop_survey.R

@@ -1,4 +1,7 @@
-library(tidyverse)
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+library(forcats)
 
 CAA_202304 <- read.csv("data/CAA_202304.csv")
 CAA_202304$Workshop <- "202304_CAA"
@@ -6,6 +9,8 @@ EAA_202308 <- read.csv("data/EAA_202308.csv")
 EAA_202308$Workshop <- "202308_EAA"
 CAADENLFl_202310 <- read.csv("data/CAADENLFl_202310.csv")
 CAADENLFl_202310$Workshop <- "202310_CAA-DE-NLFl"
+Reuvensdagen_202311 <- read.csv("data/Reuvensdagen_202311.csv")
+Reuvensdagen_202311$Workshop <- "202311_Reuvensdagen"
 CAAUK_202311 <- read.csv("data/CAAUK_202311.csv")
 CAAUK_202311$Workshop <- "202311_CAA-UK"
 Leiden_202312 <- read.csv("data/Leiden_202312.csv")
@@ -13,7 +18,7 @@ Leiden_202312$Workshop <- "202312_Leiden"
 Aarhus_202401 <- read.csv("data/Aarhus_online_202401.csv")
 Aarhus_202401$Workshop <- "202401_Aarhus"
 
-pre_surveys <-rbind(CAA_202304, EAA_202308, CAADENLFl_202310, CAAUK_202311, Leiden_202312, Aarhus_202401)
+pre_surveys <-rbind(CAA_202304, EAA_202308, CAADENLFl_202310, Reuvensdagen_202311, CAAUK_202311, Leiden_202312, Aarhus_202401)
 
 write.csv(pre_surveys, "data/pre_workshop_surveys.csv")
 
@@ -26,7 +31,7 @@ ggsave("export/preworkshop/ABM_known_pie.png")
 
 ggplot(pre_surveys, aes(x = `Do.you.know.what.Agent.Based.Modeling.is.`)) +
   geom_bar() + xlab("I know what ABM is") + facet_grid(~Workshop)
-ggsave("export/preworkshop/ABM_known_facet.png", width = 10)
+ggsave("export/preworkshop/ABM_known_facet.png", width = 12)
 
 # Theory on ABM
 
@@ -49,7 +54,7 @@ pre_surveys %>%
   ggplot(aes(x = `How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`)) +
[…]

### 2024-02-27 paper.Rmd
update paper

@@
 title: "Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists"
-author: "Ronald Visser"
-date: "`r Sys.Date()`"
-output: pdf_document
+author:
+  - name: Ronald M. Visser \orcidlink{0000-0001-6966-1729}
+    affil-id: 1
+  - name: Annemarie Jutte \orcidlink{0000-0000-0000-0000}
+    affil-id: 1
+  - name: Doug Rocks-Macqueen \orcidlink{0000-0000-0000-0000}
+    affil-id: 2
+  - name: Iza Romanowska \orcidlink{0000-0000-0000-0000}
+    affil-id: 3
+  - name: Tom Brughmans \orcidlink{0000-0000-0000-0000}
+    affil-id: 3
+  - name: Laura van der Knaap \orcidlink{0000-0000-0000-0000}
+    affil-id: 4
+  - name: Karsten Lambers \orcidlink{0000-0000-0000-0000}
+    affil-id: 4
+  - name: Kenneth Aitchison \orcidlink{0000-0000-0000-0000}
+    affil-id: 2
+affiliations:
+  - id: 1
+    name: Saxion University of Applied Sciences,
+  - id: 2
+    name: Landward Research Teoranta.
+  - id: 3
+    name: Aarhus University.
+  - id: 4
+    name: Leiden University.
+date: "`r format(Sys.time(), '%d %B %Y')`"
+geometry:
+  - margin=1in
+output:
+  pdf_document:
+    number_sections: yes
+    #template: default.tex
+bibliography: references.bib
+csl: journal-of-computer-applications-in-archaeology.csl
 ---
@@
 knitr::opts_chunk$set(echo = FALSE)
+library(ggplot2)
 ```
@@
 
-Tutorials developed
+During the project we have developed tutorials.
+The tutorials were first designed using storyboards which were later converted into tutorials using NetLOGO Web (<https://www.netlogoweb.org/launch>) and Javascript ().
+The tutorials were developed together and we worked together using GitHub (<https://github.com/lljvdk/EU-ABMA/>).
+During the project initial testing was done with all authors and issues in GitHub were used to tackle bugs.
 
-Testing on conferences
+The tutorial were tested during various online and in-person conferences.
+In addition, we also had students working with the tutorials at Saxion Universty of applied sciences and Leiden University.
+We used the feedback of the participants to improve the tutorials.
+The feedback was obtained structurally using two surveys using Qualtrics.
+The participants of the events were asked to answer questions of a survey before participating.
+We used the answers to understand what the background of the participants was and to get an estimate of their level of knowledge in relation to ABM (see appendix ...
+for the questions).
+After the participants worked with the tutorials during the event they were asked to answer a second survey.
+Questions of this survey were aimed at the measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix ...
+for the questions).
+For some events participants had to register beforehand and we were able to send the pre-workshop survey by email.
+At other events no registration was possible or necessary and the pre-workshop survey was given at the start of the event.
+The post-workshop survey was distributed using QR-codes or links at the end of the workshop during the event.
+After each event we have improved the tutorials based on the received feedback.
 
-Survey before and after the workshop
+| Conference or event          | Period                  | Online, in person or hybrid  | Number of participants | Number of registrations |
+|---------------|---------------|---------------|---------------|---------------|
+| CAA Amsterdam                | April 2023              | In person                    |                        |                         |
+| EAA Belfast                  | August 2023             | In person                    |                        | No registration         |
+| CAA-DE/NL-Fl Online workshop | October 2023            | Online                       |                        |                         |
+| Reuvensdagen                 | November 2023           | In person                    |                        | No registration         |
+| CAA-UK                       | November 2023           | Hybrid                       |                        |                         |
+| Leiden (course)              | December 2023           | In person                    |                        |                         |
+| Aarhus                       | January/February 2024   | Online                       | 178                    | 400                     |
+| Saxion                       | March 2023-January 2024 | In person                    | 18                     |                         |
 
-Qualtrics
+: Conferences, events and situations were the workshops given and the tutorials were tested.
 
-R + packages
-
-Open data: Github / Zenodo
-
-Website
+The surveys were analysed using R [@rcoreteam2023].
+The following packages were used for analyses: ggplot2 [@wickham2016], dplyr [@wickham2023a], tidyr [@wickham2023], forcats[@wickham2023b] , lubridate [@grolemund2011] and stringr [@wickham2023c].
+The date and the code is available at <https://github.com/RonaldVisser/ABM_tutorials> and (...
+insert Zenodo reference...)
 
@@
 
+The participants of the workshop came from many different countries all over the world.
+
+```{r}
+#pre_surveys <- readRDS("data/pre_workshop_surveys.rds")
+#ggplot(pre_surveys, aes(x = `What.is.you.nationality....Nationality...Text`)) +
+#  geom_bar() +  coord_flip() + xlab("Nationality")
+
+```
+
+```{r}
+#ggplot(pre_surveys, aes(x = `What.is.your.age.`, fill = `What.is.your.gender....Selected.Choice`)) +
+#  geom_bar() + xlab("Age") + facet_grid(~Workshop) +
+#  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+
+```
+
 ## After tutorial
@@
 
+Website
+
 # Conclusion
@@
 Future aspects
+
+# Acknowledgements
+
+The roles of this paper according to CReDiT (Contributor Roles Taxonomy):
+
+Kenneth Aitchison
+
+Tom Brughmans
+
+Annemarie Jutte
+
+Laura van der Knaap
+
+Karsten Lambers
+
+Doug Rocks-Macqueen (Software)
+
+Iza Romanowska
+
+Ronald M. Visser (Writing -- original draft, Formal Analysis, Visualization)
+
+# References

### 2024-02-27 Analyse_post_workshop_survey.R

@@ -1,3 +1,4 @@
+### analyses of the surveys after the workshops
 library(ggplot2)
 library(dplyr)
 library(lubridate)

### 2024-02-27 Analyse_pre_workshop_survey.R

@@ -2,6 +2,7 @@ library(ggplot2)
 library(dplyr)
 library(tidyr)
 library(forcats)
+library(stringr)
 
 CAA_202304 <- read.csv("data/CAA_202304.csv")
 CAA_202304$Workshop <- "202304_CAA"
@@ -20,7 +21,38 @@ Aarhus_202401$Workshop <- "202401_Aarhus"
 
 pre_surveys <-rbind(CAA_202304, EAA_202308, CAADENLFl_202310, Reuvensdagen_202311, CAAUK_202311, Leiden_202312, Aarhus_202401)
 
+# correct different spellings of nationalities
+replace_nationality <- c("USA" = "American", "United States of America" = "American", "American " = "American",
+                         "UK (Scotland)" = "British", "UK" = "British", "British / Danish" = "British",
+                         "British/Canadian" = "British", "British/Canadian " = "British", "Brazil" = "Brazilian",
+                         "Spain" = "Spanish", "Polish " = "Polish", "nl" = "Dutch",
+                         "NL" = "Dutch", "Nl" = "Dutch", "Netherlands" = "Dutch",
+                         "Nederlands" = "Dutch", "Dutch " = "Dutch", "Ditch" = "Dutch",
+                         "Dutch and Israeli" = "Dutch", "israel" = "Israeli", "italian" = "Italian",
+                         "Italy" = "Italian", "Hungary" = "Hungarian", "Hungarian " = "Hungarian",
+                         "Hungary " = "Hungarian", "greek" = "Greek", "Germany" = "German",
+                         "german" = "German","China" = "Chinese", "Belgium" = "Belgian",
+                         "Belgium " = "Belgian", "Belg" = "Belgian", "czech" = "Czech",
+                         "cz" = "Czech", "Argentinian/Venezuelan/Italian" = "Argentinian", "Austrian" = "Austrian",
+                         "New Zealand" = "New Zealander", "INDIAN" = "Indian", "CZ" = "Czech",
+                         "Belgianian" = "Belgian", "Albanian " = "Albanian", "British(Scotland)" = "British",
+                         "Brazilianian" = "Brazilian", "Dutchand Israeli" = "Dutch", "Macedonian " = "Macedonian",
+                         "New Zealanderer" = "New Zealander", "British " = "British", "Egyptian " = "Egyptian",
[…]

### 2024-02-28 paper.Rmd
update paper

@@
 author:
-  - name: Ronald M. Visser \orcidlink{0000-0001-6966-1729}
+  - name: Ronald M. Visser
+    orcid: 0000-0001-6966-1729
     affil-id: 1
-  - name: Annemarie Jutte \orcidlink{0000-0000-0000-0000}
+  - name: Annemarie Jutte
+    orcid: 0000-0000-0000-0000
     affil-id: 1
-  - name: Doug Rocks-Macqueen \orcidlink{0000-0000-0000-0000}
+  - name: Doug Rocks-Macqueen 
+    orcid: 0000-0000-0000-0000
     affil-id: 2
-  - name: Iza Romanowska \orcidlink{0000-0000-0000-0000}
+  - name: Iza Romanowska 
+    orcid: 0000-0000-0000-0000
     affil-id: 3
-  - name: Tom Brughmans \orcidlink{0000-0000-0000-0000}
+  - name: Tom Brughmans
+    orcid: 0000-0000-0000-0000
     affil-id: 3
-  - name: Laura van der Knaap \orcidlink{0000-0000-0000-0000}
+  - name: Laura van der Knaap 
+    orcid: 0000-0000-0000-0000
     affil-id: 4
-  - name: Karsten Lambers \orcidlink{0000-0000-0000-0000}
+  - name: Karsten Lambers 
+    orcid: 0000-0000-0000-0000
     affil-id: 4
-  - name: Kenneth Aitchison \orcidlink{0000-0000-0000-0000}
+  - name: Kenneth Aitchison 
+    orcid: 0000-0000-0000-0000
     affil-id: 2
@@
   - id: 1
-    name: Saxion University of Applied Sciences,
+    name: Saxion University of Applied Sciences.
   - id: 2
@@
     number_sections: yes
-    #template: default.tex
+    keep_tex: true
+    fig_caption: yes
+    latex_engine: pdflatex
+  html_document:
+    number_sections: yes
+  word_document:
+    number_sections: yes
 bibliography: references.bib
 csl: journal-of-computer-applications-in-archaeology.csl
+knit: (function(inputFile, encoding) {
+  rmarkdown::render(inputFile, encoding = encoding,
+  output_format = "all") })
 ---
@@
 library(ggplot2)
+library(dplyr)
+library(tidyr)
+library(forcats)
 ```
@@
 
-The participants of the workshop came from many different countries all over the world.
+A large proportion of the participants of the workshops gave us information using the survey before the workshops 172 of (...total number...) participants.
+The respondents came from many different countries all over the world (see Figure \@ref(fig:nationality).
+The number of female respondents slightly outnumber the male ones and a small group did not share their gender and two were non-binary \@ref(fig:gender-age).
+The ages of the respondents ranged from below 20 to over 70, reaching people in different age classes, although the majority was aged between 20 and 40.
 
-```{r}
-#pre_surveys <- readRDS("data/pre_workshop_surveys.rds")
-#ggplot(pre_surveys, aes(x = `What.is.you.nationality....Nationality...Text`)) +
-#  geom_bar() +  coord_flip() + xlab("Nationality")
+```{r nationality, fig.cap="The nationality of the participants that filled in the survey.", echo=FALSE}
+pre_surveys <- readRDS("data/pre_workshop_surveys.rds")
+ggplot(pre_surveys, aes(x = `What.is.you.nationality....Nationality...Text`)) +
+  geom_bar() +  coord_flip() + xlab("Nationality")
 
@@
 
-```{r}
-#ggplot(pre_surveys, aes(x = `What.is.your.age.`, fill = `What.is.your.gender....Selected.Choice`)) +
-#  geom_bar() + xlab("Age") + facet_grid(~Workshop) +
-#  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+```{r gender-age, fig.cap="The gender and age of the respondents for each workshop", echo=FALSE}
+ggplot(pre_surveys, aes(x = `What.is.your.age.`, fill = `What.is.your.gender....Selected.Choice`)) +
+  geom_bar() + labs(x= "Age", y="Count", fill = "Gender") + facet_grid(~Workshop) +   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+
+```
+
+Most of the respondents had several skills with computer applications in archaeology (see Figure @ref(fig:computer-skills)), and many of them did have some knowledge of ABM, although a large group was completely new to the subject.
+The majority had never applied ABM before participating in the workshops with 14 having applied ABM.
+It is interesting to note that many respondents did know what kind of software was available for ABM (see Figure \@ref(fig:abm-knowledge)).
+
+```{r, computer-skills, fig.cap="The computer skills of the respondents."}
+pre_surveys %>%
+  select(`Which.computer.skills.do.you.have....Selected.Choice`) %>%
+  separate_rows(`Which.computer.skills.do.you.have....Selected.Choice`, sep = ',\\s*') %>%
+  filter(`Which.computer.skills.do.you.have....Selected.Choice`!="") %>%
+  group_by(`Which.computer.skills.do.you.have....Selected.Choice`) %>%
+  ggplot(aes(x=`Which.computer.skills.do.you.have....Selected.Choice`)) + geom_bar() +
+    xlab("Computer skills") + coord_flip()
+
+```
+
+```{r, abm-knowledge, fig.cap="Respondents knowledge and experience with ABM.", fig.show="hold", out.width="50%"}
+pre_surveys %>% 
+  filter(`Do.you.know.what.Agent.Based.Modeling.is.`!="Not shared") %>% 
+  ggplot(aes(x = `Do.you.know.what.Agent.Based.Modeling.is.`)) +
+    geom_bar(na.rm = TRUE) + xlab("I know what ABM is") + facet_grid(~Workshop)
+pre_surveys %>%
+  select(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`) %>%
+  filter(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`!="") %>%
+  separate_rows(Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice, sep = ',\\s*') %>%
+  group_by(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`) %>%
+  ggplot(aes(x=`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`)) + geom_bar() +
+  xlab("ABM Software known") + coord_flip()
 

### 2024-02-28 paper.tex
update paper

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Lessons learned: creating tutorials teaching Agent-Based
Modelling for Archaeologists}
\author{true \and true \and true \and true \and true \and true \and true \and true}
\date{28 February 2024}

\begin{document}
\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

ABM application increased in Archaeology

No formal training

EU project

\hypertarget{method}{%
\section{Method}\label{method}}

During the project we have developed tutorials. The tutorials were first
designed using storyboards which were later converted into tutorials
using NetLOGO Web (\url{https://www.netlogoweb.org/launch}) and
Javascript (). The tutorials were developed together and we worked
together using GitHub (\url{https://github.com/lljvdk/EU-ABMA/}). During
the project initial testing was done with all authors and issues in
GitHub were used to tackle bugs.

The tutorial were tested during various online and in-person
conferences. In addition, we also had students working with the
tutorials at Saxion Universty of applied sciences and Leiden University.
We used the feedback of the participants to improve the tutorials. The
feedback was obtained structurally using two surveys using Qualtrics.
The participants of the events were asked to answer questions of a
survey before participating. We used the answers to understand what the
background of the participants was and to get an estimate of their level
of knowledge in relation to ABM (see appendix \ldots{} for the
questions). After the participants worked with the tutorials during the
event they were asked to answer a second survey. Questions of this
survey were aimed at the measuring the effectiveness of the tutorials
and getting feedback on the workshop and tutorials (see appendix
\ldots{} for the questions). For some events participants had to
register beforehand and we were able to send the pre-workshop survey by
email. At other events no registration was possible or necessary and the
pre-workshop survey was given at the start of the event. The
post-workshop survey was distributed using QR-codes or links at the end
of the workshop during the event. After each event we have improved the
tutorials based on the received feedback.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}
\caption{Conferences, events and situations were the workshops given and
the tutorials were tested.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Conference or event
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Period
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Online, in person or hybrid
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Number of participants
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Number of registrations
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Conference or event
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Period
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Online, in person or hybrid
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Number of participants
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Number of registrations
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
CAA Amsterdam & April 2023 & In person & & \\
EAA Belfast & August 2023 & In person & & No registration \\
CAA-DE/NL-Fl Online workshop & October 2023 & Online & & \\
Reuvensdagen & November 2023 & In person & & No registration \\
CAA-UK & November 2023 & Hybrid & & \\
Leiden (course) & December 2023 & In person & & \\
Aarhus & January/February 2024 & Online & 178 & 400 \\
Saxion & March 2023-January 2024 & In person & 18 & \\
\end{longtable}

The surveys were analysed using R (R Core Team 2023). The following
packages were used for analyses: ggplot2 (Wickham 2016), dplyr (Wickham
et al. 2023), tidyr (Wickham, Vaughan and Girlich 2023), forcats(Wickham
2023a) , lubridate (Grolemund and Wickham 2011) and stringr (Wickham
2023b). The date and the code is available at
\url{https://github.com/RonaldVisser/ABM_tutorials} and (\ldots{} insert
Zenodo reference\ldots)

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{tutorials}{%
\subsection{Tutorials}\label{tutorials}}

\hypertarget{before-tutorial}{%
\subsection{Before tutorial}\label{before-tutorial}}

A large proportion of the participants of the workshops gave us
information using the survey before the workshops 172 of (\ldots total
number\ldots) participants. The respondents came from many different
countries all over the world (see Figure @ref(fig:nationality). The
number of female respondents slightly outnumber the male ones and a
small group did not share their gender and two were non-binary
@ref(fig:gender-age). The ages of the respondents ranged from below 20
to over 70, reaching people in different age classes, although the
majority was aged between 20 and 40.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/nationality-1.pdf}
\caption{The nationality of the participants that filled in the survey.}
\end{figure}

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/gender-age-1.pdf}
\caption{The gender and age of the respondents for each workshop}
\end{figure}

Most of the respondents had several skills with computer applications in
archaeology (see Figure (\textbf{ref?})(fig:computer-skills)), and many
of them did have some knowledge of ABM, although a large group was
completely new to the subject. The majority had never applied ABM before
participating in the workshops with 14 having applied ABM. It is
interesting to note that many respondents did know what kind of software
was available for ABM (see Figure @ref(fig:abm-knowledge)).

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/computer-skills-1.pdf}
\caption{The computer skills of the respondents.}
\end{figure}

\begin{figure}
\includegraphics[width=0.5\linewidth]{paper_files/figure-latex/abm-knowledge-1} \includegraphics[width=0.5\linewidth]{paper_files/figure-latex/abm-knowledge-2} \caption{Respondents knowledge and experience with ABM.}\label{fig:abm-knowledge}
\end{figure}

\hypertarget{after-tutorial}{%
\subsection{After tutorial}\label{after-tutorial}}

\hypertarget{final-tutorials}{%
\subsection{Final tutorials}\label{final-tutorials}}

\hypertarget{dissemination-of-tutorials}{%
\subsection{Dissemination of
tutorials}\label{dissemination-of-tutorials}}

Website

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

Lessons learned

Future aspects

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

The roles of this paper according to CReDiT (Contributor Roles
Taxonomy):

Kenneth Aitchison

Tom Brughmans

Annemarie Jutte

Laura van der Knaap

Karsten Lambers

Doug Rocks-Macqueen (Software)

Iza Romanowska

Ronald M. Visser (Writing -- original draft, Formal Analysis,
Visualization)

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-grolemund2011}{}}%
Grolemund, G and Wickham, H. 2011
\href{https://www.jstatsoft.org/v40/i03/}{Dates and times made easy with
lubridate}. \emph{Journal of Statistical Software} 40(3): 125.

\leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2023}{}}%
R Core Team. 2023 \emph{\href{https://www.R-project.org/}{R: A language
and environment for statistical computing}}.

\leavevmode\vadjust pre{\hypertarget{ref-wickham2016}{}}%
Wickham, H. 2016. \emph{\href{http://ggplot2.org}{ggplot2: Elegant
graphics for data analysis}}. Springer-Verlag New York.

\leavevmode\vadjust pre{\hypertarget{ref-wickham2023b}{}}%
Wickham, H. 2023a \emph{\href{https://forcats.tidyverse.org/}{Forcats:
Tools for working with categorical variables (factors)}}.

\leavevmode\vadjust pre{\hypertarget{ref-wickham2023c}{}}%
Wickham, H. 2023b \emph{\href{https://stringr.tidyverse.org}{Stringr:
Simple, consistent wrappers for common string operations}}.

\leavevmode\vadjust pre{\hypertarget{ref-wickham2023a}{}}%
Wickham, H, François, R, Henry, L, Müller, K and Vaughan, D. 2023
\emph{\href{https://dplyr.tidyverse.org}{Dplyr: A grammar of data
manipulation}}.

\leavevmode\vadjust pre{\hypertarget{ref-wickham2023}{}}%
Wickham, H, Vaughan, D and Girlich, M. 2023
\emph{\href{https://tidyr.tidyverse.org}{Tidyr: Tidy messy data}}.

\end{CSLReferences}

\end{document}

### 2024-02-28 Analyse_pre_workshop_survey.R

@@ -50,9 +50,15 @@ pre_surveys$What.is.your.age.[pre_surveys$What.is.your.age.=="Over 71 years"] <-
 # gender fill blanks
 pre_surveys$What.is.your.gender....Selected.Choice[pre_surveys$What.is.your.gender....Selected.Choice==""] <- "Not shared"
 
+pre_surveys$Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice <- gsub("Other, please describe:", "Other", pre_surveys$Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice)
+pre_surveys$Have.you.ever.applied.Agent.Based.Modelling.[pre_surveys$Have.you.ever.applied.Agent.Based.Modelling.==""] <- NA
+pre_surveys$Which.computer.skills.do.you.have....Selected.Choice <- gsub("Other, please describe:", "Other", pre_surveys$Which.computer.skills.do.you.have....Selected.Choice)
+pre_surveys$Which.of.the.option.below.applies.to.you.[pre_surveys$Which.of.the.option.below.applies.to.you.==""] <- "Not Shared"
+pre_surveys$Do.you.know.what.Agent.Based.Modeling.is.[pre_surveys$Do.you.know.what.Agent.Based.Modeling.is.==""] <- "Not shared"
+
 write.csv(pre_surveys, "data/pre_workshop_surveys.csv")
 saveRDS(pre_surveys, "data/pre_workshop_surveys.rds")
-readRDS("data/pre_workshop_surveys.rds")
+pre_surveys <- readRDS("data/pre_workshop_surveys.rds")
 
 # do you know ABM
 
@@ -61,8 +67,10 @@ ggplot(pre_surveys, aes(x = 1, fill = `Do.you.know.what.Agent.Based.Modeling.is.
   scale_fill_discrete(name = paste0("I know what ABM is (n=", sum(!is.na(pre_surveys$Do.you.know.what.Agent.Based.Modeling.is.)), ")"))
 ggsave("export/preworkshop/ABM_known_pie.png")
 
-ggplot(pre_surveys, aes(x = `Do.you.know.what.Agent.Based.Modeling.is.`)) +
-  geom_bar() + xlab("I know what ABM is") + facet_grid(~Workshop)
+pre_surveys %>%
+  filter(`Do.you.know.what.Agent.Based.Modeling.is.`!="Not shared") %>%
+  ggplot(aes(x = `Do.you.know.what.Agent.Based.Modeling.is.`)) +
[…]

### 2024-02-29 paper.Rmd
paper update

@@
   - name: Annemarie Jutte
-    orcid: 0000-0000-0000-0000
+    orcid: 
     affil-id: 1
   - name: Doug Rocks-Macqueen 
-    orcid: 0000-0000-0000-0000
+    orcid: 
     affil-id: 2
   - name: Iza Romanowska 
-    orcid: 0000-0000-0000-0000
+    orcid: 
     affil-id: 3
   - name: Tom Brughmans
-    orcid: 0000-0000-0000-0000
+    orcid: 
     affil-id: 3
   - name: Laura van der Knaap 
-    orcid: 0000-0000-0000-0000
+    orcid: 
     affil-id: 4
   - name: Karsten Lambers 
-    orcid: 0000-0000-0000-0000
+    orcid: 
     affil-id: 4
   - name: Kenneth Aitchison 
-    orcid: 0000-0000-0000-0000
+    orcid: 
     affil-id: 2
@@
 output:
-  pdf_document:
+  bookdown::pdf_document2:
     number_sections: yes
@@
     fig_caption: yes
-    latex_engine: pdflatex
   html_document:
     number_sections: yes
+    fig_caption: yes
   word_document:
@@
 The surveys were analysed using R [@rcoreteam2023].
-The following packages were used for analyses: ggplot2 [@wickham2016], dplyr [@wickham2023a], tidyr [@wickham2023], forcats[@wickham2023b] , lubridate [@grolemund2011] and stringr [@wickham2023c].
+The following packages were used for analyses: ggplot2 [@wickham2016], dplyr [@wickham2023a], tidyr [@wickham2023], forcats[@wickham2023b], lubridate [@grolemund2011] and stringr [@wickham2023c].
 The date and the code is available at <https://github.com/RonaldVisser/ABM_tutorials> and (...
@@
 The respondents came from many different countries all over the world (see Figure \@ref(fig:nationality).
-The number of female respondents slightly outnumber the male ones and a small group did not share their gender and two were non-binary \@ref(fig:gender-age).
+The number of female respondents slightly outnumber the male ones and a small group did not share their gender and two were non-binary (see Figure \@ref(fig:gender-age)).
 The ages of the respondents ranged from below 20 to over 70, reaching people in different age classes, although the majority was aged between 20 and 40.
+It seems that the different events also had a slightly different distribution of both gender and age.
+For example, more older people attended the workshop at the CAA conference in April 2023 and more male people were present during the workshop at the Reuvensdagen in November 2023.
+
+students???
 
@@
 
-Most of the respondents had several skills with computer applications in archaeology (see Figure @ref(fig:computer-skills)), and many of them did have some knowledge of ABM, although a large group was completely new to the subject.
+Most of the respondents had several skills with computer applications in archaeology (see Figure \@ref(fig:computer-skills)), and many of them did have some knowledge of ABM, although a large group was completely new to the subject.
 The majority had never applied ABM before participating in the workshops with 14 having applied ABM.
@@
 
-```{r, computer-skills, fig.cap="The computer skills of the respondents."}
+```{r, computer-skills, fig.cap="The computer skills of the respondents.", out.width="50%"}
 pre_surveys %>%
@@
     xlab("Computer skills") + coord_flip()
-
 ```
@@
   xlab("ABM Software known") + coord_flip()
-
 ```

### 2024-02-29 paper.tex
paper update

@@
 
-\title{Lessons learned: creating tutorials teaching Agent-Based
-Modelling for Archaeologists}
+\title{Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists}
 \author{true \and true \and true \and true \and true \and true \and true \and true}
@@
 
+{
+\setcounter{tocdepth}{2}
+\tableofcontents
+}
 \hypertarget{introduction}{%
@@
 
-During the project we have developed tutorials. The tutorials were first
-designed using storyboards which were later converted into tutorials
-using NetLOGO Web (\url{https://www.netlogoweb.org/launch}) and
-Javascript (). The tutorials were developed together and we worked
-together using GitHub (\url{https://github.com/lljvdk/EU-ABMA/}). During
-the project initial testing was done with all authors and issues in
-GitHub were used to tackle bugs.
+During the project we have developed tutorials.
+The tutorials were first designed using storyboards which were later converted into tutorials using NetLOGO Web (\url{https://www.netlogoweb.org/launch}) and Javascript ().
+The tutorials were developed together and we worked together using GitHub (\url{https://github.com/lljvdk/EU-ABMA/}).
+During the project initial testing was done with all authors and issues in GitHub were used to tackle bugs.
 
-The tutorial were tested during various online and in-person
-conferences. In addition, we also had students working with the
-tutorials at Saxion Universty of applied sciences and Leiden University.
-We used the feedback of the participants to improve the tutorials. The
-feedback was obtained structurally using two surveys using Qualtrics.
-The participants of the events were asked to answer questions of a
-survey before participating. We used the answers to understand what the
-background of the participants was and to get an estimate of their level
-of knowledge in relation to ABM (see appendix \ldots{} for the
-questions). After the participants worked with the tutorials during the
-event they were asked to answer a second survey. Questions of this
-survey were aimed at the measuring the effectiveness of the tutorials
-and getting feedback on the workshop and tutorials (see appendix
-\ldots{} for the questions). For some events participants had to
-register beforehand and we were able to send the pre-workshop survey by
-email. At other events no registration was possible or necessary and the
-pre-workshop survey was given at the start of the event. The
-post-workshop survey was distributed using QR-codes or links at the end
-of the workshop during the event. After each event we have improved the
-tutorials based on the received feedback.
+The tutorial were tested during various online and in-person conferences.
+In addition, we also had students working with the tutorials at Saxion Universty of applied sciences and Leiden University.
+We used the feedback of the participants to improve the tutorials.
+The feedback was obtained structurally using two surveys using Qualtrics.
+The participants of the events were asked to answer questions of a survey before participating.
+We used the answers to understand what the background of the participants was and to get an estimate of their level of knowledge in relation to ABM (see appendix \ldots{} for the questions).
+After the participants worked with the tutorials during the event they were asked to answer a second survey.
+Questions of this survey were aimed at the measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix \ldots{} for the questions).
+For some events participants had to register beforehand and we were able to send the pre-workshop survey by email.
+At other events no registration was possible or necessary and the pre-workshop survey was given at the start of the event.
+The post-workshop survey was distributed using QR-codes or links at the end of the workshop during the event.
+After each event we have improved the tutorials based on the received feedback.
 
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}
-\caption{Conferences, events and situations were the workshops given and
-the tutorials were tested.}\tabularnewline
+\caption{Conferences, events and situations were the workshops given and the tutorials were tested.}\tabularnewline
 \toprule\noalign{}
@@
 
-The surveys were analysed using R (R Core Team 2023). The following
-packages were used for analyses: ggplot2 (Wickham 2016), dplyr (Wickham
-et al. 2023), tidyr (Wickham, Vaughan and Girlich 2023), forcats(Wickham
-2023a) , lubridate (Grolemund and Wickham 2011) and stringr (Wickham
-2023b). The date and the code is available at
-\url{https://github.com/RonaldVisser/ABM_tutorials} and (\ldots{} insert
-Zenodo reference\ldots)
+The surveys were analysed using R (R Core Team 2023).
+The following packages were used for analyses: ggplot2 (Wickham 2016), dplyr (Wickham et al.
+2023), tidyr (Wickham, Vaughan and Girlich 2023), forcats(Wickham 2023a), lubridate (Grolemund and Wickham 2011) and stringr (Wickham 2023b).
+The date and the code is available at \url{https://github.com/RonaldVisser/ABM_tutorials} and (\ldots{} insert Zenodo reference\ldots)
 
@@
 
-A large proportion of the participants of the workshops gave us
-information using the survey before the workshops 172 of (\ldots total
-number\ldots) participants. The respondents came from many different
-countries all over the world (see Figure @ref(fig:nationality). The
-number of female respondents slightly outnumber the male ones and a
-small group did not share their gender and two were non-binary
-@ref(fig:gender-age). The ages of the respondents ranged from below 20
-to over 70, reaching people in different age classes, although the
-majority was aged between 20 and 40.
+A large proportion of the participants of the workshops gave us information using the survey before the workshops 172 of (\ldots total number\ldots) participants.
+The respondents came from many different countries all over the world (see Figure \ref{fig:nationality}.
+The number of female respondents slightly outnumber the male ones and a small group did not share their gender and two were non-binary (see Figure \ref{fig:gender-age}).
+The ages of the respondents ranged from below 20 to over 70, reaching people in different age classes, although the majority was aged between 20 and 40.
 
@@
 \includegraphics{paper_files/figure-latex/nationality-1.pdf}
-\caption{The nationality of the participants that filled in the survey.}
+\caption{\label{fig:nationality}The nationality of the participants that filled in the survey.}
 \end{figure}
@@
 \includegraphics{paper_files/figure-latex/gender-age-1.pdf}
-\caption{The gender and age of the respondents for each workshop}
+\caption{\label{fig:gender-age}The gender and age of the respondents for each workshop}
 \end{figure}
 
-Most of the respondents had several skills with computer applications in
-archaeology (see Figure (\textbf{ref?})(fig:computer-skills)), and many
-of them did have some knowledge of ABM, although a large group was
-completely new to the subject. The majority had never applied ABM before
-participating in the workshops with 14 having applied ABM. It is
-interesting to note that many respondents did know what kind of software
-was available for ABM (see Figure @ref(fig:abm-knowledge)).
+Most of the respondents had several skills with computer applications in archaeology (see Figure \ref{fig:computer-skills}), and many of them did have some knowledge of ABM, although a large group was completely new to the subject.
+The majority had never applied ABM before participating in the workshops with 14 having applied ABM.
+It is interesting to note that many respondents did know what kind of software was available for ABM (see Figure \ref{fig:abm-knowledge}).
 
 \begin{figure}
-\centering
-\includegraphics{paper_files/figure-latex/computer-skills-1.pdf}
-\caption{The computer skills of the respondents.}
+\includegraphics[width=0.5\linewidth]{paper_files/figure-latex/computer-skills-1} \caption{The computer skills of the respondents.}\label{fig:computer-skills}
 \end{figure}
@@
 \hypertarget{dissemination-of-tutorials}{%
-\subsection{Dissemination of
-tutorials}\label{dissemination-of-tutorials}}
+\subsection{Dissemination of tutorials}\label{dissemination-of-tutorials}}
 
@@
 
-The roles of this paper according to CReDiT (Contributor Roles
-Taxonomy):
+The roles of this paper according to CReDiT (Contributor Roles Taxonomy):
 
@@
 
-Ronald M. Visser (Writing -- original draft, Formal Analysis,
-Visualization)
+Ronald M. Visser (Writing -- original draft, Formal Analysis, Visualization)
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-grolemund2011}{}}%
-Grolemund, G and Wickham, H. 2011
-\href{https://www.jstatsoft.org/v40/i03/}{Dates and times made easy with
-lubridate}. \emph{Journal of Statistical Software} 40(3): 125.
+Grolemund, G and Wickham, H. 2011 \href{https://www.jstatsoft.org/v40/i03/}{Dates and times made easy with lubridate}. \emph{Journal of Statistical Software} 40(3): 125.
 
 \leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2023}{}}%
-R Core Team. 2023 \emph{\href{https://www.R-project.org/}{R: A language
-and environment for statistical computing}}.
+R Core Team. 2023 \emph{\href{https://www.R-project.org/}{R: A language and environment for statistical computing}}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-wickham2016}{}}%
-Wickham, H. 2016. \emph{\href{http://ggplot2.org}{ggplot2: Elegant
-graphics for data analysis}}. Springer-Verlag New York.
+Wickham, H. 2016. \emph{\href{http://ggplot2.org}{ggplot2: Elegant graphics for data analysis}}. Springer-Verlag New York.
 
 \leavevmode\vadjust pre{\hypertarget{ref-wickham2023b}{}}%
-Wickham, H. 2023a \emph{\href{https://forcats.tidyverse.org/}{Forcats:
-Tools for working with categorical variables (factors)}}.
+Wickham, H. 2023a \emph{\href{https://forcats.tidyverse.org/}{Forcats: Tools for working with categorical variables (factors)}}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-wickham2023c}{}}%
-Wickham, H. 2023b \emph{\href{https://stringr.tidyverse.org}{Stringr:
-Simple, consistent wrappers for common string operations}}.
+Wickham, H. 2023b \emph{\href{https://stringr.tidyverse.org}{Stringr: Simple, consistent wrappers for common string operations}}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-wickham2023a}{}}%
-Wickham, H, François, R, Henry, L, Müller, K and Vaughan, D. 2023
-\emph{\href{https://dplyr.tidyverse.org}{Dplyr: A grammar of data
-manipulation}}.
+Wickham, H, François, R, Henry, L, Müller, K and Vaughan, D. 2023 \emph{\href{https://dplyr.tidyverse.org}{Dplyr: A grammar of data manipulation}}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-wickham2023}{}}%
-Wickham, H, Vaughan, D and Girlich, M. 2023
-\emph{\href{https://tidyr.tidyverse.org}{Tidyr: Tidy messy data}}.
+Wickham, H, Vaughan, D and Girlich, M. 2023 \emph{\href{https://tidyr.tidyverse.org}{Tidyr: Tidy messy data}}.
 

### 2024-02-29 paper.Rmd
Paper update

@@
   - name: Ronald M. Visser
-    orcid: 0000-0001-6966-1729
-    affil-id: 1
+    orcid_id: 0000-0001-6966-1729
+    role: 
+      - Writing - Original Draft Preparation
+      - Formal Analysis
+      - Visualization
+      - Writing - Review & Editing
+    affiliation: Saxion University of Applied Sciences
   - name: Annemarie Jutte
-    orcid: 
-    affil-id: 1
+    orcid_id: 0000
+    role: 
+    - Writing - Original Draft Preparation
+    - Writing - Review & Editing
+    affiliation: Saxion University of Applied Sciences
   - name: Doug Rocks-Macqueen 
-    orcid: 
-    affil-id: 2
+    orcid_id: 0000
+    role: 
+      - Software
+      - Writing - Original Draft Preparation
+      - Writing - Review & Editing
+    affiliation: Landward Research Teoranta
   - name: Iza Romanowska 
-    orcid: 
-    affil-id: 3
+    orcid_id: 00000
+    role: 
+     - Writing - Original Draft Preparation
+     - Writing - Review & Editing
+    affiliation: Aarhus University
   - name: Tom Brughmans
-    orcid: 
-    affil-id: 3
+    role: 
+      - Writing - Original Draft Preparation
+      - Writing - Review & Editing
+    orcid_id: 0000
+    affiliation: Aarhus University
   - name: Laura van der Knaap 
-    orcid: 
-    affil-id: 4
+    role: 
+      - Software
+      - Writing - Original Draft Preparation
+      - Writing - Review & Editing
+    orcid_id: 0000
+    affiliation: Leiden University
   - name: Karsten Lambers 
-    orcid: 
-    affil-id: 4
+    role: 
+      - Writing - Original Draft Preparation
+      - Writing - Review & Editing
+    orcid_id: 0000
+    affiliation: Leiden University
   - name: Kenneth Aitchison 
-    orcid: 
-    affil-id: 2
-affiliations:
-  - id: 1
-    name: Saxion University of Applied Sciences.
-  - id: 2
-    name: Landward Research Teoranta.
-  - id: 3
-    name: Aarhus University.
-  - id: 4
-    name: Leiden University.
-date: "`r format(Sys.time(), '%d %B %Y')`"
-geometry:
-  - margin=1in
+    role: 
+      - Writing - Original Draft Preparation
+      - Writing - Review & Editing
+      - Project administration
+    orcid_id: 0000
+    affiliation: Landward Research Teoranta
+  - name: Adam Pažout
+    role: 
+      - Writing - Original Draft Preparation
+      - Writing - Review & Editing
+    orcid_id: 0000
+    affiliation: Aarhus University
+#date: "`r format(Sys.time(), '%d %B %Y')`"
 output:
+  distill::distill_article:
+    number_sections: yes
+    fig_caption: yes
   bookdown::pdf_document2:
@@
     fig_caption: yes
-  html_document:
+  bookdown::word_document2:
     number_sections: yes
     fig_caption: yes
-  word_document:
-    number_sections: yes
 bibliography: references.bib
 csl: journal-of-computer-applications-in-archaeology.csl
-knit: (function(inputFile, encoding) {
-  rmarkdown::render(inputFile, encoding = encoding,
-  output_format = "all") })
 ---
@@
 library(forcats)
+# read data
+pre_surveys <- readRDS("data/pre_workshop_surveys.rds")
+post_surveys <- readRDS("data/post_workshop_surveys.rds")
+# factore levels for charts
+factor_levels <- c("Excellent", "Good", "Sufficient", "Insufficient", "Not shared")
+
 ```
@@
 ABM application increased in Archaeology
+
+Other teaching material available SPOC [@scherjon2019]
+
+New handbook [@romanowska2021]
 
@@
 The tutorial were tested during various online and in-person conferences.
-In addition, we also had students working with the tutorials at Saxion Universty of applied sciences and Leiden University.
+The in-person conferences were attended by several people from the project team.
+During the online events we were always present with several people and enabled participants to go to break-out rooms if they needed any help or wanted to discuss things.
+In addition, we also had students working with the tutorials at Saxion University of applied sciences and Leiden University.
 We used the feedback of the participants to improve the tutorials.
@@
 |---------------|---------------|---------------|---------------|---------------|
-| CAA Amsterdam                | April 2023              | In person                   |                        |                         |
-| EAA Belfast                  | August 2023             | In person                   |                        | No registration         |
-| CAA-DE/NL-Fl Online workshop | October 2023            | Online                      |                        |                         |
-| Reuvensdagen                 | November 2023           | In person                   |                        | No registration         |
-| CAA-UK                       | November 2023           | Hybrid                      |                        |                         |
-| Leiden (course)              | December 2023           | In person                   |                        |                         |
-| Aarhus                       | January/February 2024   | Online                      | 178                    | 400                     |
-| Saxion                       | March 2023-January 2024 | In person                   | 18                     |                         |
+| CAA Amsterdam                | April 2023              | In person                   | 36                     | 53                      |
+| EAA Belfast                  | August 2023             | In person                   | 21                     | No registration         |
+| CAA-DE/NL-Fl Online workshop | October 2023            | Online                      | 40                     | 76                      |
+| Reuvensdagen                 | November 2023           | In person                   | 13                     | No registration         |
+| CAA-UK                       | November 2023           | Hybrid                      | 53                     |                         |
+| Leiden (course)              | December 2023           | In person                   | 11                     | No registration         |
+| Aarhus                       | January/February 2024   | Online                      | 176                    | \>500                   |
+| Saxion                       | March 2023-January 2024 | In person                   | 18                     | No registration         |
+| **Total**                    |                         |                             | **368**                | **\>628**               |
 
@@
 
+Over the course of the project three groups of students worked on the project during the Smart Solutions Semester at Saxion University of Applied Sciences.
+This is an interdisciplinary semester in which students of at least three different study-programmes or disciplines work together on a complex problem/project (<https://www.saxion.edu/business-and-research/collaborate-with-saxion/smart-solutions>).
+The backgrounds of the various students were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media & Game Technologies, and ICT.
+The diverse groups of students contributed with new ideas, developing educational material, testing the tutorials and developing a style for the website and other materials.
+
 # Results
 
-## Tutorials
+## Workhops and tutorials
 
-## Before tutorial
+## Before the workhops
 
-A large proportion of the participants of the workshops gave us information using the survey before the workshops 172 of (...total number...) participants.
+A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants).
 The respondents came from many different countries all over the world (see Figure \@ref(fig:nationality).
@@
 
-students???
-
 ```{r nationality, fig.cap="The nationality of the participants that filled in the survey.", echo=FALSE}
-pre_surveys <- readRDS("data/pre_workshop_surveys.rds")
 ggplot(pre_surveys, aes(x = `What.is.you.nationality....Nationality...Text`)) +
@@
 
-```{r gender-age, fig.cap="The gender and age of the respondents for each workshop", echo=FALSE}
+```{r gender-age, fig.cap="The gender and age of the respondents for each workshop", fig.width=11, echo=FALSE}
 ggplot(pre_surveys, aes(x = `What.is.your.age.`, fill = `What.is.your.gender....Selected.Choice`)) +
@@
 
+As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply this or had never applied this before.
+The respondents were also asked how they rated the available knowledge on ABM (Figure \@ref(fig:available-theory)).
+While the largest group of them had no opinion on the subject, a very large proportion of the responded answered that they rated the available theory on ABM as limited and a smaller group as sufficient.
+A minority rated the available theory on ABM as more than sufficient.
+This clearly shows the need for better educational material.
+
+```{r, available-theory, fig.cap="Respondents optionion on the quality of theory on ABM faceted out by event.", fig.width=11}
+pre_surveys %>%
+  select(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`, Workshop) %>%
+  filter(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`!="") %>%
+  mutate(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.` =
+           fct_relevel(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`,
+                       "No opinion", "Limited", "Sufficient", "More than sufficient", "Very good")) %>%
+  ggplot(aes(x = `How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`)) +
+  geom_bar() + xlab("Quality of theory on ABM") + coord_flip() +
+  facet_grid(~Workshop)
+```
+
 ## After tutorial
+
+A large proportion of the participants of the workshops gave us information using the survey after the workshops (171 of 368 participants), a similar proportion as those responding to the survey before the workshop.
+
+The respondents
+
+```{r, rating-workshop, fig.cap="Respondents rating of the workshop in general faceted for each event.", fig.width=11}
+post_surveys %>%
+  mutate(How.would.you.rate.the.workshop.in.general. = fct_relevel(How.would.you.rate.the.workshop.in.general., factor_levels)) %>%
+  ggplot(aes(x=How.would.you.rate.the.workshop.in.general.)) + geom_bar() +
+  xlab("Rating of the workshop") + facet_grid(~Workshop) +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+
+```
+
+The majority of the respondents were enthusiastic about the teaching material, with the majority rating it as excellent or good.
+The teachers were even rated better than the teaching material, with more respondents rating them as excellent (see Figure \@ref(fig:rating-teaching)).
+
+```{r, rating-teaching, fig.cap="Respondents rating of the teaching material in general faceted for each event.", fig.width=11, fig.show='hold', out.height="50%"}
+post_surveys %>%
+  mutate(How.would.you.rate.the.teaching.material. = fct_relevel(How.would.you.rate.the.teaching.material., factor_levels)) %>%
+  ggplot(aes(x=How.would.you.rate.the.teaching.material.)) + geom_bar() +
+  xlab("Rating of the teaching material") + facet_grid(~Workshop) +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+post_surveys %>%
+  mutate(How.would.you.rate.the.trainers.teachers. = fct_relevel(How.would.you.rate.the.trainers.teachers., factor_levels)) %>%
+  ggplot(aes(x=How.would.you.rate.the.trainers.teachers.)) + geom_bar() +
+  xlab("Rating of the teacher") + facet_grid(~Workshop) +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+
+```
+
+Two more open questions were asked to the respondents. The first was aimed at getting positive feedback and the other was aimed at getting feedback to improve the tutorials.
+
+The respondents generally liked the interactive step-by-step way of going through the tutorials, but also valued the interaction with the teachers very much.
+The easy and intuitive introduction into ABM and NetLOGO were also mentioned quite often.
+This was also observed during the workshops by us, most participants were working in the own pace and did not need much help.
+From a didactic point of view, the self-paced going through the tutorials is very efficient for teachers.
+
+We also received feedback on possible improvements.
+For the first workshops we mostly received feedback on the number of bugs.
+During the CAA-conference in Amsterdam we did the first test of the tutorials and we also instructed the participants that this was work in progress.
+The number of references to bugs reduced to zero by the end and people started providing other feedback, for example the wish to see more examples or expanding the tutorials more.
+Some wanted more cooperation, which is possible, but for some online events harder, although we provided break-out rooms.
+It is also interesting to note that more and more people did not see any room for improvement.
+
+The majority of the respondents that wanted to apply ABM in the future for research (144 of the 164 that answered this question, see Figure \@ref(fig:future-abm)).
+Some of the respondents were already using ABM.
+A large group was not sure yet in how to apply ABM and wanted to read more on the subject or play a bit with the possibilities.
+Many respondents also shared the context in which they were thinking to apply ABM to.
+Various were thinking about movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms.
+Others thought of demography, social networks, migration or settlement distributions patterns.
+The natural environment and the interaction with humans in the past was also mentioned by some, and often in relation to GIS or how to replace GIS with ABM.
+The archaeological periods that the participants were interested in were very diverse ranging from the paleolithic to the medieval period.
+
+```{r, future-abm, fig.cap="The respondents reaction to the question if they thing that they will apply ABM in the future."}
+post_surveys %>%
+  filter(!is.na(Do.you.think.that.you.will.apply.ABM.to.your.own.research..If.yes..can.you.elaborate....Selected.Choice)) %>%
+  ggplot(aes(x = 1, fill = `Do.you.think.that.you.will.apply.ABM.to.your.own.research..If.yes..can.you.elaborate....Selected.Choice`)) +
+  geom_bar(na.rm = TRUE) + coord_polar(theta = "y", start=0) + theme_void() +
+  scale_fill_discrete(name = paste0("Will apply ABM in the future (n=", sum(!is.na(post_surveys$Do.you.think.that.you.will.apply.ABM.to.your.own.research..If.yes..can.you.elaborate....Selected.Choice)), ")"),
+                      labels = c("No", "Yes"))
+```
 
@@
 Website
+
+Github
 
@@
 
-The roles of this paper according to CReDiT (Contributor Roles Taxonomy):
+We would like to thank the student groups at Saxion University of Applied Sciences that worked with us:
 
-Kenneth Aitchison
+-   September 2022-February 2023: Dany Dragoi, Liam van den Bosch, Marko Stojkovic, Max van Duinen, Nora van den Engel, Roan Man, Stefan Oostingh and their tutor: Mark Spanjer.
 
-Tom Brughmans
+-   February 2022-July 2023: Alice Overgaauw, Johan Broersma, Mandy Hazenberg, Paulina Fulneczek, Ties Heesink and their tutor: Jan Willem Huson.
 
-Annemarie Jutte
+-   September 2023-February 2024: Arnfinn Sijbrant, Eva Baan, Jip Mulder, Robert Aalpoel, Sem Lucas, Sterre Regts and their tutor: Jan Willem Huson.
 
-Laura van der Knaap
-
-Karsten Lambers
-
-Doug Rocks-Macqueen (Software)
-
-Iza Romanowska
-
-Ronald M. Visser (Writing -- original draft, Formal Analysis, Visualization)
+We had help from many people during the various workshops and we would like to thank them for their help (in alphabetical order of their first name): Adéla Sobotkova, Alice Overgaauw, Eduardo Herrera Malatesta, Jens Emil Bødstrup Christoffersen, Johan Broersma, Magnus Lindholm Nielsen, Mandy Hazenberg, María Coto Sarmiento, Paulina Fulneczek, Petra Hermankova, Ties Heesink.
+We also want to thank all the participants of our workshops.
 

### 2024-02-29 paper.tex
Paper update

@@
 \title{Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists}
-\author{true \and true \and true \and true \and true \and true \and true \and true}
-\date{28 February 2024}
+\author{true \and true \and true \and true \and true \and true \and true \and true \and true}
+\date{29 February 2024}
 
@@
 Leiden (course) & December 2023 & In person & & \\
-Aarhus & January/February 2024 & Online & 178 & 400 \\
+Aarhus & January/February 2024 & Online & 176 & \textgreater500 \\
 Saxion & March 2023-January 2024 & In person & 18 & \\
@@
 The ages of the respondents ranged from below 20 to over 70, reaching people in different age classes, although the majority was aged between 20 and 40.
+It seems that the different events also had a slightly different distribution of both gender and age.
+For example, more older people attended the workshop at the CAA conference in April 2023 and more male people were present during the workshop at the Reuvensdagen in November 2023.
+
+students???
 

### 2024-02-29 Analyse_post_workshop_survey.R

@@ -4,7 +4,7 @@ library(dplyr)
 library(lubridate)
 library(forcats)
 
-post_surveys <- read.csv("data/ABM_post_workshop_survey.csv")
+post_surveys <- read.csv("data/post_workshop_surveys.csv")
 
 post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-04-03"]  <- "202304_CAA"
 post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-08-31"]  <- "202308_EAA"
@@ -19,15 +19,20 @@ post_surveys$Workshop[date(post_surveys$Start.Date)>="2024-01-22"]  <- "202401_A
 
 #correct typos in the survey
 post_surveys$How.would.you.rate.the.teaching.material.[post_surveys$How.would.you.rate.the.teaching.material.=="Sufficent"] <- "Sufficient"
+post_surveys$How.would.you.rate.the.teaching.material.[post_surveys$How.would.you.rate.the.teaching.material.==""] <- "Not shared"
 post_surveys$How.would.you.rate.the.workshop.in.general.[post_surveys$How.would.you.rate.the.workshop.in.general.=="Sufficent"] <- "Sufficient"
 post_surveys$How.would.you.rate.the.trainers.teachers.[post_surveys$How.would.you.rate.the.trainers.teachers.=="Sufficent"] <- "Sufficient"
+post_surveys$How.would.you.rate.the.trainers.teachers.[post_surveys$How.would.you.rate.the.trainers.teachers.==""] <- "Not shared"
+post_surveys$Do.you.think.that.you.will.apply.ABM.to.your.own.research..If.yes..can.you.elaborate....Selected.Choice[post_surveys$Do.you.think.that.you.will.apply.ABM.to.your.own.research..If.yes..can.you.elaborate....Selected.Choice==""] <- NA
 
 drop_privacy <- c("IP.Address","Recipient.Last.Name","Recipient.First.Name","Recipient.Email","External.Data.Reference", "Location.Latitude", "Location.Longitude","User.Language")
 post_surveys <- post_surveys[,!(names(post_surveys) %in% drop_privacy)]
 
 write.csv(post_surveys, "data/post_workshop_surveys.csv")
+saveRDS(post_surveys, "data/post_workshop_surveys.rds")
+post_surveys <- readRDS("data/post_workshop_surveys.rds")
 
[…]

### 2024-03-01 paper.Rmd
Update paper

@@
 csl: journal-of-computer-applications-in-archaeology.csl
+knit: (function(inputFile, encoding) {
+  rmarkdown::render(inputFile, encoding = encoding,
+  output_format = "all") })
 ---
@@
 
-EU project
+EU project <https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054>
 

### 2024-03-01 paper.tex
Update paper

@@
 \author{true \and true \and true \and true \and true \and true \and true \and true \and true}
-\date{29 February 2024}
+\date{}
 
@@
 
+Other teaching material available SPOC (Scherjon, Romanowska and Lambers 2019)
+
+New handbook (Romanowska, Wren and Crabtree 2021)
+
 No formal training
 
-EU project
+EU project \url{https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054}
 
@@
 \endlastfoot
-CAA Amsterdam & April 2023 & In person & & \\
-EAA Belfast & August 2023 & In person & & No registration \\
-CAA-DE/NL-Fl Online workshop & October 2023 & Online & & \\
-Reuvensdagen & November 2023 & In person & & No registration \\
-CAA-UK & November 2023 & Hybrid & & \\
-Leiden (course) & December 2023 & In person & & \\
+CAA Amsterdam & April 2023 & In person & 36 & 53 \\
+EAA Belfast & August 2023 & In person & 21 & No registration \\
+CAA-DE/NL-Fl Online workshop & October 2023 & Online & 40 & 76 \\
+Reuvensdagen & November 2023 & In person & 13 & No registration \\
+CAA-UK & November 2023 & Hybrid & 53 & \\
+Leiden (course) & December 2023 & In person & 11 & No registration \\
 Aarhus & January/February 2024 & Online & 176 & \textgreater500 \\
-Saxion & March 2023-January 2024 & In person & 18 & \\
+Saxion & March 2023-January 2024 & In person & 18 & No registration \\
+\textbf{Total} & & & \textbf{368} & \textbf{\textgreater628} \\
 \end{longtable}
@@
 
+Over the course of the project three groups of students worked on the project during the Smart Solutions Semester at Saxion University of Applied Sciences.
+This is an interdisciplinary semester in which students of at least three different study-programmes or disciplines work together on a complex problem/project (\url{https://www.saxion.edu/business-and-research/collaborate-with-saxion/smart-solutions}).
+The backgrounds of the various students were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media \& Game Technologies, and ICT.
+The diverse groups of students contributed with new ideas, developing educational material, testing the tutorials and developing a style for the website and other materials.
+
 \hypertarget{results}{%
@@
 
-\hypertarget{tutorials}{%
-\subsection{Tutorials}\label{tutorials}}
+\hypertarget{workhops-and-tutorials}{%
+\subsection{Workhops and tutorials}\label{workhops-and-tutorials}}
 
-\hypertarget{before-tutorial}{%
-\subsection{Before tutorial}\label{before-tutorial}}
+\hypertarget{before-the-workhops}{%
+\subsection{Before the workhops}\label{before-the-workhops}}
 
-A large proportion of the participants of the workshops gave us information using the survey before the workshops 172 of (\ldots total number\ldots) participants.
+A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants).
 The respondents came from many different countries all over the world (see Figure \ref{fig:nationality}.
@@
 For example, more older people attended the workshop at the CAA conference in April 2023 and more male people were present during the workshop at the Reuvensdagen in November 2023.
-
-students???
 
@@
 
+As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply this or had never applied this before.
+The respondents were also asked how they rated the available knowledge on ABM (Figure \ref{fig:available-theory}).
+While the largest group of them had no opinion on the subject, a very large proportion of the responded answered that they rated the available theory on ABM as limited and a smaller group as sufficient.
+A minority rated the available theory on ABM as more than sufficient.
+This clearly shows the need for better educational material.
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/figure-latex/available-theory-1.pdf}
+\caption{\label{fig:available-theory}Respondents optionion on the quality of theory on ABM faceted out by event.}
+\end{figure}
+
 \hypertarget{after-tutorial}{%
 \subsection{After tutorial}\label{after-tutorial}}
+
+A large proportion of the participants of the workshops gave us information using the survey after the workshops (171 of 368 participants), a similar proportion as those responding to the survey before the workshop.
+
+The respondents
+
+\begin{verbatim}
+## Warning: There was 1 warning in `mutate()`.
+## i In argument: `How.would.you.rate.the.workshop.in.general. =
+##   fct_relevel(How.would.you.rate.the.workshop.in.general., factor_levels)`.
+## Caused by warning:
+## ! 1 unknown level in `f`: Not shared
+\end{verbatim}
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/figure-latex/rating-workshop-1.pdf}
+\caption{\label{fig:rating-workshop}Respondents rating of the workshop in general faceted for each event.}
+\end{figure}
+
+The majority of the respondents were enthusiastic about the teaching material, with the majority rating it as excellent or good.
+The teachers were even rated better than the teaching material, with more respondents rating them as excellent (see Figure \ref{fig:rating-teaching}).
+
+\begin{verbatim}
+## Warning: There was 1 warning in `mutate()`.
+## i In argument: `How.would.you.rate.the.teaching.material. =
+##   fct_relevel(How.would.you.rate.the.teaching.material., factor_levels)`.
+## Caused by warning:
+## ! 1 unknown level in `f`: Insufficient
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: There was 1 warning in `mutate()`.
+## i In argument: `How.would.you.rate.the.trainers.teachers. =
+##   fct_relevel(How.would.you.rate.the.trainers.teachers., factor_levels)`.
+## Caused by warning:
+## ! 1 unknown level in `f`: Insufficient
+\end{verbatim}
+
+\begin{figure}
+\includegraphics[height=0.5\textheight]{paper_files/figure-latex/rating-teaching-1} \includegraphics[height=0.5\textheight]{paper_files/figure-latex/rating-teaching-2} \caption{Respondents rating of the teaching material in general faceted for each event.}\label{fig:rating-teaching}
+\end{figure}
+
+Two more open questions were asked to the respondents. The first was aimed at getting positive feedback and the other was aimed at getting feedback to improve the tutorials.
+
+The respondents generally liked the interactive step-by-step way of going through the tutorials, but also valued the interaction with the teachers very much.
+The easy and intuitive introduction into ABM and NetLOGO were also mentioned quite often.
+This was also observed during the workshops by us, most participants were working in the own pace and did not need much help.
+From a didactic point of view, the self-paced going through the tutorials is very efficient for teachers.
+
+We also received feedback on possible improvements.
+For the first workshops we mostly received feedback on the number of bugs.
+During the CAA-conference in Amsterdam we did the first test of the tutorials and we also instructed the participants that this was work in progress.
+The number of references to bugs reduced to zero by the end and people started providing other feedback, for example the wish to see more examples or expanding the tutorials more.
+Some wanted more cooperation, which is possible, but for some online events harder, although we provided break-out rooms.
+It is also interesting to note that more and more people did not see any room for improvement.
+
+The majority of the respondents that wanted to apply ABM in the future for research (144 of the 164 that answered this question, see Figure \ref{fig:future-abm}).
+Some of the respondents were already using ABM.
+A large group was not sure yet in how to apply ABM and wanted to read more on the subject or play a bit with the possibilities.
+Many respondents also shared the context in which they were thinking to apply ABM to.
+Various were thinking about movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms.
+Others thought of demography, social networks, migration or settlement distributions patterns.
+The natural environment and the interaction with humans in the past was also mentioned by some, and often in relation to GIS or how to replace GIS with ABM.
+The archaeological periods that the participants were interested in were very diverse ranging from the paleolithic to the medieval period.
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/figure-latex/future-abm-1.pdf}
+\caption{\label{fig:future-abm}The respondents reaction to the question if they thing that they will apply ABM in the future.}
+\end{figure}
 
@@
 
-The roles of this paper according to CReDiT (Contributor Roles Taxonomy):
+We would like to thank the student groups at Saxion University of Applied Sciences that worked with us:
 
-Kenneth Aitchison
+\begin{itemize}
+\item
+  September 2022-February 2023: Dany Dragoi, Liam van den Bosch, Marko Stojkovic, Max van Duinen, Nora van den Engel, Roan Man, Stefan Oostingh and their tutor: Mark Spanjer.
+\item
+  February 2022-July 2023: Alice Overgaauw, Johan Broersma, Mandy Hazenberg, Paulina Fulneczek, Ties Heesink and their tutor: Jan Willem Huson.
+\item
+  September 2023-February 2024: Arnfinn Sijbrant, Eva Baan, Jip Mulder, Robert Aalpoel, Sem Lucas, Sterre Regts and their tutor: Jan Willem Huson.
+\end{itemize}
 
-Tom Brughmans
-
-Annemarie Jutte
-
-Laura van der Knaap
-
-Karsten Lambers
-
-Doug Rocks-Macqueen (Software)
-
-Iza Romanowska
-
-Ronald M. Visser (Writing -- original draft, Formal Analysis, Visualization)
+We had help from many people during the various workshops and we would like to thank them for their help (in alphabetical order of their first name): Adéla Sobotkova, Alice Overgaauw, Eduardo Herrera Malatesta, Jens Emil Bødstrup Christoffersen, Johan Broersma, Magnus Lindholm Nielsen, Mandy Hazenberg, María Coto Sarmiento, Paulina Fulneczek, Petra Hermankova, Ties Heesink.
+We also want to thank all the participants of our workshops.
 
@@
 R Core Team. 2023 \emph{\href{https://www.R-project.org/}{R: A language and environment for statistical computing}}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-romanowska2021}{}}%
+Romanowska, I, Wren, CD and Crabtree, SA.
+2021.
+\emph{\href{https://www.sfipress.org/books/agent-based-modeling-archaeology}{Agent-based modeling for archaeology: Simulating the complexity of societies}}.
+Santa Fe: Santa Fe Institute Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-scherjon2019}{}}%
+Scherjon, F, Romanowska, I and Lambers, K.
+2019 Digitally Teaching Digital Skills: Lessons Drawn from a Small Private Online Course (SPOC) on {`}Modelling and Simulation in Archaeology{'} at Leiden University.
+\emph{Journal of Computer Applications in Archaeology} 2(1): 7988.
+DOI: https://doi.org/\href{https://doi.org/10.5334/jcaa.26}{10.5334/jcaa.26}.
 

### 2024-03-01 paper.Rmd
minor edits

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
 library(ggplot2)
@@
 
-```{r, computer-skills, fig.cap="The computer skills of the respondents.", out.width="50%"}
+```{r, computer-skills, fig.cap="The computer skills of the respondents.", out.width="50%", echo=FALSE}
 pre_surveys %>%
@@
 
-```{r, abm-knowledge, fig.cap="Respondents knowledge and experience with ABM.", fig.show="hold", out.width="50%"}
+```{r, abm-knowledge, fig.cap="Respondents knowledge and experience with ABM.", fig.show="hold", out.width="50%", echo=FALSE}
 pre_surveys %>% 
@@
 
-```{r, available-theory, fig.cap="Respondents optionion on the quality of theory on ABM faceted out by event.", fig.width=11}
+```{r, available-theory, fig.cap="Respondents optionion on the quality of theory on ABM faceted out by event.", fig.width=11, echo=FALSE}
 pre_surveys %>%
@@
 
-```{r, rating-workshop, fig.cap="Respondents rating of the workshop in general faceted for each event.", fig.width=11}
+```{r, rating-workshop, fig.cap="Respondents rating of the workshop in general faceted for each event.", fig.width=11, echo=FALSE}
 post_surveys %>%
@@
 
-```{r, rating-teaching, fig.cap="Respondents rating of the teaching material in general faceted for each event.", fig.width=11, fig.show='hold', out.height="50%"}
+```{r, rating-teaching, fig.cap="Respondents rating of the teaching material in general faceted for each event.", fig.width=11, fig.show='hold', out.height="50%", echo=FALSE}
 post_surveys %>%
@@
 
-```{r, future-abm, fig.cap="The respondents reaction to the question if they thing that they will apply ABM in the future."}
+```{r, future-abm, fig.cap="The respondents reaction to the question if they thing that they will apply ABM in the future.", out.height="50%", echo=FALSE}
 post_surveys %>%

### 2024-03-01 paper.tex
minor edits

@@
 
-\begin{verbatim}
-## Warning: There was 1 warning in `mutate()`.
-## i In argument: `How.would.you.rate.the.workshop.in.general. =
-##   fct_relevel(How.would.you.rate.the.workshop.in.general., factor_levels)`.
-## Caused by warning:
-## ! 1 unknown level in `f`: Not shared
-\end{verbatim}
-
 \begin{figure}
@@
 The teachers were even rated better than the teaching material, with more respondents rating them as excellent (see Figure \ref{fig:rating-teaching}).
-
-\begin{verbatim}
-## Warning: There was 1 warning in `mutate()`.
-## i In argument: `How.would.you.rate.the.teaching.material. =
-##   fct_relevel(How.would.you.rate.the.teaching.material., factor_levels)`.
-## Caused by warning:
-## ! 1 unknown level in `f`: Insufficient
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: There was 1 warning in `mutate()`.
-## i In argument: `How.would.you.rate.the.trainers.teachers. =
-##   fct_relevel(How.would.you.rate.the.trainers.teachers., factor_levels)`.
-## Caused by warning:
-## ! 1 unknown level in `f`: Insufficient
-\end{verbatim}
 
@@
 \begin{figure}
-\centering
-\includegraphics{paper_files/figure-latex/future-abm-1.pdf}
-\caption{\label{fig:future-abm}The respondents reaction to the question if they thing that they will apply ABM in the future.}
+\includegraphics[height=0.5\textheight]{paper_files/figure-latex/future-abm-1} \caption{The respondents reaction to the question if they thing that they will apply ABM in the future.}\label{fig:future-abm}
 \end{figure}

### 2024-03-01 paper.Rmd
minor edits

@@
 
-We had help from many people during the various workshops and we would like to thank them for their help (in alphabetical order of their first name): Adéla Sobotkova, Alice Overgaauw, Eduardo Herrera Malatesta, Jens Emil Bødstrup Christoffersen, Johan Broersma, Magnus Lindholm Nielsen, Mandy Hazenberg, María Coto Sarmiento, Paulina Fulneczek, Petra Hermankova, Ties Heesink.
+We had help from many people during the various workshops and we would like to thank them for their help (in alphabetical order of their first name): Adéla Sobotkova, Agnes Schneider, Alice Overgaauw, Eduardo Herrera Malatesta, Jens Emil Bødstrup Christoffersen, Johan Broersma, Magnus Lindholm Nielsen, Mandy Hazenberg, María Coto Sarmiento, Paulina Fulneczek, Petra Hermankova, Ties Heesink.
 We also want to thank all the participants of our workshops.

### 2024-03-01 paper.Rmd
minor edits

@@
     affiliation: Aarhus University
-#date: "`r format(Sys.time(), '%d %B %Y')`"
+date: "`r format(Sys.time(), '%d %B %Y')`"
 output:

### 2024-03-01 paper.Rmd
Added digcomp 2.2 and the tutorials to the paper, from the DigComp document

@@
 
+The learning and teaching material for ABM for archaeologists assumes that the users have completed secondary education.
+It is suitable for archaeologists with at least some background in archaeology and preferably after at least the first year of a Bachelor study in Archaeology (preferably above EQF levels 4 or 5).
+In addition, the tutorials are aimed at professionals working in archaeology and some experience with computer.
+We also assume that learners are proficient in English and have at least the Reading B1 level, but Reading B2 is recommended [@councilofeurope2020].
+The tutorials are developed aiming to develop learners in the following competence area's: 1.
+Information and data literacy, 2.
+Communication and collaboration, 3.
+Digital content creation and, 5.
+Problem solving [@europeancommissionjointresearchcentre2022].
+
 The tutorial were tested during various online and in-person conferences.
@@
 
-## Before the workhops
+## Final tutorials
+
+The set of developed tutorials consists of the following tutorials:
+
+-   Tutorial 1: Introduction to ABM
+
+-   Tutorial 2: Beginning with NetLogo
+
+-   Tutorial 3: Expanded ABM skills
+
+-   Tutorial 4: Intermediate ABM
+
+-   Tutorial 5: How to Model
+
+Each tutorials consists of a different number of lessons that guide the learner in a self-pace manner through the lessons.
+
+In the first tutorial the learner will learn what simulations and agent-based models are and how they can help in archaeological research.
+This tutorial consists of 4 lessons.
+The first two lesson introduce the learner into simulation in general and Agent-Based Modelling in specific.
+Various concepts related to ABM are introduced and the first concepts within NetLogo are explained.
+The third lesson aims to explain how ABMs are used in archaeological research.
+The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode.
+This tutorial build a foundation for working in NetLogo and with ABM.
+All activity stays within the intermediate level, although higher levels might be touch upon.
+
+In tutorial 2 the learner will learn the basics of NetLogo by making your first simulation on the Out of Africal dispersal of homo sapiens.
+The leaerner will work with basic NetLogo syntax and learn how to set up a simulation and visualize their outcomes.
+This tutorial consists of 9 lessons.
+This tutorial intends to guide the learner from the intermediate to the advanced level of proficiency.
+The learning curve is relatively gentle.
+This is achieved by learning to work with the NetLogo web interface.
+The primitives are introduced and initialization phase of a simple ABM is gone through.
+The world within NetLogo's world (including dimensions, coordinates, origin) and how to alter it are explained and experience in a hands-on approach.
+The learners learn about simple simulation loops and how to use primitives.
+The advanced level is achieved with the custom procedures and variables.
+In addition, if-statements are introduced and more complex versions thereof.
+Exporting information using plots is also explained.
+
+In the third tutorial, the learner will build a simple trade model, picking up more advanced NetLogo coding along the way, such as loops, lists and reporters.
+The learner will be introduced to some techniques like modular code development and debugging which will become important with this increased coding complexity.
+This tutorial consists of 7 lessons.
+This tutorial intends to guide the learner from the intermediate to the advanced and highly specialised level of proficiency.
+This is achieved by improving development through using modular code, pseudocode and annotating the code.
+In addition, custom agent breeds are introduced.
+Visualization with labels and reporters, plots and monitors are learned.
+The use of loops is further explained.
+Debugging is learned in lesson 7.
+This can be very advanced, since it involves knowledge of the NetLogo-language combined with problem solving of unexpected behaviour.
+
+In tutorial 4 the learner will work with Sugarscape simulations to further expand your ABM and NetLogo skills.
+The learner will learn how to set up more complex interactions between agents and the environment.
+Furthermore, the basics of setting up good experiments and validating models will be explained.
+This tutorial consists of 8 lessons.
+This tutorial intends to guide the learner from the (intermediate and) advanced level to the highly specialised level of proficiency.
+In this tutorial more complex agent-environment interaction is further learned, including learning ways to visualize the environment and how to give more agency to patches.
+Creating toy landscapes are central to lesson 5.
+The learner also learns the basics of setting up experiments in NetLogo and how to use monitors and flexible plots to understand the results.
+The validation of agent-based models is an important concept of designing a good experiment.
+The finale step, is to compare simulation results with the archaeological record.
+The last lesson is aimed at learning syntax to make lists more dynamic and how to refactor code.
+
+In the final and fifth tutorial the learner will learn more about how to actually incorporate agent-based modelling in research.
+This tutorial focusses less on programming in NetLogo and more on the model development process.
+The learner will learn about the different phases in modelling and how to export data from models.
+This tutorial consists of 5 lessons.
+This tutorial touches upon various levels, but mainly between the intermediate and highly specialised level of proficiency.
+This tutorial is approaches more theoretical aspects in a practical environment.
+The learner leans about the model development process starting with the conceptual phase of the model development.
+Good and bad research questions for modelling are explained and the difference between them.
+The learner needs to understand how to pick the right modelling technique.
+The learner should understand the importance of properly conceptualizing a model before starting the technical phase of the model development.
+In this phase parametrisation, designing experiments and the analyses and interpretation of the models are important.
+Learning about the dissemination phase of the model development is aimed at understanding (the importance of) publishing models and replication.
+In addition, learning how to export results from NetLogo using basic export primitives is explained.
+The BehaviorSpace is explained to enable the learners to analyse the output of models.
+
+## Before the workshops
 
@@
 
-## After tutorial
+## During the workshop
+
+During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials.
+Some particpants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general.
+
+## After workshop
 
 A large proportion of the participants of the workshops gave us information using the survey after the workshops (171 of 368 participants), a similar proportion as those responding to the survey before the workshop.
-
-The respondents
 
@@
 
-## Final tutorials
+## The Digital Competence Framework for Citizens (2.2)
+
+The tutorials are aimed at developing learners in the following competence area's: 1.
+Information and data literacy, 2.
+Communication and collaboration, 3.
+Digital content creation and, 5.
+Problem solving [@europeancommissionjointresearchcentre2022].
+In these competence area's various competences are developed towards an advanced or even specialised level.
+
+For competence area 1.
+Information and data literacy: the tutorials mainly address 1.2 Evaluating Data, Information And Digital Content and 1.3 Managing Data, Information And Digital Content.
+The users learn about working with data, being critacal of their digital content and how to model and manage the data in the context of an ABM.
+For tutorial 2 the user is expected to be able to lookup information using the NetLogo(Web) dictionary (<http://ccl.northwestern.edu/netlogo/bind/> or <http://ccl.northwestern.edu/netlogo/docs/dictionary.html>).
+
+For Competence Area 2.
+Communication and collaboration the tutorials mainly address 2.1 Interacting Through Digital Technologies, because the users are constantly interacting with digital technologies, but communications is not relevant for the tutorials.
+However, during the workshops given in the course of the project, participants interacted with each other and the teachers.
+
+Competence Area 3.
+Digital content creation is one of the main competence area's for the ABM-teaching material.
+In the process of learning ABM the users are constantly 3.1 Developing Digital Content and 3.2 Integrating And Re-Elaborating Digital Content.
+The material does not touch upon 3.3 Copyright And Licences, but we assume that users have a general understanding.
+A very important aspect of learning ABM is 3.4 Programming.
+The users are exploring the possibilities and chances of programming an ABM.
+The users are expected to learn both syntax and more general concepts such as modular code development, loops, lists and commenting and documenting code.
+
+Competence Area 5.
+Problem solving is very important and is closely tied to 3.4 Programming, since programming involves a lot of problem solving.
+In addition, the creation of a highly complex ABM consists of problem solving all the time, including debugging.
+All competencies are addressed: 5.1 Solving Technical Problems, is common when writing code, learning about coding, error handling and going from pseudocode to real code.
+The competence 5.2 Identifying Needs And Technological Responses, is a central issue of the course, while 5.3 Creatively Using Digital Technology is an important aspect of learning to think in models and develop models themselves.
+In tutorial 2 various aspects of 5.4 Identifying Digital Competence Gaps, are relevant, since this tutorial will force the user to remember code and commands and forces the user to assess their knowledge and skills.
+
+## Digital Skills Passport
 

### 2024-03-01 paper.tex
minor edits
Added digcomp 2.2 and the tutorials to the paper, from the DigComp document

@@
 \author{true \and true \and true \and true \and true \and true \and true \and true \and true}
-\date{}
+\date{01 March 2024}
 
@@
 During the project initial testing was done with all authors and issues in GitHub were used to tackle bugs.
+
+The learning and teaching material for ABM for archaeologists assumes that the users have completed secondary education.
+It is suitable for archaeologists with at least some background in archaeology and preferably after at least the first year of a Bachelor study in Archaeology (preferably above EQF levels 4 or 5).
+In addition, the tutorials are aimed at professionals working in archaeology and some experience with computer.
+We also assume that learners are proficient in English and have at least the Reading B1 level, but Reading B2 is recommended (Council of Europe 2020).
+The tutorials are developed aiming to develop learners in the following competence area's: 1.
+Information and data literacy, 2.
+Communication and collaboration, 3.
+Digital content creation and, 5.
+Problem solving (European Commission, Joint Research Centre et al.
+2022).
 
@@
 
-\hypertarget{before-the-workhops}{%
-\subsection{Before the workhops}\label{before-the-workhops}}
+\hypertarget{final-tutorials}{%
+\subsection{Final tutorials}\label{final-tutorials}}
+
+The set of developed tutorials consists of the following tutorials:
+
+\begin{itemize}
+\item
+  Tutorial 1: Introduction to ABM
+\item
+  Tutorial 2: Beginning with NetLogo
+\item
+  Tutorial 3: Expanded ABM skills
+\item
+  Tutorial 4: Intermediate ABM
+\item
+  Tutorial 5: How to Model
+\end{itemize}
+
+Each tutorials consists of a different number of lessons that guide the learner in a self-pace manner through the lessons.
+
+In the first tutorial the learner will learn what simulations and agent-based models are and how they can help in archaeological research.
+This tutorial consists of 4 lessons.
+The first two lesson introduce the learner into simulation in general and Agent-Based Modelling in specific.
+Various concepts related to ABM are introduced and the first concepts within NetLogo are explained.
+The third lesson aims to explain how ABMs are used in archaeological research.
+The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode.
+This tutorial build a foundation for working in NetLogo and with ABM.
+All activity stays within the intermediate level, although higher levels might be touch upon.
+
+In tutorial 2 the learner will learn the basics of NetLogo by making your first simulation on the Out of Africal dispersal of homo sapiens.
+The leaerner will work with basic NetLogo syntax and learn how to set up a simulation and visualize their outcomes.
+This tutorial consists of 9 lessons.
+This tutorial intends to guide the learner from the intermediate to the advanced level of proficiency.
+The learning curve is relatively gentle.
+This is achieved by learning to work with the NetLogo web interface.
+The primitives are introduced and initialization phase of a simple ABM is gone through.
+The world within NetLogo's world (including dimensions, coordinates, origin) and how to alter it are explained and experience in a hands-on approach.
+The learners learn about simple simulation loops and how to use primitives.
+The advanced level is achieved with the custom procedures and variables.
+In addition, if-statements are introduced and more complex versions thereof.
+Exporting information using plots is also explained.
+
+In the third tutorial, the learner will build a simple trade model, picking up more advanced NetLogo coding along the way, such as loops, lists and reporters.
+The learner will be introduced to some techniques like modular code development and debugging which will become important with this increased coding complexity.
+This tutorial consists of 7 lessons.
+This tutorial intends to guide the learner from the intermediate to the advanced and highly specialised level of proficiency.
+This is achieved by improving development through using modular code, pseudocode and annotating the code.
+In addition, custom agent breeds are introduced.
+Visualization with labels and reporters, plots and monitors are learned.
+The use of loops is further explained.
+Debugging is learned in lesson 7.
+This can be very advanced, since it involves knowledge of the NetLogo-language combined with problem solving of unexpected behaviour.
+
+In tutorial 4 the learner will work with Sugarscape simulations to further expand your ABM and NetLogo skills.
+The learner will learn how to set up more complex interactions between agents and the environment.
+Furthermore, the basics of setting up good experiments and validating models will be explained.
+This tutorial consists of 8 lessons.
+This tutorial intends to guide the learner from the (intermediate and) advanced level to the highly specialised level of proficiency.
+In this tutorial more complex agent-environment interaction is further learned, including learning ways to visualize the environment and how to give more agency to patches.
+Creating toy landscapes are central to lesson 5.
+The learner also learns the basics of setting up experiments in NetLogo and how to use monitors and flexible plots to understand the results.
+The validation of agent-based models is an important concept of designing a good experiment.
+The finale step, is to compare simulation results with the archaeological record.
+The last lesson is aimed at learning syntax to make lists more dynamic and how to refactor code.
+
+In the final and fifth tutorial the learner will learn more about how to actually incorporate agent-based modelling in research.
+This tutorial focusses less on programming in NetLogo and more on the model development process.
+The learner will learn about the different phases in modelling and how to export data from models.
+This tutorial consists of 5 lessons.
+This tutorial touches upon various levels, but mainly between the intermediate and highly specialised level of proficiency.
+This tutorial is approaches more theoretical aspects in a practical environment.
+The learner leans about the model development process starting with the conceptual phase of the model development.
+Good and bad research questions for modelling are explained and the difference between them.
+The learner needs to understand how to pick the right modelling technique.
+The learner should understand the importance of properly conceptualizing a model before starting the technical phase of the model development.
+In this phase parametrisation, designing experiments and the analyses and interpretation of the models are important.
+Learning about the dissemination phase of the model development is aimed at understanding (the importance of) publishing models and replication.
+In addition, learning how to export results from NetLogo using basic export primitives is explained.
+The BehaviorSpace is explained to enable the learners to analyse the output of models.
+
+\hypertarget{before-the-workshops}{%
+\subsection{Before the workshops}\label{before-the-workshops}}
 
@@
 
-\hypertarget{after-tutorial}{%
-\subsection{After tutorial}\label{after-tutorial}}
+\hypertarget{during-the-workshop}{%
+\subsection{During the workshop}\label{during-the-workshop}}
+
+During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials.
+Some particpants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general.
+
+\hypertarget{after-workshop}{%
+\subsection{After workshop}\label{after-workshop}}
 
 A large proportion of the participants of the workshops gave us information using the survey after the workshops (171 of 368 participants), a similar proportion as those responding to the survey before the workshop.
-
-The respondents
 
@@
 
-\hypertarget{final-tutorials}{%
-\subsection{Final tutorials}\label{final-tutorials}}
+\hypertarget{the-digital-competence-framework-for-citizens-2.2}{%
+\subsection{The Digital Competence Framework for Citizens (2.2)}\label{the-digital-competence-framework-for-citizens-2.2}}
+
+The tutorials are aimed at developing learners in the following competence area's: 1.
+Information and data literacy, 2.
+Communication and collaboration, 3.
+Digital content creation and, 5.
+Problem solving (European Commission, Joint Research Centre et al.
+2022).
+In these competence area's various competences are developed towards an advanced or even specialised level.
+
+For competence area 1.
+Information and data literacy: the tutorials mainly address 1.2 Evaluating Data, Information And Digital Content and 1.3 Managing Data, Information And Digital Content.
+The users learn about working with data, being critacal of their digital content and how to model and manage the data in the context of an ABM.
+For tutorial 2 the user is expected to be able to lookup information using the NetLogo(Web) dictionary (\url{http://ccl.northwestern.edu/netlogo/bind/} or \url{http://ccl.northwestern.edu/netlogo/docs/dictionary.html}).
+
+For Competence Area 2.
+Communication and collaboration the tutorials mainly address 2.1 Interacting Through Digital Technologies, because the users are constantly interacting with digital technologies, but communications is not relevant for the tutorials.
+However, during the workshops given in the course of the project, participants interacted with each other and the teachers.
+
+Competence Area 3.
+Digital content creation is one of the main competence area's for the ABM-teaching material.
+In the process of learning ABM the users are constantly 3.1 Developing Digital Content and 3.2 Integrating And Re-Elaborating Digital Content.
+The material does not touch upon 3.3 Copyright And Licences, but we assume that users have a general understanding.
+A very important aspect of learning ABM is 3.4 Programming.
+The users are exploring the possibilities and chances of programming an ABM.
+The users are expected to learn both syntax and more general concepts such as modular code development, loops, lists and commenting and documenting code.
+
+Competence Area 5.
+Problem solving is very important and is closely tied to 3.4 Programming, since programming involves a lot of problem solving.
+In addition, the creation of a highly complex ABM consists of problem solving all the time, including debugging.
+All competencies are addressed: 5.1 Solving Technical Problems, is common when writing code, learning about coding, error handling and going from pseudocode to real code.
+The competence 5.2 Identifying Needs And Technological Responses, is a central issue of the course, while 5.3 Creatively Using Digital Technology is an important aspect of learning to think in models and develop models themselves.
+In tutorial 2 various aspects of 5.4 Identifying Digital Competence Gaps, are relevant, since this tutorial will force the user to remember code and commands and forces the user to assess their knowledge and skills.
+
+\hypertarget{digital-skills-passport}{%
+\subsection{Digital Skills Passport}\label{digital-skills-passport}}
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-councilofeurope2020}{}}%
+Council of Europe.
+2020.
+\emph{\href{http://www.coe.int/lang-cefr}{Common european framework of reference for languages: Learning, teaching, assessment.
+Companion volume}}.
+Strasbourg: Council of Europe Publishing.
+
+\leavevmode\vadjust pre{\hypertarget{ref-europeancommissionjointresearchcentre2022}{}}%
+European Commission, Joint Research Centre, Vuorikari, R, Kluzer, S and Punie, Y.
+2022.
+\emph{\href{https://data.europa.eu/doi/10.2760/115376}{DigComp 2.2: The Digital Competence Framework for Citizens - With new examples of knowledge, skills and attitudes}}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-grolemund2011}{}}%

### 2024-03-01 paper.Rmd
minor edits

@@
   - name: Iza Romanowska 
-    orcid_id: 00000
+    orcid_id: 0000-0002-9487-2111
     role: 
@@
 
+This project was funded as an Erasmus+ project (<https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054>) and we thank the Erasmus+ program for enabling us to make these tutorials.
+
 We would like to thank the student groups at Saxion University of Applied Sciences that worked with us:

### 2024-03-05 paper.Rmd
minor edits and ORCID's added

@@
       - Writing - Review & Editing
-    orcid_id: 0000
+    orcid_id: 0000-0002-1589-7768
     affiliation: Aarhus University
@@
       - Writing - Review & Editing
-    orcid_id: 0000
+    orcid_id: 0000-0001-6432-0925
     affiliation: Leiden University
@@
 
+Funding acknowledgements (need for the part of the APC that I will pay): The Carlsberg Foundation\'s *Young Researcher Fellowship* (CF21-0382).
+
 # References

### 2024-03-05 paper.tex
minor edits
minor edits and ORCID's added

@@
 \author{true \and true \and true \and true \and true \and true \and true \and true \and true}
-\date{01 March 2024}
+\date{05 March 2024}
 
@@
 
+This project was funded as an Erasmus+ project (\url{https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054}) and we thank the Erasmus+ program for enabling us to make these tutorials.
+
 We would like to thank the student groups at Saxion University of Applied Sciences that worked with us:
@@
 We also want to thank all the participants of our workshops.
+
+Funding acknowledgements (need for the part of the APC that I will pay): The Carlsberg Foundation\textquotesingle s \emph{Young Researcher Fellowship} (CF21-0382).
 

### 2024-03-18 paper.Rmd
minor edits analysis-script and reprodicing figures

@@
 | Conference or event          | Period                  | Online, in person or hybrid | Number of participants | Number of registrations |
-|---------------|---------------|---------------|---------------|---------------|
+|------------------------------|-------------------------|-----------------------------|------------------------|-------------------------|
 | CAA Amsterdam                | April 2023              | In person                   | 36                     | 53                      |
@@
 
-Funding acknowledgements (need for the part of the APC that I will pay): The Carlsberg Foundation\'s *Young Researcher Fellowship* (CF21-0382).
+Funding acknowledgements (need for the part of the APC that I will pay): The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382).
 

### 2024-03-18 Analyse_post_workshop_survey.R

@@ -67,14 +67,14 @@ post_surveys %>%
   mutate(How.would.you.rate.the.trainers.teachers. = fct_relevel(How.would.you.rate.the.trainers.teachers., factor_levels)) %>%
   ggplot(aes(x=How.would.you.rate.the.trainers.teachers.)) + geom_bar() +
   xlab("Rating of the teacher")
-ggsave("export/postworkshop/rating_teaching_material.png")
+ggsave("export/postworkshop/rating_teachers.png")
 
 post_surveys %>%
   mutate(How.would.you.rate.the.trainers.teachers. = fct_relevel(How.would.you.rate.the.trainers.teachers., factor_levels)) %>%
   ggplot(aes(x=How.would.you.rate.the.trainers.teachers.)) + geom_bar() +
   xlab("Rating of the teacher") + facet_grid(~Workshop) +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
-ggsave("export/postworkshop/rating_workshop_facet.png", width = 10)
+ggsave("export/postworkshop/rating_teachers_facet.png", width = 10)
 
 # Apply ABM in the future?
 

### 2024-03-19 paper.Rmd
Update section on website

@@
 
-The tutorial were tested during various online and in-person conferences.
+The tutorials were tested during various online and in-person conferences.
 The in-person conferences were attended by several people from the project team.
 During the online events we were always present with several people and enabled participants to go to break-out rooms if they needed any help or wanted to discuss things.
-In addition, we also had students working with the tutorials at Saxion University of applied sciences and Leiden University.
+In addition, we also had students working with the tutorials at Saxion University of Applied Sciences and Leiden University.
 We used the feedback of the participants to improve the tutorials.
-The feedback was obtained structurally using two surveys using Qualtrics.
-The participants of the events were asked to answer questions of a survey before participating.
+The feedback was obtained structurally from two surveys using Qualtrics.
+The participants of the events were asked to answer questions of the first survey before participating.
 We used the answers to understand what the background of the participants was and to get an estimate of their level of knowledge in relation to ABM (see appendix ...
@@
 
-Website
+For the further dissemination of the tutorials a website has been launced^[https://ABMArchaeologists.github.io/ABMA_Website/#/].
+The tutorials can be found on this website and there is a link that can be used for feedback.
+Additionally, the website contains: basic explanations of agent-based modelling, a recording from one of the workshops, videos from an online course at Leiden University [@spoc2019] and external resources people can use to learn agent-based modelling.
 

### 2024-03-19 paper.Rmd
Update dissemination section

@@
 
-For the further dissemination of the tutorials a website has been launced^[https://ABMArchaeologists.github.io/ABMA_Website/#/].
-The tutorials can be found on this website and there is a link that can be used for feedback.
-Additionally, the website contains: basic explanations of agent-based modelling, a recording from one of the workshops, videos from an online course at Leiden University [@spoc2019] and external resources people can use to learn agent-based modelling.
+The tutorials are publically hosted on GitHub ^[https://ABMArchaeologists.github.io/ABMA_Website/#/].
+Anyone interested can download the tutorials and their source code.
+To increase the accessibility and visibility of the tutorials, a website has been launced^[https://ABMArchaeologists.github.io/ABMA_Website/#/].
+On this website, the tutorials can be followed without the need to download any software.
 
-Github
+Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University [@spoc2019] and external resources people can use to learn agent-based modelling.
 

### 2024-03-19 paper.Rmd
Update link
Credits Saxion students

@@
 
-The tutorials are publically hosted on GitHub ^[https://ABMArchaeologists.github.io/ABMA_Website/#/].
+The tutorials are publically hosted on GitHub^[https://github.com/ABMArchaeologists/ABMA_tutorials/].
 Anyone interested can download the tutorials and their source code.
@@
 Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University [@spoc2019] and external resources people can use to learn agent-based modelling.
+The website was mainly developed by one of the student groups from Saxion University of Applied Sciences, see the acknowledgements.
 
@@
 
+This last group, we want to thank for providing us with the website on which the tutorials are hosted.
+
 We had help from many people during the various workshops and we would like to thank them for their help (in alphabetical order of their first name): Adéla Sobotkova, Agnes Schneider, Alice Overgaauw, Eduardo Herrera Malatesta, Jens Emil Bødstrup Christoffersen, Johan Broersma, Magnus Lindholm Nielsen, Mandy Hazenberg, María Coto Sarmiento, Paulina Fulneczek, Petra Hermankova, Ties Heesink.

### 2024-03-20 paper.Rmd
minor corrections

@@
     orcid_id: 0000-0001-6966-1729
-    role: 
-      - Writing - Original Draft Preparation
-      - Formal Analysis
-      - Visualization
-      - Writing - Review & Editing
+    roles: [writing, editing, analysis, visualization]
     affiliation: Saxion University of Applied Sciences
   - name: Annemarie Jutte
-    orcid_id: 0000
-    role: 
-    - Writing - Original Draft Preparation
-    - Writing - Review & Editing
+    orcid_id: 0009-0004-9322-0674
+    roles: [writing, editing, software]
     affiliation: Saxion University of Applied Sciences
@@
     orcid_id: 0000
-    role: 
-      - Software
-      - Writing - Original Draft Preparation
-      - Writing - Review & Editing
+    roles: [writing, editing, software]
     affiliation: Landward Research Teoranta
@@
     orcid_id: 0000-0002-9487-2111
-    role: 
-     - Writing - Original Draft Preparation
-     - Writing - Review & Editing
+    roles: [writing, editing]
     affiliation: Aarhus University
   - name: Tom Brughmans
-    role: 
-      - Writing - Original Draft Preparation
-      - Writing - Review & Editing
+    roles: [writing, editing]
     orcid_id: 0000-0002-1589-7768
@@
   - name: Laura van der Knaap 
-    role: 
-      - Software
-      - Writing - Original Draft Preparation
-      - Writing - Review & Editing
+    roles: [writing, editing, software]
     orcid_id: 0000
@@
   - name: Karsten Lambers 
-    role: 
-      - Writing - Original Draft Preparation
-      - Writing - Review & Editing
+    roles: [writing, editing]
     orcid_id: 0000-0001-6432-0925
@@
   - name: Kenneth Aitchison 
-    role: 
-      - Writing - Original Draft Preparation
-      - Writing - Review & Editing
-      - Project administration
-    orcid_id: 0000
+    roles: [writing, editing, software, project administration]
+    orcid_id: 0000-0001-6594-1408
     affiliation: Landward Research Teoranta
   - name: Adam Pažout
-    role: 
-      - Writing - Original Draft Preparation
-      - Writing - Review & Editing
+    roles: [writing, editing]
     orcid_id: 0000
     affiliation: Aarhus University
-date: "`r format(Sys.time(), '%d %B %Y')`"
+date: '`r format(Sys.Date(), "%B %d, %Y")`'
 output:
@@
     fig_caption: yes
+    pandoc_args:
+      --lua-filter=scholarly-metadata.lua
   bookdown::word_document2:
@@
 The tutorials were first designed using storyboards which were later converted into tutorials using NetLOGO Web (<https://www.netlogoweb.org/launch>) and Javascript ().
-The tutorials were developed together and we worked together using GitHub (<https://github.com/lljvdk/EU-ABMA/>).
+The tutorials were developed together and we worked together using GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials>).
 During the project initial testing was done with all authors and issues in GitHub were used to tackle bugs.
@@
 | Conference or event          | Period                  | Online, in person or hybrid | Number of participants | Number of registrations |
-|------------------------------|-------------------------|-----------------------------|------------------------|-------------------------|
+|---------------|---------------|---------------|---------------|---------------|
 | CAA Amsterdam                | April 2023              | In person                   | 36                     | 53                      |
@@
 The following packages were used for analyses: ggplot2 [@wickham2016], dplyr [@wickham2023a], tidyr [@wickham2023], forcats[@wickham2023b], lubridate [@grolemund2011] and stringr [@wickham2023c].
-The date and the code is available at <https://github.com/RonaldVisser/ABM_tutorials> and (...
+The date and the code is available at <https://github.com/ABMArchaeologists/ABMA_paper> and (...
 insert Zenodo reference...)
@@
 Information and data literacy: the tutorials mainly address 1.2 Evaluating Data, Information And Digital Content and 1.3 Managing Data, Information And Digital Content.
-The users learn about working with data, being critacal of their digital content and how to model and manage the data in the context of an ABM.
-For tutorial 2 the user is expected to be able to lookup information using the NetLogo(Web) dictionary (<http://ccl.northwestern.edu/netlogo/bind/> or <http://ccl.northwestern.edu/netlogo/docs/dictionary.html>).
+The users learn about working with data, being critical of their digital content and how to model and manage the data in the context of an ABM.
+For tutorial 2 the user is expected to be able to find information using the NetLogo(Web) dictionary (<http://ccl.northwestern.edu/netlogo/bind/> or <http://ccl.northwestern.edu/netlogo/docs/dictionary.html>).
 
@@
 
-The tutorials are publically hosted on GitHub^[https://github.com/ABMArchaeologists/ABMA_tutorials/].
+The tutorials are publicly hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials/>).
 Anyone interested can download the tutorials and their source code.
-To increase the accessibility and visibility of the tutorials, a website has been launced^[https://ABMArchaeologists.github.io/ABMA_Website/#/].
+To increase the accessibility and visibility of the tutorials, a website has been launced ([https://ABMArchaeologists.github.io/ABMA_Website](https://ABMArchaeologists.github.io/ABMA_Website/#/){.uri}).
 On this website, the tutorials can be followed without the need to download any software.
 
-Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University [@spoc2019] and external resources people can use to learn agent-based modelling.
+Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University [@scherjon2019] and external resources people can use to learn agent-based modelling.
 The website was mainly developed by one of the student groups from Saxion University of Applied Sciences, see the acknowledgements.
@@
 
-Funding acknowledgements (need for the part of the APC that I will pay): The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382).
+Funding acknowledgements for the publication: The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382).
 

### 2024-03-20 paper.tex
minor corrections

@@
 \author{true \and true \and true \and true \and true \and true \and true \and true \and true}
-\date{05 March 2024}
+\date{March 20, 2024}
 
@@
 The tutorials were first designed using storyboards which were later converted into tutorials using NetLOGO Web (\url{https://www.netlogoweb.org/launch}) and Javascript ().
-The tutorials were developed together and we worked together using GitHub (\url{https://github.com/lljvdk/EU-ABMA/}).
+The tutorials were developed together and we worked together using GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials}).
 During the project initial testing was done with all authors and issues in GitHub were used to tackle bugs.
@@
 
-The tutorial were tested during various online and in-person conferences.
+The tutorials were tested during various online and in-person conferences.
 The in-person conferences were attended by several people from the project team.
 During the online events we were always present with several people and enabled participants to go to break-out rooms if they needed any help or wanted to discuss things.
-In addition, we also had students working with the tutorials at Saxion University of applied sciences and Leiden University.
+In addition, we also had students working with the tutorials at Saxion University of Applied Sciences and Leiden University.
 We used the feedback of the participants to improve the tutorials.
-The feedback was obtained structurally using two surveys using Qualtrics.
-The participants of the events were asked to answer questions of a survey before participating.
+The feedback was obtained structurally from two surveys using Qualtrics.
+The participants of the events were asked to answer questions of the first survey before participating.
 We used the answers to understand what the background of the participants was and to get an estimate of their level of knowledge in relation to ABM (see appendix \ldots{} for the questions).
@@
 2023), tidyr (Wickham, Vaughan and Girlich 2023), forcats(Wickham 2023a), lubridate (Grolemund and Wickham 2011) and stringr (Wickham 2023b).
-The date and the code is available at \url{https://github.com/RonaldVisser/ABM_tutorials} and (\ldots{} insert Zenodo reference\ldots)
+The date and the code is available at \url{https://github.com/ABMArchaeologists/ABMA_paper} and (\ldots{} insert Zenodo reference\ldots)
 
@@
 Information and data literacy: the tutorials mainly address 1.2 Evaluating Data, Information And Digital Content and 1.3 Managing Data, Information And Digital Content.
-The users learn about working with data, being critacal of their digital content and how to model and manage the data in the context of an ABM.
-For tutorial 2 the user is expected to be able to lookup information using the NetLogo(Web) dictionary (\url{http://ccl.northwestern.edu/netlogo/bind/} or \url{http://ccl.northwestern.edu/netlogo/docs/dictionary.html}).
+The users learn about working with data, being critical of their digital content and how to model and manage the data in the context of an ABM.
+For tutorial 2 the user is expected to be able to find information using the NetLogo(Web) dictionary (\url{http://ccl.northwestern.edu/netlogo/bind/} or \url{http://ccl.northwestern.edu/netlogo/docs/dictionary.html}).
 
@@
 
-Website
+The tutorials are publicly hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials/}).
+Anyone interested can download the tutorials and their source code.
+To increase the accessibility and visibility of the tutorials, a website has been launced (\href{https://ABMArchaeologists.github.io/ABMA_Website/\#/}{https://ABMArchaeologists.github.io/ABMA\_Website}).
+On this website, the tutorials can be followed without the need to download any software.
 
-Github
+Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University (Scherjon, Romanowska and Lambers 2019) and external resources people can use to learn agent-based modelling.
+The website was mainly developed by one of the student groups from Saxion University of Applied Sciences, see the acknowledgements.
 
@@
 
-Funding acknowledgements (need for the part of the APC that I will pay): The Carlsberg Foundation\textquotesingle s \emph{Young Researcher Fellowship} (CF21-0382).
+Funding acknowledgements for the publication: The Carlsberg Foundation's \emph{Young Researcher Fellowship} (CF21-0382).
 

### 2024-03-22 paper.Rmd
Updated paper based on feedback Karsten Lambers

@@
     fig_caption: yes
-    pandoc_args:
-      --lua-filter=scholarly-metadata.lua
   bookdown::word_document2:
@@
 
-Other teaching material available SPOC [@scherjon2019]
+We were able to draw on previous experiences with a small private online course (SPOC) on 'Modelling and Simulation in Archaeology' that some of us designed for graduate teaching at the Faculty of Archaeology, Leiden University, where we taught this course in 2016/17 and again in 2018/19 [@scherjon2019].
+For this course we developed new online teaching materials comprising pre-recorded lectures, practical exercises, reading assignments and exams.
+While we greatly benefited from that experience, the target group of the SPOC was graduate students with prior experience in digital archaeology, whereas the ERASMUS+ project was aimed at a much broader target group of archaeology students and professionals with only basic digital skills.
 
@@
 
-During the project we have developed tutorials.
+During the project we developed online tutorials.
 The tutorials were first designed using storyboards which were later converted into tutorials using NetLOGO Web (<https://www.netlogoweb.org/launch>) and Javascript ().
-The tutorials were developed together and we worked together using GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials>).
-During the project initial testing was done with all authors and issues in GitHub were used to tackle bugs.
+The tutorials were developed as a team and we worked together using GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials>).
+During the project, initial testing was done by all authors and issues in GitHub were used to tackle bugs.
 
 The learning and teaching material for ABM for archaeologists assumes that the users have completed secondary education.
-It is suitable for archaeologists with at least some background in archaeology and preferably after at least the first year of a Bachelor study in Archaeology (preferably above EQF levels 4 or 5).
-In addition, the tutorials are aimed at professionals working in archaeology and some experience with computer.
+It is suitable for archaeologists with at least some background in archaeology, preferably after the first year of a Bachelor study in Archaeology (above EQF levels 4 or 5).
+In addition, the tutorials are aimed at professionals working in archaeology who have some experience with computer applications.
 We also assume that learners are proficient in English and have at least the Reading B1 level, but Reading B2 is recommended [@councilofeurope2020].
-The tutorials are developed aiming to develop learners in the following competence area's: 1.
+The tutorials are developed aiming to develop learners' skills in the following competence area's: 1.
 Information and data literacy, 2.
@@
 
-The tutorials were tested during various online and in-person conferences.
+After initial debugging by the project team, the tutorials were tested 'in the wild' during various online and in-person conferences.
 The in-person conferences were attended by several people from the project team.
@@
 In addition, we also had students working with the tutorials at Saxion University of Applied Sciences and Leiden University.
-We used the feedback of the participants to improve the tutorials.
-The feedback was obtained structurally from two surveys using Qualtrics.
+We We obtained structured feedback from the participants and used it to improve the tutorials prior to the next test.
+The feedback was obtained from two surveys using Qualtrics.
 The participants of the events were asked to answer questions of the first survey before participating.
@@
 for the questions).
-After the participants worked with the tutorials during the event they were asked to answer a second survey.
-Questions of this survey were aimed at the measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix ...
+At the end of the event the participants were asked to answer a second survey.
+Questions of this survey were aimed at measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix ...
 for the questions).
@@
 The post-workshop survey was distributed using QR-codes or links at the end of the workshop during the event.
-After each event we have improved the tutorials based on the received feedback.
+After each event we improved the tutorials based on the received feedback.
 
 | Conference or event          | Period                  | Online, in person or hybrid | Number of participants | Number of registrations |
-|---------------|---------------|---------------|---------------|---------------|
+|------------------------------|-------------------------|-----------------------------|------------------------|-------------------------|
 | CAA Amsterdam                | April 2023              | In person                   | 36                     | 53                      |
@@
 
-: Conferences, events and situations were the workshops given and the tutorials were tested.
+: Conferences, events and situations were workshops were held at which the tutorials were tested.
 
@@
 The backgrounds of the various students were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media & Game Technologies, and ICT.
-The diverse groups of students contributed with new ideas, developing educational material, testing the tutorials and developing a style for the website and other materials.
+The diverse groups of students contributed new ideas, developing educational material, testing the tutorials and developing a style for the website and other materials.
 
@@
 
-In the first tutorial the learner will learn what simulations and agent-based models are and how they can help in archaeological research.
+In the first tutorial the user will learn what simulations and agent-based models are and how they can help in archaeological research.
 This tutorial consists of 4 lessons.
-The first two lesson introduce the learner into simulation in general and Agent-Based Modelling in specific.
+The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific.
 Various concepts related to ABM are introduced and the first concepts within NetLogo are explained.
@@
 This tutorial build a foundation for working in NetLogo and with ABM.
-All activity stays within the intermediate level, although higher levels might be touch upon.
+All activity stays within the intermediate level, although higher levels might be touched upon.
 
-In tutorial 2 the learner will learn the basics of NetLogo by making your first simulation on the Out of Africal dispersal of homo sapiens.
-The leaerner will work with basic NetLogo syntax and learn how to set up a simulation and visualize their outcomes.
+In tutorial 2 the user will learn the basics of NetLogo by making your their simulation on the Out of Africal dispersal of homo sapiens.
+The learner will work with basic NetLogo syntax and learn how to set up a simulation and visualize its outcomes.
 This tutorial consists of 9 lessons.
@@
 
-In tutorial 4 the learner will work with Sugarscape simulations to further expand your ABM and NetLogo skills.
+In tutorial 4 the learner will work with Sugarscape simulations to further expand their ABM and NetLogo skills.
 The learner will learn how to set up more complex interactions between agents and the environment.
@@
 This tutorial intends to guide the learner from the (intermediate and) advanced level to the highly specialised level of proficiency.
-In this tutorial more complex agent-environment interaction is further learned, including learning ways to visualize the environment and how to give more agency to patches.
-Creating toy landscapes are central to lesson 5.
+In this tutorial more complex agent-environment interaction is further learned, including ways to visualize the environment and how to give more agency to patches.
+Creating toy landscapes is central to lesson 5.
 The learner also learns the basics of setting up experiments in NetLogo and how to use monitors and flexible plots to understand the results.
@@
 This tutorial touches upon various levels, but mainly between the intermediate and highly specialised level of proficiency.
-This tutorial is approaches more theoretical aspects in a practical environment.
+This tutorial approaches more theoretical aspects in a practical environment.
 The learner leans about the model development process starting with the conceptual phase of the model development.
@@
 The learner should understand the importance of properly conceptualizing a model before starting the technical phase of the model development.
-In this phase parametrisation, designing experiments and the analyses and interpretation of the models are important.
+In this phase parametrisation, designing experiments and the analysis and interpretation of the models are important.
 Learning about the dissemination phase of the model development is aimed at understanding (the importance of) publishing models and replication.
@@
 The respondents came from many different countries all over the world (see Figure \@ref(fig:nationality).
-The number of female respondents slightly outnumber the male ones and a small group did not share their gender and two were non-binary (see Figure \@ref(fig:gender-age)).
+The number of female respondents slightly outnumbers the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \@ref(fig:gender-age)).
 The ages of the respondents ranged from below 20 to over 70, reaching people in different age classes, although the majority was aged between 20 and 40.
@@
 
-```{r gender-age, fig.cap="The gender and age of the respondents for each workshop", fig.width=11, echo=FALSE}
+```{r gender-age, fig.cap="The gender and age distribution of the respondents for each workshop", fig.width=11, echo=FALSE}
 ggplot(pre_surveys, aes(x = `What.is.your.age.`, fill = `What.is.your.gender....Selected.Choice`)) +
@@
 
-As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply this or had never applied this before.
+As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before.
 The respondents were also asked how they rated the available knowledge on ABM (Figure \@ref(fig:available-theory)).
@@
 Some particpants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general.
+Individual progress varied depending on skills and preferences of the participants.
+While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later instalments.
 
-## After workshop
+## After the workshop
 
@@
 
-Two more open questions were asked to the respondents. The first was aimed at getting positive feedback and the other was aimed at getting feedback to improve the tutorials.
+Two more open questions were asked to the respondents.
+The first was aimed at learning what aspects the participants liked in the tutorials, and the other was aimed at getting feedback to improve the tutorials.
 
@@
 The number of references to bugs reduced to zero by the end and people started providing other feedback, for example the wish to see more examples or expanding the tutorials more.
-Some wanted more cooperation, which is possible, but for some online events harder, although we provided break-out rooms.
-It is also interesting to note that more and more people did not see any room for improvement.
+Some wanted more cooperation, which is possible, but harder for some online events, although we provided break-out rooms.
+It is also interesting to note that over time more and more people did not see any room for improvement.
 
-The majority of the respondents that wanted to apply ABM in the future for research (144 of the 164 that answered this question, see Figure \@ref(fig:future-abm)).
+The majority of the respondents wanted to apply ABM in the future for research (144 of the 164 that answered this question, see Figure \@ref(fig:future-abm)).
 Some of the respondents were already using ABM.
-A large group was not sure yet in how to apply ABM and wanted to read more on the subject or play a bit with the possibilities.
-Many respondents also shared the context in which they were thinking to apply ABM to.
-Various were thinking about movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms.
+A large group was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities.
+Many respondents also shared the context to which they were thinking to apply ABM.
+Various participants mentioned movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms.
 Others thought of demography, social networks, migration or settlement distributions patterns.
 The natural environment and the interaction with humans in the past was also mentioned by some, and often in relation to GIS or how to replace GIS with ABM.
-The archaeological periods that the participants were interested in were very diverse ranging from the paleolithic to the medieval period.
+The archaeological periods that the participants were interested in were very diverse, ranging from the Paleolithic to the Medieval period.
 
@@
 
-The tutorials are aimed at developing learners in the following competence area's: 1.
+As mentioned above, the following competence area's were identified as relevant for the tutorials: 1.
 Information and data literacy, 2.
@@
 For Competence Area 2.
-Communication and collaboration the tutorials mainly address 2.1 Interacting Through Digital Technologies, because the users are constantly interacting with digital technologies, but communications is not relevant for the tutorials.
+Communication and collaboration the tutorials mainly address 2.1 Interacting Through Digital Technologies, because the users are constantly interacting with digital technologies, but communication is not relevant for the tutorials.
 However, during the workshops given in the course of the project, participants interacted with each other and the teachers.
@@
 Competence Area 3.
-Digital content creation is one of the main competence area's for the ABM-teaching material.
+Digital content creation is one of the main competence areas for the ABM-teaching material.
 In the process of learning ABM the users are constantly 3.1 Developing Digital Content and 3.2 Integrating And Re-Elaborating Digital Content.
@@
 The competence 5.2 Identifying Needs And Technological Responses, is a central issue of the course, while 5.3 Creatively Using Digital Technology is an important aspect of learning to think in models and develop models themselves.
-In tutorial 2 various aspects of 5.4 Identifying Digital Competence Gaps, are relevant, since this tutorial will force the user to remember code and commands and forces the user to assess their knowledge and skills.
+In tutorial 2 various aspects of 5.4 Identifying Digital Competence Gaps, are relevant, since this tutorial will force the user to remember code and commands and to assess their knowledge and skills.
 

### 2024-03-22 paper.tex
Updated paper based on feedback Karsten Lambers

@@
 \author{true \and true \and true \and true \and true \and true \and true \and true \and true}
-\date{March 20, 2024}
+\date{March 22, 2024}
 
@@
 
-Other teaching material available SPOC (Scherjon, Romanowska and Lambers 2019)
+We were able to draw on previous experiences with a small private online course (SPOC) on `Modelling and Simulation in Archaeology' that some of us designed for graduate teaching at the Faculty of Archaeology, Leiden University, where we taught this course in 2016/17 and again in 2018/19 (Scherjon, Romanowska and Lambers 2019).
+For this course we developed new online teaching materials comprising pre-recorded lectures, practical exercises, reading assignments and exams.
+While we greatly benefited from that experience, the target group of the SPOC was graduate students with prior experience in digital archaeology, whereas the ERASMUS+ project was aimed at a much broader target group of archaeology students and professionals with only basic digital skills.
 
@@
 
-During the project we have developed tutorials.
+During the project we developed online tutorials.
 The tutorials were first designed using storyboards which were later converted into tutorials using NetLOGO Web (\url{https://www.netlogoweb.org/launch}) and Javascript ().
-The tutorials were developed together and we worked together using GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials}).
-During the project initial testing was done with all authors and issues in GitHub were used to tackle bugs.
+The tutorials were developed as a team and we worked together using GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials}).
+During the project, initial testing was done by all authors and issues in GitHub were used to tackle bugs.
 
 The learning and teaching material for ABM for archaeologists assumes that the users have completed secondary education.
-It is suitable for archaeologists with at least some background in archaeology and preferably after at least the first year of a Bachelor study in Archaeology (preferably above EQF levels 4 or 5).
-In addition, the tutorials are aimed at professionals working in archaeology and some experience with computer.
+It is suitable for archaeologists with at least some background in archaeology, preferably after the first year of a Bachelor study in Archaeology (above EQF levels 4 or 5).
+In addition, the tutorials are aimed at professionals working in archaeology who have some experience with computer applications.
 We also assume that learners are proficient in English and have at least the Reading B1 level, but Reading B2 is recommended (Council of Europe 2020).
-The tutorials are developed aiming to develop learners in the following competence area's: 1.
+The tutorials are developed aiming to develop learners' skills in the following competence area's: 1.
 Information and data literacy, 2.
@@
 
-The tutorials were tested during various online and in-person conferences.
+After initial debugging by the project team, the tutorials were tested `in the wild' during various online and in-person conferences.
 The in-person conferences were attended by several people from the project team.
@@
 In addition, we also had students working with the tutorials at Saxion University of Applied Sciences and Leiden University.
-We used the feedback of the participants to improve the tutorials.
-The feedback was obtained structurally from two surveys using Qualtrics.
+We We obtained structured feedback from the participants and used it to improve the tutorials prior to the next test.
+The feedback was obtained from two surveys using Qualtrics.
 The participants of the events were asked to answer questions of the first survey before participating.
 We used the answers to understand what the background of the participants was and to get an estimate of their level of knowledge in relation to ABM (see appendix \ldots{} for the questions).
-After the participants worked with the tutorials during the event they were asked to answer a second survey.
-Questions of this survey were aimed at the measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix \ldots{} for the questions).
+At the end of the event the participants were asked to answer a second survey.
+Questions of this survey were aimed at measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix \ldots{} for the questions).
 For some events participants had to register beforehand and we were able to send the pre-workshop survey by email.
@@
 The post-workshop survey was distributed using QR-codes or links at the end of the workshop during the event.
-After each event we have improved the tutorials based on the received feedback.
+After each event we improved the tutorials based on the received feedback.
 
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}
-\caption{Conferences, events and situations were the workshops given and the tutorials were tested.}\tabularnewline
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2256}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1880}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2180}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1805}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1880}}@{}}
+\caption{Conferences, events and situations were workshops were held at which the tutorials were tested.}\tabularnewline
 \toprule\noalign{}
@@
 The backgrounds of the various students were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media \& Game Technologies, and ICT.
-The diverse groups of students contributed with new ideas, developing educational material, testing the tutorials and developing a style for the website and other materials.
+The diverse groups of students contributed new ideas, developing educational material, testing the tutorials and developing a style for the website and other materials.
 
@@
 The respondents came from many different countries all over the world (see Figure \ref{fig:nationality}.
-The number of female respondents slightly outnumber the male ones and a small group did not share their gender and two were non-binary (see Figure \ref{fig:gender-age}).
+The number of female respondents slightly outnumbers the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \ref{fig:gender-age}).
 The ages of the respondents ranged from below 20 to over 70, reaching people in different age classes, although the majority was aged between 20 and 40.
@@
 \includegraphics{paper_files/figure-latex/gender-age-1.pdf}
-\caption{\label{fig:gender-age}The gender and age of the respondents for each workshop}
+\caption{\label{fig:gender-age}The gender and age distribution of the respondents for each workshop}
 \end{figure}
@@
 
-As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply this or had never applied this before.
+As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before.
 The respondents were also asked how they rated the available knowledge on ABM (Figure \ref{fig:available-theory}).
@@
 Some particpants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general.
+Individual progress varied depending on skills and preferences of the participants.
+While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later instalments.
 
-\hypertarget{after-workshop}{%
-\subsection{After workshop}\label{after-workshop}}
+\hypertarget{after-the-workshop}{%
+\subsection{After the workshop}\label{after-the-workshop}}
 
@@
 The number of references to bugs reduced to zero by the end and people started providing other feedback, for example the wish to see more examples or expanding the tutorials more.
-Some wanted more cooperation, which is possible, but for some online events harder, although we provided break-out rooms.
-It is also interesting to note that more and more people did not see any room for improvement.
+Some wanted more cooperation, which is possible, but harder for some online events, although we provided break-out rooms.
+It is also interesting to note that over time more and more people did not see any room for improvement.
 
-The majority of the respondents that wanted to apply ABM in the future for research (144 of the 164 that answered this question, see Figure \ref{fig:future-abm}).
+The majority of the respondents wanted to apply ABM in the future for research (144 of the 164 that answered this question, see Figure \ref{fig:future-abm}).
 Some of the respondents were already using ABM.
-A large group was not sure yet in how to apply ABM and wanted to read more on the subject or play a bit with the possibilities.
-Many respondents also shared the context in which they were thinking to apply ABM to.
-Various were thinking about movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms.
+A large group was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities.
+Many respondents also shared the context to which they were thinking to apply ABM.
+Various participants mentioned movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms.
 Others thought of demography, social networks, migration or settlement distributions patterns.
 The natural environment and the interaction with humans in the past was also mentioned by some, and often in relation to GIS or how to replace GIS with ABM.
-The archaeological periods that the participants were interested in were very diverse ranging from the paleolithic to the medieval period.
+The archaeological periods that the participants were interested in were very diverse, ranging from the Paleolithic to the Medieval period.
 

### 2024-04-03 paper.Rmd
new version paper

@@
     affiliation: Saxion University of Applied Sciences
+  - name: Karsten Lambers 
+    roles: [writing, editing]
+    orcid_id: 0000-0001-6432-0925
+    affiliation: Leiden University
+  - name: Kenneth Aitchison 
+    roles: [writing, editing, software, project administration]
+    orcid_id: 0000-0001-6594-1408
+    affiliation: Landward Research Teoranta
   - name: Annemarie Jutte
@@
     affiliation: Landward Research Teoranta
+  - name: Laura van der Knaap 
+    roles: [writing, editing, software]
+    orcid_id: 0000
+    affiliation: Leiden University
   - name: Iza Romanowska 
@@
     affiliation: Aarhus University
-  - name: Laura van der Knaap 
-    roles: [writing, editing, software]
-    orcid_id: 0000
-    affiliation: Leiden University
-  - name: Karsten Lambers 
-    roles: [writing, editing]
-    orcid_id: 0000-0001-6432-0925
-    affiliation: Leiden University
-  - name: Kenneth Aitchison 
-    roles: [writing, editing, software, project administration]
-    orcid_id: 0000-0001-6594-1408
-    affiliation: Landward Research Teoranta
   - name: Adam Pažout
@@
 
-ABM application increased in Archaeology
+While Agent-Based Modelling is a predictive tool that is in high demand from archaeologists, there are significant sectoral problems with access to vocational training for this skill.
+A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 76% of modellers were primarily self-taught with some degree of secondary support i.e.
+27% relied on peer support, only 7% had any sort of formal training at a workshop or summer school while only 14% received training as part of a degree.
+
+A consortium of Landward Research Teoranta, Aarhus Universitet, Universiteit Leiden and Saxion University of Applied Sciences recognised this need, and sought to address and overcome this training deficiency through the development and delivery of a project to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.
+
+This project, Agent-Based Modelling for Archaeologists (<https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054>), was delivered with the support of the Erasmus+ programme of the European Union.
+
+The project sought to create Open (Open Access and Open Source) Educational Resources (OERs) for training in Agent-Based Modelling for Archaeologists.
+These OERs were to be HTML and JavaScript based so they could be used on any device with a browser, even if it is not connected to the internet.
+These are hands-on practical exercises, taking learners through the process of creating the digital models, and modular so they can be independent or incorporated into other teaching tools like learning management systems.
+They are deliberately 'method agnostic' so they can be incorporated into any tools from self-teaching to MOOCs, and are supported by the creation of support materials to encourage use of the OERs.
+
+The activities the project implemented were:
+
+-   Development of a hands-on vocational training programme, aligned with the recently published *Agent-Based Modelling for Archaeology: Simulating the Complexity of Societies* [@romanowska2021a].
+
+-   Conversion of the training programme into modular learning resources that are interactive, based on HTML and JavaScript standards.
+
+-   Creation of support materials to encourage use of the OERs.
+
+-   Creation of how-to guides to help trainers incorporate OERs into their teaching
+
+-   Creation of a code and learning materials repository that will allow the project to continue sustainability after the European Union support has ended and to facilitate the creation of an open-source community to manage and improve the OERs in the future.
+
+-   Promotion through a series of multiplier events.
+
+The expected results and outcomes, for the target audience of professional archaeologists, were:
+
+1.  increased level of digital competence, specifically in advanced computer modelling skills that are highly transferable to other industries (computer programming and simulation development);
+
+2.
+improved transnational mobility through matching of the skills being delivered with European and Sectoral Frameworks, leading to individual workers being better enabled to demonstrate their skills to employers across Europe;
+
+3.  increased opportunities for professional development by having access to professional training that can be incorporated into formal or non-formal/informal learning.
+
+For the secondary target audience of vocational training providers in archaeology e.g. employers, colleges and not-for-profit organisations, the expected results and outcomes were:
+
+1.  increased quality of education and training in the areas of ABM that are free and open;
+
+2.  access to teaching materials for skills that are in high demand and aligned with employers\' needs.
 
@@
 While we greatly benefited from that experience, the target group of the SPOC was graduate students with prior experience in digital archaeology, whereas the ERASMUS+ project was aimed at a much broader target group of archaeology students and professionals with only basic digital skills.
-
-New handbook [@romanowska2021]
-
-No formal training
-
-EU project <https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054>
 
@@
 
-Lessons learned
+Things to address
 
-Future aspects
+-   demand for tutorials exceeded expectations
+
+-   tutorials helpful in understanding ABM
+
+-   international cooperation
+
+-   Digital competency framework / Skills passport
+
+-   OER: open learning
+
+-   Future aspects:
+
+    -   application in educational programmes
+
+    -   community driven updates (?)
 
@@
 
-This project was funded as an Erasmus+ project (<https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054>) and we thank the Erasmus+ program for enabling us to make these tutorials.
+This project was undertaken with the support of the Erasmus+ programme of the European Union (<https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054>).
 

### 2024-04-03 paper.tex
new version paper

@@
 \author{true \and true \and true \and true \and true \and true \and true \and true \and true}
-\date{March 22, 2024}
+\date{April 03, 2024}
 
@@
 
-ABM application increased in Archaeology
+While Agent-Based Modelling is a predictive tool that is in high demand from archaeologists, there are significant sectoral problems with access to vocational training for this skill.
+A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 76\% of modellers were primarily self-taught with some degree of secondary support i.e.~27\% relied on peer support, only 7\% had any sort of formal training at a workshop or summer school while only 14\% received training as part of a degree.
+
+A consortium of Landward Research Teoranta, Aarhus Universitet, Universiteit Leiden and Saxion University of Applied Sciences recognised this need, and sought to address and overcome this training deficiency through the development and delivery of a project to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.
+
+This project, Agent-Based Modelling for Archaeologists (\url{https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054}), was delivered with the support of the Erasmus+ programme of the European Union.
+
+The project sought to create Open (Open Access and Open Source) Educational Resources (OERs) for training in Agent-Based Modelling for Archaeologists.
+These OERs were to be HTML and JavaScript based so they could be used on any device with a browser, even if it is not connected to the internet.
+These are hands-on practical exercises, taking learners through the process of creating the digital models, and modular so they can be independent or incorporated into other teaching tools like learning management systems.
+They are deliberately `method agnostic' so they can be incorporated into any tools from self-teaching to MOOCs, and are supported by the creation of support materials to encourage use of the OERs.
+
+The activities the project implemented were:
+
+\begin{itemize}
+\item
+  Development of a hands-on vocational training programme, aligned with the recently published \emph{Agent-Based Modelling for Archaeology: Simulating the Complexity of Societies} (Romanowska, Wren and Crabtree 2021).
+\item
+  Conversion of the training programme into modular learning resources that are interactive, based on HTML and JavaScript standards.
+\item
+  Creation of support materials to encourage use of the OERs.
+\item
+  Creation of how-to guides to help trainers incorporate OERs into their teaching
+\item
+  Creation of a code and learning materials repository that will allow the project to continue sustainability after the European Union support has ended and to facilitate the creation of an open-source community to manage and improve the OERs in the future.
+\item
+  Promotion through a series of multiplier events.
+\end{itemize}
+
+The expected results and outcomes, for the target audience of professional archaeologists, were:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  increased level of digital competence, specifically in advanced computer modelling skills that are highly transferable to other industries (computer programming and simulation development);
+\item
+  improved transnational mobility through matching of the skills being delivered with European and Sectoral Frameworks, leading to individual workers being better enabled to demonstrate their skills to employers across Europe;
+\item
+  increased opportunities for professional development by having access to professional training that can be incorporated into formal or non-formal/informal learning.
+\end{enumerate}
+
+For the secondary target audience of vocational training providers in archaeology e.g.~employers, colleges and not-for-profit organisations, the expected results and outcomes were:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  increased quality of education and training in the areas of ABM that are free and open;
+\item
+  access to teaching materials for skills that are in high demand and aligned with employers\textquotesingle{} needs.
+\end{enumerate}
 
@@
 While we greatly benefited from that experience, the target group of the SPOC was graduate students with prior experience in digital archaeology, whereas the ERASMUS+ project was aimed at a much broader target group of archaeology students and professionals with only basic digital skills.
-
-New handbook (Romanowska, Wren and Crabtree 2021)
-
-No formal training
-
-EU project \url{https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054}
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2256}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1880}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2180}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1805}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1880}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}
 \caption{Conferences, events and situations were workshops were held at which the tutorials were tested.}\tabularnewline
@@
 
-Lessons learned
+Things to address
 
-Future aspects
+\begin{itemize}
+\item
+  demand for tutorials exceeded expectations
+\item
+  tutorials helpful in understanding ABM
+\item
+  international cooperation
+\item
+  Digital competency framework / Skills passport
+\item
+  OER: open learning
+\item
+  Future aspects:
+
+  \begin{itemize}
+  \item
+    application in educational programmes
+  \item
+    community driven updates (?)
+  \end{itemize}
+\end{itemize}
 
@@
 
-This project was funded as an Erasmus+ project (\url{https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054}) and we thank the Erasmus+ program for enabling us to make these tutorials.
+This project was undertaken with the support of the Erasmus+ programme of the European Union (\url{https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054}).
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-davies2018}{}}%
+Davies, B and Romanowska, I. 2018 An emergent community? Agent-based modelers in archaeology. \emph{The SAA Archaeological Record} 18(2): 27--32.
+
 \leavevmode\vadjust pre{\hypertarget{ref-europeancommissionjointresearchcentre2022}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-romanowska2021}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-romanowska2021a}{}}%
 Romanowska, I, Wren, CD and Crabtree, SA.

### 2024-04-10 paper.Rmd
paper edit

@@
 
-2.  access to teaching materials for skills that are in high demand and aligned with employers\' needs.
+2.  access to teaching materials for skills that are in high demand and aligned with employers' needs.
 
@@
 
+The final tutorals are available for download at <https://github.com/ABMArchaeologists/ABMA_tutorials> and are citeable [@rocks-macqueen2024].
 The set of developed tutorials consists of the following tutorials:
@@
 
-The tutorials are publicly hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials/>).
+The tutorials are publicly hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials/>) and are citeable [@jutte2024].
 Anyone interested can download the tutorials and their source code.

### 2024-04-10 paper.tex
paper edit

@@
 \author{true \and true \and true \and true \and true \and true \and true \and true \and true}
-\date{April 03, 2024}
+\date{April 10, 2024}
 
@@
 \item
-  access to teaching materials for skills that are in high demand and aligned with employers\textquotesingle{} needs.
+  access to teaching materials for skills that are in high demand and aligned with employers' needs.
 \end{enumerate}
@@
 
+The final tutorals are available for download at \url{https://github.com/ABMArchaeologists/ABMA_tutorials} and are citeable (Rocks-Macqueen et al.
+2024).
 The set of developed tutorials consists of the following tutorials:
@@
 
-The tutorials are publicly hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials/}).
+The tutorials are publicly hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials/}) and are citeable (Jutte et al.
+2024).
 Anyone interested can download the tutorials and their source code.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-jutte2024}{}}%
+Jutte, A, Aalpoel, R, Baan, E, Regts, S, Lucas, S, Mulder, J, Sijbrant, A, Visser, RM, Rocks-Macqueen, D, Knaap, L van der, Romanowska, I, Brughmans, T, Lambers, K and Aitchison, K.
+2024.
+\emph{Agent based modelling for archaeologists: Website source}.
+Zenodo.
+DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.10931759}{10.5281/zenodo.10931759}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2023}{}}%
 R Core Team. 2023 \emph{\href{https://www.R-project.org/}{R: A language and environment for statistical computing}}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rocks-macqueen2024}{}}%
+Rocks-Macqueen, D, Knaap, L van der, Romanowska, I, Jutte, A, Visser, RM, Brughmans, T, Lambers, K and Aitchison, K.
+2024.
+\emph{Agent based modelling for archaeologists: tutorials}.
+Zenodo.
+DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.10931587}{10.5281/zenodo.10931587}.
 

### 2024-04-23 paper.Rmd
name correction Alice
added reference to DigComp

@@
 In these competence area's various competences are developed towards an advanced or even specialised level.
+A more extensive description of the Digital Competence Framework for Citizens in relation to Agent-Based Modelling for archaeologists is also available [@visser2024].
 
@@
 
--   February 2022-July 2023: Alice Overgaauw, Johan Broersma, Mandy Hazenberg, Paulina Fulneczek, Ties Heesink and their tutor: Jan Willem Huson.
+-   February 2022-July 2023: Alice Poorthuis, Johan Broersma, Mandy Hazenberg, Paulina Fulneczek, Ties Heesink and their tutor: Jan Wille
 

### 2024-04-23 paper.tex
name correction Alice
added reference to DigComp

@@
 \author{true \and true \and true \and true \and true \and true \and true \and true \and true}
-\date{April 10, 2024}
+\date{April 23, 2024}
 
@@
 In these competence area's various competences are developed towards an advanced or even specialised level.
+A more extensive description of the Digital Competence Framework for Citizens in relation to Agent-Based Modelling for archaeologists is also available (Visser et al.
+2024).
 
@@
 \item
-  February 2022-July 2023: Alice Overgaauw, Johan Broersma, Mandy Hazenberg, Paulina Fulneczek, Ties Heesink and their tutor: Jan Willem Huson.
+  February 2022-July 2023: Alice Poorthuis, Johan Broersma, Mandy Hazenberg, Paulina Fulneczek, Ties Heesink and their tutor: Jan Wille
 \item
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-visser2024}{}}%
+Visser, R, Lambers, K, Aitchison, K, Jutte, A, Rocks-Macqueen, D, Knaap, L van der, Romanowska, I and Brughmans, T.
+2024 \emph{The Digital Competence Framework for Citizens (2.2) and Agent-Based modelling for Archaeologists}.
+DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.11046488}{10.5281/zenodo.11046488}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-wickham2016}{}}%

### 2024-04-25 paper.Rmd
first batch of edits
I made some edits and added a couple of comments. Search for "!!!" to quickly jump to a comment.

--- 2024-04-10 paper.Rmd
@@
 
-While Agent-Based Modelling is a predictive tool that is in high demand from archaeologists, there are significant sectoral problems with access to vocational training for this skill.
+Computational modelling has become a standard research technique across most scientific disciplines including archaeology.
+Among the many computational modelling techniques the discipline has found agent-based modelling to be particularly well suited to the questions asked by archaeologists.
+Agent-based modelling is a simulation technique where autonomous agents interact with each other and with the environment around.
+Their aggregated behaviour gives rise to system-level outcomes thus enabling researchers to understand how *micro* (individual) *mechanisms* lead to *macro* (system-level) *patterns*.
+
+While Agent-Based Modelling importance in archaeological practice is steadily growing \_\_\_ !!!please add the following reference here: <https://www.researchgate.net/publication/376410267_Bibliometric_Analysis_of_Agent-Based_Simulation_in_Archaeology_People_Topics_and_Future_Prospects> \_\_\_, there are significant bariers to its wider adoption.
+One of the sectorial problems is the access to vocational training required to gain the necessary computer programming and modelling skills.
 A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 76% of modellers were primarily self-taught with some degree of secondary support i.e.
-27% relied on peer support, only 7% had any sort of formal training at a workshop or summer school while only 14% received training as part of a degree.
+27% relied on peer support, only 7% had any sort of formal training at a workshop or summer school while 14% received training as part of a degree.
 
-A consortium of Landward Research Teoranta, Aarhus Universitet, Universiteit Leiden and Saxion University of Applied Sciences recognised this need, and sought to address and overcome this training deficiency through the development and delivery of a project to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.
+A consortium of Landward Research Teoranta, Aarhus Universitet, Universiteit Leiden and Saxion University of Applied Sciences recognised this need, and sought to address the limited opportunities that archaeologists have to access high quality training through a project aiming to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.
 
-This project, Agent-Based Modelling for Archaeologists (<https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054>), was delivered with the support of the Erasmus+ programme of the European Union.
+[!!!COMMENT: this sentence should go into the acknowledgements] This project, Agent-Based Modelling for Archaeologists (<https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054>), was delivered with the support of the Erasmus+ programme of the European Union.
 
@@
 These OERs were to be HTML and JavaScript based so they could be used on any device with a browser, even if it is not connected to the internet.
-These are hands-on practical exercises, taking learners through the process of creating the digital models, and modular so they can be independent or incorporated into other teaching tools like learning management systems.
-They are deliberately 'method agnostic' so they can be incorporated into any tools from self-teaching to MOOCs, and are supported by the creation of support materials to encourage use of the OERs.
+The full programme consists of five hands-on practical tutorials, taking learners through the process of creating the digital models.
+The modular architecture ensures that they can be independent or incorporated into other teaching tools like learning management systems.
+They are also deliberately 'method agnostic' so they can be incorporated into any tools from self-teaching to MOOCs, and are supported by a set of support materials to encourage use of the OERs.
 
@@
 
+-   Development of the framework for skills acquisition aligned with the Digital Skills Passport !!!\_\_ add ref here!
+
 -   Promotion through a series of multiplier events.
@@
 
-1.  increased level of digital competence, specifically in advanced computer modelling skills that are highly transferable to other industries (computer programming and simulation development);
+1.  a surge in professionals who can build, apply and critique archaeological simulation thus opening new research avenues for the discipline as a whole;
 
 2.
-improved transnational mobility through matching of the skills being delivered with European and Sectoral Frameworks, leading to individual workers being better enabled to demonstrate their skills to employers across Europe;
+increased level of digital competence, specifically in advanced computer skills that are highly transferable to other industries.
+This includes technical skills such as programming but also "soft" skills related to computational thinking;
 
-3.  increased opportunities for professional development by having access to professional training that can be incorporated into formal or non-formal/informal learning.
+3.
+!!!
+[Comment: This is very unclear.
+Currently it reads as if taking the course will make people emigrage to other countries.
+I don't think that is in any way true.
+Also, most readers would not know what European or Sectoral Framework we are referring to] improved transnational mobility through matching of the skills being delivered with European and Sectoral Frameworks, leading to individual workers being better enabled to demonstrate their skills to employers across Europe;
+
+4.  increased opportunities for professional development by having access to professional training that can be incorporated into formal or non-formal/informal learning.
 
@@
 
-1.  increased quality of education and training in the areas of ABM that are free and open;
+1.  increased quality of training in Agent-based Modelling that is free and open;
 
@@
 For this course we developed new online teaching materials comprising pre-recorded lectures, practical exercises, reading assignments and exams.
-While we greatly benefited from that experience, the target group of the SPOC was graduate students with prior experience in digital archaeology, whereas the ERASMUS+ project was aimed at a much broader target group of archaeology students and professionals with only basic digital skills.
+While we greatly benefited from that experience, the target group of the SPOC was graduate students with prior experience in digital archaeology, whereas the ERASMUS+ project was aimed at a much broader target group of archaeology students and professionals whose computational literacy could not be assumed.
 
@@
 
-During the project we developed online tutorials.
-The tutorials were first designed using storyboards which were later converted into tutorials using NetLOGO Web (<https://www.netlogoweb.org/launch>) and Javascript ().
-The tutorials were developed as a team and we worked together using GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials>).
+The project team developed five online tutorials.
+The tutorials were first designed using storyboards and later converted into interactive guided walk-throughs using NetLOGO Web (<https://www.netlogoweb.org/launch>) and Javascript and hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials>).
 During the project, initial testing was done by all authors and issues in GitHub were used to tackle bugs.
 
-The learning and teaching material for ABM for archaeologists assumes that the users have completed secondary education.
-It is suitable for archaeologists with at least some background in archaeology, preferably after the first year of a Bachelor study in Archaeology (above EQF levels 4 or 5).
+The ABMA learning materials assume that participants have learning skills at least at a secondary education level.
+It is suitable for participants with at least some background in archaeology, such as those typical among students after the first year of a Bachelor study in Archaeology, i.e., above EQF levels 4 or 5.
 In addition, the tutorials are aimed at professionals working in archaeology who have some experience with computer applications.
 We also assume that learners are proficient in English and have at least the Reading B1 level, but Reading B2 is recommended [@councilofeurope2020].
-The tutorials are developed aiming to develop learners' skills in the following competence area's: 1.
+The tutorials were also developed with a secondary aim to develop learners' general skills in the following competence area's: 1.
 Information and data literacy, 2.
@@
 
-After initial debugging by the project team, the tutorials were tested 'in the wild' during various online and in-person conferences.
-The in-person conferences were attended by several people from the project team.
-During the online events we were always present with several people and enabled participants to go to break-out rooms if they needed any help or wanted to discuss things.
-In addition, we also had students working with the tutorials at Saxion University of Applied Sciences and Leiden University.
-We We obtained structured feedback from the participants and used it to improve the tutorials prior to the next test.
-The feedback was obtained from two surveys using Qualtrics.
-The participants of the events were asked to answer questions of the first survey before participating.
-We used the answers to understand what the background of the participants was and to get an estimate of their level of knowledge in relation to ABM (see appendix ...
+After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events all of which were run by members of the project team.
+Recognising the difficulty of delivering practical training in an online environment during the online events several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things.
+Throughout the project we sought feedback from the participants and used it to improve the tutorials prior to the following events.
+The feedback was obtained through two surveys using Qualtrics.
+The participants of the events were asked to respond to two surveys: one before the start of the event and one completed after the end of the event.
+The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix ...
 for the questions).
@@
 At other events no registration was possible or necessary and the pre-workshop survey was given at the start of the event.
-The post-workshop survey was distributed using QR-codes or links at the end of the workshop during the event.
-After each event we improved the tutorials based on the received feedback.
+The post-workshop survey was distributed using QR-codes or links at the end of the event.
 
@@
 
-: Conferences, events and situations were workshops were held at which the tutorials were tested.
+: Conferences, events and workshops at which the tutorials were delivered and tested.
 
@@
 The diverse groups of students contributed new ideas, developing educational material, testing the tutorials and developing a style for the website and other materials.
+!!!
+COMMENT- this section needs to be more specific.
+Why did these student work on the project?
+At what capacity?
+What did they contribute?
+etc
 

### 2024-04-25 paper.Rmd
first batch of edits
I've edited the first part of the paper and added a few comments - search !!! to jump directly to them.

--- 2024-04-25 paper.Rmd
@@
 
-## Final tutorials
+## ABM tutorials
 
-The final tutorals are available for download at <https://github.com/ABMArchaeologists/ABMA_tutorials> and are citeable [@rocks-macqueen2024].
+The final set of ABMA tutorals are available for download at <https://github.com/ABMArchaeologists/ABMA_tutorials> and are citeable [@rocks-macqueen2024].
 The set of developed tutorials consists of the following tutorials:
@@
 
-Each tutorials consists of a different number of lessons that guide the learner in a self-pace manner through the lessons.
+Each tutorials consists of a number of individual lessons, each of which guides the learner in a self-pace manner through an increasingly challenging set of exercises that build the skills and knowledge necessary to become a confident agent-based modeller.
 
-In the first tutorial the user will learn what simulations and agent-based models are and how they can help in archaeological research.
+In the first tutorial the user learn what simulations and agent-based models are and how they can help in archaeological research.
 This tutorial consists of 4 lessons.

### 2024-04-25 paper.Rmd
Final batch up edits
Please search for !!! to see comments

@@
 
-: Conferences, events and workshops at which the tutorials were delivered and tested.
+: Events at which the tutorials were delivered and tested.
 
@@
 The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific.
-Various concepts related to ABM are introduced and the first concepts within NetLogo are explained.
-The third lesson aims to explain how ABMs are used in archaeological research.
+Various concepts related to ABM, such as the definitions of model and simulation, are introduced and the participants make their first steps in NetLogo syntax.
+The third lesson explains how ABMs are used in archaeological research.
 The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode.
-This tutorial build a foundation for working in NetLogo and with ABM.
-All activity stays within the intermediate level, although higher levels might be touched upon.
+This tutorial builds a foundation for working in NetLogo and with ABM.
+All activity stays within the entree level of difficulty, although some higher level concepts are lightly touched upon.
 
-In tutorial 2 the user will learn the basics of NetLogo by making your their simulation on the Out of Africal dispersal of homo sapiens.
-The learner will work with basic NetLogo syntax and learn how to set up a simulation and visualize its outcomes.
+In tutorial 2 the user learns the basics of programming in NetLogo and develops their first simulation based on a previously published model of the Out of Africa dispersal of *Homo sapiens !!!
+Pls insert the following reference*
+
+Young, David A., and Robert L. Bettinger. 1992. “The Numic Spread: A Computer Simulation.” *American Antiquity* 57 (1): 85–99. <https://doi.org/10.2307/2694836>.
+
+.
+The learner works through the basics of NetLogo syntax and learn how to set up a simulation and visualize its outcomes.
 This tutorial consists of 9 lessons.
-This tutorial intends to guide the learner from the intermediate to the advanced level of proficiency.
+This tutorial intends to guide the learner from the introduction to the beginner level of proficiency.
 The learning curve is relatively gentle.
-This is achieved by learning to work with the NetLogo web interface.
-The primitives are introduced and initialization phase of a simple ABM is gone through.
-The world within NetLogo's world (including dimensions, coordinates, origin) and how to alter it are explained and experience in a hands-on approach.
-The learners learn about simple simulation loops and how to use primitives.
-The advanced level is achieved with the custom procedures and variables.
-In addition, if-statements are introduced and more complex versions thereof.
-Exporting information using plots is also explained.
+This is achieved by working with the NetLogo web interface that enables highlighting specific words and buttons and giving visual cues as to how to proceed.
+The tutorial intoduces several most commonly used built-in functions in Netlogo (called *primitives*) are introduces the general structure of all models cosisting of the initialization phase and the main simulation loop.
+It walks the participants through te most important NetLogo settings, such as dimensions, coordinates and the origin point of the grid on which the simulation runs.
+The learners gain their first stripes in programming by using simple loops and functions as well as conditional statements.
+The next level is achieved by introducing the topic of function definition and different types of variables.
+Exporting data produced by the model through plots is also explained.
 
-In the third tutorial, the learner will build a simple trade model, picking up more advanced NetLogo coding along the way, such as loops, lists and reporters.
-The learner will be introduced to some techniques like modular code development and debugging which will become important with this increased coding complexity.
+In the third tutorial, the learner builds a simple trade model, again based on a published simulation !!! pls insert the following reference:
+
+Romanowska, Iza.
+2018.
+“‪Using Agent-Based Modelling to Infer Economic Processes in the Past‬.” In *‪Cuantificar Las Economías Antiguas: Problemas y Métodos*, Publicacions i Edicions de la Universitat de Barcelona, 107–18.
+Barcelona: Universitat de Barcelona.
+<https://scholar.google.com/citations?view_op=view_citation&hl=en&user=tdcyWfoAAAAJ&citation_for_view=tdcyWfoAAAAJ:YOwf2qJgpHMC>.
+
+Here learners are expanding their NetLogo programming skills with more complexity added to such syntactic structures as loops, lists and reporters.
+The learner is introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity.
 This tutorial consists of 7 lessons.
-This tutorial intends to guide the learner from the intermediate to the advanced and highly specialised level of proficiency.
-This is achieved by improving development through using modular code, pseudocode and annotating the code.
-In addition, custom agent breeds are introduced.
-Visualization with labels and reporters, plots and monitors are learned.
-The use of loops is further explained.
-Debugging is learned in lesson 7.
-This can be very advanced, since it involves knowledge of the NetLogo-language combined with problem solving of unexpected behaviour.
+This tutorial intends to guide the learner from the beginner to the intermediate level of proficiency.
+This is achieved by introducing standards of code development such as modular code, pseudocode, debugging and annotating the code.
+In addition, the tutorial introduces custom agent breeds, visualization with labels and reporters, plots and monitors.
+This takes the learner to a more advanced level, since it develops knowledge of the programming language combined with diagnosing unexpected behaviour and problem solving.
 
-In tutorial 4 the learner will work with Sugarscape simulations to further expand their ABM and NetLogo skills.
-The learner will learn how to set up more complex interactions between agents and the environment.
-Furthermore, the basics of setting up good experiments and validating models will be explained.
+In tutorial 4 the learner works with a seminal Sugarscape model !!! pls insert the following ref.
+
+Epstein, Joshua M., and Robert L. Axtell. 1996. *Growing Artificial Societies. Social Science from the Bottom Up*. Boston: MIT Press.
+
+to further expand their agent-based modelling and NetLogo skills.
+The learner explores how to set up more complex interactions between agents and the environment.
+Furthermore, the principles of setting up good experiments and validating models is explained.
 This tutorial consists of 8 lessons.
-This tutorial intends to guide the learner from the (intermediate and) advanced level to the highly specialised level of proficiency.
-In this tutorial more complex agent-environment interaction is further learned, including ways to visualize the environment and how to give more agency to patches.
+This tutorial intends to solidify the learner's understanding at the intermediate level of proficiency by introducing yet more complex agent-environment interaction, alternative ways to visualize the environment and modelling spatial dynamics.
 Creating toy landscapes is central to lesson 5.
-The learner also learns the basics of setting up experiments in NetLogo and how to use monitors and flexible plots to understand the results.
-The validation of agent-based models is an important concept of designing a good experiment.
-The finale step, is to compare simulation results with the archaeological record.
-The last lesson is aimed at learning syntax to make lists more dynamic and how to refactor code.
+The learner dives into the topic of experiment design at a practical level but also with regards to the scientific principles, such as validation.
+Thus they are guided through the process of setting up experiments in NetLogo and collecting results using monitors and plots.
+The topic of validation of agent-based models is critical for most archaelogical applications and thus several ways to compare simulation results with the archaeological record are explored.
 
-In the final and fifth tutorial the learner will learn more about how to actually incorporate agent-based modelling in research.
-This tutorial focusses less on programming in NetLogo and more on the model development process.
-The learner will learn about the different phases in modelling and how to export data from models.
+In the final and fifth tutorial the participant learn smore about how to incorporate agent-based modelling in archaeological research.
+This tutorial focuses less on programming in NetLogo and more on the principle and standards of computer-based modelling.
+The learner learns about the different phases of the model development process.
 This tutorial consists of 5 lessons.
-This tutorial touches upon various levels, but mainly between the intermediate and highly specialised level of proficiency.
+This tutorial summarizes the material delivered in the previous lessons into one coherent framework consistent with an intermediate level of proficiency.
 This tutorial approaches more theoretical aspects in a practical environment.
-The learner leans about the model development process starting with the conceptual phase of the model development.
-Good and bad research questions for modelling are explained and the difference between them.
-The learner needs to understand how to pick the right modelling technique.
-The learner should understand the importance of properly conceptualizing a model before starting the technical phase of the model development.
-In this phase parametrisation, designing experiments and the analysis and interpretation of the models are important.
-Learning about the dissemination phase of the model development is aimed at understanding (the importance of) publishing models and replication.
-In addition, learning how to export results from NetLogo using basic export primitives is explained.
-The BehaviorSpace is explained to enable the learners to analyse the output of models.
+The learner revisits all phases of the model development process starting with the conceptual phase and finishing with the analysis and interpretation of results.
+There is a particular focus on discussing what kind of research questions are suitable for modelling and how to pick the right modelling technique.
+The importance of properly conceptualizing a model before starting the technical phase of the model development is highlighted as it is a common issue for less experienced modellers.
+The experiment design phase, including parametrisation, validation and the analysis and interpretation of the models is explained at length since it often prove challenging for students.
+On a practical side, the BehaviorSpace - NetLogo's experiment environment - is explained to enable the learners to batch run their models.
+Finally the dissemination phase of the model development is also included to ensure that the standards in model publication are widely known.
+Finally, the learners explores the fundamental importance of replication for advancing scientific knowledge.
 
@@
 A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants).
-The respondents came from many different countries all over the world (see Figure \@ref(fig:nationality).
-The number of female respondents slightly outnumbers the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \@ref(fig:gender-age)).
-The ages of the respondents ranged from below 20 to over 70, reaching people in different age classes, although the majority was aged between 20 and 40.
+The respondents came from !!!
+COMMENT PLS ADD THE NUMBER countries that represented almost all continents (see Figure \@ref(fig:nationality).
+There was a clear skew towards participants from !!!
+COMMENT what part of the world?
+Europe?
+South America.
+The number of women and men in our sample were close to equal (see Figure \@ref(fig:gender-age)).
+The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years old indicating that the audience was predominantly composed of early career archaeologists.
 It seems that the different events also had a slightly different distribution of both gender and age.
-For example, more older people attended the workshop at the CAA conference in April 2023 and more male people were present during the workshop at the Reuvensdagen in November 2023.
+For example, more older people attended the workshop at the CAA conference in April 2023 and more men were present during the workshop at the Reuvensdagen in November 2023.
+COMMENT - in the last sentence, are these significant differences?
+And if so, please provide the possible reasons for these differences.
 
@@
 
-Most of the respondents had several skills with computer applications in archaeology (see Figure \@ref(fig:computer-skills)), and many of them did have some knowledge of ABM, although a large group was completely new to the subject.
-The majority had never applied ABM before participating in the workshops with 14 having applied ABM.
-It is interesting to note that many respondents did know what kind of software was available for ABM (see Figure \@ref(fig:abm-knowledge)).
+Most of the respondents reported having a level of computer literacy already including particular computational skills common in archaeology, for example !!!
+COMMENT pls add the most popular ones (see Figure \@ref(fig:computer-skills)), and !!!
+COMMENT PLS insert the % number here had some prior knowledge of ABM, while !!!
+COMMENT PLS insert the % number here was completely new to the subject.
+The majority !!!
+COMMENT PLS insert the % number here had never applied ABM to their research before participating in the workshops with only 14 participants who had experience of developing archaeological ABM.
+It is interesting to note that many respondents (!!!
+COMMENT PLS insert the % number here) did know what kind of software was available for ABM (see Figure \@ref(fig:abm-knowledge)).
 
@@
 As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before.
-The respondents were also asked how they rated the available knowledge on ABM (Figure \@ref(fig:available-theory)).
-While the largest group of them had no opinion on the subject, a very large proportion of the responded answered that they rated the available theory on ABM as limited and a smaller group as sufficient.
-A minority rated the available theory on ABM as more than sufficient.
-This clearly shows the need for better educational material.
+The respondents were also asked how they rated the availability of ABM educational resources (Figure \@ref(fig:available-theory)).
+While *!!!
+COMMENT PLS insert the % number here* of them had no opinion on the subject, a !!!
+COMMENT PLS insert the % number here proportion of the responded answered that they rated the available theory on ABM as limited and only a smaller group !!!
+COMMENT PLS insert the % number here as sufficient or as more than sufficient.
+This clearly shows the need for more and better educational material.
 
@@
 During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials.
-Some particpants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general.
+Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general.
 Individual progress varied depending on skills and preferences of the participants.
-While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later instalments.
+While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later installments.
 
@@
 
-A large proportion of the participants of the workshops gave us information using the survey after the workshops (171 of 368 participants), a similar proportion as those responding to the survey before the workshop.
+A similar proportion of the participants of the workshops gave us information after the workshops (171 of 368 participants).
 
@@
 
-The majority of the respondents were enthusiastic about the teaching material, with the majority rating it as excellent or good.
-The teachers were even rated better than the teaching material, with more respondents rating them as excellent (see Figure \@ref(fig:rating-teaching)).
+*!!!
+COMMENT PLS insert the % number here* of the respondents were enthusiastic about the teaching material, with the majority rating it as excellent or good.
+The teachers were rated even higher than the teaching material, with more respondents giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).
 
@@
 
-Two more open questions were asked to the respondents.
-The first was aimed at learning what aspects the participants liked in the tutorials, and the other was aimed at getting feedback to improve the tutorials.
+Two open questions were asked to the respondents: *!!! COMMENT PLS spell out the questions*
 
@@
 The easy and intuitive introduction into ABM and NetLOGO were also mentioned quite often.
-This was also observed during the workshops by us, most participants were working in the own pace and did not need much help.
-From a didactic point of view, the self-paced going through the tutorials is very efficient for teachers.
+This was also observed during the workshops as most participants were working in the own pace and did not need much help.
+From a didactic point of view, the self-guided and self-paced way of going through the tutorials is very efficient for teachers and limits the frustration of students that may otherwise go too fast or fall behind the class when the tempo is dictated by the teacher.
 
 We also received feedback on possible improvements.
-For the first workshops we mostly received feedback on the number of bugs.
-During the CAA-conference in Amsterdam we did the first test of the tutorials and we also instructed the participants that this was work in progress.
-The number of references to bugs reduced to zero by the end and people started providing other feedback, for example the wish to see more examples or expanding the tutorials more.
-Some wanted more cooperation, which is possible, but harder for some online events, although we provided break-out rooms.
-It is also interesting to note that over time more and more people did not see any room for improvement.
+During the first, big event CAA-conference in Amsterdam in April 2023 we did the first test of the tutorials and we also instructed the participants that this was work in progress.
+For this and the following few workshops we mostly received feedback about bugs causing the tutorial to crash or creating unexpected behaviour.
+The number of references to bugs reduced to zero by the end of the project and participants started providing other feedback, for example the wish to see more examples or expanding the tutorials further.
+Some wanted to introduce more cooperation between participants, which we did not consider primarily aiming the tutorials for individual work.
+A more cooperative aspect of the education materials is possible, even if harder for online events, and we hope that a wider classroom use of the ABMA materials may deliver solutions.
+It was also interesting to note that over time more and more people did not see any additional room for improvement indicating that the level, content and format of the tutorials met their needs.
 
-The majority of the respondents wanted to apply ABM in the future for research (144 of the 164 that answered this question, see Figure \@ref(fig:future-abm)).
-Some of the respondents were already using ABM.
-A large group was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities.
-Many respondents also shared the context to which they were thinking to apply ABM.
-Various participants mentioned movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms.
+We also asked the participants about their intentions in terms of ABM and their professional life.
+The majority of the respondents - 88%, wanted to apply ABM in the future in their research (Figure \@ref(fig:future-abm)).
+Some of the respondents, *!!!
+COMMENT PLS insert the % number here,* were already using ABM in their work.
+A large group *!!!
+COMMENT PLS insert the % number here* was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities.
+Many respondents also shared the context to which they were thinking to apply ABM mentioning topics such as movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms.
 Others thought of demography, social networks, migration or settlement distributions patterns.
-The natural environment and the interaction with humans in the past was also mentioned by some, and often in relation to GIS or how to replace GIS with ABM.
+The natural environment and the interaction with humans in the past was also mentioned frequently, often in relation to GIS and the question of how complimentary these two computational techniques are.
 The archaeological periods that the participants were interested in were very diverse, ranging from the Paleolithic to the Medieval period.
@@
 Problem solving [@europeancommissionjointresearchcentre2022].
-In these competence area's various competences are developed towards an advanced or even specialised level.
+In these competence area's various competences are developed towards an advanced or specialised level.
 
@@
 Information and data literacy: the tutorials mainly address 1.2 Evaluating Data, Information And Digital Content and 1.3 Managing Data, Information And Digital Content.
-The users learn about working with data, being critical of their digital content and how to model and manage the data in the context of an ABM.
+In the ABMA educational materials, users learn about working with data, being critical of their digital content and how to model and manage the data in the context of an ABM.
 For tutorial 2 the user is expected to be able to find information using the NetLogo(Web) dictionary (<http://ccl.northwestern.edu/netlogo/bind/> or <http://ccl.northwestern.edu/netlogo/docs/dictionary.html>).
@@
 In the process of learning ABM the users are constantly 3.1 Developing Digital Content and 3.2 Integrating And Re-Elaborating Digital Content.
-The material does not touch upon 3.3 Copyright And Licences, but we assume that users have a general understanding.
+The material does not touch upon 3.3 Copyright And Licences, as they fall out of the scope of this project.
 A very important aspect of learning ABM is 3.4 Programming.
-The users are exploring the possibilities and chances of programming an ABM.
-The users are expected to learn both syntax and more general concepts such as modular code development, loops, lists and commenting and documenting code.
+The users are guidend through the process of developing the necessary programming skills to build a simulation.
+While the programming language used - NetLogo - is specific to ABM, its syntax follow general programming patterns and the learning is thus transferable to other programming languages.
+The users are expected to learn both syntax and more general computational concepts such as modular code development, testing and documenting code.
 
 Competence Area 5.
-Problem solving is very important and is closely tied to 3.4 Programming, since programming involves a lot of problem solving.
-In addition, the creation of a highly complex ABM consists of problem solving all the time, including debugging.
+Problem solving is very important and is closely tied to 3.4 Programming, since programming involves a lot of problem solving but also the general principles of model development that have a prominent place in the educational materials.
+In addition, the creation of a highly complex ABM consists of problem solving all the time, including development of research questions, conceptualisation of theories, experiment design and validating against data - all of which require creative abilities and high level understanding of the scientific process.
 All competencies are addressed: 5.1 Solving Technical Problems, is common when writing code, learning about coding, error handling and going from pseudocode to real code.
-The competence 5.2 Identifying Needs And Technological Responses, is a central issue of the course, while 5.3 Creatively Using Digital Technology is an important aspect of learning to think in models and develop models themselves.
-In tutorial 2 various aspects of 5.4 Identifying Digital Competence Gaps, are relevant, since this tutorial will force the user to remember code and commands and to assess their knowledge and skills.
+The competence 5.2 Identifying Needs And Technological Responses, is a central issue of the course, while 5.3 Creatively Using Digital Technology is an important aspect of so-called model thinking necessary to develop robust and insightful models.
+In tutorial 2 various aspects of 5.4 Identifying Digital Competence Gaps, are relevant, since this tutorial force the user to assess their knowledge and skills.
 
@@
 Anyone interested can download the tutorials and their source code.
-To increase the accessibility and visibility of the tutorials, a website has been launced ([https://ABMArchaeologists.github.io/ABMA_Website](https://ABMArchaeologists.github.io/ABMA_Website/#/){.uri}).
-On this website, the tutorials can be followed without the need to download any software.
+To increase the accessibility and visibility of the tutorials, they are also hosted on a static website ([https://ABMArchaeologists.github.io/ABMA_Website](https://ABMArchaeologists.github.io/ABMA_Website/#/){.uri}) can be followed without the need to download any software.
 
 Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University [@scherjon2019] and external resources people can use to learn agent-based modelling.
-The website was mainly developed by one of the student groups from Saxion University of Applied Sciences, see the acknowledgements.
+The website was mainly developed by a student group from Saxion University of Applied Sciences, see the acknowledgements.
 
@@
 
-Funding acknowledgements for the publication: The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382).
+Funding acknowledgements: the publication was financed by The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382); IR was supported by the European Union’s Horizon 2020 Research and
+Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513 and the
+Aarhus University Research Foundation.
 

### 2024-04-25 paper.Rmd
updates

@@
 
-While Agent-Based Modelling importance in archaeological practice is steadily growing \_\_\_ !!!please add the following reference here: <https://www.researchgate.net/publication/376410267_Bibliometric_Analysis_of_Agent-Based_Simulation_in_Archaeology_People_Topics_and_Future_Prospects> \_\_\_, there are significant bariers to its wider adoption.
+While Agent-Based Modelling importance in archaeological practice is steadily growing [@romanowska2023] there are significant bariers to its wider adoption.
 One of the sectorial problems is the access to vocational training required to gain the necessary computer programming and modelling skills.
 A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 76% of modellers were primarily self-taught with some degree of secondary support i.e.
-27% relied on peer support, only 7% had any sort of formal training at a workshop or summer school while 14% received training as part of a degree.
+27% relied on peer support, only 7% had any sort of formal training at a workshop or summer school while only 14% received training as part of a degree.
 
@@
 
-[!!!COMMENT: this sentence should go into the acknowledgements] This project, Agent-Based Modelling for Archaeologists (<https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054>), was delivered with the support of the Erasmus+ programme of the European Union.
+This project, Agent-Based Modelling for Archaeologists, was delivered with the support of the Erasmus+ programme of the European Union.
 
@@
 
--   Development of the framework for skills acquisition aligned with the Digital Skills Passport !!!\_\_ add ref here!
+-   Development of the framework for skills acquisition aligned with the Digital Skills Passport [DOUG: reference]?
 
@@
 3.
-!!!
-[Comment: This is very unclear.
-Currently it reads as if taking the course will make people emigrage to other countries.
-I don't think that is in any way true.
-Also, most readers would not know what European or Sectoral Framework we are referring to] improved transnational mobility through matching of the skills being delivered with European and Sectoral Frameworks, leading to individual workers being better enabled to demonstrate their skills to employers across Europe;
+improved transnational mobility through matching of the skills being delivered with European and Sectoral Frameworks for education, leading to individual workers being better enabled to demonstrate their skills to employers across Europe;
 
@@
 
-: Events at which the tutorials were delivered and tested.
+: Conferences, events and situations were workshops were held at which the tutorials were tested.
 
@@
 The diverse groups of students contributed new ideas, developing educational material, testing the tutorials and developing a style for the website and other materials.
-!!!
-COMMENT- this section needs to be more specific.
-Why did these student work on the project?
-At what capacity?
-What did they contribute?
-etc
 
@@
 
-## ABM tutorials
+## Final tutorials
 
@@
 
-In tutorial 2 the user learns the basics of programming in NetLogo and develops their first simulation based on a previously published model of the Out of Africa dispersal of *Homo sapiens !!!
-Pls insert the following reference*
-
-Young, David A., and Robert L. Bettinger. 1992. “The Numic Spread: A Computer Simulation.” *American Antiquity* 57 (1): 85–99. <https://doi.org/10.2307/2694836>.
-
-.
-The learner works through the basics of NetLogo syntax and learn how to set up a simulation and visualize its outcomes.
+In tutorial 2 the user will learn the basics of NetLogo by making your their simulation on the Out of Africal dispersal of *Homo sapiens* [@young1995].The learner works through the basics of NetLogo syntax and learn how to set up a simulation and visualize its outcomes.
 This tutorial consists of 9 lessons.
@@
 
-In the third tutorial, the learner builds a simple trade model, again based on a published simulation !!! pls insert the following reference:
-
-Romanowska, Iza.
-2018.
-“‪Using Agent-Based Modelling to Infer Economic Processes in the Past‬.” In *‪Cuantificar Las Economías Antiguas: Problemas y Métodos*, Publicacions i Edicions de la Universitat de Barcelona, 107–18.
-Barcelona: Universitat de Barcelona.
-<https://scholar.google.com/citations?view_op=view_citation&hl=en&user=tdcyWfoAAAAJ&citation_for_view=tdcyWfoAAAAJ:YOwf2qJgpHMC>.
-
+In the third tutorial, the learner builds a simple trade model, again based on a published simulation [@romanowska2018].
 Here learners are expanding their NetLogo programming skills with more complexity added to such syntactic structures as loops, lists and reporters.
@@
 
-In tutorial 4 the learner works with a seminal Sugarscape model !!! pls insert the following ref.
-
-Epstein, Joshua M., and Robert L. Axtell. 1996. *Growing Artificial Societies. Social Science from the Bottom Up*. Boston: MIT Press.
-
-to further expand their agent-based modelling and NetLogo skills.
+In tutorial 4 the learner works with a seminal Sugarscape model [@epstein1996] to further expand their agent-based modelling and NetLogo skills.
 The learner explores how to set up more complex interactions between agents and the environment.
@@
 A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants).
-The respondents came from !!!
-COMMENT PLS ADD THE NUMBER countries that represented almost all continents (see Figure \@ref(fig:nationality).
-There was a clear skew towards participants from !!!
-COMMENT what part of the world?
-Europe?
-South America.
-The number of women and men in our sample were close to equal (see Figure \@ref(fig:gender-age)).
+The respondents came from at least 40 countries that represented almost all continents (see Figure \@ref(fig:nationality).
+There was a clear skew towards participants from Europe.
+The number of female respondents slightly outnumbers the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \@ref(fig:gender-age)).
 The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years old indicating that the audience was predominantly composed of early career archaeologists.
@@
 For example, more older people attended the workshop at the CAA conference in April 2023 and more men were present during the workshop at the Reuvensdagen in November 2023.
-COMMENT - in the last sentence, are these significant differences?
-And if so, please provide the possible reasons for these differences.
+This might be due the the differences of audiences at the conferences.
+For the Reuvensdagen the number of participants of the workshop was relatively low and this variance might be due to chance.
 
@@
 
-Most of the respondents reported having a level of computer literacy already including particular computational skills common in archaeology, for example !!!
-COMMENT pls add the most popular ones (see Figure \@ref(fig:computer-skills)), and !!!
-COMMENT PLS insert the % number here had some prior knowledge of ABM, while !!!
-COMMENT PLS insert the % number here was completely new to the subject.
-The majority !!!
-COMMENT PLS insert the % number here had never applied ABM to their research before participating in the workshops with only 14 participants who had experience of developing archaeological ABM.
-It is interesting to note that many respondents (!!!
-COMMENT PLS insert the % number here) did know what kind of software was available for ABM (see Figure \@ref(fig:abm-knowledge)).
+Most of the respondents reported having a level of computer literacy already including particular computational skills common in archaeology, for example working with word processors, GIS and spreadsheets (see Figure \@ref(fig:computer-skills)), and 111 of the respondents had some prior knowledge of ABM, while 59 were completely new to the subject.
+The majority (158) had never applied ABM to their research before participating in the workshops with only 14 participants who had experience of developing archaeological ABM.
+It is interesting to note that many respondents (108) did know what kind of software was available for ABM (see Figure \@ref(fig:abm-knowledge)).
 
@@
 As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before.
-The respondents were also asked how they rated the availability of ABM educational resources (Figure \@ref(fig:available-theory)).
-While *!!!
-COMMENT PLS insert the % number here* of them had no opinion on the subject, a !!!
-COMMENT PLS insert the % number here proportion of the responded answered that they rated the available theory on ABM as limited and only a smaller group !!!
-COMMENT PLS insert the % number here as sufficient or as more than sufficient.
+The respondents were also asked how they rated the available knowledge on ABM (Figure \@ref(fig:available-theory)).
+While 64 of them had no opinion on the subject, a proportion of the respondents (28) answered that they rated the available theory on ABM as limited and only a small group (39) as sufficient or better.
 This clearly shows the need for more and better educational material.
@@
 
-*!!!
-COMMENT PLS insert the % number here* of the respondents were enthusiastic about the teaching material, with the majority rating it as excellent or good.
+of the respondents were enthusiastic about the teaching material, with the majority rating it as excellent or good.
 The teachers were rated even higher than the teaching material, with more respondents giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).
+
+The majority of the respondents (153) were enthusiastic about the teaching material, with the majority rating it as excellent or good.
+The teachers were even rated better than the teaching material, with more respondents (165) giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).
 
@@
 
-Two open questions were asked to the respondents: *!!! COMMENT PLS spell out the questions*
+Two more open questions were asked to the respondents.
+The first was aimed at learning what aspects the participants liked in the tutorials (What aspects did you like most?), and the other was aimed at getting feedback to improve the tutorials (What would you like to see changed or what could be better/different?).
 
@@
 The majority of the respondents - 88%, wanted to apply ABM in the future in their research (Figure \@ref(fig:future-abm)).
-Some of the respondents, *!!!
-COMMENT PLS insert the % number here,* were already using ABM in their work.
-A large group *!!!
-COMMENT PLS insert the % number here* was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities.
+Some of the respondents were already using ABM in their work.
+A large group (around 20-30%) was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities.
 Many respondents also shared the context to which they were thinking to apply ABM mentioning topics such as movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms.
@@
 Problem solving [@europeancommissionjointresearchcentre2022].
-In these competence area's various competences are developed towards an advanced or specialised level.
+In these competence area's various competences are developed towards an advanced or even specialised level.
+A more extensive description of the Digital Competence Framework for Citizens in relation to Agent-Based Modelling for archaeologists is also available [@visser2024].
 
@@
 
+This project was undertaken with the support of the Erasmus+ programme of the European Union (<https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054>).
+
 We would like to thank the student groups at Saxion University of Applied Sciences that worked with us:
@@
 
-Funding acknowledgements: the publication was financed by The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382); IR was supported by the European Union’s Horizon 2020 Research and
-Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513 and the
-Aarhus University Research Foundation.
+Funding acknowledgements: the publication was financed by The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382); IR was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513 and the Aarhus University Research Foundation.
 

### 2024-04-25 paper.tex
first batch of edits
I made some edits and added a couple of comments. Search for "!!!" to quickly jump to a comment.
updates

@@
 \author{true \and true \and true \and true \and true \and true \and true \and true \and true}
-\date{April 23, 2024}
+\date{April 25, 2024}
 
@@
 
-While Agent-Based Modelling is a predictive tool that is in high demand from archaeologists, there are significant sectoral problems with access to vocational training for this skill.
+Computational modelling has become a standard research technique across most scientific disciplines including archaeology.
+Among the many computational modelling techniques the discipline has found agent-based modelling to be particularly well suited to the questions asked by archaeologists.
+Agent-based modelling is a simulation technique where autonomous agents interact with each other and with the environment around.
+Their aggregated behaviour gives rise to system-level outcomes thus enabling researchers to understand how \emph{micro} (individual) \emph{mechanisms} lead to \emph{macro} (system-level) \emph{patterns}.
+
+While Agent-Based Modelling importance in archaeological practice is steadily growing (Romanowska and Scherjon 2023) there are significant bariers to its wider adoption.
+One of the sectorial problems is the access to vocational training required to gain the necessary computer programming and modelling skills.
 A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 76\% of modellers were primarily self-taught with some degree of secondary support i.e.~27\% relied on peer support, only 7\% had any sort of formal training at a workshop or summer school while only 14\% received training as part of a degree.
 
-A consortium of Landward Research Teoranta, Aarhus Universitet, Universiteit Leiden and Saxion University of Applied Sciences recognised this need, and sought to address and overcome this training deficiency through the development and delivery of a project to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.
+A consortium of Landward Research Teoranta, Aarhus Universitet, Universiteit Leiden and Saxion University of Applied Sciences recognised this need, and sought to address the limited opportunities that archaeologists have to access high quality training through a project aiming to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.
 
-This project, Agent-Based Modelling for Archaeologists (\url{https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054}), was delivered with the support of the Erasmus+ programme of the European Union.
+This project, Agent-Based Modelling for Archaeologists, was delivered with the support of the Erasmus+ programme of the European Union.
 
@@
 These OERs were to be HTML and JavaScript based so they could be used on any device with a browser, even if it is not connected to the internet.
-These are hands-on practical exercises, taking learners through the process of creating the digital models, and modular so they can be independent or incorporated into other teaching tools like learning management systems.
-They are deliberately `method agnostic' so they can be incorporated into any tools from self-teaching to MOOCs, and are supported by the creation of support materials to encourage use of the OERs.
+The full programme consists of five hands-on practical tutorials, taking learners through the process of creating the digital models.
+The modular architecture ensures that they can be independent or incorporated into other teaching tools like learning management systems.
+They are also deliberately `method agnostic' so they can be incorporated into any tools from self-teaching to MOOCs, and are supported by a set of support materials to encourage use of the OERs.
 
@@
 \item
+  Development of the framework for skills acquisition aligned with the Digital Skills Passport {[}DOUG: reference{]}?
+\item
   Promotion through a series of multiplier events.
@@
 \item
-  increased level of digital competence, specifically in advanced computer modelling skills that are highly transferable to other industries (computer programming and simulation development);
+  a surge in professionals who can build, apply and critique archaeological simulation thus opening new research avenues for the discipline as a whole;
 \item
-  improved transnational mobility through matching of the skills being delivered with European and Sectoral Frameworks, leading to individual workers being better enabled to demonstrate their skills to employers across Europe;
+  increased level of digital competence, specifically in advanced computer skills that are highly transferable to other industries.
+This includes technical skills such as programming but also ``soft'' skills related to computational thinking;
+\item
+  improved transnational mobility through matching of the skills being delivered with European and Sectoral Frameworks for education, leading to individual workers being better enabled to demonstrate their skills to employers across Europe;
 \item
@@
 \item
-  increased quality of education and training in the areas of ABM that are free and open;
+  increased quality of training in Agent-based Modelling that is free and open;
 \item
@@
 
-During the project we developed online tutorials.
-The tutorials were first designed using storyboards which were later converted into tutorials using NetLOGO Web (\url{https://www.netlogoweb.org/launch}) and Javascript ().
-The tutorials were developed as a team and we worked together using GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials}).
+The project team developed five online tutorials.
+The tutorials were first designed using storyboards and later converted into interactive guided walk-throughs using NetLOGO Web (\url{https://www.netlogoweb.org/launch}) and Javascript and hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials}).
 During the project, initial testing was done by all authors and issues in GitHub were used to tackle bugs.
 
-The learning and teaching material for ABM for archaeologists assumes that the users have completed secondary education.
-It is suitable for archaeologists with at least some background in archaeology, preferably after the first year of a Bachelor study in Archaeology (above EQF levels 4 or 5).
+The ABMA learning materials assume that participants have learning skills at least at a secondary education level.
+It is suitable for participants with at least some background in archaeology, such as those typical among students after the first year of a Bachelor study in Archaeology, i.e., above EQF levels 4 or 5.
 In addition, the tutorials are aimed at professionals working in archaeology who have some experience with computer applications.
 We also assume that learners are proficient in English and have at least the Reading B1 level, but Reading B2 is recommended (Council of Europe 2020).
-The tutorials are developed aiming to develop learners' skills in the following competence area's: 1.
+The tutorials were also developed with a secondary aim to develop learners' general skills in the following competence area's: 1.
 Information and data literacy, 2.
@@
 
-After initial debugging by the project team, the tutorials were tested `in the wild' during various online and in-person conferences.
-The in-person conferences were attended by several people from the project team.
-During the online events we were always present with several people and enabled participants to go to break-out rooms if they needed any help or wanted to discuss things.
-In addition, we also had students working with the tutorials at Saxion University of Applied Sciences and Leiden University.
-We We obtained structured feedback from the participants and used it to improve the tutorials prior to the next test.
-The feedback was obtained from two surveys using Qualtrics.
-The participants of the events were asked to answer questions of the first survey before participating.
-We used the answers to understand what the background of the participants was and to get an estimate of their level of knowledge in relation to ABM (see appendix \ldots{} for the questions).
+After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events all of which were run by members of the project team.
+Recognising the difficulty of delivering practical training in an online environment during the online events several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things.
+Throughout the project we sought feedback from the participants and used it to improve the tutorials prior to the following events.
+The feedback was obtained through two surveys using Qualtrics.
+The participants of the events were asked to respond to two surveys: one before the start of the event and one completed after the end of the event.
+The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix \ldots{} for the questions).
 At the end of the event the participants were asked to answer a second survey.
@@
 At other events no registration was possible or necessary and the pre-workshop survey was given at the start of the event.
-The post-workshop survey was distributed using QR-codes or links at the end of the workshop during the event.
-After each event we improved the tutorials based on the received feedback.
+The post-workshop survey was distributed using QR-codes or links at the end of the event.
 
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2256}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1880}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2180}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1805}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1880}}@{}}
 \caption{Conferences, events and situations were workshops were held at which the tutorials were tested.}\tabularnewline
@@
 
-The final tutorals are available for download at \url{https://github.com/ABMArchaeologists/ABMA_tutorials} and are citeable (Rocks-Macqueen et al.
+The final set of ABMA tutorals are available for download at \url{https://github.com/ABMArchaeologists/ABMA_tutorials} and are citeable (Rocks-Macqueen et al.
 2024).
@@
 
-Each tutorials consists of a different number of lessons that guide the learner in a self-pace manner through the lessons.
+Each tutorials consists of a number of individual lessons, each of which guides the learner in a self-pace manner through an increasingly challenging set of exercises that build the skills and knowledge necessary to become a confident agent-based modeller.
 
-In the first tutorial the user will learn what simulations and agent-based models are and how they can help in archaeological research.
+In the first tutorial the user learn what simulations and agent-based models are and how they can help in archaeological research.
 This tutorial consists of 4 lessons.
 The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific.
-Various concepts related to ABM are introduced and the first concepts within NetLogo are explained.
-The third lesson aims to explain how ABMs are used in archaeological research.
+Various concepts related to ABM, such as the definitions of model and simulation, are introduced and the participants make their first steps in NetLogo syntax.
+The third lesson explains how ABMs are used in archaeological research.
 The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode.
-This tutorial build a foundation for working in NetLogo and with ABM.
-All activity stays within the intermediate level, although higher levels might be touched upon.
+This tutorial builds a foundation for working in NetLogo and with ABM.
+All activity stays within the entree level of difficulty, although some higher level concepts are lightly touched upon.
 
-In tutorial 2 the user will learn the basics of NetLogo by making your their simulation on the Out of Africal dispersal of homo sapiens.
-The learner will work with basic NetLogo syntax and learn how to set up a simulation and visualize its outcomes.
+In tutorial 2 the user will learn the basics of NetLogo by making your their simulation on the Out of Africal dispersal of \emph{Homo sapiens} (Young and Bettinger 1995).The learner works through the basics of NetLogo syntax and learn how to set up a simulation and visualize its outcomes.
 This tutorial consists of 9 lessons.
-This tutorial intends to guide the learner from the intermediate to the advanced level of proficiency.
+This tutorial intends to guide the learner from the introduction to the beginner level of proficiency.
 The learning curve is relatively gentle.
-This is achieved by learning to work with the NetLogo web interface.
-The primitives are introduced and initialization phase of a simple ABM is gone through.
-The world within NetLogo's world (including dimensions, coordinates, origin) and how to alter it are explained and experience in a hands-on approach.
-The learners learn about simple simulation loops and how to use primitives.
-The advanced level is achieved with the custom procedures and variables.
-In addition, if-statements are introduced and more complex versions thereof.
-Exporting information using plots is also explained.
+This is achieved by working with the NetLogo web interface that enables highlighting specific words and buttons and giving visual cues as to how to proceed.
+The tutorial intoduces several most commonly used built-in functions in Netlogo (called \emph{primitives}) are introduces the general structure of all models cosisting of the initialization phase and the main simulation loop.
+It walks the participants through te most important NetLogo settings, such as dimensions, coordinates and the origin point of the grid on which the simulation runs.
+The learners gain their first stripes in programming by using simple loops and functions as well as conditional statements.
+The next level is achieved by introducing the topic of function definition and different types of variables.
+Exporting data produced by the model through plots is also explained.
 
-In the third tutorial, the learner will build a simple trade model, picking up more advanced NetLogo coding along the way, such as loops, lists and reporters.
-The learner will be introduced to some techniques like modular code development and debugging which will become important with this increased coding complexity.
+In the third tutorial, the learner builds a simple trade model, again based on a published simulation (Romanowska 2018).
+Here learners are expanding their NetLogo programming skills with more complexity added to such syntactic structures as loops, lists and reporters.
+The learner is introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity.
 This tutorial consists of 7 lessons.
-This tutorial intends to guide the learner from the intermediate to the advanced and highly specialised level of proficiency.
-This is achieved by improving development through using modular code, pseudocode and annotating the code.
-In addition, custom agent breeds are introduced.
-Visualization with labels and reporters, plots and monitors are learned.
-The use of loops is further explained.
-Debugging is learned in lesson 7.
-This can be very advanced, since it involves knowledge of the NetLogo-language combined with problem solving of unexpected behaviour.
+This tutorial intends to guide the learner from the beginner to the intermediate level of proficiency.
+This is achieved by introducing standards of code development such as modular code, pseudocode, debugging and annotating the code.
+In addition, the tutorial introduces custom agent breeds, visualization with labels and reporters, plots and monitors.
+This takes the learner to a more advanced level, since it develops knowledge of the programming language combined with diagnosing unexpected behaviour and problem solving.
 
-In tutorial 4 the learner will work with Sugarscape simulations to further expand their ABM and NetLogo skills.
-The learner will learn how to set up more complex interactions between agents and the environment.
-Furthermore, the basics of setting up good experiments and validating models will be explained.
+In tutorial 4 the learner works with a seminal Sugarscape model (Epstein and Axtell 1996) to further expand their agent-based modelling and NetLogo skills.
+The learner explores how to set up more complex interactions between agents and the environment.
+Furthermore, the principles of setting up good experiments and validating models is explained.
 This tutorial consists of 8 lessons.
-This tutorial intends to guide the learner from the (intermediate and) advanced level to the highly specialised level of proficiency.
-In this tutorial more complex agent-environment interaction is further learned, including ways to visualize the environment and how to give more agency to patches.
+This tutorial intends to solidify the learner's understanding at the intermediate level of proficiency by introducing yet more complex agent-environment interaction, alternative ways to visualize the environment and modelling spatial dynamics.
 Creating toy landscapes is central to lesson 5.
-The learner also learns the basics of setting up experiments in NetLogo and how to use monitors and flexible plots to understand the results.
-The validation of agent-based models is an important concept of designing a good experiment.
-The finale step, is to compare simulation results with the archaeological record.
-The last lesson is aimed at learning syntax to make lists more dynamic and how to refactor code.
+The learner dives into the topic of experiment design at a practical level but also with regards to the scientific principles, such as validation.
+Thus they are guided through the process of setting up experiments in NetLogo and collecting results using monitors and plots.
+The topic of validation of agent-based models is critical for most archaelogical applications and thus several ways to compare simulation results with the archaeological record are explored.
 
-In the final and fifth tutorial the learner will learn more about how to actually incorporate agent-based modelling in research.
-This tutorial focusses less on programming in NetLogo and more on the model development process.
-The learner will learn about the different phases in modelling and how to export data from models.
+In the final and fifth tutorial the participant learn smore about how to incorporate agent-based modelling in archaeological research.
+This tutorial focuses less on programming in NetLogo and more on the principle and standards of computer-based modelling.
+The learner learns about the different phases of the model development process.
 This tutorial consists of 5 lessons.
-This tutorial touches upon various levels, but mainly between the intermediate and highly specialised level of proficiency.
+This tutorial summarizes the material delivered in the previous lessons into one coherent framework consistent with an intermediate level of proficiency.
 This tutorial approaches more theoretical aspects in a practical environment.
-The learner leans about the model development process starting with the conceptual phase of the model development.
-Good and bad research questions for modelling are explained and the difference between them.
-The learner needs to understand how to pick the right modelling technique.
-The learner should understand the importance of properly conceptualizing a model before starting the technical phase of the model development.
-In this phase parametrisation, designing experiments and the analysis and interpretation of the models are important.
-Learning about the dissemination phase of the model development is aimed at understanding (the importance of) publishing models and replication.
-In addition, learning how to export results from NetLogo using basic export primitives is explained.
-The BehaviorSpace is explained to enable the learners to analyse the output of models.
+The learner revisits all phases of the model development process starting with the conceptual phase and finishing with the analysis and interpretation of results.
+There is a particular focus on discussing what kind of research questions are suitable for modelling and how to pick the right modelling technique.
+The importance of properly conceptualizing a model before starting the technical phase of the model development is highlighted as it is a common issue for less experienced modellers.
+The experiment design phase, including parametrisation, validation and the analysis and interpretation of the models is explained at length since it often prove challenging for students.
+On a practical side, the BehaviorSpace - NetLogo's experiment environment - is explained to enable the learners to batch run their models.
+Finally the dissemination phase of the model development is also included to ensure that the standards in model publication are widely known.
+Finally, the learners explores the fundamental importance of replication for advancing scientific knowledge.
 
@@
 A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants).
-The respondents came from many different countries all over the world (see Figure \ref{fig:nationality}.
+The respondents came from at least 40 countries that represented almost all continents (see Figure \ref{fig:nationality}.
+There was a clear skew towards participants from Europe.
 The number of female respondents slightly outnumbers the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \ref{fig:gender-age}).
-The ages of the respondents ranged from below 20 to over 70, reaching people in different age classes, although the majority was aged between 20 and 40.
+The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years old indicating that the audience was predominantly composed of early career archaeologists.
 It seems that the different events also had a slightly different distribution of both gender and age.
-For example, more older people attended the workshop at the CAA conference in April 2023 and more male people were present during the workshop at the Reuvensdagen in November 2023.
+For example, more older people attended the workshop at the CAA conference in April 2023 and more men were present during the workshop at the Reuvensdagen in November 2023.
+This might be due the the differences of audiences at the conferences.
+For the Reuvensdagen the number of participants of the workshop was relatively low and this variance might be due to chance.
 
@@
 
-Most of the respondents had several skills with computer applications in archaeology (see Figure \ref{fig:computer-skills}), and many of them did have some knowledge of ABM, although a large group was completely new to the subject.
-The majority had never applied ABM before participating in the workshops with 14 having applied ABM.
-It is interesting to note that many respondents did know what kind of software was available for ABM (see Figure \ref{fig:abm-knowledge}).
+Most of the respondents reported having a level of computer literacy already including particular computational skills common in archaeology, for example working with word processors, GIS and spreadsheets (see Figure \ref{fig:computer-skills}), and 111 of the respondents had some prior knowledge of ABM, while 59 were completely new to the subject.
+The majority (158) had never applied ABM to their research before participating in the workshops with only 14 participants who had experience of developing archaeological ABM.
+It is interesting to note that many respondents (108) did know what kind of software was available for ABM (see Figure \ref{fig:abm-knowledge}).
 
@@
 The respondents were also asked how they rated the available knowledge on ABM (Figure \ref{fig:available-theory}).
-While the largest group of them had no opinion on the subject, a very large proportion of the responded answered that they rated the available theory on ABM as limited and a smaller group as sufficient.
-A minority rated the available theory on ABM as more than sufficient.
-This clearly shows the need for better educational material.
+While 64 of them had no opinion on the subject, a proportion of the respondents (28) answered that they rated the available theory on ABM as limited and only a small group (39) as sufficient or better.
+This clearly shows the need for more and better educational material.
 
@@
 
-The majority of the respondents were enthusiastic about the teaching material, with the majority rating it as excellent or good.
-The teachers were even rated better than the teaching material, with more respondents rating them as excellent (see Figure \ref{fig:rating-teaching}).
+of the respondents were enthusiastic about the teaching material, with the majority rating it as excellent or good.
+The teachers were rated even higher than the teaching material, with more respondents giving them the \emph{excellent} score (see Figure \ref{fig:rating-teaching}).
+
+The majority of the respondents (153) were enthusiastic about the teaching material, with the majority rating it as excellent or good.
+The teachers were even rated better than the teaching material, with more respondents (165) giving them the \emph{excellent} score (see Figure \ref{fig:rating-teaching}).
 
@@
 Two more open questions were asked to the respondents.
-The first was aimed at learning what aspects the participants liked in the tutorials, and the other was aimed at getting feedback to improve the tutorials.
+The first was aimed at learning what aspects the participants liked in the tutorials (What aspects did you like most?), and the other was aimed at getting feedback to improve the tutorials (What would you like to see changed or what could be better/different?).
 
@@
 The easy and intuitive introduction into ABM and NetLOGO were also mentioned quite often.
-This was also observed during the workshops by us, most participants were working in the own pace and did not need much help.
-From a didactic point of view, the self-paced going through the tutorials is very efficient for teachers.
+This was also observed during the workshops as most participants were working in the own pace and did not need much help.
+From a didactic point of view, the self-guided and self-paced way of going through the tutorials is very efficient for teachers and limits the frustration of students that may otherwise go too fast or fall behind the class when the tempo is dictated by the teacher.
 
 We also received feedback on possible improvements.
-For the first workshops we mostly received feedback on the number of bugs.
-During the CAA-conference in Amsterdam we did the first test of the tutorials and we also instructed the participants that this was work in progress.
-The number of references to bugs reduced to zero by the end and people started providing other feedback, for example the wish to see more examples or expanding the tutorials more.
-Some wanted more cooperation, which is possible, but harder for some online events, although we provided break-out rooms.
-It is also interesting to note that over time more and more people did not see any room for improvement.
+During the first, big event CAA-conference in Amsterdam in April 2023 we did the first test of the tutorials and we also instructed the participants that this was work in progress.
+For this and the following few workshops we mostly received feedback about bugs causing the tutorial to crash or creating unexpected behaviour.
+The number of references to bugs reduced to zero by the end of the project and participants started providing other feedback, for example the wish to see more examples or expanding the tutorials further.
+Some wanted to introduce more cooperation between participants, which we did not consider primarily aiming the tutorials for individual work.
+A more cooperative aspect of the education materials is possible, even if harder for online events, and we hope that a wider classroom use of the ABMA materials may deliver solutions.
+It was also interesting to note that over time more and more people did not see any additional room for improvement indicating that the level, content and format of the tutorials met their needs.
 
-The majority of the respondents wanted to apply ABM in the future for research (144 of the 164 that answered this question, see Figure \ref{fig:future-abm}).
-Some of the respondents were already using ABM.
-A large group was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities.
-Many respondents also shared the context to which they were thinking to apply ABM.
-Various participants mentioned movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms.
+We also asked the participants about their intentions in terms of ABM and their professional life.
+The majority of the respondents - 88\%, wanted to apply ABM in the future in their research (Figure \ref{fig:future-abm}).
+Some of the respondents were already using ABM in their work.
+A large group (around 20-30\%) was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities.
+Many respondents also shared the context to which they were thinking to apply ABM mentioning topics such as movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms.
 Others thought of demography, social networks, migration or settlement distributions patterns.
-The natural environment and the interaction with humans in the past was also mentioned by some, and often in relation to GIS or how to replace GIS with ABM.
+The natural environment and the interaction with humans in the past was also mentioned frequently, often in relation to GIS and the question of how complimentary these two computational techniques are.
 The archaeological periods that the participants were interested in were very diverse, ranging from the Paleolithic to the Medieval period.
@@
 Information and data literacy: the tutorials mainly address 1.2 Evaluating Data, Information And Digital Content and 1.3 Managing Data, Information And Digital Content.
-The users learn about working with data, being critical of their digital content and how to model and manage the data in the context of an ABM.
+In the ABMA educational materials, users learn about working with data, being critical of their digital content and how to model and manage the data in the context of an ABM.
 For tutorial 2 the user is expected to be able to find information using the NetLogo(Web) dictionary (\url{http://ccl.northwestern.edu/netlogo/bind/} or \url{http://ccl.northwestern.edu/netlogo/docs/dictionary.html}).
@@
 Anyone interested can download the tutorials and their source code.
-To increase the accessibility and visibility of the tutorials, a website has been launced (\href{https://ABMArchaeologists.github.io/ABMA_Website/\#/}{https://ABMArchaeologists.github.io/ABMA\_Website}).
-On this website, the tutorials can be followed without the need to download any software.
+To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (\href{https://ABMArchaeologists.github.io/ABMA_Website/\#/}{https://ABMArchaeologists.github.io/ABMA\_Website}) can be followed without the need to download any software.
 
 Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University (Scherjon, Romanowska and Lambers 2019) and external resources people can use to learn agent-based modelling.
-The website was mainly developed by one of the student groups from Saxion University of Applied Sciences, see the acknowledgements.
+The website was mainly developed by a student group from Saxion University of Applied Sciences, see the acknowledgements.
 
@@
 
+This project was undertaken with the support of the Erasmus+ programme of the European Union (\url{https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054}).
+
 We would like to thank the student groups at Saxion University of Applied Sciences that worked with us:
@@
 \item
-  February 2022-July 2023: Alice Poorthuis, Johan Broersma, Mandy Hazenberg, Paulina Fulneczek, Ties Heesink and their tutor: Jan Wille
+  February 2022-July 2023: Alice Overgaauw, Johan Broersma, Mandy Hazenberg, Paulina Fulneczek, Ties Heesink and their tutor: Jan Willem Huson.
 \item
@@
 
-Funding acknowledgements for the publication: The Carlsberg Foundation's \emph{Young Researcher Fellowship} (CF21-0382).
+Funding acknowledgements: the publication was financed by The Carlsberg Foundation's \emph{Young Researcher Fellowship} (CF21-0382); IR was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513 and the Aarhus University Research Foundation.
 
@@
 Davies, B and Romanowska, I. 2018 An emergent community? Agent-based modelers in archaeology. \emph{The SAA Archaeological Record} 18(2): 27--32.
+
+\leavevmode\vadjust pre{\hypertarget{ref-epstein1996}{}}%
+Epstein, JM and Axtell, RL.
+1996.
+\emph{Growing Artificial Societies: Social Science from the Bottom Up}.
+The MIT Press.
+DOI: https://doi.org/\href{https://doi.org/10.7551/mitpress/3374.001.0001}{10.7551/mitpress/3374.001.0001}.
 
@@
 DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.10931587}{10.5281/zenodo.10931587}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-romanowska2018}{}}%
+Romanowska, I.
+2018 \emph{Using agent-based modelling to infer economic processes in the past}.
+In: Remesal Rodriguez, J, Revilla Calvo, V, and Bermudez Lorenzo, JM (eds.).
+Barcelona, Spain: Universitat de Barcelona * Servei de Publicacions.
+pp.
+107--118.
+
+\leavevmode\vadjust pre{\hypertarget{ref-romanowska2023}{}}%
+Romanowska, I and Scherjon, F.
+2023 \emph{Bibliometric analysis of agent-based simulation in archaeology: People, topics, and future prospects}.
+In: Kalayci, T, Lambers, K, and Klinkenberg, V (eds.).
+Analecta praehistorica leidensia.
+Leiden: Sidestone Press.
+pp.
+95--112.
+DOI: https://doi.org/\href{https://doi.org/10.59641/f48820ir}{10.59641/f48820ir}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-young1995}{}}%
+Young, DA and Bettinger, RL.
+1995 Simulating the global human expansion in the late pleistocene.
+\emph{Journal of Archaeological Science} 22(1): 89--92.
+DOI: https://doi.org/\href{https://doi.org/10.1016/S0305-4403(95)80165-0}{10.1016/S0305-4403(95)80165-0}.
+
 \end{CSLReferences}

### 2024-05-15 paper.Rmd
correction

@@
 
-This project was undertaken with the support of the Erasmus+ programme of the European Union (<https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054>).
-
 We would like to thank the student groups at Saxion University of Applied Sciences that worked with us:

### 2024-05-22 paper.Rmd
updates paper

@@
 3.
-improved transnational mobility through matching of the skills being delivered with European and Sectoral Frameworks for education, leading to individual workers being better enabled to demonstrate their skills to employers across Europe;
+improved opportunities for archaeologists to work in other European countries (transnational mobility) through matching the skills being delivered with the European Commission's Digital Competence Framework for Citizens (2.2) and the sectoral Skills Passport for Archaeology, which will mean that it will be easier for individual workers to demonstrate their skills to employers across Europe;
 
@@
 This is an interdisciplinary semester in which students of at least three different study-programmes or disciplines work together on a complex problem/project (<https://www.saxion.edu/business-and-research/collaborate-with-saxion/smart-solutions>).
-The backgrounds of the various students were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media & Game Technologies, and ICT.
-The diverse groups of students contributed new ideas, developing educational material, testing the tutorials and developing a style for the website and other materials.
+Within the semester students have to come up with new ideas or solutions to help a client with a wicked problem.
+The backgrounds of the various students we worked with were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media & Game Technologies, and ICT.
+The aim of involving interdisciplinary groups in the project was to involve students in current research projects, to enable them to learn new things and share their innovative ideas and solutions with us.
+The students groups were supplementary to the deliverables of the project, but their new ideas were benificial to the project.
+Their help during workshops was useful, but they also were engaged in developing educational material [@aalpoel2024], testing the tutorials, developing a style for the website [@jutte2024] and other materials.
 

### 2024-05-22 paper.tex
correction
updates paper

@@
 \author{true \and true \and true \and true \and true \and true \and true \and true \and true}
-\date{April 25, 2024}
+\date{May 22, 2024}
 
@@
 \item
-  improved transnational mobility through matching of the skills being delivered with European and Sectoral Frameworks for education, leading to individual workers being better enabled to demonstrate their skills to employers across Europe;
+  improved opportunities for archaeologists to work in other European countries (transnational mobility) through matching the skills being delivered with the European Commission's Digital Competence Framework for Citizens (2.2) and the sectoral Skills Passport for Archaeology, which will mean that it will be easier for individual workers to demonstrate their skills to employers across Europe;
 \item
@@
 This is an interdisciplinary semester in which students of at least three different study-programmes or disciplines work together on a complex problem/project (\url{https://www.saxion.edu/business-and-research/collaborate-with-saxion/smart-solutions}).
-The backgrounds of the various students were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media \& Game Technologies, and ICT.
-The diverse groups of students contributed new ideas, developing educational material, testing the tutorials and developing a style for the website and other materials.
+Within the semester students have to come up with new ideas or solutions to help a client with a wicked problem.
+The backgrounds of the various students we worked with were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media \& Game Technologies, and ICT.
+The aim of involving interdisciplinary groups in the project was to involve students in current research projects, to enable them to learn new things and share their innovative ideas and solutions with us.
+The students groups were supplementary to the deliverables of the project, but their new ideas were benificial to the project.
+Their help during workshops was useful, but they also were engaged in developing educational material (Aalpoel et al.
+2024), testing the tutorials, developing a style for the website (Jutte et al.
+2024) and other materials.
 
@@
 
-This project was undertaken with the support of the Erasmus+ programme of the European Union (\url{https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054}).
-
 We would like to thank the student groups at Saxion University of Applied Sciences that worked with us:
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-aalpoel2024}{}}%
+Aalpoel, R, Baan, E, Lucas, S, Mulder, J, Regts, S, Sijbrant, A, Jutte, A and Visser, R.
+2024 \emph{Teaching material Agent-Based Modelling for Archaeologists (EN/NL)} DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.10950153}{10.5281/zenodo.10950153}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-councilofeurope2020}{}}%

### 2024-07-19 paper.Rmd
Edits to paper, including additions by Dough

@@
   - name: Doug Rocks-Macqueen 
-    orcid_id: 0000
+    orcid_id: 0000-0001-9021-0260
     roles: [writing, editing, software]
@@
     roles: [writing, editing]
-    orcid_id: 0000
+    orcid_id: 0000-0001-7745-5634
     affiliation: Aarhus University
@@
 factor_levels <- c("Excellent", "Good", "Sufficient", "Insufficient", "Not shared")
+# readable event names
+event_names <- as_labeller(c("202304_CAA" = "CAA", "202308_EAA" = "EAA", "202310_CAA-DE-NLFl" = "CAA-DE-NLFl", "202311_Reuvensdagen" = "Reuvensdagen", "202311_CAA_UK" = "CAA_UK", "202312_Leiden" = "Leiden", "202401_Aarhus" = "Aarhus"))
 
@@
 
--   Development of a hands-on vocational training programme, aligned with the recently published *Agent-Based Modelling for Archaeology: Simulating the Complexity of Societies* [@romanowska2021a].
+-   development of a hands-on vocational training programme, aligned with the recently published *Agent-Based Modelling for Archaeology: Simulating the Complexity of Societies* [@romanowska2021a];
 
--   Conversion of the training programme into modular learning resources that are interactive, based on HTML and JavaScript standards.
+-   conversion of the training programme into modular learning resources that are interactive, based on HTML and JavaScript standards;
 
--   Creation of support materials to encourage use of the OERs.
+-   creation of support materials to encourage use of the OERs;
 
--   Creation of how-to guides to help trainers incorporate OERs into their teaching
+-   creation of how-to guides to help trainers incorporate OERs into their teaching;
 
--   Creation of a code and learning materials repository that will allow the project to continue sustainability after the European Union support has ended and to facilitate the creation of an open-source community to manage and improve the OERs in the future.
+-   creation of a code and learning materials repository that will allow the project to continue sustainability after the European Union support has ended and to facilitate the creation of an open-source community to manage and improve the OERs in the future;
 
--   Development of the framework for skills acquisition aligned with the Digital Skills Passport [DOUG: reference]?
+-   development of the framework for skills acquisition aligned with the Digital Skills Passport;
 
--   Promotion through a series of multiplier events.
+-   promotion through a series of multiplier events.
 
@@
 The participants of the events were asked to respond to two surveys: one before the start of the event and one completed after the end of the event.
-The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix ...
-for the questions).
+The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix 1 for the questions).
 At the end of the event the participants were asked to answer a second survey.
-Questions of this survey were aimed at measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix ...
-for the questions).
+Questions of this survey were aimed at measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix 2 for the questions).
 For some events participants had to register beforehand and we were able to send the pre-workshop survey by email.
@@
 | Reuvensdagen                 | November 2023           | In person                   | 13                     | No registration         |
-| CAA-UK                       | November 2023           | Hybrid                      | 53                     |                         |
+| CAA-UK                       | November 2023           | Hybrid                      | 53                     | No registration         |
 | Leiden (course)              | December 2023           | In person                   | 11                     | No registration         |
@@
 
-```{r nationality, fig.cap="The nationality of the participants that filled in the survey.", echo=FALSE}
+```{r nationality, fig.cap="The nationality of the participants that filled in the survey.", fig.height=5, echo=FALSE}
 ggplot(pre_surveys, aes(x = `What.is.you.nationality....Nationality...Text`)) +
-  geom_bar() +  coord_flip() + xlab("Nationality")
+  geom_bar() +  coord_flip() + xlab("Nationality") + scale_x_discrete(limits=rev)
 
@@
 ggplot(pre_surveys, aes(x = `What.is.your.age.`, fill = `What.is.your.gender....Selected.Choice`)) +
-  geom_bar() + labs(x= "Age", y="Count", fill = "Gender") + facet_grid(~Workshop) +   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+  geom_bar() + labs(x= "Age", y="Count", fill = "Gender") + facet_grid(~Workshop, labeller = event_names) +   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 
@@
   ggplot(aes(x = `Do.you.know.what.Agent.Based.Modeling.is.`)) +
-    geom_bar(na.rm = TRUE) + xlab("I know what ABM is") + facet_grid(~Workshop)
+    geom_bar(na.rm = TRUE) + xlab("I know what ABM is") + facet_grid(~Workshop, labeller = event_names)
 pre_surveys %>%
@@
   geom_bar() + xlab("Quality of theory on ABM") + coord_flip() +
-  facet_grid(~Workshop)
+  facet_grid(~Workshop, labeller = event_names)
 ```
@@
   ggplot(aes(x=How.would.you.rate.the.workshop.in.general.)) + geom_bar() +
-  xlab("Rating of the workshop") + facet_grid(~Workshop) +
+  xlab("Rating of the workshop") + facet_grid(~Workshop, labeller = event_names) +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
@@
   ggplot(aes(x=How.would.you.rate.the.teaching.material.)) + geom_bar() +
-  xlab("Rating of the teaching material") + facet_grid(~Workshop) +
+  xlab("Rating of the teaching material") + facet_grid(~Workshop, labeller = event_names) +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
@@
   ggplot(aes(x=How.would.you.rate.the.trainers.teachers.)) + geom_bar() +
-  xlab("Rating of the teacher") + facet_grid(~Workshop) +
+  xlab("Rating of the teacher") + facet_grid(~Workshop, labeller = event_names) +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
@@
 
-## The Digital Competence Framework for Citizens (2.2)
+## Integration with Learning and Skills Tracking Frameworks
+
+Beyond the creation of the OERs the project also worked on aligning the learning outcomes of the OERs with the Digital Competence Framework for Citizens and the Digital Archaeological Skills Passport.
+
+### The Digital Competence Framework for Citizens (2.2)
 
@@
 
-## Digital Skills Passport
+### Digital Skills Passport
+
+An archaeological skills passport is a tool for tracking and verifying the archaeological skills that an individual has.
+The first passport was created as part of *The Ardnamurchan Transitions Project*, a long running investigation of the archaeology and history of the Ardnamurchan Peninsula, Western Scotland, run by the Universities of Leicester and Manchester, with Archaeology Scotland [@cobb2020].
+Since then the passport has spread around the world to countries such as Germany [@karl2019] and the United States. While there are many different versions, in its simplest form it is a log booklet listing skills that others can verify by signing off on their level of competencies.
+
+A digital skills passport is the same concept as a physical skills passport but it is digitised to overcome some of the drawbacks of a physical skills passport:
+
+-   losing a copy means losing a record;
+
+-   inability to easily share the log;
+
+-   costs of production;
+
+-   ability to verify those signing off on skill levels.
+
+A trans-Atlantic consortium has been established to create a digital archaeological skills passport.
+It consists of the professional body in the United States of America, the *Register of Professional Archaeologists* and five archaeological companies in Europe, working and based in Ireland, the United Kingdom, the Netherlands, Germany and France.
+This is the passport that the ABMA integrate the OERs with.
+
+Agent Based Modelling (ABM) will be placed within the transferable skills category of the Digital Archaeological Skills Passport.
+This will be facilitated by the creation of a larger category of Analytical Skills and then a sub-category of modelling:
+
+```         
+38	Analytical Skills
+  38.1 Modelling
+	  38.1.1 Computational Modelling
+	    38.1.1.1 Agent Based Modelling
+	      38.1.1.1.1 Software mastery
+	        38.1.1.1.1.1 Creation of models and understanding of key components
+	         38.1.1.1.1.2 Integration of other fields - GIS, network science
+	      38.1.1.1.2 Analysis and interpretation of results
+	 38.1.2 Mathematical Modelling
+etc.
+```
+
+The digital skills passport has not launched yet, but when it is, it will be found at this url: <https://archaeologypassport.com>.
 

### 2024-07-19 paper.tex
Edits to paper, including additions by Dough

@@
 \author{true \and true \and true \and true \and true \and true \and true \and true \and true}
-\date{May 22, 2024}
+\date{July 19, 2024}
 
@@
 \item
-  Development of a hands-on vocational training programme, aligned with the recently published \emph{Agent-Based Modelling for Archaeology: Simulating the Complexity of Societies} (Romanowska, Wren and Crabtree 2021).
+  development of a hands-on vocational training programme, aligned with the recently published \emph{Agent-Based Modelling for Archaeology: Simulating the Complexity of Societies} (Romanowska, Wren and Crabtree 2021);
 \item
-  Conversion of the training programme into modular learning resources that are interactive, based on HTML and JavaScript standards.
+  conversion of the training programme into modular learning resources that are interactive, based on HTML and JavaScript standards;
 \item
-  Creation of support materials to encourage use of the OERs.
+  creation of support materials to encourage use of the OERs;
 \item
-  Creation of how-to guides to help trainers incorporate OERs into their teaching
+  creation of how-to guides to help trainers incorporate OERs into their teaching;
 \item
-  Creation of a code and learning materials repository that will allow the project to continue sustainability after the European Union support has ended and to facilitate the creation of an open-source community to manage and improve the OERs in the future.
+  creation of a code and learning materials repository that will allow the project to continue sustainability after the European Union support has ended and to facilitate the creation of an open-source community to manage and improve the OERs in the future;
 \item
-  Development of the framework for skills acquisition aligned with the Digital Skills Passport {[}DOUG: reference{]}?
+  development of the framework for skills acquisition aligned with the Digital Skills Passport;
 \item
-  Promotion through a series of multiplier events.
+  promotion through a series of multiplier events.
 \end{itemize}
@@
 The participants of the events were asked to respond to two surveys: one before the start of the event and one completed after the end of the event.
-The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix \ldots{} for the questions).
+The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix 1 for the questions).
 At the end of the event the participants were asked to answer a second survey.
-Questions of this survey were aimed at measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix \ldots{} for the questions).
+Questions of this survey were aimed at measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix 2 for the questions).
 For some events participants had to register beforehand and we were able to send the pre-workshop survey by email.
@@
 Reuvensdagen & November 2023 & In person & 13 & No registration \\
-CAA-UK & November 2023 & Hybrid & 53 & \\
+CAA-UK & November 2023 & Hybrid & 53 & No registration \\
 Leiden (course) & December 2023 & In person & 11 & No registration \\
@@
 
+\hypertarget{integration-with-learning-and-skills-tracking-frameworks}{%
+\subsection{Integration with Learning and Skills Tracking Frameworks}\label{integration-with-learning-and-skills-tracking-frameworks}}
+
+Beyond the creation of the OERs the project also worked on aligning the learning outcomes of the OERs with the Digital Competence Framework for Citizens and the Digital Archaeological Skills Passport.
+
 \hypertarget{the-digital-competence-framework-for-citizens-2.2}{%
-\subsection{The Digital Competence Framework for Citizens (2.2)}\label{the-digital-competence-framework-for-citizens-2.2}}
+\subsubsection{The Digital Competence Framework for Citizens (2.2)}\label{the-digital-competence-framework-for-citizens-2.2}}
 
@@
 \hypertarget{digital-skills-passport}{%
-\subsection{Digital Skills Passport}\label{digital-skills-passport}}
+\subsubsection{Digital Skills Passport}\label{digital-skills-passport}}
+
+An archaeological skills passport is a tool for tracking and verifying the archaeological skills that an individual has.
+The first passport was created as part of \emph{The Ardnamurchan Transitions Project}, a long running investigation of the archaeology and history of the Ardnamurchan Peninsula, Western Scotland, run by the Universities of Leicester and Manchester, with Archaeology Scotland (Cobb and Croucher 2020).
+Since then the passport has spread around the world to countries such as Germany (Karl et al.
+2019) and the United States.~While there are many different versions, in its simplest form it is a log booklet listing skills that others can verify by signing off on their level of competencies.~
+
+A digital skills passport is the same concept as a physical skills passport but it is digitised to overcome some of the drawbacks of a physical skills passport:
+
+\begin{itemize}
+\item
+  losing a copy means losing a record;
+\item
+  inability to easily share the log;
+\item
+  costs of production;
+\item
+  ability to verify those signing off on skill levels.~
+\end{itemize}
+
+A trans-Atlantic consortium has been established to create a digital archaeological skills passport.
+It consists of the professional body in the United States of America, the \emph{Register of Professional Archaeologists} and five archaeological companies in Europe, working and based in Ireland, the United Kingdom, the Netherlands, Germany and France.
+This is the passport that the ABMA integrate the OERs with.~
+
+Agent Based Modelling (ABM) will be placed within the transferable skills category of the Digital Archaeological Skills Passport.
+This will be facilitated by the creation of a larger category of Analytical Skills and then a sub-category of modelling:
+
+\begin{verbatim}
+38  Analytical Skills
+  38.1 Modelling
+      38.1.1 Computational Modelling
+        38.1.1.1 Agent Based Modelling
+          38.1.1.1.1 Software mastery
+            38.1.1.1.1.1 Creation of models and understanding of key components
+             38.1.1.1.1.2 Integration of other fields - GIS, network science
+          38.1.1.1.2 Analysis and interpretation of results
+     38.1.2 Mathematical Modelling
+etc.
+\end{verbatim}
+
+The digital skills passport has not launched yet, but when it is, it will be found at this url: \url{https://archaeologypassport.com}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-cobb2020}{}}%
+Cobb, H and Croucher, K.
+2020 \emph{Pedagogy, political economy, and training}.
+In: Cobb, H and Croucher, K (eds.).
+Oxford University Press.
+p.
+0.
+DOI: https://doi.org/\href{https://doi.org/10.1093/oso/9780198784258.003.0004}{10.1093/oso/9780198784258.003.0004}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-councilofeurope2020}{}}%
@@
 DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.10931759}{10.5281/zenodo.10931759}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-karl2019}{}}%
+Karl, R, Möller, K, Connolly, D, Trausmuth, T, Krierer, KR, Rocks-Macqueen, D, Aitchison, K, Edeso, M, Pintucci, A and Marciniak, A.
+2019 Der archäologische Grabungs-Kompetenz-Pass {\textendash} ein Mittel zur Dokumentation durch Praxis erworbener archäologischer Fertigkeiten.
+\emph{Archäologische Informationen} 42: 237--249.
+DOI: https://doi.org/\href{https://doi.org/10.11588/ai.2019.0.69362}{10.11588/ai.2019.0.69362}.
 

### 2024-07-19 Analyse_pre_workshop_survey.R

@@ -38,7 +38,7 @@ replace_nationality <- c("USA" = "American", "United States of America" = "Ameri
                          "Belgianian" = "Belgian", "Albanian " = "Albanian", "British(Scotland)" = "British",
                          "Brazilianian" = "Brazilian", "Dutchand Israeli" = "Dutch", "Macedonian " = "Macedonian",
                          "New Zealanderer" = "New Zealander", "British " = "British", "Egyptian " = "Egyptian",
-                         "English" = "British")
+                         "English" = "British", "British(Scotland)" = "British")
 pre_surveys$What.is.you.nationality....Nationality...Text <- str_replace_all(pre_surveys$What.is.you.nationality....Nationality...Text, replace_nationality)
 pre_surveys$What.is.you.nationality....Nationality...Text[pre_surveys$What.is.you.nationality....Nationality...Text==""] <- "_not shared_"
 
@@ -194,7 +194,7 @@ ggsave("export/preworkshop/age_classes_gender_facet.png", width = 12)
 #nationality
 
 ggplot(pre_surveys, aes(x = `What.is.you.nationality....Nationality...Text`)) +
-  geom_bar() +  coord_flip() + xlab("Nationality")
+  geom_bar() +  coord_flip() + xlab("Nationality") + scale_x_discrete(limits=rev)
 ggsave("export/preworkshop/nationality.png")
 
 ggplot(pre_surveys, aes(x = `What.is.you.nationality....Nationality...Text`)) +
@@ -202,6 +202,12 @@ ggplot(pre_surveys, aes(x = `What.is.you.nationality....Nationality...Text`)) +
   facet_grid(~Workshop)
 ggsave("export/preworkshop/nationality_facet.png", width = 12)
 
+pre_surveys %>%
+  count(`What.is.you.nationality....Nationality...Text`) %>%
+  mutate(`What.is.you.nationality....Nationality...Text` = fct_reorder(`What.is.you.nationality....Nationality...Text`,n)) %>%
+  ggplot(aes(x = `What.is.you.nationality....Nationality...Text`, y= n)) +
+  geom_col() + coord_flip() + xlab("Nationality")
+ggsave("export/preworkshop/nationality_ordered_n.png")
 
 # student
 

### 2024-07-19 paper.tex
conclusion added

@@
 
-Things to address
+The process of creating and testing Open Educational Resources for Agent-Based Modelling for Archaeologists showed us many things.
+The demand for tutorials exceeded our expectations, showing the importance of creating these materials.
+The testing of tutorials during workshops at international conferences proved very useful in gaining direct feedback.
+This greatly helped improving the material.
+The tutorials have proven their use in guiding learners into understanding ABM in an archaeological context.
+The international cooperation improved the quality of the material and also the accessibility of the material for various audiences.
+The integration in the Digital competency framework and Skills passport increased the didactic and pedagogic quality even while the latter is not yet published.
 
-\begin{itemize}
-\item
-  demand for tutorials exceeded expectations
-\item
-  tutorials helpful in understanding ABM
-\item
-  international cooperation
-\item
-  Digital competency framework / Skills passport
-\item
-  OER: open learning
-\item
-  Future aspects:
-
-  \begin{itemize}
-  \item
-    application in educational programmes
-  \item
-    community driven updates (?)
-  \end{itemize}
-\end{itemize}
+Future aspects of these open learning materials will be the further application and integration within the educational programmes in our universities and hopefully in others too.
+We have made the material openly available and we hope that the ABM-community will stay involved and that future updates will be community driven.
 

### 2024-09-18 paper.Rmd
updates and corrections

@@
 
-Computational modelling has become a standard research technique across most scientific disciplines including archaeology.
+Computational modelling has become a standard research technique across most scientific disciplines, including archaeology.
 Among the many computational modelling techniques the discipline has found agent-based modelling to be particularly well suited to the questions asked by archaeologists.
-Agent-based modelling is a simulation technique where autonomous agents interact with each other and with the environment around.
+Agent-based modelling is a simulation technique where autonomous agents interact with each other and with their environment.
 Their aggregated behaviour gives rise to system-level outcomes thus enabling researchers to understand how *micro* (individual) *mechanisms* lead to *macro* (system-level) *patterns*.
@@
 A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 76% of modellers were primarily self-taught with some degree of secondary support i.e.
-27% relied on peer support, only 7% had any sort of formal training at a workshop or summer school while only 14% received training as part of a degree.
+27% relied on peer support, only 7% had any sort of formal training at a workshop or summer school, while only 14% received training as part of a degree.
 
@@
 Throughout the project we sought feedback from the participants and used it to improve the tutorials prior to the following events.
-The feedback was obtained through two surveys using Qualtrics.
+The feedback was obtained through two surveys using Qualtrics (<https://www.qualtrics.com/>).
 The participants of the events were asked to respond to two surveys: one before the start of the event and one completed after the end of the event.
@@
 All activity stays within the entree level of difficulty, although some higher level concepts are lightly touched upon.
+
+![Screenhot of lesson 3 in tutorial 1 showing the population dynamics of the Long House Valley in the Artificial Anasazi model created by Axtell et al.
+(2002).](paper_files/Screenshot_ABMA_tutorial_1_lesson_3.png)
 
@@
   group_by(`Which.computer.skills.do.you.have....Selected.Choice`) %>%
-  ggplot(aes(x=`Which.computer.skills.do.you.have....Selected.Choice`)) + geom_bar() +
+    mutate(count_skills = n()) %>%
+  ggplot(aes(x=reorder(`Which.computer.skills.do.you.have....Selected.Choice`, count_skills))) + geom_bar() +
     xlab("Computer skills") + coord_flip()
@@
 
-The respondents generally liked the interactive step-by-step way of going through the tutorials, but also valued the interaction with the teachers very much.
+The respondents generally liked the interactive step-by-step way of going through the tutorials, but also valued the interaction with the teachers strongly.
 The easy and intuitive introduction into ABM and NetLOGO were also mentioned quite often.
@@
 We also received feedback on possible improvements.
-During the first, big event CAA-conference in Amsterdam in April 2023 we did the first test of the tutorials and we also instructed the participants that this was work in progress.
+During the first workshop, the CAA-conference in Amsterdam in April 2023, we did the first test of the tutorials and we also instructed the participants that this was work in progress.
 For this and the following few workshops we mostly received feedback about bugs causing the tutorial to crash or creating unexpected behaviour.
@@
 
-The tutorials are publicly hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials/>) and are citeable [@jutte2024].
+The tutorials are publicly hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials/>) and are citeable [@rocks-macqueen2024].
 Anyone interested can download the tutorials and their source code.
 To increase the accessibility and visibility of the tutorials, they are also hosted on a static website ([https://ABMArchaeologists.github.io/ABMA_Website](https://ABMArchaeologists.github.io/ABMA_Website/#/){.uri}) can be followed without the need to download any software.
+The source code of the website is also openly available [@jutte2024].
 
@@
 
-Things to address
+The process of creating and testing Open Educational Resources for Agent-Based Modelling for Archaeologists showed us many things.
+The demand for tutorials exceeded our expectations, showing the importance of creating these materials.
+The testing of tutorials during workshops at international an national conferences proved very useful in gaining direct feedback.
+This greatly helped improving the material.
+The tutorials have proven their use in guiding learners into understanding ABM in an archaeological context.
+The international cooperation improved the quality of the material and also the accessibility of the material for various audiences.
+The integration in the Digital competency framework and Skills passport increased the didactic and pedagogic quality even while the latter is not yet published.
 
--   demand for tutorials exceeded expectations
-
--   tutorials helpful in understanding ABM
-
--   international cooperation
-
--   Digital competency framework / Skills passport
-
--   OER: open learning
-
--   Future aspects:
-
-    -   application in educational programmes
-
-    -   community driven updates (?)
+Future aspects of these open learning materials will be the further application and integration within the educational programmes in our universities and hopefully in others too.
+We have made the material openly available and we hope that the ABM-community will stay involved and that future updates will be community driven.
 
@@
 
-We had help from many people during the various workshops and we would like to thank them for their help (in alphabetical order of their first name): Adéla Sobotkova, Agnes Schneider, Alice Overgaauw, Eduardo Herrera Malatesta, Jens Emil Bødstrup Christoffersen, Johan Broersma, Magnus Lindholm Nielsen, Mandy Hazenberg, María Coto Sarmiento, Paulina Fulneczek, Petra Hermankova, Ties Heesink.
+We had help from many people during the various workshops and we would like to thank them for their help (in alphabetical order of their first name): Adéla Sobotkova, Agnes Schneider, Alice Poorthuis, Eduardo Herrera Malatesta, Jens Emil Bødstrup Christoffersen, Johan Broersma, Magnus Lindholm Nielsen, Mandy Hazenberg, María Coto Sarmiento, Paulina Fulneczek, Petra Hermankova, Ties Heesink.
 We also want to thank all the participants of our workshops.
 
-Funding acknowledgements: the publication was financed by The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382); IR was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513 and the Aarhus University Research Foundation.
+Funding acknowledgements: the publication was financed by The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382); IR was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513, the Aarhus University Research Foundation and FWO grant number W001220N.
 

### 2024-09-18 paper.tex
updates and corrections

@@
 \author{true \and true \and true \and true \and true \and true \and true \and true \and true}
-\date{July 19, 2024}
+\date{September 18, 2024}
 
@@
 
-Computational modelling has become a standard research technique across most scientific disciplines including archaeology.
+Computational modelling has become a standard research technique across most scientific disciplines, including archaeology.
 Among the many computational modelling techniques the discipline has found agent-based modelling to be particularly well suited to the questions asked by archaeologists.
-Agent-based modelling is a simulation technique where autonomous agents interact with each other and with the environment around.
+Agent-based modelling is a simulation technique where autonomous agents interact with each other and with their environment.
 Their aggregated behaviour gives rise to system-level outcomes thus enabling researchers to understand how \emph{micro} (individual) \emph{mechanisms} lead to \emph{macro} (system-level) \emph{patterns}.
@@
 One of the sectorial problems is the access to vocational training required to gain the necessary computer programming and modelling skills.
-A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 76\% of modellers were primarily self-taught with some degree of secondary support i.e.~27\% relied on peer support, only 7\% had any sort of formal training at a workshop or summer school while only 14\% received training as part of a degree.
+A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 76\% of modellers were primarily self-taught with some degree of secondary support i.e.~27\% relied on peer support, only 7\% had any sort of formal training at a workshop or summer school, while only 14\% received training as part of a degree.
 
@@
 Throughout the project we sought feedback from the participants and used it to improve the tutorials prior to the following events.
-The feedback was obtained through two surveys using Qualtrics.
+The feedback was obtained through two surveys using Qualtrics (\url{https://www.qualtrics.com/}).
 The participants of the events were asked to respond to two surveys: one before the start of the event and one completed after the end of the event.
@@
 All activity stays within the entree level of difficulty, although some higher level concepts are lightly touched upon.
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/Screenshot_ABMA_tutorial_1_lesson_3.png}
+\caption{Screenhot of lesson 3 in tutorial 1 showing the population dynamics of the Long House Valley in the Artificial Anasazi model created by Axtell et al.~(2002).}
+\end{figure}
 
@@
 
-The tutorials are publicly hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials/}) and are citeable (Jutte et al.
+The tutorials are publicly hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials/}) and are citeable (Rocks-Macqueen et al.
 2024).
@@
 To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (\href{https://ABMArchaeologists.github.io/ABMA_Website/\#/}{https://ABMArchaeologists.github.io/ABMA\_Website}) can be followed without the need to download any software.
+The source code of the website is also openly available (Jutte et al.
+2024).
 
@@
 The demand for tutorials exceeded our expectations, showing the importance of creating these materials.
-The testing of tutorials during workshops at international conferences proved very useful in gaining direct feedback.
+The testing of tutorials during workshops at international an national conferences proved very useful in gaining direct feedback.
 This greatly helped improving the material.
@@
 
-We had help from many people during the various workshops and we would like to thank them for their help (in alphabetical order of their first name): Adéla Sobotkova, Agnes Schneider, Alice Overgaauw, Eduardo Herrera Malatesta, Jens Emil Bødstrup Christoffersen, Johan Broersma, Magnus Lindholm Nielsen, Mandy Hazenberg, María Coto Sarmiento, Paulina Fulneczek, Petra Hermankova, Ties Heesink.
+We had help from many people during the various workshops and we would like to thank them for their help (in alphabetical order of their first name): Adéla Sobotkova, Agnes Schneider, Alice Poorthuis, Eduardo Herrera Malatesta, Jens Emil Bødstrup Christoffersen, Johan Broersma, Magnus Lindholm Nielsen, Mandy Hazenberg, María Coto Sarmiento, Paulina Fulneczek, Petra Hermankova, Ties Heesink.
 We also want to thank all the participants of our workshops.
 
-Funding acknowledgements: the publication was financed by The Carlsberg Foundation's \emph{Young Researcher Fellowship} (CF21-0382); IR was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513 and the Aarhus University Research Foundation.
+Funding acknowledgements: the publication was financed by The Carlsberg Foundation's \emph{Young Researcher Fellowship} (CF21-0382); IR was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513, the Aarhus University Research Foundation and FWO grant number W001220N.
 

### 2024-09-18 Analyse_pre_workshop_survey.R

@@ -144,7 +144,8 @@ pre_surveys %>%
   separate_rows(`Which.computer.skills.do.you.have....Selected.Choice`, sep = ',\\s*') %>%
   filter(`Which.computer.skills.do.you.have....Selected.Choice`!="") %>%
   group_by(`Which.computer.skills.do.you.have....Selected.Choice`) %>%
-  ggplot(aes(x=`Which.computer.skills.do.you.have....Selected.Choice`)) + geom_bar() +
+  mutate(count_skills = n()) %>%
+  ggplot(aes(x=reorder(`Which.computer.skills.do.you.have....Selected.Choice`, count_skills))) + geom_bar() +
     xlab("Computer skills") + coord_flip()
 ggsave("export/preworkshop/computer_skills.png")
 

### 2024-09-18 paper.Rmd
Figure and language corrections

@@
 The following packages were used for analyses: ggplot2 [@wickham2016], dplyr [@wickham2023a], tidyr [@wickham2023], forcats[@wickham2023b], lubridate [@grolemund2011] and stringr [@wickham2023c].
-The date and the code is available at <https://github.com/ABMArchaeologists/ABMA_paper> and (...
+For the sake of reproducibility [@marwick2017] this paper was written in R Markdown (<https://rmarkdown.rstudio.com/>), with data and code available at <https://github.com/ABMArchaeologists/ABMA_paper> (...
 insert Zenodo reference...)
@@
 ```{r nationality, fig.cap="The nationality of the participants that filled in the survey.", fig.height=5, echo=FALSE}
-ggplot(pre_surveys, aes(x = `What.is.you.nationality....Nationality...Text`)) +
-  geom_bar() +  coord_flip() + xlab("Nationality") + scale_x_discrete(limits=rev)
+pre_surveys %>% 
+  group_by(`What.is.you.nationality....Nationality...Text`) %>% 
+  mutate(count_nationality = n()) %>%
+  ggplot(aes(x=reorder(`What.is.you.nationality....Nationality...Text`, count_nationality))) + geom_bar() +  coord_flip() + xlab("Nationality")
 
@@
 
-```{r, abm-knowledge, fig.cap="Respondents knowledge and experience with ABM.", fig.show="hold", out.width="50%", echo=FALSE}
+```{r, abm-knowledge, fig.cap="Respondents knowledge of ABM per event knowledge of the various softwares for ABM, before participating in the workshops.", fig.show="hold", out.width="50%", echo=FALSE}
 pre_surveys %>%
   filter(`Do.you.know.what.Agent.Based.Modeling.is.`!="Not shared") %>%
-  ggplot(aes(x = `Do.you.know.what.Agent.Based.Modeling.is.`)) +
-    geom_bar(na.rm = TRUE) + xlab("I know what ABM is") + facet_grid(~Workshop, labeller = event_names)
+  select(Do.you.know.what.Agent.Based.Modeling.is., Workshop) %>%
+  group_by(`Do.you.know.what.Agent.Based.Modeling.is.`, Workshop) %>%
+  mutate(count_response = n()) %>%
+  distinct() %>%
+  ggplot(aes(fill = `Do.you.know.what.Agent.Based.Modeling.is.`, x=Workshop, y = count_response)) +
+  geom_bar(na.rm = TRUE, position="fill", stat="identity") +
+  geom_text(aes(label=count_response), position = position_fill(vjust = 0.5), size = 2.5, colour = "grey41") + scale_x_discrete(labels = c("CAA", "EAA", "CAA-DE-NLFl", "Reuvensdagen", "CAA_UK", "Leiden", "Aarhus")) +
+  xlab("") + scale_fill_discrete(name = "ABM known") + ylab("Percentage") +  scale_y_continuous(labels = scales::label_percent(scale = 100, prefix = "", suffix = "")) + coord_flip()
 pre_surveys %>%
@@
   group_by(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`) %>%
-  ggplot(aes(x=`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`)) + geom_bar() +
-  xlab("ABM Software known") + coord_flip()
+    mutate(count_software = n()) %>%
+  ggplot(aes(x=reorder(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`, count_software))) + geom_bar() + scale_x_discrete(labels = ) + xlab("ABM Software known") + coord_flip()
 ```
@@
 We also asked the participants about their intentions in terms of ABM and their professional life.
-The majority of the respondents - 88%, wanted to apply ABM in the future in their research (Figure \@ref(fig:future-abm)).
+The majority of the respondents - 88%, wanted to apply ABM in the future in their research.
 Some of the respondents were already using ABM in their work.
@@
 The archaeological periods that the participants were interested in were very diverse, ranging from the Paleolithic to the Medieval period.
-
-```{r, future-abm, fig.cap="The respondents reaction to the question if they thing that they will apply ABM in the future.", out.height="50%", echo=FALSE}
-post_surveys %>%
-  filter(!is.na(Do.you.think.that.you.will.apply.ABM.to.your.own.research..If.yes..can.you.elaborate....Selected.Choice)) %>%
-  ggplot(aes(x = 1, fill = `Do.you.think.that.you.will.apply.ABM.to.your.own.research..If.yes..can.you.elaborate....Selected.Choice`)) +
-  geom_bar(na.rm = TRUE) + coord_polar(theta = "y", start=0) + theme_void() +
-  scale_fill_discrete(name = paste0("Will apply ABM in the future (n=", sum(!is.na(post_surveys$Do.you.think.that.you.will.apply.ABM.to.your.own.research..If.yes..can.you.elaborate....Selected.Choice)), ")"),
-                      labels = c("No", "Yes"))
-```
 

### 2024-09-18 paper.tex
Figure and language corrections

@@
 2023), tidyr (Wickham, Vaughan and Girlich 2023), forcats(Wickham 2023a), lubridate (Grolemund and Wickham 2011) and stringr (Wickham 2023b).
-The date and the code is available at \url{https://github.com/ABMArchaeologists/ABMA_paper} and (\ldots{} insert Zenodo reference\ldots)
+For the sake of reproducibility (Marwick 2017) this paper was written in R Markdown (\url{https://rmarkdown.rstudio.com/}), with data and code available at \url{https://github.com/ABMArchaeologists/ABMA_paper} (\ldots{} insert Zenodo reference\ldots)
 
@@
 \begin{figure}
-\includegraphics[width=0.5\linewidth]{paper_files/figure-latex/abm-knowledge-1} \includegraphics[width=0.5\linewidth]{paper_files/figure-latex/abm-knowledge-2} \caption{Respondents knowledge and experience with ABM.}\label{fig:abm-knowledge}
+\includegraphics[width=0.5\linewidth]{paper_files/figure-latex/abm-knowledge-1} \includegraphics[width=0.5\linewidth]{paper_files/figure-latex/abm-knowledge-2} \caption{Respondents knowledge of ABM per event knowledge of the various softwares for ABM, before participating in the workshops.}\label{fig:abm-knowledge}
 \end{figure}
@@
 We also asked the participants about their intentions in terms of ABM and their professional life.
-The majority of the respondents - 88\%, wanted to apply ABM in the future in their research (Figure \ref{fig:future-abm}).
+The majority of the respondents - 88\%, wanted to apply ABM in the future in their research.
 Some of the respondents were already using ABM in their work.
@@
 The archaeological periods that the participants were interested in were very diverse, ranging from the Paleolithic to the Medieval period.
-
-\begin{figure}
-\includegraphics[height=0.5\textheight]{paper_files/figure-latex/future-abm-1} \caption{The respondents reaction to the question if they thing that they will apply ABM in the future.}\label{fig:future-abm}
-\end{figure}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-marwick2017}{}}%
+Marwick, B.
+2017 Computational Reproducibility in Archaeological Research: Basic Principles and a Case Study of Their Implementation.
+\emph{Journal of Archaeological Method and Theory} 24(2): 424--450.
+DOI: https://doi.org/\href{https://doi.org/10.1007/s10816-015-9272-9}{10.1007/s10816-015-9272-9}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2023}{}}%

### 2024-09-18 Analyse_pre_workshop_survey.R

@@ -4,6 +4,7 @@ library(tidyr)
 library(forcats)
 library(stringr)
 
+# importing data
 CAA_202304 <- read.csv("data/CAA_202304.csv")
 CAA_202304$Workshop <- "202304_CAA"
 EAA_202308 <- read.csv("data/EAA_202308.csv")
@@ -73,6 +74,19 @@ pre_surveys %>%
     geom_bar(na.rm = TRUE) + xlab("I know what ABM is") + facet_grid(~Workshop)
 ggsave("export/preworkshop/ABM_known_facet.png", width = 12)
 
+pre_surveys %>%
+  filter(`Do.you.know.what.Agent.Based.Modeling.is.`!="Not shared") %>%
+  select(Do.you.know.what.Agent.Based.Modeling.is., Workshop) %>%
+  group_by(`Do.you.know.what.Agent.Based.Modeling.is.`, Workshop) %>%
+  mutate(count_response = n()) %>%
+  distinct() %>%
+  ggplot(aes(fill = `Do.you.know.what.Agent.Based.Modeling.is.`, x=Workshop, y = count_response)) +
+  geom_bar(na.rm = TRUE, position="fill", stat="identity") +
+  geom_text(aes(label=count_response), position = position_fill(vjust = 0.5)) +
+  xlab("") +  scale_fill_discrete(name = "I know what ABM is") + ylab("Percentage") +  coord_flip()
+ggsave("export/preworkshop/ABM_known_percent_stacked.png", width = 12)
+
+
 # Theory on ABM
 
 pre_surveys %>%
@@ -105,7 +119,8 @@ pre_surveys %>%
   filter(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`!="") %>%
   separate_rows(Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice, sep = ',\\s*') %>%
   group_by(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`) %>%
-  ggplot(aes(x=`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`)) + geom_bar() +
+  mutate(count_software = n()) %>%
+  ggplot(aes(x=reorder(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`, count_software))) + geom_bar() +
   xlab("ABM Software known") + coord_flip()
 ggsave("export/preworkshop/ABM_software_known.png")
 
[…]

### 2024-09-20 paper.Rmd
correction

@@
     orcid_id: 0000-0002-1589-7768
-    affiliation: Aarhus University
+    affiliation: Aarhus 
+  - name: Jeroen Linssen
+    orcid_id: 0000-0002-2626-1837
+    roles: [writing, editing]
+    affiliation: Saxion University of Applied Sciences
   - name: Adam Pažout

### 2024-10-18 paper.Rmd
updates to paper

--- 2024-09-18 paper.Rmd
@@
 title: "Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists"
-author:
+authors:
   - name: Ronald M. Visser
@@
     affiliation: Aarhus University
+  - name: Jeroen Linssen
+    orcid_id: 0000-0002-2626-1837
+    roles: [writing, editing]
+    affiliation: Saxion University of Applied Sciences
   - name: Adam Pažout
@@
 ```
+
+**Abstract**
+
+Agent-based modelling (ABM) is one of the most applied computational modelling research techniques in archaeology, because it is well suited to the questions asked by archaeologists.
+However, the access to vocational training for ABM is limited.
+In this paper we present the Open Educational Resources (OER) that we produced in an international cooperation.
+For this purpose we created online and open tutorials for ABM for archaeologists.
+In addition, how-to guides and support material was created, both for teachers and students.
+The material adheres to international (European) standards in relation to education and the Digital Skills Passport for archaeology.
+The teaching material is well tested using a diverse group of archaeologists during various international conferences and meetings.
+It was shown that the knowledge and skills of the participants in relation to ABM improved.
+In addition, the demand for the teaching material and the workshops was high, ensuring that the tutorials on ABM will be used in the future.
 
@@
 
-In the first tutorial the user learn what simulations and agent-based models are and how they can help in archaeological research.
+In the first tutorial the user learn what simulations and agent-based models are and how they can help in archaeological research (\@ref(fig:screenshot-lesson-3)).
 This tutorial consists of 4 lessons.
@@
 
-![Screenhot of lesson 3 in tutorial 1 showing the population dynamics of the Long House Valley in the Artificial Anasazi model created by Axtell et al.
-(2002).](paper_files/Screenshot_ABMA_tutorial_1_lesson_3.png)
+```{r, screenshot-lesson-3, echo = FALSE, out.width="100%", fig.cap = "Screenshot of lesson 3 in tutorial 1 showing the population dynamics of the Long House Valley in the Artificial Anasazi model created by Axtell et al.
+(2002)."}
+knitr::include_graphics("paper_files/ScreenshotABMAtutorial1lesson3.png")
+```
 
@@
 
-During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials.
+During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials \@ref(fig:workshop-impression).
 Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general.
@@
 While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later installments.
+
+```{r workshop-impression, echo = FALSE, out.width="100%", fig.cap = "Impression of various workshops (CAA 2023 in Amsterdam and EAA 2023 in Belfast) with participants working."}
+knitr::include_graphics("paper_files/CollageWorkshops.jpg")
+```
 

### 2024-10-18 paper.tex
correction
updates to paper

@@
 \title{Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists}
-\author{true \and true \and true \and true \and true \and true \and true \and true \and true}
-\date{September 18, 2024}
+\author{}
+\date{\vspace{-2.5em}October 18, 2024}
 
@@
 }
+\textbf{Abstract}
+
+Agent-based modelling (ABM) is one of the most applied computational modelling research techniques in archaeology, because it is well suited to the questions asked by archaeologists.
+However, the access to vocational training for ABM is limited.
+In this paper we present the Open Educational Resources (OER) that we produced in an international cooperation.
+For this purpose we created online and open tutorials for ABM for archaeologists.
+In addition, how-to guides and support material was created, both for teachers and students.
+The material adheres to international (European) standards in relation to education and the Digital Skills Passport for archaeology.
+The teaching material is well tested using a diverse group of archaeologists during various international conferences and meetings.
+It was shown that the knowledge and skills of the participants in relation to ABM improved.
+In addition, the demand for the teaching material and the workshops was high, ensuring that the tutorials on ABM will be used in the future.
+
 \hypertarget{introduction}{%
@@
 
-In the first tutorial the user learn what simulations and agent-based models are and how they can help in archaeological research.
+In the first tutorial the user learn what simulations and agent-based models are and how they can help in archaeological research (\ref{fig:screenshot-lesson-3}).
 This tutorial consists of 4 lessons.
@@
 \begin{figure}
-\centering
-\includegraphics{paper_files/Screenshot_ABMA_tutorial_1_lesson_3.png}
-\caption{Screenhot of lesson 3 in tutorial 1 showing the population dynamics of the Long House Valley in the Artificial Anasazi model created by Axtell et al.~(2002).}
+\includegraphics[width=1\linewidth]{paper_files/ScreenshotABMAtutorial1lesson3} \caption{Screenshot of lesson 3 in tutorial 1 showing the population dynamics of the Long House Valley in the Artificial Anasazi model created by Axtell et al.
+(2002).}\label{fig:screenshot-lesson-3}
 \end{figure}
@@
 
-During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials.
+During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials \ref{fig:workshop-impression}.
 Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general.
@@
 While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later installments.
+
+\begin{figure}
+\includegraphics[width=1\linewidth]{paper_files/CollageWorkshops} \caption{Impression of various workshops (CAA 2023 in Amsterdam and EAA 2023 in Belfast) with participants working.}\label{fig:workshop-impression}
+\end{figure}
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-aalpoel2024}{}}%
-Aalpoel, R, Baan, E, Lucas, S, Mulder, J, Regts, S, Sijbrant, A, Jutte, A and Visser, R.
+Aalpoel, R, Baan, E, Lucas, S, Mulder, J, Regts, S, Sijbrant, A, Jutte, A and Visser, RM.
 2024 \emph{Teaching material Agent-Based Modelling for Archaeologists (EN/NL)} DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.10950153}{10.5281/zenodo.10950153}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-visser2024}{}}%
-Visser, R, Lambers, K, Aitchison, K, Jutte, A, Rocks-Macqueen, D, Knaap, L van der, Romanowska, I and Brughmans, T.
+Visser, RM, Lambers, K, Aitchison, K, Jutte, A, Rocks-Macqueen, D, Knaap, L van der, Romanowska, I and Brughmans, T.
 2024 \emph{The Digital Competence Framework for Citizens (2.2) and Agent-Based modelling for Archaeologists}.

### 2024-11-06 paper.Rmd
Corrections (Karsten/Kenneth)

@@
 event_names <- as_labeller(c("202304_CAA" = "CAA", "202308_EAA" = "EAA", "202310_CAA-DE-NLFl" = "CAA-DE-NLFl", "202311_Reuvensdagen" = "Reuvensdagen", "202311_CAA_UK" = "CAA_UK", "202312_Leiden" = "Leiden", "202401_Aarhus" = "Aarhus"))
+events <- read.csv("data/events.csv", sep = ";")
 
@@
 Computational modelling has become a standard research technique across most scientific disciplines, including archaeology.
-Among the many computational modelling techniques the discipline has found agent-based modelling to be particularly well suited to the questions asked by archaeologists.
+Among the many computational modelling techniques, the discipline has found agent-based modelling to be particularly well suited to the questions asked by archaeologists.
 Agent-based modelling is a simulation technique where autonomous agents interact with each other and with their environment.
-Their aggregated behaviour gives rise to system-level outcomes thus enabling researchers to understand how *micro* (individual) *mechanisms* lead to *macro* (system-level) *patterns*.
+Their aggregated behaviour gives rise to system-level outcomes, thus enabling researchers to understand how *micro* (individual) *mechanisms* lead to *macro* (system-level) *patterns*.
 
@@
 One of the sectorial problems is the access to vocational training required to gain the necessary computer programming and modelling skills.
-A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 76% of modellers were primarily self-taught with some degree of secondary support i.e.
+A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 76% of modellers were primarily self-taught with some degree of secondary support, i.e.
 27% relied on peer support, only 7% had any sort of formal training at a workshop or summer school, while only 14% received training as part of a degree.
 
-A consortium of Landward Research Teoranta, Aarhus Universitet, Universiteit Leiden and Saxion University of Applied Sciences recognised this need, and sought to address the limited opportunities that archaeologists have to access high quality training through a project aiming to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.
+A consortium of Landward Research Teoranta, Aarhus Universitet, Leiden University and Saxion University of Applied Sciences recognised this need, and sought to address the limited opportunities that archaeologists have to access high quality training through a project aiming to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.
 
@@
 
--   development of a hands-on vocational training programme, aligned with the recently published *Agent-Based Modelling for Archaeology: Simulating the Complexity of Societies* [@romanowska2021a];
+-   development of a hands-on vocational training programme, aligned with the recently published textbook *Agent-Based Modelling for Archaeology: Simulating the Complexity of Societies* [@romanowska2021a];
 
@@
 3.
-improved opportunities for archaeologists to work in other European countries (transnational mobility) through matching the skills being delivered with the European Commission's Digital Competence Framework for Citizens (2.2) and the sectoral Skills Passport for Archaeology, which will mean that it will be easier for individual workers to demonstrate their skills to employers across Europe;
+improved opportunities for archaeologists to work in other European countries (transnational mobility) through matching the skills being delivered with the European Commission's Digital Competence Framework for Citizens (2.2) and the sectoral Skills Passport for Archaeology, which means that it will be easier for individual workers to demonstrate their skills to employers across Europe;
 
@@
 The ABMA learning materials assume that participants have learning skills at least at a secondary education level.
-It is suitable for participants with at least some background in archaeology, such as those typical among students after the first year of a Bachelor study in Archaeology, i.e., above EQF levels 4 or 5.
+They are suitable for participants with at least some background in archaeology, such as those typical among students after the first year of a Bachelor study in Archaeology, i.e., above EQF levels 4 or 5.
 In addition, the tutorials are aimed at professionals working in archaeology who have some experience with computer applications.
 We also assume that learners are proficient in English and have at least the Reading B1 level, but Reading B2 is recommended [@councilofeurope2020].
-The tutorials were also developed with a secondary aim to develop learners' general skills in the following competence area's: 1.
+The tutorials were also developed with a secondary aim to develop learners' general skills in the following competence areas: 1.
 Information and data literacy, 2.
@@
 
-After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events all of which were run by members of the project team.
+After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events all of which were run by members of the project team (Table \@ref(tab:events)).
 Recognising the difficulty of delivering practical training in an online environment during the online events several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things.
@@
 For some events participants had to register beforehand and we were able to send the pre-workshop survey by email.
-At other events no registration was possible or necessary and the pre-workshop survey was given at the start of the event.
+At other events no registration was possible or necessary and the pre-workshop survey was shared at the start of the event.
 The post-workshop survey was distributed using QR-codes or links at the end of the event.
 
-| Conference or event          | Period                  | Online, in person or hybrid | Number of participants | Number of registrations |
-|---------------|---------------|---------------|---------------|---------------|
-| CAA Amsterdam                | April 2023              | In person                   | 36                     | 53                      |
-| EAA Belfast                  | August 2023             | In person                   | 21                     | No registration         |
-| CAA-DE/NL-Fl Online workshop | October 2023            | Online                      | 40                     | 76                      |
-| Reuvensdagen                 | November 2023           | In person                   | 13                     | No registration         |
-| CAA-UK                       | November 2023           | Hybrid                      | 53                     | No registration         |
-| Leiden (course)              | December 2023           | In person                   | 11                     | No registration         |
-| Aarhus                       | January/February 2024   | Online                      | 176                    | \>500                   |
-| Saxion                       | March 2023-January 2024 | In person                   | 18                     | No registration         |
-| **Total**                    |                         |                             | **368**                | **\>628**               |
-
-: Conferences, events and situations were workshops were held at which the tutorials were tested.
+```{r, events, echo=FALSE, tab.cap= "Conferences, events and situations where workshops were held at which the tutorials were tested." }
+knitr::kable(events, align = "l", col.names = c("Conference or event", "Period", "Online, in person or hybrid", "Number of participants", "Number of registrations")) 
+```
 
@@
 
-The final set of ABMA tutorals are available for download at <https://github.com/ABMArchaeologists/ABMA_tutorials> and are citeable [@rocks-macqueen2024].
-The set of developed tutorials consists of the following tutorials:
+The final set of ABMA tutorials is available for download at <https://github.com/ABMArchaeologists/ABMA_tutorials> and is citeable [@rocks-macqueen2024]. The set consists of the following tutorials:
 
@@
 
-Each tutorials consists of a number of individual lessons, each of which guides the learner in a self-pace manner through an increasingly challenging set of exercises that build the skills and knowledge necessary to become a confident agent-based modeller.
+Each tutorials consists of a number of individual lessons, each of which guides the learner in a self-paced manner through an increasingly challenging set of exercises that build the skills and knowledge necessary to become a confident agent-based modeller.
 
-In the first tutorial the user learn what simulations and agent-based models are and how they can help in archaeological research (\@ref(fig:screenshot-lesson-3)).
-This tutorial consists of 4 lessons.
+In the first tutorial the users learn what simulations and agent-based models are and how they can help in archaeological research (Figure \@ref(fig:screenshot-lesson-3)).
+This tutorial consists of four lessons and makes use of the Artificial Anasazi model [@axtell2002].
 The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific.
@@
 This tutorial builds a foundation for working in NetLogo and with ABM.
-All activity stays within the entree level of difficulty, although some higher level concepts are lightly touched upon.
+All activity stays within the entry level of difficulty, although some higher level concepts are lightly touched upon.
 
@@
 
-In tutorial 2 the user will learn the basics of NetLogo by making your their simulation on the Out of Africal dispersal of *Homo sapiens* [@young1995].The learner works through the basics of NetLogo syntax and learn how to set up a simulation and visualize its outcomes.
-This tutorial consists of 9 lessons.
+In tutorial 2 the user will learn the basics of NetLogo by making their simulation on the Out of Africal dispersal of *Homo sapiens* [@young1995].The learner works through the basics of NetLogo syntax and learns how to set up a simulation and visualize its outcomes.
+This tutorial consists of nine lessons.
 This tutorial intends to guide the learner from the introduction to the beginner level of proficiency.
@@
 This is achieved by working with the NetLogo web interface that enables highlighting specific words and buttons and giving visual cues as to how to proceed.
-The tutorial intoduces several most commonly used built-in functions in Netlogo (called *primitives*) are introduces the general structure of all models cosisting of the initialization phase and the main simulation loop.
-It walks the participants through te most important NetLogo settings, such as dimensions, coordinates and the origin point of the grid on which the simulation runs.
+The tutorial introduces several most commonly used built-in functions in Netlogo (called *primitives*) and introduces the general structure of all models consisting of the initialization phase and the main simulation loop.
+It walks the participants through the most important NetLogo settings, such as dimensions, coordinates and the origin point of the grid on which the simulation runs.
 The learners gain their first stripes in programming by using simple loops and functions as well as conditional statements.
@@
 In the third tutorial, the learner builds a simple trade model, again based on a published simulation [@romanowska2018].
-Here learners are expanding their NetLogo programming skills with more complexity added to such syntactic structures as loops, lists and reporters.
-The learner is introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity.
-This tutorial consists of 7 lessons.
-This tutorial intends to guide the learner from the beginner to the intermediate level of proficiency.
+Here learners expand their NetLogo programming skills with more complexity added to such syntactic structures as loops, lists and reporters.
+They are introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity.
+This tutorial consists of seven lessons.
+It intends to guide the learner from the beginner to the intermediate level of proficiency.
 This is achieved by introducing standards of code development such as modular code, pseudocode, debugging and annotating the code.
@@
 
-In tutorial 4 the learner works with a seminal Sugarscape model [@epstein1996] to further expand their agent-based modelling and NetLogo skills.
+In tutorial 4 the learner works with the seminal Sugarscape model [@epstein1996] to further expand their agent-based modelling and NetLogo skills.
 The learner explores how to set up more complex interactions between agents and the environment.
 Furthermore, the principles of setting up good experiments and validating models is explained.
-This tutorial consists of 8 lessons.
-This tutorial intends to solidify the learner's understanding at the intermediate level of proficiency by introducing yet more complex agent-environment interaction, alternative ways to visualize the environment and modelling spatial dynamics.
+This tutorial consists of eigth lessons.
+It intends to solidify the learner's understanding at the intermediate level of proficiency by introducing yet more complex agent-environment interaction, alternative ways to visualize the environment and modelling spatial dynamics.
 Creating toy landscapes is central to lesson 5.
@@
 
-In the final and fifth tutorial the participant learn smore about how to incorporate agent-based modelling in archaeological research.
+In the final and fifth tutorial the participant learns more about how to incorporate agent-based modelling in archaeological research.
 This tutorial focuses less on programming in NetLogo and more on the principle and standards of computer-based modelling.
 The learner learns about the different phases of the model development process.
-This tutorial consists of 5 lessons.
-This tutorial summarizes the material delivered in the previous lessons into one coherent framework consistent with an intermediate level of proficiency.
+This tutorial consists of five lessons.
+It summarizes the material delivered in the previous lessons into one coherent framework consistent with an intermediate level of proficiency.
 This tutorial approaches more theoretical aspects in a practical environment.
@@
 The importance of properly conceptualizing a model before starting the technical phase of the model development is highlighted as it is a common issue for less experienced modellers.
-The experiment design phase, including parametrisation, validation and the analysis and interpretation of the models is explained at length since it often prove challenging for students.
+The experiment design phase, including parametrisation, validation and the analysis and interpretation of the models is explained at length since it often proves challenging for students.
 On a practical side, the BehaviorSpace - NetLogo's experiment environment - is explained to enable the learners to batch run their models.
-Finally the dissemination phase of the model development is also included to ensure that the standards in model publication are widely known.
-Finally, the learners explores the fundamental importance of replication for advancing scientific knowledge.
+Finally the dissemination phase of the model development is also included to ensure that the standards in model publication are widely known and the learners explore the fundamental importance of replication for advancing scientific knowledge.
 
@@
 A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants).
-The respondents came from at least 40 countries that represented almost all continents (see Figure \@ref(fig:nationality).
+The respondents came from at least 40 countries that represented almost all continents (see Figure \@ref(fig:nationality)).
 There was a clear skew towards participants from Europe.
-The number of female respondents slightly outnumbers the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \@ref(fig:gender-age)).
+The number of female respondents slightly outnumbered the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \@ref(fig:gender-age)).
 The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years old indicating that the audience was predominantly composed of early career archaeologists.
@@
 For example, more older people attended the workshop at the CAA conference in April 2023 and more men were present during the workshop at the Reuvensdagen in November 2023.
-This might be due the the differences of audiences at the conferences.
+This might be due to the differences of audiences at the conferences.
 For the Reuvensdagen the number of participants of the workshop was relatively low and this variance might be due to chance.
@@
 
-```{r gender-age, fig.cap="The gender and age distribution of the respondents for each workshop", fig.width=11, echo=FALSE}
+```{r gender-age, fig.cap="The gender and age distribution of the respondents for each workshop.", fig.width=11, echo=FALSE}
 ggplot(pre_surveys, aes(x = `What.is.your.age.`, fill = `What.is.your.gender....Selected.Choice`)) +
@@
 
-```{r, available-theory, fig.cap="Respondents optionion on the quality of theory on ABM faceted out by event.", fig.width=11, echo=FALSE}
+```{r, available-theory, fig.cap="Respondents opinion on the quality of theory on ABM faceted out by event.", fig.width=11, echo=FALSE}
 pre_surveys %>%
@@
 
-During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials \@ref(fig:workshop-impression).
+During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials (Figure \@ref(fig:workshop-impression)).
 Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general.
@@
 
-of the respondents were enthusiastic about the teaching material, with the majority rating it as excellent or good.
-The teachers were rated even higher than the teaching material, with more respondents giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).
-
-The majority of the respondents (153) were enthusiastic about the teaching material, with the majority rating it as excellent or good.
-The teachers were even rated better than the teaching material, with more respondents (165) giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).
+The majority of the respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good.
+The teachers were even rated better than the teaching materials, with more respondents (165) giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).
 
@@
 The respondents generally liked the interactive step-by-step way of going through the tutorials, but also valued the interaction with the teachers strongly.
-The easy and intuitive introduction into ABM and NetLOGO were also mentioned quite often.
-This was also observed during the workshops as most participants were working in the own pace and did not need much help.
+The easy and intuitive introduction in ABM and NetLOGO were also mentioned quite often.
+This was also observed during the workshops as most participants were working at their own pace and did not need much help.
 From a didactic point of view, the self-guided and self-paced way of going through the tutorials is very efficient for teachers and limits the frustration of students that may otherwise go too fast or fall behind the class when the tempo is dictated by the teacher.
@@
 We also received feedback on possible improvements.
-During the first workshop, the CAA-conference in Amsterdam in April 2023, we did the first test of the tutorials and we also instructed the participants that this was work in progress.
+During the first workshop, the CAA conference in Amsterdam in April 2023, we did the first test of the tutorials and we also instructed the participants that this was work in progress.
 For this and the following few workshops we mostly received feedback about bugs causing the tutorial to crash or creating unexpected behaviour.
 The number of references to bugs reduced to zero by the end of the project and participants started providing other feedback, for example the wish to see more examples or expanding the tutorials further.
-Some wanted to introduce more cooperation between participants, which we did not consider primarily aiming the tutorials for individual work.
-A more cooperative aspect of the education materials is possible, even if harder for online events, and we hope that a wider classroom use of the ABMA materials may deliver solutions.
+Some wanted to introduce more cooperation between participants, which we did not consider primarily because the tutorials are aiming for individual work.
+A more cooperative aspect of the education materials is possible, even if harder for online events, and we hope that a wider classroom use of the ABMA materials will deliver solutions.
 It was also interesting to note that over time more and more people did not see any additional room for improvement indicating that the level, content and format of the tutorials met their needs.
@@
 A large group (around 20-30%) was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities.
-Many respondents also shared the context to which they were thinking to apply ABM mentioning topics such as movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms.
-Others thought of demography, social networks, migration or settlement distributions patterns.
+Many respondents also shared the context to which they were thinking to apply ABM, mentioning topics such as movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms.
+Others thought of demography, social networks, migration or settlement distribution patterns.
 The natural environment and the interaction with humans in the past was also mentioned frequently, often in relation to GIS and the question of how complimentary these two computational techniques are.
@@
 
-As mentioned above, the following competence area's were identified as relevant for the tutorials: 1.
+As mentioned above, the following competence areas were identified as relevant for the tutorials: 1.
 Information and data literacy, 2.
@@
 Problem solving [@europeancommissionjointresearchcentre2022].
-In these competence area's various competences are developed towards an advanced or even specialised level.
-A more extensive description of the Digital Competence Framework for Citizens in relation to Agent-Based Modelling for archaeologists is also available [@visser2024].
+In these competence areas various competences are developed towards an advanced or even specialised level.
+A more extensive description of the Digital Competence Framework for Citizens in relation to Agent-Based Modelling for archaeologists is available elsewhere [@visser2024].
 
@@
 Competence Area 3.
-Digital content creation is one of the main competence areas for the ABM-teaching material.
+Digital content creation is one of the main competence areas for the ABM teaching material.
 In the process of learning ABM the users are constantly 3.1 Developing Digital Content and 3.2 Integrating And Re-Elaborating Digital Content.
@@
 A very important aspect of learning ABM is 3.4 Programming.
-The users are guidend through the process of developing the necessary programming skills to build a simulation.
-While the programming language used - NetLogo - is specific to ABM, its syntax follow general programming patterns and the learning is thus transferable to other programming languages.
+The users are guided through the process of developing the necessary programming skills to build a simulation.
+While the programming language used - NetLogo - is specific to ABM, its syntax follows general programming patterns and the learning is thus transferable to other programming languages.
 The users are expected to learn both syntax and more general computational concepts such as modular code development, testing and documenting code.
@@
 Problem solving is very important and is closely tied to 3.4 Programming, since programming involves a lot of problem solving but also the general principles of model development that have a prominent place in the educational materials.
-In addition, the creation of a highly complex ABM consists of problem solving all the time, including development of research questions, conceptualisation of theories, experiment design and validating against data - all of which require creative abilities and high level understanding of the scientific process.
+In addition, the creation of a highly complex ABM consists of problem solving all the time, including development of research questions, conceptualisation of theories, experiment design and validating against data - all of which require creative abilities and a high level understanding of the scientific process.
 All competencies are addressed: 5.1 Solving Technical Problems, is common when writing code, learning about coding, error handling and going from pseudocode to real code.
 The competence 5.2 Identifying Needs And Technological Responses, is a central issue of the course, while 5.3 Creatively Using Digital Technology is an important aspect of so-called model thinking necessary to develop robust and insightful models.
-In tutorial 2 various aspects of 5.4 Identifying Digital Competence Gaps, are relevant, since this tutorial force the user to assess their knowledge and skills.
+In tutorial 2 various aspects of 5.4 Identifying Digital Competence Gaps, are relevant, since this tutorial forces the user to assess their knowledge and skills.
 
@@
 
-The tutorials are publicly hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials/>) and are citeable [@rocks-macqueen2024].
+The ABMA tutorials are publicly hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials/>) and are citeable [@rocks-macqueen2024].
 Anyone interested can download the tutorials and their source code.
-To increase the accessibility and visibility of the tutorials, they are also hosted on a static website ([https://ABMArchaeologists.github.io/ABMA_Website](https://ABMArchaeologists.github.io/ABMA_Website/#/){.uri}) can be followed without the need to download any software.
+To increase the accessibility and visibility of the tutorials, they are also hosted on a static website ([https://ABMArchaeologists.github.io/ABMA_Website](https://ABMArchaeologists.github.io/ABMA_Website/#/){.uri}) and can be followed without the need to download any software.
 The source code of the website is also openly available [@jutte2024].
 
-Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University [@scherjon2019] and external resources people can use to learn agent-based modelling.
+Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University [@scherjon2019] and external resources that people can use to learn agent-based modelling.
 The website was mainly developed by a student group from Saxion University of Applied Sciences, see the acknowledgements.
@@
 The demand for tutorials exceeded our expectations, showing the importance of creating these materials.
-The testing of tutorials during workshops at international an national conferences proved very useful in gaining direct feedback.
-This greatly helped improving the material.
+The testing of tutorials during workshops at international and national conferences proved very useful in gaining direct feedback.
+This greatly helped with improving the materials.
 The tutorials have proven their use in guiding learners into understanding ABM in an archaeological context.
@@
 This project was undertaken with the support of the Erasmus+ programme of the European Union (<https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054>).
+The publication was financed by The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382); IR was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513, the Aarhus University Research Foundation and FWO grant number W001220N.
 
@@
 
-Funding acknowledgements: the publication was financed by The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382); IR was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513, the Aarhus University Research Foundation and FWO grant number W001220N.
-
 # References

### 2024-11-06 paper.tex
Corrections (Karsten/Kenneth)

@@
 \author{}
-\date{\vspace{-2.5em}October 18, 2024}
+\date{\vspace{-2.5em}November 06, 2024}
 
@@
 Computational modelling has become a standard research technique across most scientific disciplines, including archaeology.
-Among the many computational modelling techniques the discipline has found agent-based modelling to be particularly well suited to the questions asked by archaeologists.
+Among the many computational modelling techniques, the discipline has found agent-based modelling to be particularly well suited to the questions asked by archaeologists.
 Agent-based modelling is a simulation technique where autonomous agents interact with each other and with their environment.
-Their aggregated behaviour gives rise to system-level outcomes thus enabling researchers to understand how \emph{micro} (individual) \emph{mechanisms} lead to \emph{macro} (system-level) \emph{patterns}.
+Their aggregated behaviour gives rise to system-level outcomes, thus enabling researchers to understand how \emph{micro} (individual) \emph{mechanisms} lead to \emph{macro} (system-level) \emph{patterns}.
 
@@
 One of the sectorial problems is the access to vocational training required to gain the necessary computer programming and modelling skills.
-A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 76\% of modellers were primarily self-taught with some degree of secondary support i.e.~27\% relied on peer support, only 7\% had any sort of formal training at a workshop or summer school, while only 14\% received training as part of a degree.
+A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 76\% of modellers were primarily self-taught with some degree of secondary support, i.e.~27\% relied on peer support, only 7\% had any sort of formal training at a workshop or summer school, while only 14\% received training as part of a degree.
 
-A consortium of Landward Research Teoranta, Aarhus Universitet, Universiteit Leiden and Saxion University of Applied Sciences recognised this need, and sought to address the limited opportunities that archaeologists have to access high quality training through a project aiming to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.
+A consortium of Landward Research Teoranta, Aarhus Universitet, Leiden University and Saxion University of Applied Sciences recognised this need, and sought to address the limited opportunities that archaeologists have to access high quality training through a project aiming to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.
 
@@
 \item
-  development of a hands-on vocational training programme, aligned with the recently published \emph{Agent-Based Modelling for Archaeology: Simulating the Complexity of Societies} (Romanowska, Wren and Crabtree 2021);
+  development of a hands-on vocational training programme, aligned with the recently published textbook \emph{Agent-Based Modelling for Archaeology: Simulating the Complexity of Societies} (Romanowska, Wren and Crabtree 2021);
 \item
@@
 \item
-  improved opportunities for archaeologists to work in other European countries (transnational mobility) through matching the skills being delivered with the European Commission's Digital Competence Framework for Citizens (2.2) and the sectoral Skills Passport for Archaeology, which will mean that it will be easier for individual workers to demonstrate their skills to employers across Europe;
+  improved opportunities for archaeologists to work in other European countries (transnational mobility) through matching the skills being delivered with the European Commission's Digital Competence Framework for Citizens (2.2) and the sectoral Skills Passport for Archaeology, which means that it will be easier for individual workers to demonstrate their skills to employers across Europe;
 \item
@@
 The ABMA learning materials assume that participants have learning skills at least at a secondary education level.
-It is suitable for participants with at least some background in archaeology, such as those typical among students after the first year of a Bachelor study in Archaeology, i.e., above EQF levels 4 or 5.
+They are suitable for participants with at least some background in archaeology, such as those typical among students after the first year of a Bachelor study in Archaeology, i.e., above EQF levels 4 or 5.
 In addition, the tutorials are aimed at professionals working in archaeology who have some experience with computer applications.
 We also assume that learners are proficient in English and have at least the Reading B1 level, but Reading B2 is recommended (Council of Europe 2020).
-The tutorials were also developed with a secondary aim to develop learners' general skills in the following competence area's: 1.
+The tutorials were also developed with a secondary aim to develop learners' general skills in the following competence areas: 1.
 Information and data literacy, 2.
@@
 
-After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events all of which were run by members of the project team.
+After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events all of which were run by members of the project team (Table \ref{tab:events}).
 Recognising the difficulty of delivering practical training in an online environment during the online events several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things.
@@
 For some events participants had to register beforehand and we were able to send the pre-workshop survey by email.
-At other events no registration was possible or necessary and the pre-workshop survey was given at the start of the event.
+At other events no registration was possible or necessary and the pre-workshop survey was shared at the start of the event.
 The post-workshop survey was distributed using QR-codes or links at the end of the event.
 
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}
-\caption{Conferences, events and situations were workshops were held at which the tutorials were tested.}\tabularnewline
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-Conference or event
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Period
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Online, in person or hybrid
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Number of participants
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Number of registrations
-\end{minipage} \\
-\midrule\noalign{}
-\endfirsthead
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-Conference or event
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Period
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Online, in person or hybrid
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Number of participants
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Number of registrations
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-CAA Amsterdam & April 2023 & In person & 36 & 53 \\
-EAA Belfast & August 2023 & In person & 21 & No registration \\
-CAA-DE/NL-Fl Online workshop & October 2023 & Online & 40 & 76 \\
-Reuvensdagen & November 2023 & In person & 13 & No registration \\
-CAA-UK & November 2023 & Hybrid & 53 & No registration \\
-Leiden (course) & December 2023 & In person & 11 & No registration \\
-Aarhus & January/February 2024 & Online & 176 & \textgreater500 \\
-Saxion & March 2023-January 2024 & In person & 18 & No registration \\
-\textbf{Total} & & & \textbf{368} & \textbf{\textgreater628} \\
-\end{longtable}
+\begin{table}
+
+\caption{\label{tab:events}Conferences, events and situations where workshops were held at which the tutorials were tested.}
+\centering
+\begin{tabular}[t]{l|l|l|l|l}
+\hline
+Conference or event & Period & Online, in person or hybrid & Number of participants & Number of registrations\\
+\hline
+CAA Amsterdam & April 2023 & In person & 36 & 53\\
+\hline
+EAA Belfast & August 2023 & In person & 21 & No registration\\
+\hline
+CAA-DE/NL-Fl Online workshop & October 2023 & Online & 40 & 76\\
+\hline
+Reuvensdagen & November 2023 & In person & 13 & No registration\\
+\hline
+CAA-UK & November 2023 & Hybrid & 53 & No registration\\
+\hline
+Leiden (course) & December 2023 & In person & 11 & No registration\\
+\hline
+Aarhus & January/February 2024 & Online & 176 & >500\\
+\hline
+Saxion & March 2023-January 2024 & In person & 18 & No registration\\
+\hline
+**Total** &  &  & **368** & **>628**\\
+\hline
+\end{tabular}
+\end{table}
 
@@
 
-The final set of ABMA tutorals are available for download at \url{https://github.com/ABMArchaeologists/ABMA_tutorials} and are citeable (Rocks-Macqueen et al.
+The final set of ABMA tutorials is available for download at \url{https://github.com/ABMArchaeologists/ABMA_tutorials} and is citeable (Rocks-Macqueen et al.
 2024).
-The set of developed tutorials consists of the following tutorials:
+The set consists of the following tutorials:
 
@@
 
-Each tutorials consists of a number of individual lessons, each of which guides the learner in a self-pace manner through an increasingly challenging set of exercises that build the skills and knowledge necessary to become a confident agent-based modeller.
+Each tutorials consists of a number of individual lessons, each of which guides the learner in a self-paced manner through an increasingly challenging set of exercises that build the skills and knowledge necessary to become a confident agent-based modeller.
 
-In the first tutorial the user learn what simulations and agent-based models are and how they can help in archaeological research (\ref{fig:screenshot-lesson-3}).
-This tutorial consists of 4 lessons.
+In the first tutorial the users learn what simulations and agent-based models are and how they can help in archaeological research (Figure \ref{fig:screenshot-lesson-3}).
+This tutorial consists of four lessons and makes use of the Artificial Anasazi model (Axtell et al.
+2002).
 The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific.
@@
 
-In tutorial 2 the user will learn the basics of NetLogo by making your their simulation on the Out of Africal dispersal of \emph{Homo sapiens} (Young and Bettinger 1995).The learner works through the basics of NetLogo syntax and learn how to set up a simulation and visualize its outcomes.
-This tutorial consists of 9 lessons.
+In tutorial 2 the user will learn the basics of NetLogo by making their simulation on the Out of Africal dispersal of \emph{Homo sapiens} (Young and Bettinger 1995).The learner works through the basics of NetLogo syntax and learns how to set up a simulation and visualize its outcomes.
+This tutorial consists of nine lessons.
 This tutorial intends to guide the learner from the introduction to the beginner level of proficiency.
@@
 This is achieved by working with the NetLogo web interface that enables highlighting specific words and buttons and giving visual cues as to how to proceed.
-The tutorial intoduces several most commonly used built-in functions in Netlogo (called \emph{primitives}) are introduces the general structure of all models cosisting of the initialization phase and the main simulation loop.
-It walks the participants through te most important NetLogo settings, such as dimensions, coordinates and the origin point of the grid on which the simulation runs.
+The tutorial introduces several most commonly used built-in functions in Netlogo (called \emph{primitives}) and introduces the general structure of all models consisting of the initialization phase and the main simulation loop.
+It walks the participants through the most important NetLogo settings, such as dimensions, coordinates and the origin point of the grid on which the simulation runs.
 The learners gain their first stripes in programming by using simple loops and functions as well as conditional statements.
@@
 In the third tutorial, the learner builds a simple trade model, again based on a published simulation (Romanowska 2018).
-Here learners are expanding their NetLogo programming skills with more complexity added to such syntactic structures as loops, lists and reporters.
-The learner is introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity.
-This tutorial consists of 7 lessons.
-This tutorial intends to guide the learner from the beginner to the intermediate level of proficiency.
+Here learners expand their NetLogo programming skills with more complexity added to such syntactic structures as loops, lists and reporters.
+They are introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity.
+This tutorial consists of seven lessons.
+It intends to guide the learner from the beginner to the intermediate level of proficiency.
 This is achieved by introducing standards of code development such as modular code, pseudocode, debugging and annotating the code.
@@
 
-In tutorial 4 the learner works with a seminal Sugarscape model (Epstein and Axtell 1996) to further expand their agent-based modelling and NetLogo skills.
+In tutorial 4 the learner works with the seminal Sugarscape model (Epstein and Axtell 1996) to further expand their agent-based modelling and NetLogo skills.
 The learner explores how to set up more complex interactions between agents and the environment.
 Furthermore, the principles of setting up good experiments and validating models is explained.
-This tutorial consists of 8 lessons.
-This tutorial intends to solidify the learner's understanding at the intermediate level of proficiency by introducing yet more complex agent-environment interaction, alternative ways to visualize the environment and modelling spatial dynamics.
+This tutorial consists of eigth lessons.
+It intends to solidify the learner's understanding at the intermediate level of proficiency by introducing yet more complex agent-environment interaction, alternative ways to visualize the environment and modelling spatial dynamics.
 Creating toy landscapes is central to lesson 5.
@@
 A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants).
-The respondents came from at least 40 countries that represented almost all continents (see Figure \ref{fig:nationality}.
+The respondents came from at least 40 countries that represented almost all continents (see Figure \ref{fig:nationality}).
 There was a clear skew towards participants from Europe.
-The number of female respondents slightly outnumbers the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \ref{fig:gender-age}).
+The number of female respondents slightly outnumbered the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \ref{fig:gender-age}).
 The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years old indicating that the audience was predominantly composed of early career archaeologists.
@@
 \includegraphics{paper_files/figure-latex/gender-age-1.pdf}
-\caption{\label{fig:gender-age}The gender and age distribution of the respondents for each workshop}
+\caption{\label{fig:gender-age}The gender and age distribution of the respondents for each workshop.}
 \end{figure}
@@
 \includegraphics{paper_files/figure-latex/available-theory-1.pdf}
-\caption{\label{fig:available-theory}Respondents optionion on the quality of theory on ABM faceted out by event.}
+\caption{\label{fig:available-theory}Respondents opinion on the quality of theory on ABM faceted out by event.}
 \end{figure}
@@
 
-During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials \ref{fig:workshop-impression}.
+During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials (Figure \ref{fig:workshop-impression}).
 Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general.
@@
 
-of the respondents were enthusiastic about the teaching material, with the majority rating it as excellent or good.
-The teachers were rated even higher than the teaching material, with more respondents giving them the \emph{excellent} score (see Figure \ref{fig:rating-teaching}).
-
-The majority of the respondents (153) were enthusiastic about the teaching material, with the majority rating it as excellent or good.
-The teachers were even rated better than the teaching material, with more respondents (165) giving them the \emph{excellent} score (see Figure \ref{fig:rating-teaching}).
+The majority of the respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good.
+The teachers were even rated better than the teaching materials, with more respondents (165) giving them the \emph{excellent} score (see Figure \ref{fig:rating-teaching}).
 
@@
 A large group (around 20-30\%) was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities.
-Many respondents also shared the context to which they were thinking to apply ABM mentioning topics such as movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms.
-Others thought of demography, social networks, migration or settlement distributions patterns.
+Many respondents also shared the context to which they were thinking to apply ABM, mentioning topics such as movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms.
+Others thought of demography, social networks, migration or settlement distribution patterns.
 The natural environment and the interaction with humans in the past was also mentioned frequently, often in relation to GIS and the question of how complimentary these two computational techniques are.
@@
 2022).
-In these competence area's various competences are developed towards an advanced or even specialised level.
-A more extensive description of the Digital Competence Framework for Citizens in relation to Agent-Based Modelling for archaeologists is also available (Visser et al.
+In these competence areas various competences are developed towards an advanced or even specialised level.
+A more extensive description of the Digital Competence Framework for Citizens in relation to Agent-Based Modelling for archaeologists is available elsewhere (Visser et al.
 2024).
@@
 
-The tutorials are publicly hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials/}) and are citeable (Rocks-Macqueen et al.
+The ABMA tutorials are publicly hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials/}) and are citeable (Rocks-Macqueen et al.
 2024).
 Anyone interested can download the tutorials and their source code.
-To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (\href{https://ABMArchaeologists.github.io/ABMA_Website/\#/}{https://ABMArchaeologists.github.io/ABMA\_Website}) can be followed without the need to download any software.
+To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (\href{https://ABMArchaeologists.github.io/ABMA_Website/\#/}{https://ABMArchaeologists.github.io/ABMA\_Website}) and can be followed without the need to download any software.
 The source code of the website is also openly available (Jutte et al.
@@
 
-Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University (Scherjon, Romanowska and Lambers 2019) and external resources people can use to learn agent-based modelling.
+Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University (Scherjon, Romanowska and Lambers 2019) and external resources that people can use to learn agent-based modelling.
 The website was mainly developed by a student group from Saxion University of Applied Sciences, see the acknowledgements.
@@
 This project was undertaken with the support of the Erasmus+ programme of the European Union (\url{https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054}).
+The publication was financed by The Carlsberg Foundation's \emph{Young Researcher Fellowship} (CF21-0382); IR was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513, the Aarhus University Research Foundation and FWO grant number W001220N.
 
@@
 
-Funding acknowledgements: the publication was financed by The Carlsberg Foundation's \emph{Young Researcher Fellowship} (CF21-0382); IR was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513, the Aarhus University Research Foundation and FWO grant number W001220N.
-
 \hypertarget{references}{%
@@
 2024 \emph{Teaching material Agent-Based Modelling for Archaeologists (EN/NL)} DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.10950153}{10.5281/zenodo.10950153}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-axtell2002}{}}%
+Axtell, RL, Epstein, JM, Dean, JS, Gumerman, GJ, Swedlund, AC, Harburger, J, Chakravarty, S, Hammond, R, Parker, J and Parker, M.
+2002 Population growth and collapse in a multiagent model of the kayenta anasazi in long house valley.
+\emph{Proceedings of the National Academy of Sciences} 99(suppl{\_}3): 7275--7279.
+DOI: https://doi.org/\href{https://doi.org/10.1073/pnas.092080799}{10.1073/pnas.092080799}.
 
@@
 Karl, R, Möller, K, Connolly, D, Trausmuth, T, Krierer, KR, Rocks-Macqueen, D, Aitchison, K, Edeso, M, Pintucci, A and Marciniak, A.
-2019 Der archäologische Grabungs-Kompetenz-Pass {\textendash} ein Mittel zur Dokumentation durch Praxis erworbener archäologischer Fertigkeiten.
+2019 Der archäologische Grabungs-Kompetenz-Pass - ein Mittel zur Dokumentation durch Praxis erworbener archäologischer Fertigkeiten.
 \emph{Archäologische Informationen} 42: 237--249.
@@
 Romanowska, I and Scherjon, F.
-2023 \emph{Bibliometric analysis of agent-based simulation in archaeology: People, topics, and future prospects}.
+2023 Bibliometric analysis of agent-based simulation in archaeology: People, topics, and future prospects.
 In: Kalayci, T, Lambers, K, and Klinkenberg, V (eds.).
+\emph{Digital archaeology: Promises and impasses}.
 Analecta praehistorica leidensia.

### 2024-11-12 paper.Rmd
minor corrections

@@
 
-While Agent-Based Modelling importance in archaeological practice is steadily growing [@romanowska2023] there are significant bariers to its wider adoption.
-One of the sectorial problems is the access to vocational training required to gain the necessary computer programming and modelling skills.
+While Agent-Based Modelling importance in archaeological practice is steadily growing [@romanowska2023] there are significant barriers to its wider adoption.
+One of the problems is the access to vocational training required to gain the necessary computer programming and modelling skills.
 A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 76% of modellers were primarily self-taught with some degree of secondary support, i.e.
@@
 
-A consortium of Landward Research Teoranta, Aarhus Universitet, Leiden University and Saxion University of Applied Sciences recognised this need, and sought to address the limited opportunities that archaeologists have to access high quality training through a project aiming to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.
+A consortium of Landward Research Teoranta, Aarhus University, Leiden University and Saxion University of Applied Sciences recognised this need, and sought to address the limited opportunities that archaeologists have to access high quality training.
+A project was set up to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.
 
@@
 The tutorials were first designed using storyboards and later converted into interactive guided walk-throughs using NetLOGO Web (<https://www.netlogoweb.org/launch>) and Javascript and hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials>).
-During the project, initial testing was done by all authors and issues in GitHub were used to tackle bugs.
+During the project, initial testing was done by all authors and the issues functionality in GitHub was used to tackle bugs.
 
@@
 After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events all of which were run by members of the project team (Table \@ref(tab:events)).
-Recognising the difficulty of delivering practical training in an online environment during the online events several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things.
+Recognising the difficulty of delivering practical training in an online environment (Microsoft Teams and Zoom) during the online events several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things.
 Throughout the project we sought feedback from the participants and used it to improve the tutorials prior to the following events.
 The feedback was obtained through two surveys using Qualtrics (<https://www.qualtrics.com/>).
-The participants of the events were asked to respond to two surveys: one before the start of the event and one completed after the end of the event.
+The participants of the events were asked to respond to two surveys: one before the start of the event and one after the end of the event.
 The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix 1 for the questions).
@@
 This is an interdisciplinary semester in which students of at least three different study-programmes or disciplines work together on a complex problem/project (<https://www.saxion.edu/business-and-research/collaborate-with-saxion/smart-solutions>).
-Within the semester students have to come up with new ideas or solutions to help a client with a wicked problem.
+Within the semester students have to come up with new ideas or solutions to help a client with a wicked, or at least challenging, problem.
 The backgrounds of the various students we worked with were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media & Game Technologies, and ICT.
 The aim of involving interdisciplinary groups in the project was to involve students in current research projects, to enable them to learn new things and share their innovative ideas and solutions with us.
-The students groups were supplementary to the deliverables of the project, but their new ideas were benificial to the project.
+The students groups were supplementary to the deliverables of the project, but their new ideas were beneficial to the project.
 Their help during workshops was useful, but they also were engaged in developing educational material [@aalpoel2024], testing the tutorials, developing a style for the website [@jutte2024] and other materials.
@@
 
-Each tutorials consists of a number of individual lessons, each of which guides the learner in a self-paced manner through an increasingly challenging set of exercises that build the skills and knowledge necessary to become a confident agent-based modeller.
+Each tutorials consists of a number of individual lessons, which guide the learner in a self-paced manner through an increasingly challenging set of exercises.
+Collectively, they build the skills and knowledge necessary to become a confident agent-based modeller.
 
@@
 The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific.
-Various concepts related to ABM, such as the definitions of model and simulation, are introduced and the participants make their first steps in NetLogo syntax.
+Various concepts related to ABM, such as the definitions of model and simulation, are introduced and the participants take their first steps in NetLogo syntax.
 The third lesson explains how ABMs are used in archaeological research.
@@
 
-In tutorial 2 the user will learn the basics of NetLogo by making their simulation on the Out of Africal dispersal of *Homo sapiens* [@young1995].The learner works through the basics of NetLogo syntax and learns how to set up a simulation and visualize its outcomes.
+In tutorial 2 the user will learn the basics of NetLogo by making their simulation on the Out of Africal dispersal of *Homo sapiens* [@young1995].
+The learner works through the basics of NetLogo syntax and learns how to set up a simulation and visualize its outcomes.
 This tutorial consists of nine lessons.
@@
 This is achieved by working with the NetLogo web interface that enables highlighting specific words and buttons and giving visual cues as to how to proceed.
-The tutorial introduces several most commonly used built-in functions in Netlogo (called *primitives*) and introduces the general structure of all models consisting of the initialization phase and the main simulation loop.
-It walks the participants through the most important NetLogo settings, such as dimensions, coordinates and the origin point of the grid on which the simulation runs.
+The tutorial introduces several commonly used built-in functions in Netlogo (called *primitives*) and introduces the general structure of all models consisting of the initialization phase and the main simulation loop.
+It walks the participants through the most important features of the NetLogo world, such as dimensions, coordinates and the origin point of the grid on which the simulation runs.
 The learners gain their first stripes in programming by using simple loops and functions as well as conditional statements.
@@
 In the third tutorial, the learner builds a simple trade model, again based on a published simulation [@romanowska2018].
-Here learners expand their NetLogo programming skills with more complexity added to such syntactic structures as loops, lists and reporters.
+Here learners expand their NetLogo programming skills with more complex syntactic structures as loops, lists and reporters.
 They are introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity.
@@
 Furthermore, the principles of setting up good experiments and validating models is explained.
-This tutorial consists of eigth lessons.
+This tutorial consists of eight lessons.
 It intends to solidify the learner's understanding at the intermediate level of proficiency by introducing yet more complex agent-environment interaction, alternative ways to visualize the environment and modelling spatial dynamics.
@@
 The learner dives into the topic of experiment design at a practical level but also with regards to the scientific principles, such as validation.
-Thus they are guided through the process of setting up experiments in NetLogo and collecting results using monitors and plots.
-The topic of validation of agent-based models is critical for most archaelogical applications and thus several ways to compare simulation results with the archaeological record are explored.
+Thus, they are guided through the process of setting up experiments in NetLogo and collecting results using monitors and plots.
+The topic of validation of agent-based models is critical for most archaeological applications and thus several ways to compare simulation results with the archaeological record are explored.
 
-In the final and fifth tutorial the participant learns more about how to incorporate agent-based modelling in archaeological research.
+In the final and fifth tutorial the learner learns more about how to incorporate agent-based modelling in archaeological research.
 This tutorial focuses less on programming in NetLogo and more on the principle and standards of computer-based modelling.
@@
 As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before.
-The respondents were also asked how they rated the available knowledge on ABM (Figure \@ref(fig:available-theory)).
+The respondents were also asked how they rated the available resources on ABM (Figure \@ref(fig:available-theory)).
 While 64 of them had no opinion on the subject, a proportion of the respondents (28) answered that they rated the available theory on ABM as limited and only a small group (39) as sufficient or better.
@@
 The majority of the respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good.
-The teachers were even rated better than the teaching materials, with more respondents (165) giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).
+The teachers were rated even better than the teaching materials, with more respondents (165) giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).
 
@@
 The first passport was created as part of *The Ardnamurchan Transitions Project*, a long running investigation of the archaeology and history of the Ardnamurchan Peninsula, Western Scotland, run by the Universities of Leicester and Manchester, with Archaeology Scotland [@cobb2020].
-Since then the passport has spread around the world to countries such as Germany [@karl2019] and the United States. While there are many different versions, in its simplest form it is a log booklet listing skills that others can verify by signing off on their level of competencies.
+Since then, the passport has spread around the world to countries such as Germany [@karl2019] and the United States. While there are many different versions, in its simplest form it is a log booklet listing skills that others can verify by signing off on their level of competencies.
 
@@
 Anyone interested can download the tutorials and their source code.
-To increase the accessibility and visibility of the tutorials, they are also hosted on a static website ([https://ABMArchaeologists.github.io/ABMA_Website](https://ABMArchaeologists.github.io/ABMA_Website/#/){.uri}) and can be followed without the need to download any software.
+To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (<https://abmarchaeologists.github.io/ABMA_website>) and can be followed without the need to download any software.
 The source code of the website is also openly available [@jutte2024].

### 2024-11-12 paper.tex
minor corrections

@@
 \author{}
-\date{\vspace{-2.5em}November 06, 2024}
+\date{\vspace{-2.5em}November 12, 2024}
 
@@
 
-While Agent-Based Modelling importance in archaeological practice is steadily growing (Romanowska and Scherjon 2023) there are significant bariers to its wider adoption.
-One of the sectorial problems is the access to vocational training required to gain the necessary computer programming and modelling skills.
+While Agent-Based Modelling importance in archaeological practice is steadily growing (Romanowska and Scherjon 2023) there are significant barriers to its wider adoption.
+One of the problems is the access to vocational training required to gain the necessary computer programming and modelling skills.
 A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 76\% of modellers were primarily self-taught with some degree of secondary support, i.e.~27\% relied on peer support, only 7\% had any sort of formal training at a workshop or summer school, while only 14\% received training as part of a degree.
 
-A consortium of Landward Research Teoranta, Aarhus Universitet, Leiden University and Saxion University of Applied Sciences recognised this need, and sought to address the limited opportunities that archaeologists have to access high quality training through a project aiming to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.
+A consortium of Landward Research Teoranta, Aarhus University, Leiden University and Saxion University of Applied Sciences recognised this need, and sought to address the limited opportunities that archaeologists have to access high quality training.
+A project was set up to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.
 
@@
 The tutorials were first designed using storyboards and later converted into interactive guided walk-throughs using NetLOGO Web (\url{https://www.netlogoweb.org/launch}) and Javascript and hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials}).
-During the project, initial testing was done by all authors and issues in GitHub were used to tackle bugs.
+During the project, initial testing was done by all authors and the issues functionality in GitHub was used to tackle bugs.
 
@@
 After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events all of which were run by members of the project team (Table \ref{tab:events}).
-Recognising the difficulty of delivering practical training in an online environment during the online events several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things.
+Recognising the difficulty of delivering practical training in an online environment (Microsoft Teams and Zoom) during the online events several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things.
 Throughout the project we sought feedback from the participants and used it to improve the tutorials prior to the following events.
 The feedback was obtained through two surveys using Qualtrics (\url{https://www.qualtrics.com/}).
-The participants of the events were asked to respond to two surveys: one before the start of the event and one completed after the end of the event.
+The participants of the events were asked to respond to two surveys: one before the start of the event and one after the end of the event.
 The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix 1 for the questions).
@@
 This is an interdisciplinary semester in which students of at least three different study-programmes or disciplines work together on a complex problem/project (\url{https://www.saxion.edu/business-and-research/collaborate-with-saxion/smart-solutions}).
-Within the semester students have to come up with new ideas or solutions to help a client with a wicked problem.
+Within the semester students have to come up with new ideas or solutions to help a client with a wicked, or at least challenging, problem.
 The backgrounds of the various students we worked with were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media \& Game Technologies, and ICT.
 The aim of involving interdisciplinary groups in the project was to involve students in current research projects, to enable them to learn new things and share their innovative ideas and solutions with us.
-The students groups were supplementary to the deliverables of the project, but their new ideas were benificial to the project.
+The students groups were supplementary to the deliverables of the project, but their new ideas were beneficial to the project.
 Their help during workshops was useful, but they also were engaged in developing educational material (Aalpoel et al.
@@
 
-In tutorial 2 the user will learn the basics of NetLogo by making their simulation on the Out of Africal dispersal of \emph{Homo sapiens} (Young and Bettinger 1995).The learner works through the basics of NetLogo syntax and learns how to set up a simulation and visualize its outcomes.
+In tutorial 2 the user will learn the basics of NetLogo by making their simulation on the Out of Africal dispersal of \emph{Homo sapiens} (Young and Bettinger 1995).
+The learner works through the basics of NetLogo syntax and learns how to set up a simulation and visualize its outcomes.
 This tutorial consists of nine lessons.
@@
 This is achieved by working with the NetLogo web interface that enables highlighting specific words and buttons and giving visual cues as to how to proceed.
-The tutorial introduces several most commonly used built-in functions in Netlogo (called \emph{primitives}) and introduces the general structure of all models consisting of the initialization phase and the main simulation loop.
-It walks the participants through the most important NetLogo settings, such as dimensions, coordinates and the origin point of the grid on which the simulation runs.
+The tutorial introduces several commonly used built-in functions in Netlogo (called \emph{primitives}) and introduces the general structure of all models consisting of the initialization phase and the main simulation loop.
+It walks the participants through the most important features of the NetLogo world, such as dimensions, coordinates and the origin point of the grid on which the simulation runs.
 The learners gain their first stripes in programming by using simple loops and functions as well as conditional statements.
@@
 In the third tutorial, the learner builds a simple trade model, again based on a published simulation (Romanowska 2018).
-Here learners expand their NetLogo programming skills with more complexity added to such syntactic structures as loops, lists and reporters.
+Here learners expand their NetLogo programming skills with more complex syntactic structures as loops, lists and reporters.
 They are introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity.
@@
 As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before.
-The respondents were also asked how they rated the available knowledge on ABM (Figure \ref{fig:available-theory}).
+The respondents were also asked how they rated the available resources on ABM (Figure \ref{fig:available-theory}).
 While 64 of them had no opinion on the subject, a proportion of the respondents (28) answered that they rated the available theory on ABM as limited and only a small group (39) as sufficient or better.
@@
 The majority of the respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good.
-The teachers were even rated better than the teaching materials, with more respondents (165) giving them the \emph{excellent} score (see Figure \ref{fig:rating-teaching}).
+The teachers were rated even better than the teaching materials, with more respondents (165) giving them the \emph{excellent} score (see Figure \ref{fig:rating-teaching}).
 
@@
 The first passport was created as part of \emph{The Ardnamurchan Transitions Project}, a long running investigation of the archaeology and history of the Ardnamurchan Peninsula, Western Scotland, run by the Universities of Leicester and Manchester, with Archaeology Scotland (Cobb and Croucher 2020).
-Since then the passport has spread around the world to countries such as Germany (Karl et al.
+Since then, the passport has spread around the world to countries such as Germany (Karl et al.
 2019) and the United States.~While there are many different versions, in its simplest form it is a log booklet listing skills that others can verify by signing off on their level of competencies.~
@@
 Anyone interested can download the tutorials and their source code.
-To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (\href{https://ABMArchaeologists.github.io/ABMA_Website/\#/}{https://ABMArchaeologists.github.io/ABMA\_Website}) and can be followed without the need to download any software.
+To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (\url{https://abmarchaeologists.github.io/ABMA_website}) and can be followed without the need to download any software.
 The source code of the website is also openly available (Jutte et al.

### 2024-11-13 paper.Rmd
correcting image captions + reference to repository

@@
 library(forcats)
+library(ggpubr)
 # read data
@@
 The following packages were used for analyses: ggplot2 [@wickham2016], dplyr [@wickham2023a], tidyr [@wickham2023], forcats[@wickham2023b], lubridate [@grolemund2011] and stringr [@wickham2023c].
-For the sake of reproducibility [@marwick2017] this paper was written in R Markdown (<https://rmarkdown.rstudio.com/>), with data and code available at <https://github.com/ABMArchaeologists/ABMA_paper> (...
-insert Zenodo reference...)
+For the sake of reproducibility [@marwick2017] this paper was written in R Markdown (<https://rmarkdown.rstudio.com/>), with data and code available at <https://github.com/ABMArchaeologists/ABMA_paper> [@visser].
 
@@
 
-```{r, abm-knowledge, fig.cap="Respondents knowledge of ABM per event knowledge of the various softwares for ABM, before participating in the workshops.", fig.show="hold", out.width="50%", echo=FALSE}
-pre_surveys %>%
+```{r, abm-knowledge, fig.cap="Respondents knowledge of ABM per event (top), and knowledge of the various softwares for ABM (bottom), before participating in the workshops.", echo=FALSE}
+knowledge_plot <- pre_surveys %>%
   filter(`Do.you.know.what.Agent.Based.Modeling.is.`!="Not shared") %>%
@@
   xlab("") + scale_fill_discrete(name = "ABM known") + ylab("Percentage") +  scale_y_continuous(labels = scales::label_percent(scale = 100, prefix = "", suffix = "")) + coord_flip()
-pre_surveys %>%
+software_plot <- pre_surveys %>%
   select(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`) %>%
@@
   ggplot(aes(x=reorder(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`, count_software))) + geom_bar() + scale_x_discrete(labels = ) + xlab("ABM Software known") + coord_flip()
+ggarrange(knowledge_plot, software_plot, nrow = 2, ncol = 1)
 ```
@@
 
-```{r, rating-teaching, fig.cap="Respondents rating of the teaching material in general faceted for each event.", fig.width=11, fig.show='hold', out.height="50%", echo=FALSE}
-post_surveys %>%
+```{r, rating-teaching, fig.cap="Respondents rating of the teaching material (top) and teachers (bottom) in general faceted for each event.", fig.width=11, fig.height=8, echo=FALSE}
+rate_material_plot <- post_surveys %>%
   mutate(How.would.you.rate.the.teaching.material. = fct_relevel(How.would.you.rate.the.teaching.material., factor_levels)) %>%
@@
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
-post_surveys %>%
+rate_teacher_plot <- post_surveys %>%
   mutate(How.would.you.rate.the.trainers.teachers. = fct_relevel(How.would.you.rate.the.trainers.teachers., factor_levels)) %>%
@@
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
-
+ggarrange(rate_material_plot, rate_teacher_plot, nrow = 2, ncol = 1)
 ```

### 2024-11-13 paper.tex
correcting image captions + reference to repository

@@
 \author{}
-\date{\vspace{-2.5em}November 12, 2024}
+\date{\vspace{-2.5em}November 13, 2024}
 
@@
 2023), tidyr (Wickham, Vaughan and Girlich 2023), forcats(Wickham 2023a), lubridate (Grolemund and Wickham 2011) and stringr (Wickham 2023b).
-For the sake of reproducibility (Marwick 2017) this paper was written in R Markdown (\url{https://rmarkdown.rstudio.com/}), with data and code available at \url{https://github.com/ABMArchaeologists/ABMA_paper} (\ldots{} insert Zenodo reference\ldots)
+For the sake of reproducibility (Marwick 2017) this paper was written in R Markdown (\url{https://rmarkdown.rstudio.com/}), with data and code available at \url{https://github.com/ABMArchaeologists/ABMA_paper} (Visser et al.
+2024a).
 
@@
 \begin{figure}
-\includegraphics[width=0.5\linewidth]{paper_files/figure-latex/abm-knowledge-1} \includegraphics[width=0.5\linewidth]{paper_files/figure-latex/abm-knowledge-2} \caption{Respondents knowledge of ABM per event knowledge of the various softwares for ABM, before participating in the workshops.}\label{fig:abm-knowledge}
+\centering
+\includegraphics{paper_files/figure-latex/abm-knowledge-1.pdf}
+\caption{\label{fig:abm-knowledge}Respondents knowledge of ABM per event (top), and knowledge of the various softwares for ABM (bottom), before participating in the workshops.}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[height=0.5\textheight]{paper_files/figure-latex/rating-teaching-1} \includegraphics[height=0.5\textheight]{paper_files/figure-latex/rating-teaching-2} \caption{Respondents rating of the teaching material in general faceted for each event.}\label{fig:rating-teaching}
+\centering
+\includegraphics{paper_files/figure-latex/rating-teaching-1.pdf}
+\caption{\label{fig:rating-teaching}Respondents rating of the teaching material (top) and teachers (bottom) in general faceted for each event.}
 \end{figure}
@@
 A more extensive description of the Digital Competence Framework for Citizens in relation to Agent-Based Modelling for archaeologists is available elsewhere (Visser et al.
-2024).
+2024b).
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-visser}{}}%
+Visser, RM, Jutte, A, Rocks-Macqueen, D, Romanowska, I, Brughmans, T, Knaap, L van der, Lambers, K, Aitchison, K, Linssen, J and Pažout, A.
+2024a \emph{ABMA: Lessons learned.
+Data and source code}.
+DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.14134218}{10.5281/zenodo.14134218}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-visser2024}{}}%
 Visser, RM, Lambers, K, Aitchison, K, Jutte, A, Rocks-Macqueen, D, Knaap, L van der, Romanowska, I and Brughmans, T.
-2024 \emph{The Digital Competence Framework for Citizens (2.2) and Agent-Based modelling for Archaeologists}.
+2024b \emph{The Digital Competence Framework for Citizens (2.2) and Agent-Based modelling for Archaeologists}.
 DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.11046488}{10.5281/zenodo.11046488}.

### 2024-11-13 paper.tex
minor corrections

@@
 For the sake of reproducibility (Marwick 2017) this paper was written in R Markdown (\url{https://rmarkdown.rstudio.com/}), with data and code available at \url{https://github.com/ABMArchaeologists/ABMA_paper} (Visser et al.
-2024a).
+2024b).
 
@@
 A more extensive description of the Digital Competence Framework for Citizens in relation to Agent-Based Modelling for archaeologists is available elsewhere (Visser et al.
-2024b).
+2024a).
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-visser2024}{}}%
+Visser, RM, Lambers, K, Aitchison, K, Jutte, A, Rocks-Macqueen, D, Knaap, L van der, Romanowska, I and Brughmans, T.
+2024a \emph{The Digital Competence Framework for Citizens (2.2) and Agent-Based modelling for Archaeologists}.
+DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.11046488}{10.5281/zenodo.11046488}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-visser}{}}%
-Visser, RM, Jutte, A, Rocks-Macqueen, D, Romanowska, I, Brughmans, T, Knaap, L van der, Lambers, K, Aitchison, K, Linssen, J and Pažout, A.
-2024a \emph{ABMA: Lessons learned.
+Visser, RM, Lambers, K, Aitchison, K, Jutte, A, Rocks-Macqueen, D, Knaap, L van der, Romanowska, I, Brughmans, T, Linssen, J and Pažout, A.
+2024b \emph{ABMA: Lessons learned.
 Data and source code}.
 DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.14134218}{10.5281/zenodo.14134218}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-visser2024}{}}%
-Visser, RM, Lambers, K, Aitchison, K, Jutte, A, Rocks-Macqueen, D, Knaap, L van der, Romanowska, I and Brughmans, T.
-2024b \emph{The Digital Competence Framework for Citizens (2.2) and Agent-Based modelling for Archaeologists}.
-DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.11046488}{10.5281/zenodo.11046488}.
 

### 2025-03-17 paper.Rmd
update paper based on reviews

@@
 ---
-title: "Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists"
+title: "Lessons learned: creating tutorials teaching agent-based modelling for archaeologists"
 authors:
@@
 
-While Agent-Based Modelling importance in archaeological practice is steadily growing [@romanowska2023] there are significant barriers to its wider adoption.
-One of the problems is the access to vocational training required to gain the necessary computer programming and modelling skills.
+While agent-based modelling importance in archaeological practice is steadily growing [@romanowska2023] there are significant barriers to its wider adoption.
+One of the problems is the access to vocational training required to gain the necessary computer programming and modelling skills [@daems2020].
 A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 76% of modellers were primarily self-taught with some degree of secondary support, i.e.
@@
 This project, Agent-Based Modelling for Archaeologists, was delivered with the support of the Erasmus+ programme of the European Union.
+
+# Methods
+
+## Open educational resources
 
@@
 -   promotion through a series of multiplier events.
+
+## Objectives and target audience
 
@@
 
-# Method
+## Online tutorials
 
 The project team developed five online tutorials.
-The tutorials were first designed using storyboards and later converted into interactive guided walk-throughs using NetLOGO Web (<https://www.netlogoweb.org/launch>) and Javascript and hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials>).
+The tutorials were first designed using storyboards and later converted into interactive guided walk-throughs using NetLogo Web (<https://www.netlogoweb.org/launch>) and Javascript and hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials>).
 During the project, initial testing was done by all authors and the issues functionality in GitHub was used to tackle bugs.
@@
 The feedback was obtained through two surveys using Qualtrics (<https://www.qualtrics.com/>).
-The participants of the events were asked to respond to two surveys: one before the start of the event and one after the end of the event.
+The participants of the events were asked to respond to one survey before the start of the event and one after the end of the event.
 The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix 1 for the questions).
@@
 
+## Dissemination of tutorials
+
+The ABMA tutorials are publicly hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials/>) and are citeable [@rocks-macqueen2024].
+Anyone interested can download the tutorials and their source code.
+To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (<https://abmarchaeologists.github.io/ABMA_website>) and can be followed without the need to download any software.
+The source code of the website is also openly available [@jutte2024].
+
+Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University [@scherjon2019] and external resources that people can use to learn agent-based modelling.
+The website was mainly developed by a student group from Saxion University of Applied Sciences [@jutte2024].
+
 # Results
 
-## Workhops and tutorials
-
-## Final tutorials
+## Tutorials
 
@@
 
-## Before the workshops
+## Tutorials and workshops
+
+The tutorials were tested during various workshops and updated afterward.
+As mentioned earlier, the participants of these workshops were asked to fill in a questionnaire both before and after the workshop.
+
+### Before the workshops
 
@@
 
-## During the workshop
+### During the workshop
 
@@
 
-## After the workshop
+### After the workshop
 
@@
 The respondents generally liked the interactive step-by-step way of going through the tutorials, but also valued the interaction with the teachers strongly.
-The easy and intuitive introduction in ABM and NetLOGO were also mentioned quite often.
+The easy and intuitive introduction in ABM and NetLogo were also mentioned quite often.
 This was also observed during the workshops as most participants were working at their own pace and did not need much help.
@@
 The archaeological periods that the participants were interested in were very diverse, ranging from the Paleolithic to the Medieval period.
+
+# Discussion
 
@@
 
-## Dissemination of tutorials
+## Iterative creation of tutorials
 
-The ABMA tutorials are publicly hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials/>) and are citeable [@rocks-macqueen2024].
-Anyone interested can download the tutorials and their source code.
-To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (<https://abmarchaeologists.github.io/ABMA_website>) and can be followed without the need to download any software.
-The source code of the website is also openly available [@jutte2024].
+*From review: Although the point can be somewhat interpreted throughout the manuscript, I would strongly suggest that the authors make a brief discussion of the experience (i.e.
+the iterative design and testing of educational material),*
 
-Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University [@scherjon2019] and external resources that people can use to learn agent-based modelling.
-The website was mainly developed by a student group from Saxion University of Applied Sciences, see the acknowledgements.
+## Computational skills in archaeology
+
+*From review: particularly in light of shortcomings and resistances to digital (and computational) skills in archaeology.*
 

### 2025-03-17 paper.tex
update paper based on reviews

@@
 \hypersetup{
-  pdftitle={Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists},
+  pdftitle={Lessons learned: creating tutorials teaching agent-based modelling for archaeologists},
   hidelinks,
@@
 
-\title{Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists}
+\title{Lessons learned: creating tutorials teaching agent-based modelling for archaeologists}
 \author{}
-\date{\vspace{-2.5em}November 13, 2024}
+\date{\vspace{-2.5em}March 17, 2025}
 
@@
 
-While Agent-Based Modelling importance in archaeological practice is steadily growing (Romanowska and Scherjon 2023) there are significant barriers to its wider adoption.
-One of the problems is the access to vocational training required to gain the necessary computer programming and modelling skills.
+While agent-based modelling importance in archaeological practice is steadily growing (Romanowska and Scherjon 2023) there are significant barriers to its wider adoption.
+One of the problems is the access to vocational training required to gain the necessary computer programming and modelling skills (Daems 2020).
 A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 76\% of modellers were primarily self-taught with some degree of secondary support, i.e.~27\% relied on peer support, only 7\% had any sort of formal training at a workshop or summer school, while only 14\% received training as part of a degree.
@@
 This project, Agent-Based Modelling for Archaeologists, was delivered with the support of the Erasmus+ programme of the European Union.
+
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
+
+\hypertarget{open-educational-resources}{%
+\subsection{Open educational resources}\label{open-educational-resources}}
 
@@
 \end{itemize}
+
+\hypertarget{objectives-and-target-audience}{%
+\subsection{Objectives and target audience}\label{objectives-and-target-audience}}
 
@@
 
-\hypertarget{method}{%
-\section{Method}\label{method}}
+\hypertarget{online-tutorials}{%
+\subsection{Online tutorials}\label{online-tutorials}}
 
 The project team developed five online tutorials.
-The tutorials were first designed using storyboards and later converted into interactive guided walk-throughs using NetLOGO Web (\url{https://www.netlogoweb.org/launch}) and Javascript and hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials}).
+The tutorials were first designed using storyboards and later converted into interactive guided walk-throughs using NetLogo Web (\url{https://www.netlogoweb.org/launch}) and Javascript and hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials}).
 During the project, initial testing was done by all authors and the issues functionality in GitHub was used to tackle bugs.
@@
 The feedback was obtained through two surveys using Qualtrics (\url{https://www.qualtrics.com/}).
-The participants of the events were asked to respond to two surveys: one before the start of the event and one after the end of the event.
+The participants of the events were asked to respond to one survey before the start of the event and one after the end of the event.
 The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix 1 for the questions).
@@
 
+\hypertarget{dissemination-of-tutorials}{%
+\subsection{Dissemination of tutorials}\label{dissemination-of-tutorials}}
+
+The ABMA tutorials are publicly hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials/}) and are citeable (Rocks-Macqueen et al.
+2024).
+Anyone interested can download the tutorials and their source code.
+To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (\url{https://abmarchaeologists.github.io/ABMA_website}) and can be followed without the need to download any software.
+The source code of the website is also openly available (Jutte et al.
+2024).
+
+Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University (Scherjon, Romanowska and Lambers 2019) and external resources that people can use to learn agent-based modelling.
+The website was mainly developed by a student group from Saxion University of Applied Sciences (Jutte et al.
+2024).
+
 \hypertarget{results}{%
@@
 
-\hypertarget{workhops-and-tutorials}{%
-\subsection{Workhops and tutorials}\label{workhops-and-tutorials}}
-
-\hypertarget{final-tutorials}{%
-\subsection{Final tutorials}\label{final-tutorials}}
+\hypertarget{tutorials}{%
+\subsection{Tutorials}\label{tutorials}}
 
@@
 
+\hypertarget{tutorials-and-workshops}{%
+\subsection{Tutorials and workshops}\label{tutorials-and-workshops}}
+
+The tutorials were tested during various workshops and updated afterward.
+As mentioned earlier, the participants of these workshops were asked to fill in a questionnaire both before and after the workshop.
+
 \hypertarget{before-the-workshops}{%
-\subsection{Before the workshops}\label{before-the-workshops}}
+\subsubsection{Before the workshops}\label{before-the-workshops}}
 
@@
 \hypertarget{during-the-workshop}{%
-\subsection{During the workshop}\label{during-the-workshop}}
+\subsubsection{During the workshop}\label{during-the-workshop}}
 
@@
 \hypertarget{after-the-workshop}{%
-\subsection{After the workshop}\label{after-the-workshop}}
+\subsubsection{After the workshop}\label{after-the-workshop}}
 
@@
 The archaeological periods that the participants were interested in were very diverse, ranging from the Paleolithic to the Medieval period.
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
@@
 
-\hypertarget{dissemination-of-tutorials}{%
-\subsection{Dissemination of tutorials}\label{dissemination-of-tutorials}}
+\hypertarget{iterative-creation-of-tutorials}{%
+\subsection{Iterative creation of tutorials}\label{iterative-creation-of-tutorials}}
 
-The ABMA tutorials are publicly hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials/}) and are citeable (Rocks-Macqueen et al.
-2024).
-Anyone interested can download the tutorials and their source code.
-To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (\url{https://abmarchaeologists.github.io/ABMA_website}) and can be followed without the need to download any software.
-The source code of the website is also openly available (Jutte et al.
-2024).
+\emph{From review: Although the point can be somewhat interpreted throughout the manuscript, I would strongly suggest that the authors make a brief discussion of the experience (i.e.~the iterative design and testing of educational material),}
 
-Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University (Scherjon, Romanowska and Lambers 2019) and external resources that people can use to learn agent-based modelling.
-The website was mainly developed by a student group from Saxion University of Applied Sciences, see the acknowledgements.
+\hypertarget{computational-skills-in-archaeology}{%
+\subsection{Computational skills in archaeology}\label{computational-skills-in-archaeology}}
+
+\emph{From review: particularly in light of shortcomings and resistances to digital (and computational) skills in archaeology.}
 
@@
 Strasbourg: Council of Europe Publishing.
+
+\leavevmode\vadjust pre{\hypertarget{ref-daems2020}{}}%
+Daems, D.
+2020 A Review and Roadmap of Online Learning Platforms and Tutorials in Digital Archaeology.
+\emph{Advances in Archaeological Practice} 8(1): 87--92.
+DOI: https://doi.org/\href{https://doi.org/10.1017/aap.2019.47}{10.1017/aap.2019.47}.
 

### 2025-03-17 Analyse_post_workshop_survey.R

@@ -4,32 +4,6 @@ library(dplyr)
 library(lubridate)
 library(forcats)
 
-post_surveys <- read.csv("data/post_workshop_surveys.csv")
-
-post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-04-03"]  <- "202304_CAA"
-post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-08-31"]  <- "202308_EAA"
-post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-10-12"]  <- "202310_CAA-DE-NLFl"
-post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-10-13"]  <- "202310_CAA-DE-NLFl"
-post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-11-16"]  <- "202311_Reuvensdagen"
-post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-11-24"]  <- "202311_CAA_UK"
-post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-12-12"]  <- "202312_Leiden"
-post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-12-13"]  <- "202312_Leiden"
-post_surveys$Workshop[date(post_surveys$Start.Date)>="2024-01-22"]  <- "202401_Aarhus"
-
-
-#correct typos in the survey
-post_surveys$How.would.you.rate.the.teaching.material.[post_surveys$How.would.you.rate.the.teaching.material.=="Sufficent"] <- "Sufficient"
-post_surveys$How.would.you.rate.the.teaching.material.[post_surveys$How.would.you.rate.the.teaching.material.==""] <- "Not shared"
-post_surveys$How.would.you.rate.the.workshop.in.general.[post_surveys$How.would.you.rate.the.workshop.in.general.=="Sufficent"] <- "Sufficient"
-post_surveys$How.would.you.rate.the.trainers.teachers.[post_surveys$How.would.you.rate.the.trainers.teachers.=="Sufficent"] <- "Sufficient"
-post_surveys$How.would.you.rate.the.trainers.teachers.[post_surveys$How.would.you.rate.the.trainers.teachers.==""] <- "Not shared"
[…]

### 2025-03-17 Analyse_pre_workshop_survey.R

@@ -4,61 +4,6 @@ library(tidyr)
 library(forcats)
 library(stringr)
 
-# importing data
-CAA_202304 <- read.csv("data/CAA_202304.csv")
-CAA_202304$Workshop <- "202304_CAA"
-EAA_202308 <- read.csv("data/EAA_202308.csv")
-EAA_202308$Workshop <- "202308_EAA"
-CAADENLFl_202310 <- read.csv("data/CAADENLFl_202310.csv")
-CAADENLFl_202310$Workshop <- "202310_CAA-DE-NLFl"
-Reuvensdagen_202311 <- read.csv("data/Reuvensdagen_202311.csv")
-Reuvensdagen_202311$Workshop <- "202311_Reuvensdagen"
-CAAUK_202311 <- read.csv("data/CAAUK_202311.csv")
-CAAUK_202311$Workshop <- "202311_CAA-UK"
-Leiden_202312 <- read.csv("data/Leiden_202312.csv")
-Leiden_202312$Workshop <- "202312_Leiden"
-Aarhus_202401 <- read.csv("data/Aarhus_online_202401.csv")
-Aarhus_202401$Workshop <- "202401_Aarhus"
-
-pre_surveys <-rbind(CAA_202304, EAA_202308, CAADENLFl_202310, Reuvensdagen_202311, CAAUK_202311, Leiden_202312, Aarhus_202401)
-
-# correct different spellings of nationalities
-replace_nationality <- c("USA" = "American", "United States of America" = "American", "American " = "American",
-                         "UK (Scotland)" = "British", "UK" = "British", "British / Danish" = "British",
-                         "British/Canadian" = "British", "British/Canadian " = "British", "Brazil" = "Brazilian",
-                         "Spain" = "Spanish", "Polish " = "Polish", "nl" = "Dutch",
-                         "NL" = "Dutch", "Nl" = "Dutch", "Netherlands" = "Dutch",
-                         "Nederlands" = "Dutch", "Dutch " = "Dutch", "Ditch" = "Dutch",
-                         "Dutch and Israeli" = "Dutch", "israel" = "Israeli", "italian" = "Italian",
-                         "Italy" = "Italian", "Hungary" = "Hungarian", "Hungarian " = "Hungarian",
-                         "Hungary " = "Hungarian", "greek" = "Greek", "Germany" = "German",
-                         "german" = "German","China" = "Chinese", "Belgium" = "Belgian",
[…]

### 2025-03-17 Preprocessing_data_post_workshop_survey.R

@@ -1,0 +1,33 @@
+### This script was used for preprocessing the data before analyses.
+### The underlying data is not share due to privacy issues in the original data.
+### The output data is used in the analyses scripts
+
+library(dplyr)
+library(lubridate)
+
+post_surveys <- read.csv("data/post_workshop_surveys.csv")
+
+post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-04-03"]  <- "202304_CAA"
+post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-08-31"]  <- "202308_EAA"
+post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-10-12"]  <- "202310_CAA-DE-NLFl"
+post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-10-13"]  <- "202310_CAA-DE-NLFl"
+post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-11-16"]  <- "202311_Reuvensdagen"
+post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-11-24"]  <- "202311_CAA_UK"
+post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-12-12"]  <- "202312_Leiden"
+post_surveys$Workshop[date(post_surveys$Start.Date)=="2023-12-13"]  <- "202312_Leiden"
+post_surveys$Workshop[date(post_surveys$Start.Date)>="2024-01-22"]  <- "202401_Aarhus"
+
+
+#correct typos in the survey
+post_surveys$How.would.you.rate.the.teaching.material.[post_surveys$How.would.you.rate.the.teaching.material.=="Sufficent"] <- "Sufficient"
+post_surveys$How.would.you.rate.the.teaching.material.[post_surveys$How.would.you.rate.the.teaching.material.==""] <- "Not shared"
+post_surveys$How.would.you.rate.the.workshop.in.general.[post_surveys$How.would.you.rate.the.workshop.in.general.=="Sufficent"] <- "Sufficient"
+post_surveys$How.would.you.rate.the.trainers.teachers.[post_surveys$How.would.you.rate.the.trainers.teachers.=="Sufficent"] <- "Sufficient"
+post_surveys$How.would.you.rate.the.trainers.teachers.[post_surveys$How.would.you.rate.the.trainers.teachers.==""] <- "Not shared"
[…]

### 2025-03-17 Preprocessing_data_pre_workshop_survey.R

@@ -1,0 +1,62 @@
+### This script was used for preprocessing the data before analyses.
+### The underlying data is not share due to privacy issues in the original data.
+### The output data is used in the analyses scripts
+
+library(dplyr)
+library(stringr)
+
+# importing data.
+CAA_202304 <- read.csv("data/CAA_202304.csv")
+CAA_202304$Workshop <- "202304_CAA"
+EAA_202308 <- read.csv("data/EAA_202308.csv")
+EAA_202308$Workshop <- "202308_EAA"
+CAADENLFl_202310 <- read.csv("data/CAADENLFl_202310.csv")
+CAADENLFl_202310$Workshop <- "202310_CAA-DE-NLFl"
+Reuvensdagen_202311 <- read.csv("data/Reuvensdagen_202311.csv")
+Reuvensdagen_202311$Workshop <- "202311_Reuvensdagen"
+CAAUK_202311 <- read.csv("data/CAAUK_202311.csv")
+CAAUK_202311$Workshop <- "202311_CAA-UK"
+Leiden_202312 <- read.csv("data/Leiden_202312.csv")
+Leiden_202312$Workshop <- "202312_Leiden"
+Aarhus_202401 <- read.csv("data/Aarhus_online_202401.csv")
+Aarhus_202401$Workshop <- "202401_Aarhus"
+
+pre_surveys <-rbind(CAA_202304, EAA_202308, CAADENLFl_202310, Reuvensdagen_202311, CAAUK_202311, Leiden_202312, Aarhus_202401)
+
+# correct different spellings of nationalities
+replace_nationality <- c("USA" = "American", "United States of America" = "American", "American " = "American",
+                         "UK (Scotland)" = "British", "UK" = "British", "British / Danish" = "British",
+                         "British/Canadian" = "British", "British/Canadian " = "British", "Brazil" = "Brazilian",
+                         "Spain" = "Spanish", "Polish " = "Polish", "nl" = "Dutch",
+                         "NL" = "Dutch", "Nl" = "Dutch", "Netherlands" = "Dutch",
+                         "Nederlands" = "Dutch", "Dutch " = "Dutch", "Ditch" = "Dutch",
+                         "Dutch and Israeli" = "Dutch", "israel" = "Israeli", "italian" = "Italian",
+                         "Italy" = "Italian", "Hungary" = "Hungarian", "Hungarian " = "Hungarian",
+                         "Hungary " = "Hungarian", "greek" = "Greek", "Germany" = "German",
[…]

### 2025-03-31 paper.Rmd
update paper based on review

@@
 While agent-based modelling importance in archaeological practice is steadily growing [@romanowska2023] there are significant barriers to its wider adoption.
-One of the problems is the access to vocational training required to gain the necessary computer programming and modelling skills [@daems2020].
+While education on things like GIS and databases have found their way into many study programs [e.g.
+@visser2016; @garstki2022], the access to vocational training required to gain the necessary computer programming and modelling skills is limited.
 A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 76% of modellers were primarily self-taught with some degree of secondary support, i.e.
 27% relied on peer support, only 7% had any sort of formal training at a workshop or summer school, while only 14% received training as part of a degree.
+While ABM is taught outside of archaeology [@augustijn2019], it remains lacking in most curricula due to for example lack of expertise in a university or pressure on the curriculum due to other subjects to be taught.
+The lack of training material on ABM for archaeologists, leaves this important method in digital archaeology at the "fringes of the archaeological discipline" [@daems2020, 90] rather than moving to the front lines.
 
@@
 
-*From review: Although the point can be somewhat interpreted throughout the manuscript, I would strongly suggest that the authors make a brief discussion of the experience (i.e.
-the iterative design and testing of educational material),*
+The chosen workflow by developing tutorials.
+testing these ourselves and then testing these during workshops and conferences has proven very valuable in many ways.
+First of all it had a very practical use.
+New users found bugs that we as a team missed or did not encounter.
+Participants used different operating systems, different web browsers and different settings on their computers, leading to other uses and other bugs.
+In addition, different people do different things on their computers.
+This behavioral differences between users helped us to understand how people would use our tutorials and enabled improvements.
+Thirdly, our presence on conferences and organizing these workshops improved dissemination of the tutorials.
+A fourth aspect is related, since the high number of people interested in these workshops has shown the need and demand for educational material for ABM.
 
@@
 
-*From review: particularly in light of shortcomings and resistances to digital (and computational) skills in archaeology.*
+Based on the high demand for our online and offline workshops, it might be possible to conclude that "resistance to digital tools" [@morgan2022, 216] might be dwindling.
+The demand for our tutorials and workshops, shows that there is an increasing group of archaeologists learning to learn a new tool.
+At the same time the tutorials help to move ABM from the fringes of archaeology in offering a easy access to ABM and programming.
+However, we also noted that some participants found the workshops interesting, but would not pursue to learn ABM any further.
+However, these will probably never apply ABM in their research, although they are now aware of the possibilities of the method, at least acknowledging the possibilities of ABM and digital archaeology in de broader sense.
+
+Teaching digital archaeology is often related to the the staff of an institute, although some institutes are incorporating digital aspects of archaeology in the regular curriculum [e.g.
+@visser2016; @lambers2024].
+The number of specialists in ABM is limited world wide and the OER's created in this project enable archaeologists an easy introduction into ABM, modelling and computational thinking.
+However, we encourage people to discuss their ideas with other people in the field, for example during CAA conferences (<https://caa-international.org/>).
+We have seen during the workshops that discussion is probably the most important aspect of model building, which is impossible to incorporate in self-paced learning.
 

### 2025-03-31 paper.tex
update paper based on review

@@
 \author{}
-\date{\vspace{-2.5em}March 17, 2025}
+\date{\vspace{-2.5em}March 31, 2025}
 
@@
 While agent-based modelling importance in archaeological practice is steadily growing (Romanowska and Scherjon 2023) there are significant barriers to its wider adoption.
-One of the problems is the access to vocational training required to gain the necessary computer programming and modelling skills (Daems 2020).
+While education on things like GIS and databases have found their way into many study programs (Garstki 2022; e.g.
+Visser, Zijverden and Alders 2016), the access to vocational training required to gain the necessary computer programming and modelling skills is limited.
 A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 76\% of modellers were primarily self-taught with some degree of secondary support, i.e.~27\% relied on peer support, only 7\% had any sort of formal training at a workshop or summer school, while only 14\% received training as part of a degree.
+While ABM is taught outside of archaeology (Augustijn et al.
+2019), it remains lacking in most curricula due to for example lack of expertise in a university or pressure on the curriculum due to other subjects to be taught.
+The lack of training material on ABM for archaeologists, leaves this important method in digital archaeology at the ``fringes of the archaeological discipline'' (Daems 2020: 90) rather than moving to the front lines.
 
@@
 
-\emph{From review: Although the point can be somewhat interpreted throughout the manuscript, I would strongly suggest that the authors make a brief discussion of the experience (i.e.~the iterative design and testing of educational material),}
+The chosen workflow by developing tutorials.
+testing these ourselves and then testing these during workshops and conferences has proven very valuable in many ways.
+First of all it had a very practical use.
+New users found bugs that we as a team missed or did not encounter.
+Participants used different operating systems, different web browsers and different settings on their computers, leading to other uses and other bugs.
+In addition, different people do different things on their computers.
+This behavioral differences between users helped us to understand how people would use our tutorials and enabled improvements.
+Thirdly, our presence on conferences and organizing these workshops improved dissemination of the tutorials.
+A fourth aspect is related, since the high number of people interested in these workshops has shown the need and demand for educational material for ABM.
 
@@
 
-\emph{From review: particularly in light of shortcomings and resistances to digital (and computational) skills in archaeology.}
+Based on the high demand for our online and offline workshops, it might be possible to conclude that ``resistance to digital tools'' (Morgan 2022: 216) might be dwindling.
+The demand for our tutorials and workshops, shows that there is an increasing group of archaeologists learning to learn a new tool.
+At the same time the tutorials help to move ABM from the fringes of archaeology in offering a easy access to ABM and programming.
+However, we also noted that some participants found the workshops interesting, but would not pursue to learn ABM any further.
+However, these will probably never apply ABM in their research, although they are now aware of the possibilities of the method, at least acknowledging the possibilities of ABM and digital archaeology in de broader sense.
+
+Teaching digital archaeology is often related to the the staff of an institute, although some institutes are incorporating digital aspects of archaeology in the regular curriculum (Lambers 2024; e.g.
+Visser, Zijverden and Alders 2016).
+The number of specialists in ABM is limited world wide and the OER's created in this project enable archaeologists an easy introduction into ABM, modelling and computational thinking.
+However, we encourage people to discuss their ideas with other people in the field, for example during CAA conferences (\url{https://caa-international.org/}).
+We have seen during the workshops that discussion is probably the most important aspect of model building, which is impossible to incorporate in self-paced learning.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-augustijn2019}{}}%
+Augustijn, E-W, Kounadi, O, Kuznecova, T and Zurita-Milla, R.
+2019 \emph{Teaching Agent-Based Modelling and Machine Learning in an integrated way.
+GeoComputation 2019} DOI: https://doi.org/\href{https://doi.org/10.17608/k6.auckland.9848804}{10.17608/k6.auckland.9848804}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-axtell2002}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-garstki2022}{}}%
+Garstki, K.
+2022 Teaching for Data Reuse and Working toward Digital Literacy in Archaeology.
+\emph{Advances in Archaeological Practice} 10(2): 177--186.
+DOI: https://doi.org/\href{https://doi.org/10.1017/aap.2022.3}{10.1017/aap.2022.3}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-grolemund2011}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-lambers2024}{}}%
+Lambers, K. 2024. \emph{\emph{\href{https://hdl.handle.net/1887/4054691}{The many challenges of digital and computational archaeology}}}. PhD thesis.
+
 \leavevmode\vadjust pre{\hypertarget{ref-marwick2017}{}}%
@@
 DOI: https://doi.org/\href{https://doi.org/10.1007/s10816-015-9272-9}{10.1007/s10816-015-9272-9}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-morgan2022}{}}%
+Morgan, C.
+2022 Current digital archaeology.
+\emph{Annual Review of Anthropology} 51(1): null.
+DOI: https://doi.org/\href{https://doi.org/10.1146/annurev-anthro-041320-114101}{10.1146/annurev-anthro-041320-114101}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-visser2016}{}}%
+Visser, RM, Zijverden, WK van and Alders, PG. 2016 \emph{Teaching digital archaeology digitally}. In: Campana, S, Scopigno, R, Carpentiero, G, and Cirollo, M (eds.). Oxford: Archaeopress. pp. 11--15.
+
 \leavevmode\vadjust pre{\hypertarget{ref-wickham2016}{}}%

### 2025-06-06 paper.Rmd
update paper and correct some figures

--- 2025-03-17 paper.Rmd
@@
 ---
-title: "Lessons learned: creating tutorials teaching agent-based modelling for archaeologists"
+title: "Lessons learned: creating tutorials to teach agent-based modelling to archaeologists"
 authors:
@@
 
-Agent-based modelling (ABM) is one of the most applied computational modelling research techniques in archaeology, because it is well suited to the questions asked by archaeologists.
+Agent-based modelling (ABM) is one of the most widely applied computational modelling research techniques in archaeology, because it is well suited to the questions asked by archaeologists.
 However, the access to vocational training for ABM is limited.
@@
 For this purpose we created online and open tutorials for ABM for archaeologists.
-In addition, how-to guides and support material was created, both for teachers and students.
-The material adheres to international (European) standards in relation to education and the Digital Skills Passport for archaeology.
-The teaching material is well tested using a diverse group of archaeologists during various international conferences and meetings.
+In addition, how-to guides and support material were created, both for teachers and students.
+The materials adhere to international (European) standards in relation to education and the Digital Skills Passport for archaeology.
+The teaching materials have been well tested using a diverse group of archaeologists during various international conferences and meetings.
 It was shown that the knowledge and skills of the participants in relation to ABM improved.
-In addition, the demand for the teaching material and the workshops was high, ensuring that the tutorials on ABM will be used in the future.
+In addition, the demand for the teaching materials and the workshops was high, ensuring that the tutorials on ABM will be used in the future.
 
@@
 
-While agent-based modelling importance in archaeological practice is steadily growing [@romanowska2023] there are significant barriers to its wider adoption.
-One of the problems is the access to vocational training required to gain the necessary computer programming and modelling skills [@daems2020].
-A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 76% of modellers were primarily self-taught with some degree of secondary support, i.e.
-27% relied on peer support, only 7% had any sort of formal training at a workshop or summer school, while only 14% received training as part of a degree.
+Although the importance of agent-based modelling in archaeological practice is steadily growing [@romanowska2023] there are significant barriers to its wider adoption.
+While education on ICT skills related to techniques like GIS and databases have found their way into many study programmes [@visser2016; @garstki2022], the access to vocational training required to gain the necessary computer programming and modelling skills for ABM is limited.
+A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 70.8% of modellers were primarily self-taught with some degree of secondary support, either by peers or academic supervisors.
+Only 13.8% received training as a formal part of a degree.
+While ABM is taught outside of archaeology [e.g.
+@augustijn2019] and various handbooks exist in different domains [@acerbi2023; @crooks2019; @wilensky2015; @hamill2016; @smaldino2023; @railsback2019], it is lacking in most archaeological curricula due to for example lack of expertise in a university or competition with other subjects to be taught in a given curriculum.
+There are several online courses available, (e.g.
+<https://acerbialberto.com/IBM-cultevo/>, <https://smaldino.com/wp/abm-courses/>, <https://forum.comses.net/t/educational-resources/9159>), but none are aimed at archaeology.
+The lack of training material on ABM for archaeologists, leaves this important method in digital archaeology at the "fringes of the archaeological discipline" [@daems2020, 90] rather than moving it to the front lines.
 
@@
 
-This project, Agent-Based Modelling for Archaeologists, was delivered with the support of the Erasmus+ programme of the European Union.
+This project, Agent-Based Modelling for Archaeologists (ABMA), was delivered with the support of the Erasmus+ programme of the European Union.
 
@@
 
-The project sought to create Open (Open Access and Open Source) Educational Resources (OERs) for training in Agent-Based Modelling for Archaeologists.
-These OERs were to be HTML and JavaScript based so they could be used on any device with a browser, even if it is not connected to the internet.
+The ABMA project sought to create Open (Open Access and Open Source) Educational Resources (OERs) for training in Agent-Based Modelling for Archaeologists.
+These OERs were to be HTML- and JavaScript-based so they could be used on any device with a browser, even if it is not connected to the internet.
 The full programme consists of five hands-on practical tutorials, taking learners through the process of creating the digital models.
 The modular architecture ensures that they can be independent or incorporated into other teaching tools like learning management systems.
-They are also deliberately 'method agnostic' so they can be incorporated into any tools from self-teaching to MOOCs, and are supported by a set of support materials to encourage use of the OERs.
+They are also deliberately 'method agnostic' so they can be incorporated into any tools from self-teaching to MOOCs, and are supported by a set of materials to encourage use of the OERs.
 
-The activities the project implemented were:
+The activities that the project implemented included:
 
@@
 
--   creation of a code and learning materials repository that will allow the project to continue sustainability after the European Union support has ended and to facilitate the creation of an open-source community to manage and improve the OERs in the future;
+-   creation of a code and learning materials repository that will allow the project to ensure sustainability after the European Union support has ended and to facilitate the creation of an open-source community to manage and improve the OERs in the future;
 
--   development of the framework for skills acquisition aligned with the Digital Skills Passport;
+-   development of a framework for skills acquisition aligned with the Digital Skills Passport;
 
@@
 
-1.  a surge in professionals who can build, apply and critique archaeological simulation thus opening new research avenues for the discipline as a whole;
+1.  a surge in professionals who can build, apply and critique archaeological simulation, thus opening new research avenues for the discipline as a whole;
 
@@
 
-For the secondary target audience of vocational training providers in archaeology e.g. employers, colleges and not-for-profit organisations, the expected results and outcomes were:
+For the secondary target audience of vocational training providers in archaeology, e.g. employers, colleges and not-for-profit organisations, the expected results and outcomes were:
 
@@
 The ABMA learning materials assume that participants have learning skills at least at a secondary education level.
-They are suitable for participants with at least some background in archaeology, such as those typical among students after the first year of a Bachelor study in Archaeology, i.e., above EQF levels 4 or 5.
+Also, they are suitable for participants with at least some background in archaeology, such as those typical among students after the first year of a Bachelor study in Archaeology, i.e., above EQF levels 4 or 5.
 In addition, the tutorials are aimed at professionals working in archaeology who have some experience with computer applications.
@@
 
-After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events all of which were run by members of the project team (Table \@ref(tab:events)).
-Recognising the difficulty of delivering practical training in an online environment (Microsoft Teams and Zoom) during the online events several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things.
+After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events, all of which were run by members of the project team (Table \@ref(tab:events)).
+Recognising the difficulty of delivering practical training in an online environment (Microsoft Teams and Zoom) during the online events, several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things.
 Throughout the project we sought feedback from the participants and used it to improve the tutorials prior to the following events.
 The feedback was obtained through two surveys using Qualtrics (<https://www.qualtrics.com/>).
-The participants of the events were asked to respond to one survey before the start of the event and one after the end of the event.
+The participants of the events were asked to respond to one survey before the start of the event and another one after the end of the event.
 The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix 1 for the questions).
@@
 The surveys were analysed using R [@rcoreteam2023].
-The following packages were used for analyses: ggplot2 [@wickham2016], dplyr [@wickham2023a], tidyr [@wickham2023], forcats[@wickham2023b], lubridate [@grolemund2011] and stringr [@wickham2023c].
+The following packages were used for analyses: ggplot2 [@wickham2016], dplyr [@wickham2023a], tidyr [@wickham2023], forcats [@wickham2023b], lubridate [@grolemund2011] and stringr [@wickham2023c].
 For the sake of reproducibility [@marwick2017] this paper was written in R Markdown (<https://rmarkdown.rstudio.com/>), with data and code available at <https://github.com/ABMArchaeologists/ABMA_paper> [@visser].
@@
 The aim of involving interdisciplinary groups in the project was to involve students in current research projects, to enable them to learn new things and share their innovative ideas and solutions with us.
-The students groups were supplementary to the deliverables of the project, but their new ideas were beneficial to the project.
-Their help during workshops was useful, but they also were engaged in developing educational material [@aalpoel2024], testing the tutorials, developing a style for the website [@jutte2024] and other materials.
+The outcomes of the students groups were supplementary to the deliverables of the project, but their new ideas were beneficial to the project.
+Their help during workshops was useful, but they also were engaged in developing educational material [@aalpoel2024], testing the tutorials, and developing a style for the website [@jutte2024] and other materials.
 
@@
 Anyone interested can download the tutorials and their source code.
-To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (<https://abmarchaeologists.github.io/ABMA_website>) and can be followed without the need to download any software.
+To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (<https://abmarchaeologists.github.io/ABMA_website>) and can be followed without the need to install any software.
 The source code of the website is also openly available [@jutte2024].
 
-Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University [@scherjon2019] and external resources that people can use to learn agent-based modelling.
+Besides the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University [@scherjon2019] and external resources that people can use to learn agent-based modelling.
 The website was mainly developed by a student group from Saxion University of Applied Sciences [@jutte2024].
@@
 
-The final set of ABMA tutorials is available for download at <https://github.com/ABMArchaeologists/ABMA_tutorials> and is citeable [@rocks-macqueen2024]. The set consists of the following tutorials:
+The final set of ABMA teaching materials comprises, next to accompanying materials (see above), the following tutorials [@rocks-macqueen2024]:
 
@@
 
-In tutorial 2 the user will learn the basics of NetLogo by making their simulation on the Out of Africal dispersal of *Homo sapiens* [@young1995].
+In tutorial 2 the user learns the basics of NetLogo by building their first own simulation modelled on the Out of Africal dispersal of *Homo sapiens* [@young1995].
 The learner works through the basics of NetLogo syntax and learns how to set up a simulation and visualize its outcomes.
-This tutorial consists of nine lessons.
-This tutorial intends to guide the learner from the introduction to the beginner level of proficiency.
+This tutorial consists of nine lessons that guide the learner from the introduction to the beginner level of proficiency.
 The learning curve is relatively gentle.
@@
 In the third tutorial, the learner builds a simple trade model, again based on a published simulation [@romanowska2018].
-Here learners expand their NetLogo programming skills with more complex syntactic structures as loops, lists and reporters.
-They are introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity.
+Here learners expand their NetLogo programming skills with more complex syntactic structures such as loops, lists and reporters.
+They are introduced to some key computational techniques like modular code development and debugging which become important alongside an increased coding complexity.
 This tutorial consists of seven lessons.
@@
 In addition, the tutorial introduces custom agent breeds, visualization with labels and reporters, plots and monitors.
-This takes the learner to a more advanced level, since it develops knowledge of the programming language combined with diagnosing unexpected behaviour and problem solving.
+This advances the proficiency level of the learner, since it develops knowledge of the programming language combined with diagnosing unexpected behaviour and problem solving.
 
@@
 
-In the final and fifth tutorial the learner learns more about how to incorporate agent-based modelling in archaeological research.
+In the fifth and final tutorial the learner learns more about how to incorporate agent-based modelling in archaeological research.
 This tutorial focuses less on programming in NetLogo and more on the principle and standards of computer-based modelling.
-The learner learns about the different phases of the model development process.
-This tutorial consists of five lessons.
-It summarizes the material delivered in the previous lessons into one coherent framework consistent with an intermediate level of proficiency.
+The tutorial consists of five lessons.
+It summarizes the material delivered in the previous lessons and relates it to the phases of the model development process: the conceptual phase, the technical phase and the dissemination phase.
 This tutorial approaches more theoretical aspects in a practical environment.
-The learner revisits all phases of the model development process starting with the conceptual phase and finishing with the analysis and interpretation of results.
-There is a particular focus on discussing what kind of research questions are suitable for modelling and how to pick the right modelling technique.
-The importance of properly conceptualizing a model before starting the technical phase of the model development is highlighted as it is a common issue for less experienced modellers.
+For the conceptual phase, particular focus is directed at a discussing what kind of research questions are suitable for modelling and how to pick the right modelling technique.
+The importance of properly conceptualizing a model before starting the technical part of model development is highlighted as it is a common issue for less experienced modellers.
 The experiment design phase, including parametrisation, validation and the analysis and interpretation of the models is explained at length since it often proves challenging for students.
@@
 
-The tutorials were tested during various workshops and updated afterward.
+The tutorials were tested during various workshops and updated afterwards.
 As mentioned earlier, the participants of these workshops were asked to fill in a questionnaire both before and after the workshop.
@@
 The number of female respondents slightly outnumbered the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \@ref(fig:gender-age)).
-The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years old indicating that the audience was predominantly composed of early career archaeologists.
+The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years of age, indicating that the audience was predominantly composed of early career archaeologists.
 It seems that the different events also had a slightly different distribution of both gender and age.
@@
 The majority (158) had never applied ABM to their research before participating in the workshops with only 14 participants who had experience of developing archaeological ABM.
-It is interesting to note that many respondents (108) did know what kind of software was available for ABM (see Figure \@ref(fig:abm-knowledge)).
+It is interesting to note that many respondents (75) could mention at least one type of software to be used for ABM (see Figure \@ref(fig:abm-knowledge)).
 
@@
   separate_rows(Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice, sep = ',\\s*') %>%
+  filter(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`!="None") %>%
   group_by(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`) %>%
@@
 As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before.
-The respondents were also asked how they rated the available resources on ABM (Figure \@ref(fig:available-theory)).
+The respondents were also asked how they rated the available theory on ABM (Figure \@ref(fig:available-theory)).
 While 64 of them had no opinion on the subject, a proportion of the respondents (28) answered that they rated the available theory on ABM as limited and only a small group (39) as sufficient or better.
@@
   geom_bar() + xlab("Quality of theory on ABM") + coord_flip() +
-  facet_grid(~Workshop, labeller = event_names)
+  facet_grid(~Workshop, labeller = event_names) + 
+  scale_x_discrete(drop=FALSE)
 ```
 
-### During the workshop
+### During the workshops
 
 During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials (Figure \@ref(fig:workshop-impression)).
-Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general.
+Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research or they discussed possibilities for the application of ABM in general.
 Individual progress varied depending on skills and preferences of the participants.
@@
 
-A similar proportion of the participants of the workshops gave us information after the workshops (171 of 368 participants).
+A similar proportion of the participants of the workshops gave us feedback after the workshops (171 of 368 participants).
 
@@
 post_surveys %>%
-  mutate(How.would.you.rate.the.workshop.in.general. = fct_relevel(How.would.you.rate.the.workshop.in.general., factor_levels)) %>%
+  mutate(How.would.you.rate.the.workshop.in.general. = factor(How.would.you.rate.the.workshop.in.general., factor_levels)) %>%
   ggplot(aes(x=How.would.you.rate.the.workshop.in.general.)) + geom_bar() +
   xlab("Rating of the workshop") + facet_grid(~Workshop, labeller = event_names) +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
+  scale_x_discrete(drop=FALSE)
 
@@
 
-The majority of the respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good.
-The teachers were rated even better than the teaching materials, with more respondents (165) giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).
+Most respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good.
+The teachers were rated even better than the teaching materials, with most respondents (165) giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).
 
@@
 rate_material_plot <- post_surveys %>%
-  mutate(How.would.you.rate.the.teaching.material. = fct_relevel(How.would.you.rate.the.teaching.material., factor_levels)) %>%
+  mutate(How.would.you.rate.the.teaching.material. = factor(How.would.you.rate.the.teaching.material., factor_levels)) %>%
   ggplot(aes(x=How.would.you.rate.the.teaching.material.)) + geom_bar() +
   xlab("Rating of the teaching material") + facet_grid(~Workshop, labeller = event_names) +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
+  scale_x_discrete(drop=FALSE)
 rate_teacher_plot <- post_surveys %>%
-  mutate(How.would.you.rate.the.trainers.teachers. = fct_relevel(How.would.you.rate.the.trainers.teachers., factor_levels)) %>%
+  mutate(How.would.you.rate.the.trainers.teachers. = factor(How.would.you.rate.the.trainers.teachers., factor_levels)) %>%
   ggplot(aes(x=How.would.you.rate.the.trainers.teachers.)) + geom_bar() +
   xlab("Rating of the teacher") + facet_grid(~Workshop, labeller = event_names) +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
+  scale_x_discrete(drop=FALSE)
 ggarrange(rate_material_plot, rate_teacher_plot, nrow = 2, ncol = 1)
@@
 The respondents generally liked the interactive step-by-step way of going through the tutorials, but also valued the interaction with the teachers strongly.
-The easy and intuitive introduction in ABM and NetLogo were also mentioned quite often.
+The easy and intuitive introduction to ABM and NetLogo were also mentioned quite often.
 This was also observed during the workshops as most participants were working at their own pace and did not need much help.
@@
 
-# Discussion
-
 ## Integration with Learning and Skills Tracking Frameworks
 
-Beyond the creation of the OERs the project also worked on aligning the learning outcomes of the OERs with the Digital Competence Framework for Citizens and the Digital Archaeological Skills Passport.
+Beyond the creation of the OERs, the project also worked on aligning the learning outcomes of the OERs with the Digital Competence Framework for Citizens and the Digital Archaeological Skills Passport.
 
@@
 
-As mentioned above, the following competence areas were identified as relevant for the tutorials: 1.
-Information and data literacy, 2.
-Communication and collaboration, 3.
-Digital content creation and, 5.
-Problem solving [@europeancommissionjointresearchcentre2022].
+As mentioned above, the following competence areas were identified as relevant for the tutorials: *1.
+Information and data literacy*, *2.
+Communication and collaboration*, *3.
+Digital content creation* and, *5.
+Problem solving* [@europeancommissionjointresearchcentre2022].
 In these competence areas various competences are developed towards an advanced or even specialised level.
@@
 
-For competence area 1.
-Information and data literacy: the tutorials mainly address 1.2 Evaluating Data, Information And Digital Content and 1.3 Managing Data, Information And Digital Content.
+For *Competence area 1.
+Information and data literacy*, the tutorials mainly address *1.2 Evaluating Data, Information And Digital Content* and *1.3 Managing Data, Information And Digital Content*.
 In the ABMA educational materials, users learn about working with data, being critical of their digital content and how to model and manage the data in the context of an ABM.
@@
 
-For Competence Area 2.
-Communication and collaboration the tutorials mainly address 2.1 Interacting Through Digital Technologies, because the users are constantly interacting with digital technologies, but communication is not relevant for the tutorials.
+For *Competence Area 2.
+Communication and collaboration* the tutorials mainly address *2.1 Interacting Through Digital Technologies*, because the users are constantly interacting with digital technologies, but communication is not relevant for the tutorials.
 However, during the workshops given in the course of the project, participants interacted with each other and the teachers.
 
-Competence Area 3.
-Digital content creation is one of the main competence areas for the ABM teaching material.
-In the process of learning ABM the users are constantly 3.1 Developing Digital Content and 3.2 Integrating And Re-Elaborating Digital Content.
-The material does not touch upon 3.3 Copyright And Licences, as they fall out of the scope of this project.
-A very important aspect of learning ABM is 3.4 Programming.
+*Competence Area 3.
+Digital content creation* is one of the main foci of the ABM teaching material.
+In the process of learning ABM the users are constantly *3.1 Developing Digital Content* and *3.2 Integrating And Re-Elaborating Digital Content*.
+The material does not touch upon *3.3 Copyright And Licences*, as they fall out of the scope of this project.
+A very important aspect of learning ABM is *3.4 Programming*.
 The users are guided through the process of developing the necessary programming skills to build a simulation.
@@
 
-Competence Area 5.
-Problem solving is very important and is closely tied to 3.4 Programming, since programming involves a lot of problem solving but also the general principles of model development that have a prominent place in the educational materials.
-In addition, the creation of a highly complex ABM consists of problem solving all the time, including development of research questions, conceptualisation of theories, experiment design and validating against data - all of which require creative abilities and a high level understanding of the scientific process.
-All competencies are addressed: 5.1 Solving Technical Problems, is common when writing code, learning about coding, error handling and going from pseudocode to real code.
-The competence 5.2 Identifying Needs And Technological Responses, is a central issue of the course, while 5.3 Creatively Using Digital Technology is an important aspect of so-called model thinking necessary to develop robust and insightful models.
-In tutorial 2 various aspects of 5.4 Identifying Digital Competence Gaps, are relevant, since this tutorial forces the user to assess their knowledge and skills.
+*Competence Area 5.
+Problem solving* is very important and is closely tied to *3.4 Programming*, since programming involves a lot of problem solving but also the general principles of model development that have a prominent place in the educational materials.
+In addition, the creation of a highly complex ABM consists of problem solving all the time, including development of research questions, conceptualisation of theories, experiment design and validating against data - all of which require creative abilities and a high level of understanding the scientific process.
+All competencies are addressed: *5.1 Solving Technical Problems*, is common when writing code, learning about coding, error handling and going from pseudocode to real code.
+The competence *5.2 Identifying Needs And Technological Responses*, is a central issue of the course, while *5.3 Creatively Using Digital Technology* is an important aspect of so-called model thinking necessary to develop robust and insightful models.
+In tutorial 2 various aspects of *5.4 Identifying Digital Competence Gaps*, are relevant, since this tutorial forces the user to assess their knowledge and skills.
 
@@
 It consists of the professional body in the United States of America, the *Register of Professional Archaeologists* and five archaeological companies in Europe, working and based in Ireland, the United Kingdom, the Netherlands, Germany and France.
-This is the passport that the ABMA integrate the OERs with.
 
@@
 
-The digital skills passport has not launched yet, but when it is, it will be found at this url: <https://archaeologypassport.com>.
+The digital skills passport has not been launched yet, but when it is, it will be found at this url: <https://archaeologypassport.com>.
+
+# Discussion
+
+This discussion focuses on the process of creating tutorials in an iterative process and how the educational materials can improve computational skills for archaeologists.
 
@@
 
-*From review: Although the point can be somewhat interpreted throughout the manuscript, I would strongly suggest that the authors make a brief discussion of the experience (i.e.
-the iterative design and testing of educational material),*
+The chosen workflow for developing the tutorials, namely testing the tutorials ourselves and then testing them during workshops and conferences proved very valuable in many ways.
+First of all, it had a very practical use; new users found bugs that we as a team missed or did not encounter.
+Participants used different operating systems, different web browsers and different settings on their computers, leading to other uses and other bugs.
+In addition, different people interact differently with software on their computers.
+These behavioural differences between users helped us to understand how people would use our tutorials and enabled improvements.
+Thirdly, our presence at conferences and the organization of workshops improved dissemination of the tutorials.
+A fourth aspect is related, since the high number of people interested in these workshops has shown the need and demand for educational materials for ABM.
 
@@
 
-*From review: particularly in light of shortcomings and resistances to digital (and computational) skills in archaeology.*
+Based on the high demand for our online and offline workshops, it is possible to conclude that "resistance to digital tools" [@morgan2022, 216] might be dwindling.
+The demand for our tutorials and workshops, shows that there is an increasing group of archaeologists willing to learn a new tool.
+At the same time the tutorials help to move ABM away from the fringes of archaeology by offering an easy access to ABM and programming.
+However, we also noted that some participants found the workshops interesting, but would not pursue further learning about ABM.
+Although, these participants will probably never apply ABM in their research, they are now aware of the possibilities of the method, at least acknowledging the possibilities of ABM and digital archaeology in the broader sense.
+
+Teaching digital archaeology is often contingent on the skills and interest of the teaching staff of a gavine higher education institute, although some institutes are incorporating digital aspects of archaeology in their regular curriculum [@visser2016; @lambers2024].
+The number of specialists in ABM is limited world wide and the OER's created in this project enable archaeologists to provide an easy introduction into ABM, modelling and computational thinking.
+However, we encourage people to discuss their ideas with other people in the field, for example during CAA conferences (<https://caa-international.org/>).
+We have seen during the workshops that discussion is probably the most important aspect of model building, which is impossible to incorporate in self-paced learning.
 
@@
 
-Future aspects of these open learning materials will be the further application and integration within the educational programmes in our universities and hopefully in others too.
+Future prospects of these open learning materials will be the further application and integration in the educational programmes in our universities and hopefully in others too.
 We have made the material openly available and we hope that the ABM-community will stay involved and that future updates will be community driven.
@@
 
--   February 2022-July 2023: Alice Poorthuis, Johan Broersma, Mandy Hazenberg, Paulina Fulneczek, Ties Heesink and their tutor: Jan Wille
+-   February 2022-July 2023: Alice Poorthuis, Johan Broersma, Mandy Hazenberg, Paulina Fulneczek, Ties Heesink and their tutor: Jan Willem Huson.
 
@@
 
+We thank our two reviewers Dries Daems and Andreas Angourakis for their valuable comments and suggestions that have improved this paper.
+We thank James Allison for his rol as recommender for PCI Archaeology, contributing to enhancing the quality of the paper.
+
 # References

### 2025-06-06 paper.tex
update paper and correct some figures

--- 2025-03-17 paper.tex
@@
 \hypersetup{
-  pdftitle={Lessons learned: creating tutorials teaching agent-based modelling for archaeologists},
+  pdftitle={Lessons learned: creating tutorials to teach agent-based modelling to archaeologists},
   hidelinks,
@@
 
-\title{Lessons learned: creating tutorials teaching agent-based modelling for archaeologists}
+\title{Lessons learned: creating tutorials to teach agent-based modelling to archaeologists}
 \author{}
-\date{\vspace{-2.5em}March 17, 2025}
+\date{\vspace{-2.5em}June 06, 2025}
 
@@
 
-While agent-based modelling importance in archaeological practice is steadily growing (Romanowska and Scherjon 2023) there are significant barriers to its wider adoption.
-One of the problems is the access to vocational training required to gain the necessary computer programming and modelling skills (Daems 2020).
-A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 76\% of modellers were primarily self-taught with some degree of secondary support, i.e.~27\% relied on peer support, only 7\% had any sort of formal training at a workshop or summer school, while only 14\% received training as part of a degree.
+Although the importance of agent-based modelling in archaeological practice is steadily growing (Romanowska and Scherjon 2023) there are significant barriers to its wider adoption.
+While education on ICT skills related to techniques like GIS and databases have found their way into many study programmes (Garstki 2022; Visser, Zijverden and Alders 2016), the access to vocational training required to gain the necessary computer programming and modelling skills for ABM is limited.
+A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 70.8\% of modellers were primarily self-taught with some degree of secondary support, either by peers or academic supervisors.
+Only 13.8\% received training as a formal part of a degree.
+While ABM is taught outside of archaeology (e.g.
+Augustijn et al.
+2019) and various handbooks exist in different domains (Acerbi, Mesoudi and Smolla 2023; Crooks et al.
+2019; Hamill and Gilbert 2016; Railsback and Grimm 2019; Smaldino 2023; Wilensky and Rand 2015), it is lacking in most archaeological curricula due to for example lack of expertise in a university or competition with other subjects to be taught in a given curriculum.
+There are several online courses available, (e.g.~\url{https://acerbialberto.com/IBM-cultevo/}, \url{https://smaldino.com/wp/abm-courses/}, \url{https://forum.comses.net/t/educational-resources/9159}), but none are aimed at archaeology.
+The lack of training material on ABM for archaeologists, leaves this important method in digital archaeology at the ``fringes of the archaeological discipline'' (Daems 2020: 90) rather than moving it to the front lines.
 
@@
 
-The project sought to create Open (Open Access and Open Source) Educational Resources (OERs) for training in Agent-Based Modelling for Archaeologists.
-These OERs were to be HTML and JavaScript based so they could be used on any device with a browser, even if it is not connected to the internet.
+The ABMA project sought to create Open (Open Access and Open Source) Educational Resources (OERs) for training in Agent-Based Modelling for Archaeologists.
+These OERs were to be HTML- and JavaScript-based so they could be used on any device with a browser, even if it is not connected to the internet.
 The full programme consists of five hands-on practical tutorials, taking learners through the process of creating the digital models.
 The modular architecture ensures that they can be independent or incorporated into other teaching tools like learning management systems.
-They are also deliberately `method agnostic' so they can be incorporated into any tools from self-teaching to MOOCs, and are supported by a set of support materials to encourage use of the OERs.
+They are also deliberately `method agnostic' so they can be incorporated into any tools from self-teaching to MOOCs, and are supported by a set of materials to encourage use of the OERs.
 
-The activities the project implemented were:
+The activities that the project implemented included:
 
@@
 \item
-  creation of a code and learning materials repository that will allow the project to continue sustainability after the European Union support has ended and to facilitate the creation of an open-source community to manage and improve the OERs in the future;
+  creation of a code and learning materials repository that will allow the project to ensure sustainability after the European Union support has ended and to facilitate the creation of an open-source community to manage and improve the OERs in the future;
 \item
-  development of the framework for skills acquisition aligned with the Digital Skills Passport;
+  development of a framework for skills acquisition aligned with the Digital Skills Passport;
 \item
@@
 \item
-  a surge in professionals who can build, apply and critique archaeological simulation thus opening new research avenues for the discipline as a whole;
+  a surge in professionals who can build, apply and critique archaeological simulation, thus opening new research avenues for the discipline as a whole;
 \item
@@
 
-For the secondary target audience of vocational training providers in archaeology e.g.~employers, colleges and not-for-profit organisations, the expected results and outcomes were:
+For the secondary target audience of vocational training providers in archaeology, e.g.~employers, colleges and not-for-profit organisations, the expected results and outcomes were:
 
@@
 
-After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events all of which were run by members of the project team (Table \ref{tab:events}).
-Recognising the difficulty of delivering practical training in an online environment (Microsoft Teams and Zoom) during the online events several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things.
+After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events, all of which were run by members of the project team (Table \ref{tab:events}).
+Recognising the difficulty of delivering practical training in an online environment (Microsoft Teams and Zoom) during the online events, several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things.
 Throughout the project we sought feedback from the participants and used it to improve the tutorials prior to the following events.
 The feedback was obtained through two surveys using Qualtrics (\url{https://www.qualtrics.com/}).
-The participants of the events were asked to respond to one survey before the start of the event and one after the end of the event.
+The participants of the events were asked to respond to one survey before the start of the event and another one after the end of the event.
 The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix 1 for the questions).
@@
 The following packages were used for analyses: ggplot2 (Wickham 2016), dplyr (Wickham et al.
-2023), tidyr (Wickham, Vaughan and Girlich 2023), forcats(Wickham 2023a), lubridate (Grolemund and Wickham 2011) and stringr (Wickham 2023b).
+2023), tidyr (Wickham, Vaughan and Girlich 2023), forcats (Wickham 2023a), lubridate (Grolemund and Wickham 2011) and stringr (Wickham 2023b).
 For the sake of reproducibility (Marwick 2017) this paper was written in R Markdown (\url{https://rmarkdown.rstudio.com/}), with data and code available at \url{https://github.com/ABMArchaeologists/ABMA_paper} (Visser et al.
@@
 The aim of involving interdisciplinary groups in the project was to involve students in current research projects, to enable them to learn new things and share their innovative ideas and solutions with us.
-The students groups were supplementary to the deliverables of the project, but their new ideas were beneficial to the project.
+The outcomes of the students groups were supplementary to the deliverables of the project, but their new ideas were beneficial to the project.
 Their help during workshops was useful, but they also were engaged in developing educational material (Aalpoel et al.
-2024), testing the tutorials, developing a style for the website (Jutte et al.
+2024), testing the tutorials, and developing a style for the website (Jutte et al.
 2024) and other materials.
@@
 Anyone interested can download the tutorials and their source code.
-To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (\url{https://abmarchaeologists.github.io/ABMA_website}) and can be followed without the need to download any software.
+To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (\url{https://abmarchaeologists.github.io/ABMA_website}) and can be followed without the need to install any software.
 The source code of the website is also openly available (Jutte et al.
@@
 
-Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University (Scherjon, Romanowska and Lambers 2019) and external resources that people can use to learn agent-based modelling.
+Besides the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University (Scherjon, Romanowska and Lambers 2019) and external resources that people can use to learn agent-based modelling.
 The website was mainly developed by a student group from Saxion University of Applied Sciences (Jutte et al.
@@
 
-The final set of ABMA tutorials is available for download at \url{https://github.com/ABMArchaeologists/ABMA_tutorials} and is citeable (Rocks-Macqueen et al.
-2024).
-The set consists of the following tutorials:
+The final set of ABMA teaching materials comprises, next to accompanying materials (see above), the following tutorials (Rocks-Macqueen et al. 2024):
 
@@
 
-In tutorial 2 the user will learn the basics of NetLogo by making their simulation on the Out of Africal dispersal of \emph{Homo sapiens} (Young and Bettinger 1995).
+In tutorial 2 the user learns the basics of NetLogo by building their first own simulation modelled on the Out of Africal dispersal of \emph{Homo sapiens} (Young and Bettinger 1995).
 The learner works through the basics of NetLogo syntax and learns how to set up a simulation and visualize its outcomes.
-This tutorial consists of nine lessons.
-This tutorial intends to guide the learner from the introduction to the beginner level of proficiency.
+This tutorial consists of nine lessons that guide the learner from the introduction to the beginner level of proficiency.
 The learning curve is relatively gentle.
@@
 In the third tutorial, the learner builds a simple trade model, again based on a published simulation (Romanowska 2018).
-Here learners expand their NetLogo programming skills with more complex syntactic structures as loops, lists and reporters.
-They are introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity.
+Here learners expand their NetLogo programming skills with more complex syntactic structures such as loops, lists and reporters.
+They are introduced to some key computational techniques like modular code development and debugging which become important alongside an increased coding complexity.
 This tutorial consists of seven lessons.
@@
 The number of female respondents slightly outnumbered the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \ref{fig:gender-age}).
-The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years old indicating that the audience was predominantly composed of early career archaeologists.
+The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years of age, indicating that the audience was predominantly composed of early career archaeologists.
 It seems that the different events also had a slightly different distribution of both gender and age.
@@
 The majority (158) had never applied ABM to their research before participating in the workshops with only 14 participants who had experience of developing archaeological ABM.
-It is interesting to note that many respondents (108) did know what kind of software was available for ABM (see Figure \ref{fig:abm-knowledge}).
+It is interesting to note that many respondents (75) could mention at least one type of software to be used for ABM (see Figure \ref{fig:abm-knowledge}).
 
@@
 As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before.
-The respondents were also asked how they rated the available resources on ABM (Figure \ref{fig:available-theory}).
+The respondents were also asked how they rated the available theory on ABM (Figure \ref{fig:available-theory}).
 While 64 of them had no opinion on the subject, a proportion of the respondents (28) answered that they rated the available theory on ABM as limited and only a small group (39) as sufficient or better.
@@
 
-\hypertarget{during-the-workshop}{%
-\subsubsection{During the workshop}\label{during-the-workshop}}
+\hypertarget{during-the-workshops}{%
+\subsubsection{During the workshops}\label{during-the-workshops}}
 
 During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials (Figure \ref{fig:workshop-impression}).
-Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general.
+Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research or they discussed possibilities for the application of ABM in general.
 Individual progress varied depending on skills and preferences of the participants.
@@
 
-The majority of the respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good.
-The teachers were rated even better than the teaching materials, with more respondents (165) giving them the \emph{excellent} score (see Figure \ref{fig:rating-teaching}).
+Most respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good.
+The teachers were rated even better than the teaching materials, with most respondents (165) giving them the \emph{excellent} score (see Figure \ref{fig:rating-teaching}).
 
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
 \hypertarget{integration-with-learning-and-skills-tracking-frameworks}{%
@@
 
-Beyond the creation of the OERs the project also worked on aligning the learning outcomes of the OERs with the Digital Competence Framework for Citizens and the Digital Archaeological Skills Passport.
+Beyond the creation of the OERs, the project also worked on aligning the learning outcomes of the OERs with the Digital Competence Framework for Citizens and the Digital Archaeological Skills Passport.
 
@@
 
-As mentioned above, the following competence areas were identified as relevant for the tutorials: 1.
-Information and data literacy, 2.
-Communication and collaboration, 3.
-Digital content creation and, 5.
-Problem solving (European Commission, Joint Research Centre et al.
+As mentioned above, the following competence areas were identified as relevant for the tutorials: \emph{1.
+Information and data literacy}, \emph{2.
+Communication and collaboration}, \emph{3.
+Digital content creation} and, \emph{5.
+Problem solving} (European Commission, Joint Research Centre et al.
 2022).
@@
 
-For competence area 1.
-Information and data literacy: the tutorials mainly address 1.2 Evaluating Data, Information And Digital Content and 1.3 Managing Data, Information And Digital Content.
+For \emph{Competence area 1.
+Information and data literacy}, the tutorials mainly address \emph{1.2 Evaluating Data, Information And Digital Content} and \emph{1.3 Managing Data, Information And Digital Content}.
 In the ABMA educational materials, users learn about working with data, being critical of their digital content and how to model and manage the data in the context of an ABM.
@@
 
-For Competence Area 2.
-Communication and collaboration the tutorials mainly address 2.1 Interacting Through Digital Technologies, because the users are constantly interacting with digital technologies, but communication is not relevant for the tutorials.
+For \emph{Competence Area 2.
+Communication and collaboration} the tutorials mainly address \emph{2.1 Interacting Through Digital Technologies}, because the users are constantly interacting with digital technologies, but communication is not relevant for the tutorials.
 However, during the workshops given in the course of the project, participants interacted with each other and the teachers.
 
-Competence Area 3.
-Digital content creation is one of the main competence areas for the ABM teaching material.
-In the process of learning ABM the users are constantly 3.1 Developing Digital Content and 3.2 Integrating And Re-Elaborating Digital Content.
-The material does not touch upon 3.3 Copyright And Licences, as they fall out of the scope of this project.
-A very important aspect of learning ABM is 3.4 Programming.
+\emph{Competence Area 3.
+Digital content creation} is one of the main foci of the ABM teaching material.
+In the process of learning ABM the users are constantly \emph{3.1 Developing Digital Content} and \emph{3.2 Integrating And Re-Elaborating Digital Content}.
+The material does not touch upon \emph{3.3 Copyright And Licences}, as they fall out of the scope of this project.
+A very important aspect of learning ABM is \emph{3.4 Programming}.
 The users are guided through the process of developing the necessary programming skills to build a simulation.
@@
 
-Competence Area 5.
-Problem solving is very important and is closely tied to 3.4 Programming, since programming involves a lot of problem solving but also the general principles of model development that have a prominent place in the educational materials.
-In addition, the creation of a highly complex ABM consists of problem solving all the time, including development of research questions, conceptualisation of theories, experiment design and validating against data - all of which require creative abilities and a high level understanding of the scientific process.
-All competencies are addressed: 5.1 Solving Technical Problems, is common when writing code, learning about coding, error handling and going from pseudocode to real code.
-The competence 5.2 Identifying Needs And Technological Responses, is a central issue of the course, while 5.3 Creatively Using Digital Technology is an important aspect of so-called model thinking necessary to develop robust and insightful models.
-In tutorial 2 various aspects of 5.4 Identifying Digital Competence Gaps, are relevant, since this tutorial forces the user to assess their knowledge and skills.
+\emph{Competence Area 5.
+Problem solving} is very important and is closely tied to \emph{3.4 Programming}, since programming involves a lot of problem solving but also the general principles of model development that have a prominent place in the educational materials.
+In addition, the creation of a highly complex ABM consists of problem solving all the time, including development of research questions, conceptualisation of theories, experiment design and validating against data - all of which require creative abilities and a high level of understanding the scientific process.
+All competencies are addressed: \emph{5.1 Solving Technical Problems}, is common when writing code, learning about coding, error handling and going from pseudocode to real code.
+The competence \emph{5.2 Identifying Needs And Technological Responses}, is a central issue of the course, while \emph{5.3 Creatively Using Digital Technology} is an important aspect of so-called model thinking necessary to develop robust and insightful models.
+In tutorial 2 various aspects of \emph{5.4 Identifying Digital Competence Gaps}, are relevant, since this tutorial forces the user to assess their knowledge and skills.
 
@@
 It consists of the professional body in the United States of America, the \emph{Register of Professional Archaeologists} and five archaeological companies in Europe, working and based in Ireland, the United Kingdom, the Netherlands, Germany and France.
-This is the passport that the ABMA integrate the OERs with.~
 
@@
 
-The digital skills passport has not launched yet, but when it is, it will be found at this url: \url{https://archaeologypassport.com}.
+The digital skills passport has not been launched yet, but when it is, it will be found at this url: \url{https://archaeologypassport.com}.
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+This discussion focuses on the process of creating tutorials in an iterative process and how the educational materials can improve computational skills for archaeologists.
 
@@
 
-\emph{From review: Although the point can be somewhat interpreted throughout the manuscript, I would strongly suggest that the authors make a brief discussion of the experience (i.e.~the iterative design and testing of educational material),}
+The chosen workflow for developing the tutorials, namely testing the tutorials ourselves and then testing them during workshops and conferences proved very valuable in many ways.
+First of all, it had a very practical use; new users found bugs that we as a team missed or did not encounter.
+Participants used different operating systems, different web browsers and different settings on their computers, leading to other uses and other bugs.
+In addition, different people interact differently with software on their computers.
+These behavioural differences between users helped us to understand how people would use our tutorials and enabled improvements.
+Thirdly, our presence at conferences and the organization of workshops improved dissemination of the tutorials.
+A fourth aspect is related, since the high number of people interested in these workshops has shown the need and demand for educational materials for ABM.
 
@@
 
-\emph{From review: particularly in light of shortcomings and resistances to digital (and computational) skills in archaeology.}
+Based on the high demand for our online and offline workshops, it is possible to conclude that ``resistance to digital tools'' (Morgan 2022: 216) might be dwindling.
+The demand for our tutorials and workshops, shows that there is an increasing group of archaeologists willing to learn a new tool.
+At the same time the tutorials help to move ABM away from the fringes of archaeology by offering an easy access to ABM and programming.
+However, we also noted that some participants found the workshops interesting, but would not pursue further learning about ABM.
+Although, these participants will probably never apply ABM in their research, they are now aware of the possibilities of the method, at least acknowledging the possibilities of ABM and digital archaeology in the broader sense.
+
+Teaching digital archaeology is often contingent on the skills and interest of the teaching staff of a gavine higher education institute, although some institutes are incorporating digital aspects of archaeology in their regular curriculum (Lambers 2024; Visser, Zijverden and Alders 2016).
+The number of specialists in ABM is limited world wide and the OER's created in this project enable archaeologists to provide an easy introduction into ABM, modelling and computational thinking.
+However, we encourage people to discuss their ideas with other people in the field, for example during CAA conferences (\url{https://caa-international.org/}).
+We have seen during the workshops that discussion is probably the most important aspect of model building, which is impossible to incorporate in self-paced learning.
 
@@
 \item
-  February 2022-July 2023: Alice Poorthuis, Johan Broersma, Mandy Hazenberg, Paulina Fulneczek, Ties Heesink and their tutor: Jan Wille
+  February 2022-July 2023: Alice Poorthuis, Johan Broersma, Mandy Hazenberg, Paulina Fulneczek, Ties Heesink and their tutor: Jan Willem Huson.
 \item
@@
 We also want to thank all the participants of our workshops.
+
+We thank our two reviewers Dries Daems and Andreas Angourakis for their valuable comments and suggestions that have improved this paper.
+We thank James Allison for his rol as recommender for PCI Archaeology, contributing to enhancing the quality of the paper.
 
@@
 2024 \emph{Teaching material Agent-Based Modelling for Archaeologists (EN/NL)} DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.10950153}{10.5281/zenodo.10950153}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-acerbi2023}{}}%
+Acerbi, A, Mesoudi, A and Smolla, M. 2023. \emph{Individual-based models of cultural evolution: a step-by-step guide using R}. London New York: Routledge.
+
+\leavevmode\vadjust pre{\hypertarget{ref-augustijn2019}{}}%
+Augustijn, E-W, Kounadi, O, Kuznecova, T and Zurita-Milla, R.
+2019 \emph{Teaching Agent-Based Modelling and Machine Learning in an integrated way.
+GeoComputation 2019} DOI: https://doi.org/\href{https://doi.org/10.17608/k6.auckland.9848804}{10.17608/k6.auckland.9848804}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-crooks2019}{}}%
+Crooks, A, Malleson, N, Manley, E and Heppenstall, AJ. 2019. \emph{Agent-based modelling \& geographical information systems: a practical primer}. Spatial analytics and GIS. London: SAGE.
+
 \leavevmode\vadjust pre{\hypertarget{ref-daems2020}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-garstki2022}{}}%
+Garstki, K.
+2022 Teaching for Data Reuse and Working toward Digital Literacy in Archaeology.
+\emph{Advances in Archaeological Practice} 10(2): 177--186.
+DOI: https://doi.org/\href{https://doi.org/10.1017/aap.2022.3}{10.1017/aap.2022.3}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-grolemund2011}{}}%
 Grolemund, G and Wickham, H. 2011 \href{https://www.jstatsoft.org/v40/i03/}{Dates and times made easy with lubridate}. \emph{Journal of Statistical Software} 40(3): 125.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hamill2016}{}}%
+Hamill, L and Gilbert, GN.
+2016.
+\emph{Agent-based modelling in economics}.
+Chichester, West Sussex, United Kingdom: Wiley.
+DOI: https://doi.org/\href{https://doi.org/10.1002/9781118945520}{10.1002/9781118945520}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-lambers2024}{}}%
+Lambers, K. 2024. \emph{\href{https://hdl.handle.net/1887/4054691}{The many challenges of digital and computational archaeology}}. Leiden: Leiden University.
+
 \leavevmode\vadjust pre{\hypertarget{ref-marwick2017}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-morgan2022}{}}%
+Morgan, C.
+2022 Current digital archaeology.
+\emph{Annual Review of Anthropology} 51(1): null.
+DOI: https://doi.org/\href{https://doi.org/10.1146/annurev-anthro-041320-114101}{10.1146/annurev-anthro-041320-114101}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2023}{}}%
 R Core Team. 2023 \emph{\href{https://www.R-project.org/}{R: A language and environment for statistical computing}}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-railsback2019}{}}%
+Railsback, SF and Grimm, V. 2019. \emph{Agent-based and individual-based modeling: a practical introduction}. Second edition. Princeton Oxford: Princeton University Press.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-smaldino2023}{}}%
+Smaldino, PE. 2023. \emph{Modeling social behavior: mathematical and agent-based models of social dynamics and cultural evolution}. Princeton: Princeton University press.
+
 \leavevmode\vadjust pre{\hypertarget{ref-visser2024}{}}%
@@
 DOI: https://doi.org/\href{https://doi.org/10.5281/zenodo.14134218}{10.5281/zenodo.14134218}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-visser2016}{}}%
+Visser, RM, Zijverden, WK van and Alders, PG. 2016 \emph{Teaching digital archaeology digitally}. In: Campana, S, Scopigno, R, Carpentiero, G, and Cirollo, M (eds.). Oxford: Archaeopress. pp. 11--15.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-wilensky2015}{}}%
+Wilensky, U and Rand, W. 2015. \emph{An introduction to agent-based modeling: modeling natural, social, and engineered complex systems with NetLogo}. Cambridge (Mass.): The MIT press.
+
 \leavevmode\vadjust pre{\hypertarget{ref-young1995}{}}%

### 2025-06-06 Analyse_post_workshop_survey.R

@@ -10,44 +10,50 @@ factor_levels <- c("Excellent", "Good", "Sufficient", "Insufficient", "Not share
 
 # rating workshop in general
 post_surveys %>%
-  mutate(How.would.you.rate.the.workshop.in.general. = fct_relevel(How.would.you.rate.the.workshop.in.general., factor_levels)) %>%
+  mutate(How.would.you.rate.the.workshop.in.general. = factor(How.would.you.rate.the.workshop.in.general., factor_levels)) %>%
   ggplot(aes(x=How.would.you.rate.the.workshop.in.general.)) + geom_bar() +
-  xlab("Rating of the workshop")
+  xlab("Rating of the workshop") +
+  scale_x_discrete(drop=FALSE)
 ggsave("export/postworkshop/rating_workshop.png")
 
 post_surveys %>%
-  mutate(How.would.you.rate.the.workshop.in.general. = fct_relevel(How.would.you.rate.the.workshop.in.general., factor_levels)) %>%
+  mutate(How.would.you.rate.the.workshop.in.general. = factor(How.would.you.rate.the.workshop.in.general., factor_levels)) %>%
   ggplot(aes(x=How.would.you.rate.the.workshop.in.general.)) + geom_bar() +
   xlab("Rating of the workshop") + facet_grid(~Workshop) +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
+  scale_x_discrete(drop=FALSE)
 ggsave("export/postworkshop/rating_workshop_facet.png", width = 10)
 
 # rating teaching material
+
 post_surveys %>%
-  mutate(How.would.you.rate.the.teaching.material. = fct_relevel(How.would.you.rate.the.teaching.material., factor_levels)) %>%
+  mutate(How.would.you.rate.the.teaching.material. = factor(How.would.you.rate.the.teaching.material., factor_levels)) %>%
   ggplot(aes(x=How.would.you.rate.the.teaching.material.)) + geom_bar() +
-  xlab("Rating of the teaching material")
+  xlab("Rating of the teaching material") +
+  scale_x_discrete(drop=FALSE)
 ggsave("export/postworkshop/rating_teaching_material.png")
 
 post_surveys %>%
[…]

### 2025-06-06 Analyse_pre_workshop_survey.R

@@ -57,12 +57,18 @@ ggsave("export/preworkshop/ABM_theory_quality_facet.png", width = 12)
 
 # ABM software
 
+# repondents knowing at least one kind of software for ABM
+pre_surveys %>%
+  select(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`) %>%
+  filter(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`!="") %>%
+  filter(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`!="None")
 
 
 pre_surveys %>%
   select(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`) %>%
   filter(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`!="") %>%
   separate_rows(Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice, sep = ',\\s*') %>%
+  filter(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`!="None") %>%
   group_by(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`) %>%
   mutate(count_software = n()) %>%
   ggplot(aes(x=reorder(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`, count_software))) + geom_bar() +

### 2025-06-20 paper.Rmd
update paper

@@
 In this paper we present the Open Educational Resources (OER) that we produced in an international cooperation.
-For this purpose we created online and open tutorials for ABM for archaeologists.
+For this purpose, we created online and open tutorials for ABM for archaeologists.
 In addition, how-to guides and support material were created, both for teachers and students.
@@
 These OERs were to be HTML- and JavaScript-based so they could be used on any device with a browser, even if it is not connected to the internet.
-The full programme consists of five hands-on practical tutorials, taking learners through the process of creating the digital models.
+The full programme consists of five hands-on practical tutorials, taking learners through the process of creating digital models.
 The modular architecture ensures that they can be independent or incorporated into other teaching tools like learning management systems.
@@
 
-1.  a surge in professionals who can build, apply and critique archaeological simulation, thus opening new research avenues for the discipline as a whole;
+1.  a surge in professionals who can build, apply, and critique archaeological simulation, thus opening new research avenues for the discipline as a whole;
 
@@
 increased level of digital competence, specifically in advanced computer skills that are highly transferable to other industries.
-This includes technical skills such as programming but also "soft" skills related to computational thinking;
+This includes technical skills such as programming, but also "soft" skills related to computational thinking;
 
@@
 The aim of involving interdisciplinary groups in the project was to involve students in current research projects, to enable them to learn new things and share their innovative ideas and solutions with us.
-The outcomes of the students groups were supplementary to the deliverables of the project, but their new ideas were beneficial to the project.
+The outcomes of the student groups were supplementary to the deliverables of the project, but their new ideas were beneficial to the project.
 Their help during workshops was useful, but they also were engaged in developing educational material [@aalpoel2024], testing the tutorials, and developing a style for the website [@jutte2024] and other materials.
@@
 
-Each tutorials consists of a number of individual lessons, which guide the learner in a self-paced manner through an increasingly challenging set of exercises.
+Each tutorial consists of a number of individual lessons, which guide the learner in a self-paced manner through an increasingly challenging set of exercises.
 Collectively, they build the skills and knowledge necessary to become a confident agent-based modeller.
@@
 This tutorial builds a foundation for working in NetLogo and with ABM.
-All activity stays within the entry level of difficulty, although some higher level concepts are lightly touched upon.
+All activity stays within the entry level of difficulty, although some higher-level concepts are lightly touched upon.
 
@@
 This tutorial approaches more theoretical aspects in a practical environment.
-For the conceptual phase, particular focus is directed at a discussing what kind of research questions are suitable for modelling and how to pick the right modelling technique.
+For the conceptual phase, particular focus is directed at discussing what kind of research questions are suitable for modelling and how to pick the right modelling technique.
 The importance of properly conceptualizing a model before starting the technical part of model development is highlighted as it is a common issue for less experienced modellers.
@@
 On a practical side, the BehaviorSpace - NetLogo's experiment environment - is explained to enable the learners to batch run their models.
-Finally the dissemination phase of the model development is also included to ensure that the standards in model publication are widely known and the learners explore the fundamental importance of replication for advancing scientific knowledge.
+Finally, the dissemination phase of model development is also included to ensure that the standards in model publication are widely known, and the learners explore the fundamental importance of replication for advancing scientific knowledge.
 
@@
 
-A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants).
+A large proportion of the participants in the workshops gave us information using the survey before the workshops (172 of 368 participants).
 The respondents came from at least 40 countries that represented almost all continents (see Figure \@ref(fig:nationality)).
 There was a clear skew towards participants from Europe.
-The number of female respondents slightly outnumbered the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \@ref(fig:gender-age)).
+The number of female respondents slightly outnumbered the male ones, and a small group did not share their gender, while two identified as non-binary (see Figure \@ref(fig:gender-age)).
 The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years of age, indicating that the audience was predominantly composed of early career archaeologists.
 It seems that the different events also had a slightly different distribution of both gender and age.
-For example, more older people attended the workshop at the CAA conference in April 2023 and more men were present during the workshop at the Reuvensdagen in November 2023.
+For example, more older people attended the workshop at the CAA conference in April 2023, and more men were present during the workshop at the Reuvensdagen in November 2023.
 This might be due to the differences of audiences at the conferences.
@@
 During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials (Figure \@ref(fig:workshop-impression)).
-Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research or they discussed possibilities for the application of ABM in general.
-Individual progress varied depending on skills and preferences of the participants.
+Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research, or they discussed possibilities for the application of ABM in general.
+Individual progress varied depending on the skills and preferences of the participants.
 While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later installments.
@@
 
-A similar proportion of the participants of the workshops gave us feedback after the workshops (171 of 368 participants).
+A similar proportion of the participants in the workshops gave us feedback after the workshops (171 of 368 participants).
 
@@
 Most respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good.
-The teachers were rated even better than the teaching materials, with most respondents (165) giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).
+The teachers were rated even better than the teaching materials, with most respondents (165) giving them an *excellent* score (see Figure \@ref(fig:rating-teaching)).
 
@@
 For *Competence area 1.
-Information and data literacy*, the tutorials mainly address *1.2 Evaluating Data, Information And Digital Content* and *1.3 Managing Data, Information And Digital Content*.
+Information and data literacy*, the tutorials mainly address *1.2 Evaluating Data, Information and Digital Content* and *1.3 Managing Data, Information And Digital Content*.
 In the ABMA educational materials, users learn about working with data, being critical of their digital content and how to model and manage the data in the context of an ABM.
@@
 Digital content creation* is one of the main foci of the ABM teaching material.
-In the process of learning ABM the users are constantly *3.1 Developing Digital Content* and *3.2 Integrating And Re-Elaborating Digital Content*.
-The material does not touch upon *3.3 Copyright And Licences*, as they fall out of the scope of this project.
+In the process of learning ABM the users are constantly *3.1 Developing Digital Content* and *3.2 Integrating and Re-Elaborating Digital Content*.
+The material does not touch upon *3.3 Copyright and Licences*, as they fall out of the scope of this project.
 A very important aspect of learning ABM is *3.4 Programming*.
@@
 All competencies are addressed: *5.1 Solving Technical Problems*, is common when writing code, learning about coding, error handling and going from pseudocode to real code.
-The competence *5.2 Identifying Needs And Technological Responses*, is a central issue of the course, while *5.3 Creatively Using Digital Technology* is an important aspect of so-called model thinking necessary to develop robust and insightful models.
+The competence *5.2 Identifying Needs and Technological Responses*, is a central issue of the course, while *5.3 Creatively Using Digital Technology* is an important aspect of so-called model thinking necessary to develop robust and insightful models.
 In tutorial 2 various aspects of *5.4 Identifying Digital Competence Gaps*, are relevant, since this tutorial forces the user to assess their knowledge and skills.
@@
 
-A digital skills passport is the same concept as a physical skills passport but it is digitised to overcome some of the drawbacks of a physical skills passport:
+A digital skills passport is the same concept as a physical skills passport, but it is digitised to overcome some of the drawbacks of a physical skills passport:
 
@@
 Thirdly, our presence at conferences and the organization of workshops improved dissemination of the tutorials.
-A fourth aspect is related, since the high number of people interested in these workshops has shown the need and demand for educational materials for ABM.
+A fourth aspect is related, since the high number of people interested in these workshops has shown the need and demand for educational materials for ABM and to improve computational skills.
 
@@
 Based on the high demand for our online and offline workshops, it is possible to conclude that "resistance to digital tools" [@morgan2022, 216] might be dwindling.
-The demand for our tutorials and workshops, shows that there is an increasing group of archaeologists willing to learn a new tool.
-At the same time the tutorials help to move ABM away from the fringes of archaeology by offering an easy access to ABM and programming.
-However, we also noted that some participants found the workshops interesting, but would not pursue further learning about ABM.
-Although, these participants will probably never apply ABM in their research, they are now aware of the possibilities of the method, at least acknowledging the possibilities of ABM and digital archaeology in the broader sense.
+The high demand was striking in Amsterdam during CAA 2023.
+The large number of registered participants forced us to expand the workshop and to split the group in two different rooms to account for all people.
+In addition, the number of registrations for the online workshop was far exceeding our expectations.
+While not everyone showed up, this does show that there is a high interest and demand.
+During the various workshops we encountered many people wanting to apply ABM to their own research, but lacking the skills or tools.
+The high demand for our tutorials and workshops shows that there is an increasing group of archaeologists willing to learn about new tools.
+At the same time, the tutorials help to move ABM away from the fringes of archaeology by offering an easy access to ABM and programming.
+Of course, some participants who found the workshops interesting may still not pursue ABM further.
+However, while they may not end up applying ABM in their own research, they are now more aware of the possibilities of ABM and digital archaeology more generally.
 
-Teaching digital archaeology is often contingent on the skills and interest of the teaching staff of a gavine higher education institute, although some institutes are incorporating digital aspects of archaeology in their regular curriculum [@visser2016; @lambers2024].
-The number of specialists in ABM is limited world wide and the OER's created in this project enable archaeologists to provide an easy introduction into ABM, modelling and computational thinking.
-However, we encourage people to discuss their ideas with other people in the field, for example during CAA conferences (<https://caa-international.org/>).
-We have seen during the workshops that discussion is probably the most important aspect of model building, which is impossible to incorporate in self-paced learning.
+Teaching digital archaeology is often contingent on the skills and interest of the teaching staff of a given higher education institute, although some institutes are incorporating digital aspects of archaeology in their regular curriculum [@visser2016; @lambers2024].
+The number of specialists in ABM is limited worldwide and the OER's created in this project offers archaeologists an easy introduction into ABM, modelling and computational thinking.
+However, we encourage people to discuss their ideas with other people in the field, for example during CAA conferences ([https://caa-international.org/](#0){.uri}).
+We have seen during the workshops that discussion can ben an important aspect of model building, one which is challenging to incorporate in self-paced learning.
 
@@
 Future prospects of these open learning materials will be the further application and integration in the educational programmes in our universities and hopefully in others too.
-We have made the material openly available and we hope that the ABM-community will stay involved and that future updates will be community driven.
+It would be interesting to test the effectiveness of the OERs in the future.
+We have made the material openly available.
+We hope that the ABM-community will stay involved and that future updates will be community driven.
 

### 2025-06-20 paper.tex
update paper

@@
 \author{}
-\date{\vspace{-2.5em}June 06, 2025}
+\date{\vspace{-2.5em}June 20, 2025}
 
@@
 \item
-  a surge in professionals who can build, apply and critique archaeological simulation, thus opening new research avenues for the discipline as a whole;
+  a surge in professionals who can build, apply, and critique archaeological simulation, thus opening new research avenues for the discipline as a whole;
 \item
   increased level of digital competence, specifically in advanced computer skills that are highly transferable to other industries.
-This includes technical skills such as programming but also ``soft'' skills related to computational thinking;
+This includes technical skills such as programming, but also ``soft'' skills related to computational thinking;
 \item
@@
 The aim of involving interdisciplinary groups in the project was to involve students in current research projects, to enable them to learn new things and share their innovative ideas and solutions with us.
-The outcomes of the students groups were supplementary to the deliverables of the project, but their new ideas were beneficial to the project.
+The outcomes of the student groups were supplementary to the deliverables of the project, but their new ideas were beneficial to the project.
 Their help during workshops was useful, but they also were engaged in developing educational material (Aalpoel et al.
@@
 
-A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants).
+A large proportion of the participants in the workshops gave us information using the survey before the workshops (172 of 368 participants).
 The respondents came from at least 40 countries that represented almost all continents (see Figure \ref{fig:nationality}).
 There was a clear skew towards participants from Europe.
-The number of female respondents slightly outnumbered the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \ref{fig:gender-age}).
+The number of female respondents slightly outnumbered the male ones, and a small group did not share their gender, while two identified as non-binary (see Figure \ref{fig:gender-age}).
 The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years of age, indicating that the audience was predominantly composed of early career archaeologists.
 It seems that the different events also had a slightly different distribution of both gender and age.
-For example, more older people attended the workshop at the CAA conference in April 2023 and more men were present during the workshop at the Reuvensdagen in November 2023.
+For example, more older people attended the workshop at the CAA conference in April 2023, and more men were present during the workshop at the Reuvensdagen in November 2023.
 This might be due to the differences of audiences at the conferences.
@@
 During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials (Figure \ref{fig:workshop-impression}).
-Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research or they discussed possibilities for the application of ABM in general.
-Individual progress varied depending on skills and preferences of the participants.
+Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research, or they discussed possibilities for the application of ABM in general.
+Individual progress varied depending on the skills and preferences of the participants.
 While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later installments.
@@
 Most respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good.
-The teachers were rated even better than the teaching materials, with most respondents (165) giving them the \emph{excellent} score (see Figure \ref{fig:rating-teaching}).
+The teachers were rated even better than the teaching materials, with most respondents (165) giving them an \emph{excellent} score (see Figure \ref{fig:rating-teaching}).
 
@@
 For \emph{Competence area 1.
-Information and data literacy}, the tutorials mainly address \emph{1.2 Evaluating Data, Information And Digital Content} and \emph{1.3 Managing Data, Information And Digital Content}.
+Information and data literacy}, the tutorials mainly address \emph{1.2 Evaluating Data, Information and Digital Content} and \emph{1.3 Managing Data, Information And Digital Content}.
 In the ABMA educational materials, users learn about working with data, being critical of their digital content and how to model and manage the data in the context of an ABM.
@@
 Digital content creation} is one of the main foci of the ABM teaching material.
-In the process of learning ABM the users are constantly \emph{3.1 Developing Digital Content} and \emph{3.2 Integrating And Re-Elaborating Digital Content}.
-The material does not touch upon \emph{3.3 Copyright And Licences}, as they fall out of the scope of this project.
+In the process of learning ABM the users are constantly \emph{3.1 Developing Digital Content} and \emph{3.2 Integrating and Re-Elaborating Digital Content}.
+The material does not touch upon \emph{3.3 Copyright and Licences}, as they fall out of the scope of this project.
 A very important aspect of learning ABM is \emph{3.4 Programming}.
@@
 All competencies are addressed: \emph{5.1 Solving Technical Problems}, is common when writing code, learning about coding, error handling and going from pseudocode to real code.
-The competence \emph{5.2 Identifying Needs And Technological Responses}, is a central issue of the course, while \emph{5.3 Creatively Using Digital Technology} is an important aspect of so-called model thinking necessary to develop robust and insightful models.
+The competence \emph{5.2 Identifying Needs and Technological Responses}, is a central issue of the course, while \emph{5.3 Creatively Using Digital Technology} is an important aspect of so-called model thinking necessary to develop robust and insightful models.
 In tutorial 2 various aspects of \emph{5.4 Identifying Digital Competence Gaps}, are relevant, since this tutorial forces the user to assess their knowledge and skills.
@@
 Based on the high demand for our online and offline workshops, it is possible to conclude that ``resistance to digital tools'' (Morgan 2022: 216) might be dwindling.
-The demand for our tutorials and workshops, shows that there is an increasing group of archaeologists willing to learn a new tool.
-At the same time the tutorials help to move ABM away from the fringes of archaeology by offering an easy access to ABM and programming.
-However, we also noted that some participants found the workshops interesting, but would not pursue further learning about ABM.
-Although, these participants will probably never apply ABM in their research, they are now aware of the possibilities of the method, at least acknowledging the possibilities of ABM and digital archaeology in the broader sense.
+The high demand was striking in Amsterdam during CAA 2023.
+The large number of registered participants forced us to expand the workshop and to split the group in two different rooms to account for all people.
+In addition, the number of registrations for the online workshop was far exceeding our expectations.
+While not everyone showed up, this does show that there is a high interest and demand.
+During the various workshops we encountered many people wanting to apply ABM to their own research, but lacking the skills or tools.
+The high demand for our tutorials and workshops shows that there is an increasing group of archaeologists willing to learn about new tools.
+At the same time, the tutorials help to move ABM away from the fringes of archaeology by offering an easy access to ABM and programming.
+Of course, some participants who found the workshops interesting may still not pursue ABM further.
+However, while they may not end up applying ABM in their own research, they are now more aware of the possibilities of ABM and digital archaeology more generally.
 
-Teaching digital archaeology is often contingent on the skills and interest of the teaching staff of a gavine higher education institute, although some institutes are incorporating digital aspects of archaeology in their regular curriculum (Lambers 2024; Visser, Zijverden and Alders 2016).
-The number of specialists in ABM is limited world wide and the OER's created in this project enable archaeologists to provide an easy introduction into ABM, modelling and computational thinking.
-However, we encourage people to discuss their ideas with other people in the field, for example during CAA conferences (\url{https://caa-international.org/}).
-We have seen during the workshops that discussion is probably the most important aspect of model building, which is impossible to incorporate in self-paced learning.
+Teaching digital archaeology is often contingent on the skills and interest of the teaching staff of a given higher education institute, although some institutes are incorporating digital aspects of archaeology in their regular curriculum (Lambers 2024; Visser, Zijverden and Alders 2016).
+The number of specialists in ABM is limited worldwide and the OER's created in this project offers archaeologists an easy introduction into ABM, modelling and computational thinking.
+However, we encourage people to discuss their ideas with other people in the field, for example during CAA conferences (\protect\hyperlink{0}{https://caa-international.org/}).
+We have seen during the workshops that discussion can ben an important aspect of model building, one which is challenging to incorporate in self-paced learning.
 

### 2025-07-24 paper.Rmd
minor updates after recommendation by PCI Archaeo (https://doi.org/10.24072/pci.archaeo.100596).

@@
 We thank our two reviewers Dries Daems and Andreas Angourakis for their valuable comments and suggestions that have improved this paper.
-We thank James Allison for his rol as recommender for PCI Archaeology, contributing to enhancing the quality of the paper.
+We thank James Allison for his rol as recommender for PCI Archaeology, contributing to enhancing the quality of the paper [@allison2025].
+A preprint version of this article has been peer-reviewed and recommended by PCI Archaeo (<https://doi.org/10.24072/pci.archaeo.100596>).
 

### 2025-07-24 paper.tex
minor updates after recommendation by PCI Archaeo (https://doi.org/10.24072/pci.archaeo.100596).

@@
 \author{}
-\date{\vspace{-2.5em}June 20, 2025}
+\date{\vspace{-2.5em}July 24, 2025}
 
@@
 We thank our two reviewers Dries Daems and Andreas Angourakis for their valuable comments and suggestions that have improved this paper.
-We thank James Allison for his rol as recommender for PCI Archaeology, contributing to enhancing the quality of the paper.
+We thank James Allison for his rol as recommender for PCI Archaeology, contributing to enhancing the quality of the paper (Allison 2025).
+A preprint version of this article has been peer-reviewed and recommended by PCI Archaeo (\url{https://doi.org/10.24072/pci.archaeo.100596}).
 
@@
 Acerbi, A, Mesoudi, A and Smolla, M. 2023. \emph{Individual-based models of cultural evolution: a step-by-step guide using R}. London New York: Routledge.
+
+\leavevmode\vadjust pre{\hypertarget{ref-allison2025}{}}%
+Allison, J.
+2025 Open Educational Resources for Agent-Based Modeling in Archaeology.
+\emph{Peer Community in Archaeology} 1: 100596.
+DOI: https://doi.org/\href{https://doi.org/10.24072/pci.archaeo.100596}{10.24072/pci.archaeo.100596}.
 

### 2025-08-26 paper.Rmd
minor corrections

@@
 (2002)."}
-knitr::include_graphics("paper_files/ScreenshotABMAtutorial1lesson3.png")
+knitr::include_graphics("figures/ScreenshotABMAtutorial1lesson3.png")
 ```
@@
 ```{r workshop-impression, echo = FALSE, out.width="100%", fig.cap = "Impression of various workshops (CAA 2023 in Amsterdam and EAA 2023 in Belfast) with participants working."}
-knitr::include_graphics("paper_files/CollageWorkshops.jpg")
+knitr::include_graphics("figures/CollageWorkshops.jpg")
 ```
@@
 This project was undertaken with the support of the Erasmus+ programme of the European Union (<https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054>).
-The publication was financed by The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382); IR was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513, the Aarhus University Research Foundation and FWO grant number W001220N.
+The Past Social Networks Project was funded by The Carlsberg Foundation's Young Researcher Fellowship (CF21-0382) awarded to Tom Brughmans.
+Iza Romanowska was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513, the Aarhus University Research Foundation and FWO grant number W001220N.
 

### 2025-08-26 paper.tex
minor corrections

@@
 \author{}
-\date{\vspace{-2.5em}July 24, 2025}
+\date{\vspace{-2.5em}August 26, 2025}
 
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{paper_files/ScreenshotABMAtutorial1lesson3} \caption{Screenshot of lesson 3 in tutorial 1 showing the population dynamics of the Long House Valley in the Artificial Anasazi model created by Axtell et al.
+\includegraphics[width=1\linewidth]{figures/ScreenshotABMAtutorial1lesson3} \caption{Screenshot of lesson 3 in tutorial 1 showing the population dynamics of the Long House Valley in the Artificial Anasazi model created by Axtell et al.
 (2002).}\label{fig:screenshot-lesson-3}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{paper_files/CollageWorkshops} \caption{Impression of various workshops (CAA 2023 in Amsterdam and EAA 2023 in Belfast) with participants working.}\label{fig:workshop-impression}
+\includegraphics[width=1\linewidth]{figures/CollageWorkshops} \caption{Impression of various workshops (CAA 2023 in Amsterdam and EAA 2023 in Belfast) with participants working.}\label{fig:workshop-impression}
 \end{figure}
@@
 This project was undertaken with the support of the Erasmus+ programme of the European Union (\url{https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054}).
-The publication was financed by The Carlsberg Foundation's \emph{Young Researcher Fellowship} (CF21-0382); IR was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513, the Aarhus University Research Foundation and FWO grant number W001220N.
+The Past Social Networks Project was funded by The Carlsberg Foundation's Young Researcher Fellowship (CF21-0382) awarded to Tom Brughmans.
+Iza Romanowska was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513, the Aarhus University Research Foundation and FWO grant number W001220N.
 

### 2025-09-01 paper.Rmd
Minor correction and generate png-files for paper graphs

@@
 date: '`r format(Sys.Date(), "%B %d, %Y")`'
+keywords: ["agent-based modelling", "education", "open educational resources", "archaeological skills", "computational skills", "simulation", "teaching"]
+abstract: "Agent-based modelling (ABM) is one of the most widely applied computational modelling research techniques in archaeology, because it is well suited to the questions asked by archaeologists.
+However, the access to vocational training for ABM is limited.
+In this paper we present the Open Educational Resources (OER) that we produced in an international cooperation.
+For this purpose, we created online and open tutorials for ABM for archaeologists.
+In addition, how-to guides and support material were created, both for teachers and students.
+The materials adhere to international (European) standards in relation to education and the Digital Skills Passport for archaeology.
+The teaching materials have been well tested using a diverse group of archaeologists during various international conferences and meetings.
+It was shown that the knowledge and skills of the participants in relation to ABM improved.
+In addition, the demand for the teaching materials and the workshops was high, ensuring that the tutorials on ABM will be used in the future."
 output:
@@
     fig_caption: yes
+    keep_md: yes
   bookdown::pdf_document2:
@@
 ```
-
-**Abstract**
-
-Agent-based modelling (ABM) is one of the most widely applied computational modelling research techniques in archaeology, because it is well suited to the questions asked by archaeologists.
-However, the access to vocational training for ABM is limited.
-In this paper we present the Open Educational Resources (OER) that we produced in an international cooperation.
-For this purpose, we created online and open tutorials for ABM for archaeologists.
-In addition, how-to guides and support material were created, both for teachers and students.
-The materials adhere to international (European) standards in relation to education and the Digital Skills Passport for archaeology.
-The teaching materials have been well tested using a diverse group of archaeologists during various international conferences and meetings.
-It was shown that the knowledge and skills of the participants in relation to ABM improved.
-In addition, the demand for the teaching materials and the workshops was high, ensuring that the tutorials on ABM will be used in the future.
 
@@
 
-```{r nationality, fig.cap="The nationality of the participants that filled in the survey.", fig.height=5, echo=FALSE}
+```{r nationality, dpi=600, fig.cap="The nationality of the participants that filled in the survey.", fig.height=5, echo=FALSE}
 pre_surveys %>% 
@@
 
-```{r gender-age, fig.cap="The gender and age distribution of the respondents for each workshop.", fig.width=11, echo=FALSE}
+```{r gender-age, dpi=600, fig.cap="The gender and age distribution of the respondents for each workshop.", fig.width=11, echo=FALSE}
 ggplot(pre_surveys, aes(x = `What.is.your.age.`, fill = `What.is.your.gender....Selected.Choice`)) +
@@
 
-```{r, computer-skills, fig.cap="The computer skills of the respondents.", out.width="50%", echo=FALSE}
+```{r, computer-skills, dpi=600, fig.cap="The computer skills of the respondents.", out.width="50%", echo=FALSE}
 pre_surveys %>%
@@
 
-```{r, abm-knowledge, fig.cap="Respondents knowledge of ABM per event (top), and knowledge of the various softwares for ABM (bottom), before participating in the workshops.", echo=FALSE}
+```{r, abm-knowledge, dpi=600, fig.cap="Respondents knowledge of ABM per event (top), and knowledge of the various softwares for ABM (bottom), before participating in the workshops.", echo=FALSE}
 knowledge_plot <- pre_surveys %>%
@@
 
-```{r, available-theory, fig.cap="Respondents opinion on the quality of theory on ABM faceted out by event.", fig.width=11, echo=FALSE}
+```{r, available-theory, dpi=600, fig.cap="Respondents opinion on the quality of theory on ABM faceted out by event.", fig.width=11, echo=FALSE}
 pre_surveys %>%
@@
 
-```{r, rating-workshop, fig.cap="Respondents rating of the workshop in general faceted for each event.", fig.width=11, echo=FALSE}
+```{r, rating-workshop, dpi=600, fig.cap="Respondents rating of the workshop in general faceted for each event.", fig.width=11, echo=FALSE}
 post_surveys %>%
@@
 
-```{r, rating-teaching, fig.cap="Respondents rating of the teaching material (top) and teachers (bottom) in general faceted for each event.", fig.width=11, fig.height=8, echo=FALSE}
+```{r, rating-teaching, dpi=600, fig.cap="Respondents rating of the teaching material (top) and teachers (bottom) in general faceted for each event.", fig.width=11, fig.height=8, echo=FALSE}
 rate_material_plot <- post_surveys %>%

### 2025-09-01 paper.tex
Minor correction and generate png-files for paper graphs

@@
   pdftitle={Lessons learned: creating tutorials to teach agent-based modelling to archaeologists},
+  pdfkeywords={agent-based modelling, education, open educational resources, archaeological skills, computational skills, simulation, teaching},
   hidelinks,
@@
 \author{}
-\date{\vspace{-2.5em}August 26, 2025}
+\date{\vspace{-2.5em}September 01, 2025}
 
@@
 \maketitle
-
-{
-\setcounter{tocdepth}{2}
-\tableofcontents
-}
-\textbf{Abstract}
-
+\begin{abstract}
 Agent-based modelling (ABM) is one of the most widely applied computational modelling research techniques in archaeology, because it is well suited to the questions asked by archaeologists.
@@
 In addition, the demand for the teaching materials and the workshops was high, ensuring that the tutorials on ABM will be used in the future.
+\end{abstract}
 
+{
+\setcounter{tocdepth}{2}
+\tableofcontents
+}
 \hypertarget{introduction}{%
